---
title: "Initial phosphoproteomics analysis/QC"
author: "Ruby Fore and Sara gosline"
date: "2024-07-29"
output: html_document
---
```{r, echo=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggfortify)
synapser::synLogin()
#corrected global proteomics: syn61695507
# corrected phospho proteomics: 'syn61695519'
fileEntity <- synapser::synGet('syn61695519')
origPhospho <- read.table(fileEntity$path, header = T)
```


### impute for NAs  - change to row mean 


```{r}
# get the indices of NA's
correctedProteomics<-origPhospho

na_indices <- which(is.na(correctedProteomics), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
correctedProteomics[na_indices] <- rowMeans(correctedProteomics, na.rm=T)[na_indices[,1]]
#na_indices
sum(is.na(correctedProteomics))
dim(correctedProteomics)

```

# reshape to long to do extra filtering and plot summary

Here I moved the data to long form to easily summarize, filter, and plot.

```{r remove outliers reshape}
library(dplyr)
#noOutliersCorrectedProteomics <- dplyr::select(correctedProteomics, -one_of(outliers))

samps<-colnames(correctedProteomics)
longProt<-correctedProteomics|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)

##quick summary of distributions 
ggplot(longProt,aes(x=condition,y=logRatio,fill=drug))+geom_violin()+theme(axis.text.x = element_text(angle = 90))
```

These look similar to the proteomics, but better as in there are fewer large fold change differences when we remove the NA values. Is it also driven by specific phosphosites?

```{r protein outliers}

longOrig<-origPhospho|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)|>
  subset(!is.na(logRatio))

ggplot(longOrig,aes(x=condition,y=logRatio,fill=drug))+geom_violin()+theme(axis.text.x = element_text(angle = 90))

bigProts<-subset(longOrig,abs(logRatio)>3)|>
  dplyr::group_by(gene)|>
  summarize(count=n())|>
  arrange(desc(count))|>
  subset(count>2)|>
  arrange(desc(count))
print(bigProts)
longOrig|>
  subset(gene%in%bigProts$gene)|>
  ggplot(aes(x=gene,y=logRatio))+geom_violin()+theme(axis.text.x = element_text(angle = 90))

```
It's more challenging to interpret this. CHGA is a marker of neuroendocrine tumors, so possibly a signal? Still we can remove a few samples to see if it helps
```{r outliers}

##what if we filter by logratio?
outliers<-subset(longProt,abs(logRatio)>5)|>
  group_by(condition)|>
  summarize(numOut=n())|>
  arrange(numOut)|>
  subset(numOut>1)
  
print(outliers)
```
I decided that maybe the samples with large LFCs are outliers and removed those. This doesn't always seem to be the case, some outliers just look like regular samples, but are a bit off.

```{r plot outliers}

##collect sample metadata, annotate our outliers
meta<-longProt|>
  dplyr::select(condition,sample,drug,time)|>
  distinct()|>
  mutate(isOutlier=condition%in%outliers$condition)|>
  tibble::column_to_rownames('condition')



pc<-prcomp(t(correctedProteomics))
autoplot(pc,data = meta,color='drug',shape='isOutlier')
autoplot(pc,data = meta,color='time',shape='isOutlier')+facet_grid(sample~drug)
```

Now we can see that these outliers are off and filter them out.

```{r remove outliers}
longProt<-longProt|>
  subset(!condition%in%outliers$condition)

meta<-meta|>
  subset(!isOutlier)

noOutliersCorrectedProteomics<-longProt|>
  dplyr::select(condition,gene,logRatio)|>
  tidyr::pivot_wider(names_from=condition,values_from=logRatio)|>
  tibble::column_to_rownames('gene')|>
  as.matrix()

pc2<-prcomp(t(noOutliersCorrectedProteomics))

autoplot(pc2,data = meta,color='drug',shape='sample')
autoplot(pc2,data = meta,color='drug',shape='sample')+facet_grid(~time)
ggsave('phosphoPCA.png')

```
Now the result is much more divided by sample, which we expect.

## Differential expression

We create a single function to compute differential expression using LIMMA.
```{r}
# function to get N genes, adj.P.value < 0.05
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  

factorInfo <- data.frame(stringr::str_split_fixed(names(data), "_", 5))

names(factorInfo) <- c('sample', 'trt', 'time','hr', 'rep')
factorInfo$sample <- as.factor(factorInfo$sample)
factorInfo$trt <-as.factor(factorInfo$trt)

factorInfo$time <- as.factor(factorInfo$time)

# rownames have already been assigned
proSet <- Biobase::ExpressionSet(as.matrix(data))

 # log2 transformation if distribution isn't normal
 ex <- Biobase::exprs(proSet)
 qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                           na.rm = TRUE
 ))
 LogC <- (qx[5] > 100) ||
   (qx[6] - qx[1] > 50 && qx[2] > 0)
 if (LogC) {
   ex[which(ex <= 0)] <- NaN
   Biobase::exprs(proSet) <- log2(ex)
 }


proSet$sample <- factorInfo$sample
proSet$trt <- factorInfo$trt
proSet$time <- factorInfo$time
 #browser()
design <- stats::model.matrix(formula, proSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
#print(cont.matrix)
fit <- limma::lmFit(proSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)
#browser()
fit <- limma::eBayes(fit)
deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
return(list(deg, nDiffExpressed))
}
```

### Compare all time points and conditions

A loop through the data frame calculates all the conditions and adds them to a new data frame.


```{r}
##created for loop to calculate all diff ex stats
t0dat<-data.frame()
ddat<-data.frame()
for(d in c('DMSO','Mirdametinib','Selumetinib','Vorinostat')){
  for(tv in c('0.5','3','24')){
    for(s in c('JH.2.002','MN.2')){
      ##get matrix of values of sample at condition
      tmat<-longProt|>
        subset(sample==s)|>
        subset(drug==d)|>
        subset(time==tv)|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
      if(nrow(tmat)==0)##there is no data for this
        next
      ##matrix of values at DMSO 0
      cmat<-longProt|>
        subset(sample==s)|>
        subset(drug=='DMSO')|>
        subset(time=='0')|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
  
      if(d!='DMSO'){    
        #matrix of values at DMSO same time
        dmat<-longProt|>
          subset(sample==s)|>
          subset(drug=='DMSO')|>
          subset(time==tv)|>
          select(gene,condition,logRatio)|>
          distinct()|>
          tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
        ## get result compared to DMSO
        res<-getNdiffExpressedGenes(cbind(dmat,tmat), formula = ~ trt  + 0, analysisContrast =paste0('trt',d,'-trtDMSO')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
      }
           ##first get restult compare to t0
        res<-getNdiffExpressedGenes(cbind(cmat,tmat), formula = ~ time  + 0, analysisContrast =paste0('time',tv,'-time0')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        t0dat<-rbind(t0dat,tres)
      
    }
  }
}

t0counts = t0dat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(t0counts)
ggplot(t0counts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)
```

Again we see large changes that occur during DMSO treatment, but not as much, particularly at 3 hours.

```{r other changes}

dcounts = ddat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(dcounts)
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)

```
Here we see a few changes at .5 hours and more at 24. the 3 hour time point is not great. Let's do an upset plot to see if any proteins are shared across conditions.

```{r upset}
library(UpSetR)
ddat|>
  subset(adj.P.Val<0.1)|>
  select(Gene,drug,time,samp)|>
  tidyr::unite(drug,time,samp,col='condition')|>
  mutate(present=1)|>
  tidyr::pivot_wider(names_from=condition,values_from=present,values_fill=0)|>
  tibble::column_to_rownames('Gene')|>
#  as.matrix()|>
    upset()
```

There are 30 phosphosites shared during mirdametinib treatment, and 19 shared under vorinostat treatment. One phosphosites is shared across all treatments?

```{r shared changes}

ddat|>
  subset(adj.P.Val<0.1)|>
  group_by(Gene)|>summarize(count=n())|>
  arrange(desc(count))
```

MAPK3 and MAPK1 exhibit changes in many conditions.

## Evaluate known target genes of various drugs

Here we can do some spot checking of known Mirdametinib target genes. First we can evaluate MEK, the known target
.
```{r}
erk<-c('MAPK1-','MAPK3-','MAPK2-')
mek<-c('MAP2K1-','MAP2K2','MAP2K4','MAP2K6')
genelist<-lapply(erk,
                 function(x) unique(longProt$gene[grep(x,longProt$gene)]))|>
  unlist()

mtargs<-subset(longProt,gene%in%genelist)

ggplot(mtargs,aes(x=time,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)
ggsave('erkPhoshpo.png',width=11)

ddat|>subset(Gene%in%genelist)|>
  mutate(sig=(adj.P.Val<0.1))|>
  ggplot(aes(x=time,y=logFC,shape=sig,col=drug))+geom_point()+facet_grid(samp~Gene)
ggsave('erkPhosphoDiffex.png',width=11)

genelist<-lapply(mek,
                 function(x) unique(longProt$gene[grep(x,longProt$gene)]))|>
  unlist()

mtargs<-subset(longProt,gene%in%genelist)

ggplot(mtargs,aes(x=time,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)
ggsave('mekPhoshpo.png',width=11)

ddat|>subset(Gene%in%genelist)|>
  mutate(sig=(adj.P.Val<0.1))|>
  ggplot(aes(x=time,y=logFC,shape=sig,col=drug))+geom_point()+facet_grid(samp~Gene)
ggsave('mekPhosphoDiffex.png',width=11)


```


This is exciting - MAPK1 is showing phosphorylation early on in Selumetinib treatment, and one Mirdametinib sample.MAPK3-T202 is showing consistent alteration.

Any changes in the transcriptional targets?

```{r mek trans targets}
genelist<-lapply(c('DUSP6','ETV5','SPRY2','SPRY4','EPHA4','ETV4','EPHA2','DUSP4','CCND1'),
                    function(x) unique(longProt$gene[grep(x,longProt$gene)]))|>
  unlist()

mtargs<-subset(longProt,gene%in%genelist)

ggplot(mtargs,aes(x=time,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)

ddat|>subset(Gene%in%genelist)|>
  mutate(sig=(adj.P.Val<0.1))|>
  ggplot(aes(x=time,y=logFC,shape=sig,col=drug))+geom_point()+facet_grid(samp~Gene)

```
Some downregulation of a phosphosite on SPRY2 in one sample

### Evaluate known vorinostat targets
here we can look at the HDACs that are affected by vorinostat
```{r hdac}
hdacs<-lapply(c('HDAC1','HDAC2','HDAC3','HDAC6'),
  function(x) unique(longProt$gene[grep(x,longProt$gene)]))|>
  unlist()
vtargs<-subset(longProt,gene%in%hdacs)

ggplot(vtargs,aes(x=time,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)

ddat|>subset(Gene%in%hdacs)|>
    mutate(sig=(adj.P.Val<0.1))|>
  ggplot(aes(x=time,y=logFC,col=drug,shape=sig))+geom_point()+facet_grid(samp~Gene)
```
