---
title: "Combination pathways"
author: "Sara Gosline"
date: "2026-04-01"
output: html_document
---

Here we take a top-down look of what combinations are highly effective and
what pathways the two drugs might share. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("pathwayChanges/plottingParameters.R")
source("pathwayCorrelationStats/correlation_functions.R")

##get the transcript data, this is slow
if (!exists('full_tmat'))
  full_tmat <- get_harmonized_gex()

```

Here we want to look at the top-ranking drug combinations to see if we 
can use the expression data to understand potential mechanisms.

## Compute Pathway correlations with single agent AUCs

Now we can look into our baseline gene expression to see if any of these drugs
are anti-correlated with known biological pathway. To do this we 
read in the gene expression data, Then calculate the correlation between untreated gene expression
and single agent AUC scores

```{r untreated gene expression}

nmat <- full_tmat[,intersect(colnames(full_tmat),
                             sing_exp$improve_sample_id)] |>
  as.data.frame() |>
  drop_na()

tmat <- get_gsea_scores(nmat)
rownames(tmat) <- sapply(rownames(tmat), function(x) tolower(gsub('HALLMARK ','',gsub('_',' ',x))))

print('computing pathway/aUC correlation')

#first focus on only durgs with 10 data points
#gvals <- md|>subset(count>10)

sing_exp <- sing_exp |> mutate(expr_time = '0hr')


cor_res <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_cor(sing_exp,tmat,drug,'0hr',48), 
           compute_cor(sing_exp,tmat,drug,'0hr',120) )
    })) #end drug iter
  

cor_sig <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_sig(sing_exp,tmat,drug,'0hr',48),
            #compute_sig(sing_exp,tmat,drug,'24hr','48'),
            compute_sig(sing_exp,tmat,drug,'0hr',120))
            #compute_sig(sing_exp,tmat,drug,'24hr','120'))
    })) #end drug iter
```

### Compute correlation of genes with single AUCs

THis is what we propose in grant

```{r full_transcript cor, eval=FALSE}


sing_exp <- sing_exp |> mutate(expr_time = '0hr')

cor_res <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_cor(sing_exp,nmat,drug,'0hr',48), 
           compute_cor(sing_exp,nmat,drug,'0hr',120) )
    })) #end drug iter
  
cor_sig <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_sig(sing_exp,nmat,drug,'0hr',48),
            #compute_sig(sing_exp,tmat,drug,'24hr','48'),
            compute_sig(sing_exp,nmat,drug,'0hr',120))
            #compute_sig(sing_exp,tmat,drug,'24hr','120'))
    })) #end drug iter


merged_res <- cor_res |> left_join(cor_sig) |>
  subset(!drug %in% c('sn-38')) ##removing because we had too few samples

##let's call a signature anything with sig <0.001 and corval < -.5
dsigs <- subset(merged_res, sig<0.001) |> subset(corval<(-0.5))

dsigs |> group_by(drug) |> summarize(ngenes=n_distinct(gene))
```

## Look for genes in diffex genes (no time to finish)

```{r get diffex}

diffex_genes <- read.csv(synapser::synGet('syn64397743')$path) |>
  select(-Version)

```


## Identify significantly correlated pathways

Here we use a significance test to see if specific correlations are more 
significant.

```{r cor sig, eval=TRUE}

print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_sig(sing_exp,tmat,drug,'0hr',48),
            #compute_sig(sing_exp,tmat,drug,'24hr','48'),
            compute_sig(sing_exp,tmat,drug,'0hr',120))
            #compute_sig(sing_exp,tmat,drug,'24hr','120'))
    })) #end drug iter


merged_res <- cor_res |> left_join(cor_sig) |>
  subset(!drug %in% c('sn-38')) ##removing because we had too few samples


subset(merged_res, sig < 0.05) |>
  head()


```




## Read in perturbed pathways see what mirda does

Check for overlap between up-regulated pathways and those correlated with drug response.

```{r combine perturbed pathways}

#make sure we ran the RNA seq analysis first
hall_preds <- readr::read_csv('pathwayChanges/GSEA_Hallmark/GSEA_results_Hallmark.csv') |>
  mutate(Feature_set = tolower(gsub('_',' ',Feature_set))) |>
  mutate(Feature_set = gsub("hallmark ", "",Feature_set))


##step 1: filter for up-regulated pathways
upregs <- subset(hall_preds, sig == TRUE & NES > 0) |>
        mutate(Pathway = stringr::str_replace_all(Feature_set,'hallmark ','')) |>
  mutate(DrugTreatment = tolower(DrugTreatment))

##step 2: filter for negatively correlated pathwasy
  combo_preds <- merged_res |>
    dplyr::rename(secondDrug = 'drug','corP' = sig) |>
      mutate(Pathway = stringr::str_replace_all(gene,'HALLMARK_','')) |>
    right_join(upregs) |>
    subset(corP < 0.05) |>
    subset(corval < 0) |>
    dplyr::select(Time,DrugTreatment,secondDrug,ttime,Pathway, pathScore='NES',pathwayCor = 'corval')
  
head(combo_preds)
```

First I create a function to compare this systematically and identify pathways

```{r function to compare}
##for a particular set of drugs, get overlapping pathwas

  
##get a pathway name, plot it across a drug treatment and secondary drug
plot_pathway_with_drugs <- function(genes, 
                                    expr_drug = 'mirdametinib', 
                                    pathwayname='',
                                    ft=''){
  ##get pathway across drug expression changes
  #genes <- gene_sets[[path_name]]
  
  tsamps <- grep(expr_drug, tolower(samples$other_id))
  ##get the DMSO for the same samples
  csamps <- intersect(grep('patient derived xenograft', samples$other_id),
                              grep('3D-MEDS',samples$model_type))

  ## get pathway across untreated samples
  gvals <- full_tmat[genes,intersect(colnames(full_tmat),samples$improve_sample_id[c(tsamps,csamps)])]
#  gvals <- drop_na(as.data.frame(gvals))
  
  meta <- data.frame(drug = c(rep(expr_drug,length(tsamps)),rep('untreated',length(csamps))),
                     sample = c(samples$common_name[tsamps],samples$common_name[csamps]))
  rownames(meta) <- samples$improve_sample_id[c(tsamps,csamps)]
  anncol <- list(drug=c(classCol[expr_drug],untreated = 'white'))
  
  pheatmap::pheatmap(gvals, cellwidth = 10,cellheight = 10,annotation_col=meta, 
                    # clustering_distance_rows = 'correlation',
                     clustering_method = 'ward.D2',main = gsub('HALLMARK_','',pathwayname),
                    annotation_colors = anncol,
                     filename = paste0(ft,'_heatmap.pdf'))
  
}
##from sing_gsea results (up-regulation?) and merged_res, from other

plot_pathway_correlation <- function(pathname,ft = ''){

  dc <- data.frame(drug=tolower(names(drug.class)),class = unlist(drug.class) )
  cc <- data.frame(drug= tolower(names(classCol)), color = unlist(classCol))
  classes <- merge(dc,cc)
  colbyclass <- classes$color
  names(colbyclass) <- classes$class
  
  cr <- merged_res |> 
    subset(gene == pathname) |>
    mutate(Signif = ifelse(sig < 0.01,'**',ifelse(sig<0.05,'*',""))) |>
    left_join(classes)
  
 # print(head(cr))
  mv <- cr |> group_by(drug) |>
    summarize(mv = mean(corval)) |>
    arrange(mv)
  
  cr$drug = factor(cr$drug, levels=unique(mv$drug))
  cr |> ggplot(aes(fill = as.factor(ttime), x = drug, y = corval))+#,group=drug)) + 
     geom_bar(position = 'dodge',stat = 'identity') + 
     ggrepel::geom_text_repel(aes(label = Signif), nudge_y = -0.001)+ #+ facet_grid(ttime~.) +
    ggtitle(paste("Drug response correlation with",gsub('HALLMARK_','',pathname)))    +
    theme_bw() +    
    scale_fill_manual(values = c('slateblue4','darkred')) +
#    scale_fill_manual(values = colbyclass) +
    coord_flip() +
    ylab('AUC correlation') + xlab('Drug')+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  
  ggsave(paste0(ft,'drugCor.pdf'),width = 6,height=7)
}

##this function takes a combination, finds pathways that
##are common in both, then plots the gene expression and AUCs
##
plot_combo_get_overlap <- function(combo, prefix = ''){
  
  combo = tolower(combo)

  ##now we can look for overlap
  c1 <- subset(combo_preds,DrugTreatment==combo[1]) |> subset(secondDrug==combo[2])
  c2 <- subset(combo_preds,DrugTreatment==combo[2]) |> subset(secondDrug==combo[1])
  
  if(nrow(c1)>0){
    ##plot genes up-regulated?
    apply(c1,1,function(y){
      rv <- intersect(which(hall_preds$Time==y[['Time']]),
              which(hall_preds$Feature_set==y[['Pathway']]))
    rv <- intersect(rv,which(tolower(hall_preds$DrugTreatment)==combo[1]))
    genes <- stringr::str_split(hall_preds$Leading_edge[rv], pattern=', ') |>
      unlist()

    title=paste0(prefix,paste(combo,collapse='_'),y[['Pathway']])
    plot_pathway_with_drugs(genes,combo[1],
                            pathwayname=y[['Pathway']],
                            ft=title)
    ##plot pathway correlation

    plot_pathway_correlation(y[['Pathway']],ft=title)

    })
  }
  
    if(nrow(c2)>0){
    ##plot genes up-regulated?
  apply(c2,1,function(y){
      rv <- intersect(which(hall_preds$Time==y[['Time']]),
              which(hall_preds$Feature_set==y[['Pathway']]))
    rv <- intersect(rv,which(tolower(hall_preds$DrugTreatment) == combo[2]))
    genes <- stringr::str_split(hall_preds$Leading_edge[rv], pattern=', ') |>
      unlist()
    
    title=paste0(prefix,paste0(prefix,paste(rev(combo),collapse='_'),y[['Pathway']]))
    plot_pathway_with_drugs(genes,combo[2],
                            pathwayname=y[['Pathway']],
                            ft=title)
    ##plot pathway correlation

    plot_pathway_correlation(y[['Pathway']],
                             ft=title)

    })
  }
  #print(paste("We found",nrow(c1),'overlapping pathways from',combo[1],'upregulation and negative corelation to',combo[2]))
  #print(paste("We found",nrow(c2),'overlapping pathways from',combo[2],'upregulation and negative corelation to',combo[1]))

  ##how many pathways are up-regulated and anti-correlated? 
    return(unique(c(c1$Pathway,c2$Pathway)))
}
  

```

Just wanted to confirm, but there is no relationship between overlapping pathways
and dAUC.

# Plotting priority drug combinations

With this data in place we can now move to the data-specific plots.

## Mirda + TNO combination

What stands out from this combination that could give us transcriptional evidence?
It looks like a single pathway


```{r mirda tno}

combo <- c("Mirdametinib",'TNO155')

res <- plot_combo_get_overlap(combo, prefix = 'fig5_')


subset(hall_preds,DrugTreatment=='Mirdametinib') |> subset(sig)|>
  ggplot(aes(x=Feature_set, y = NES, fill=Time))+ geom_bar(position='dodge',stat='identity') + coord_flip() + ggtitle('Pathways altered in mirdametininb treatment') +
  scale_fill_manual(values=c('slateblue4','darkred'))

#plot_pathway_correlation(pathname='HALLMARK_KRAS_SIGNALING_UP',ft='kras_only_up')

##plot genes in pathway in treated samples?


```


## Mirda + vorin combination

```{r mirda vorin}

combo <- c("Mirdametinib",'Vorinostat')
res <- plot_combo_get_overlap(combo, prefix = '')

sig_paths <- subset(hall_preds, sig)

subset(sig_paths,DrugTreatment %in% c("Vorinostat",'Mirdametinib')) |>
  subset(NES > 0) |>
  ggplot(aes(x = Feature_set, y = NES, fill = Time)) +
  geom_bar(position = 'dodge',stat = 'identity') + coord_flip() + ggtitle('Pathways altered in treatment') +
  facet_grid(~DrugTreatment) +
  scale_fill_manual(values = c('slateblue4','darkred')) +
    theme_bw() 

ggsave('mir_vorin_altered_path.pdf',width=7, height=6)

merged_res|>subset(drug %in%c('tno155','mirdametinib')) |> subset(corval<0)|> subset(sig<0.05) |> ggplot(aes(x=gene, y = corval,fill=as.factor(ttime))) + geom_bar(position='dodge',stat='identity') + facet_grid(~drug) + coord_flip() + theme_bw() +   scale_fill_manual(values=c('slateblue4','darkred'))

ggsave('mir_tno_cor.pdf',width=7,height=3)
```


## EGFR enrichment 

Now we can try to plot the EGFR enrichmetn in the CCLE data

```{R get ccle data, MOA, and enrichment}

##read in data
#TODO: put these files on synapse
r1 <- readRDS(synGet('syn74935645')$path)
r2 <- readRDS(synGet('syn74935650')$path)

cmat <- rbind(r1,r2) |>
  as.data.frame() |>
  tibble::remove_rownames() |>
  tibble::column_to_rownames('CCLE_ID') |>
  t()

cmat <- cmat[-which(duplicated(rownames(cmat))),]

gmat <- get_gsea_scores(cmat)
rownames(gmat) <- sapply(rownames(gmat), function(x) tolower(gsub("HALLMARK ","",gsub("_"," ",x))))

##get MOA for later
##is it worth it to incorporate drug class? 
gmt.drug <- leapR::read_gene_sets(synGet('syn74955107')$path)

gmt.drug$matrix <- t(gmt.drug$matrix)

moa.df <- do.call(rbind,
                  lapply(c(1:nrow(gmt.drug$matrix)), function(i){
  dl <- setdiff(gmt.drug$matrix[i,],c('',NULL,NA))
  data.frame(moa = rep(gmt.drug$names[i]),drug=dl)
}))

##correlation
prism <- read.csv(file = synGet('syn74955119')$path, check.names=FALSE) ##these are the names that map to moa

p_exp <- prism[,2:ncol(prism)] |>
  tidyr::pivot_longer(-CCLE_ID,names_to='drug',values_to='auc')|>
  dplyr::rename(improve_sample_id = 'CCLE_ID')|>
  mutate(treat_time = 120, expr_time = '0hr')

cor_res <- do.call(rbind,lapply(unique(p_exp$drug), function(drug){
      compute_cor(p_exp,gmat,drug,'0hr',120)
                        })) #e  

cor_res$drug <- sapply(cor_res$drug,function(x) gsub('-','.',x))

##and calculate significance
print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(p_exp$drug), function(drug){
      compute_sig(p_exp,gmat,drug,'0hr',120)
    })) #end drug iter

cor_sig$drug <- sapply(cor_sig$drug,function(x) gsub('-','.',x))


merged_res <- cor_res |> left_join(cor_sig) |>
  left_join(moa.df)

##enrichment
library(leapR)

cormat <- cor_res[,c('gene','drug','corval')] |> 
  pivot_wider(names_from='gene', values_from='corval') |>
  tibble::column_to_rownames('drug') |>
  as.matrix()


##now we can run a KS test on each pathway to see
##what drug mechanisms are more negatively correlated with each
##class
eres <- do.call(rbind,lapply(colnames(cormat),function(p){
    leapR(gmt.drug,'enrichment_in_order',
       eset=SummarizedExperiment::SummarizedExperiment(assays=list(cors=cormat)), 
       assay_name='cors',
       primary_columns=p) |>
    tibble::rownames_to_column('drugClass')|>
    mutate(pathway=p)}))

```

### Now plot results for mirdametininb only

```{R plot mirda}

combined <- eres |>
  dplyr::mutate(Pathway = gsub("HALLMARK_",'',pathway)) |>
  right_join(dplyr::rename(sig_paths,pathway='Feature_set')) |>
  subset(BH_pvalue < (0.01)) |>
  subset(ingroup_mean < (-1))

subset(combined,DrugTreatment=='Mirdametinib')|>
  ggplot(aes(x=sort_by(Pathway, ingroup_mean), y =ingroup_mean, fill = drugClass)) +
  geom_bar(position='dodge',stat='identity') +
  coord_flip() +
  #facet_grid(~DrugTreatment) +
 # scale_fill_manual(values =colorRampPalette(bs3)(length(unique(combined$drugClass))))+
  xlab('Pathway') + ylab('Mean correlation with AUC across class') +
  scale_fill_viridis_d()+
  theme_bw() + 
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))

ggsave('mir_cor_class.pdf',width=6, height=9)
```
