---
title: "global-phospho-proteomics-DiffExpression"
author: "Ruby Fore"
date: "2024-07-29"
output: html_document
---
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggfortify)
```


```{r}
synapser::synLogin()
#corrected global proteomics: syn61695507
# corrected phospho proteomics: 'syn61695519'
fileEntity <- synapser::synGet('syn61695507')
correctedProteomics <- read.table(fileEntity$path, header = T)

```


### impute for NAs  - change to row mean 


```{r}
# get the indices of NA's
na_indices <- which(is.na(correctedProteomics), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
correctedProteomics[na_indices] <- rowMeans(correctedProteomics, na.rm=T)[na_indices[,1]]
#na_indices
sum(is.na(correctedProteomics))
dim(correctedProteomics)

```

# prior outliers

```{r}
outliers <- c( "JH.2.002_DMSO_3_hr_rep3" , "JH.2.002_Vorinostat_3_hr_rep1", "MN.2_Mirdametinib_24_hr_rep2")
```

# reshape to long to do extra filtering and plot summary

Here I moved the data to long form to easily summarize, filter, and plot.

```{r remove outliers reshape}
library(dplyr)
#noOutliersCorrectedProteomics <- dplyr::select(correctedProteomics, -one_of(outliers))

samps<-colnames(correctedProteomics)
longProt<-correctedProteomics|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)

##quick summary of distributions 
ggplot(longProt,aes(x=condition,y=logRatio,fill=drug))+geom_violin()+theme(axis.text.x = element_text(angle = 90))


##what if we filter by logratio?
outliers<-subset(longProt,abs(logRatio)>5)|>
  group_by(condition)|>
  summarize(numOut=n())|>
  arrange(numOut)|>
  subset(numOut>1)
  
print(outliers)
```
I decided that maybe the samples with large LFCs are outliers and removed those. This doesn't always seem to be the case, some outliers just look like regular samples. 

```{r plot outliers}

##collect sample metadata, annotate our outliers
meta<-longProt|>
  dplyr::select(condition,sample,drug,time)|>
  distinct()|>
  mutate(isOutlier=condition%in%outliers$condition)|>
  tibble::column_to_rownames('condition')



pc<-prcomp(t(correctedProteomics))
autoplot(pc,data = meta,color='drug',shape='isOutlier')
autoplot(pc,data = meta,color='time',shape='isOutlier')+facet_grid(sample~drug)
```

Now we can see that these outliers are off and filter them out.

```{r remove outliers}
longProt<-longProt|>
  subset(!condition%in%outliers$condition)
meta<-meta|>
  subset(!isOutlier)

noOutliersCorrectedProteomics<-longProt|>
  dplyr::select(condition,gene,logRatio)|>
  tidyr::pivot_wider(names_from=condition,values_from=logRatio)|>
  tibble::column_to_rownames('gene')|>
  as.matrix()

pc2<-prcomp(t(noOutliersCorrectedProteomics))

autoplot(pc2,data = meta,color='drug',shape='sample')
autoplot(pc2,data = meta,color='drug',shape='sample')+facet_grid(~time)

```
Now the result is much more divided by sample, which we expect.

### Differential expression

Here is Ruby's funnction to do the differential expression

```{r}

# ##lets put all matrices we need in a list
# 
# DMSO0 <- dplyr::select(noOutliersCorrectedProteomics, contains('DMSO_0_'))
# DMSO0.5 <- dplyr::select(noOutliersCorrectedProteomics, contains('DMSO_0.5'))
# DMSO3 <- dplyr::select(noOutliersCorrectedProteomics, contains('DMSO_3'))
# DMSO24 <- dplyr::select(noOutliersCorrectedProteomics, contains('DMSO_24'))
# 
# Mirda0.5 <- dplyr::select(noOutliersCorrectedProteomics, contains('Mirdametinib_0.5'))
# Mirda3 <- dplyr::select(noOutliersCorrectedProteomics, contains('Mirdametinib_3'))
# Mirda24 <- dplyr::select(noOutliersCorrectedProteomics, contains('Mirdametinib_24'))
# 
# Vorino0.5 <- dplyr::select(noOutliersCorrectedProteomics, contains('Vorinostat_0.5'))
# Vorino3 <- dplyr::select(noOutliersCorrectedProteomics, contains('Vorinostat_3'))
# Vorino24 <- dplyr::select(noOutliersCorrectedProteomics, contains('Vorinostat_24'))
# 
# Selu0.5 <- dplyr::select(noOutliersCorrectedProteomics, contains('Selumetinib_0.5'))
# Tramet0.5 <- dplyr::select(noOutliersCorrectedProteomics, contains('Trametinib_0.5'))
# 




```

```{r}
# function to get N genes, adj.P.value < 0.05
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  

factorInfo <- data.frame(stringr::str_split_fixed(names(data), "_", 5))

names(factorInfo) <- c('sample', 'trt', 'time','hr', 'rep')
factorInfo$sample <- as.factor(factorInfo$sample)
factorInfo$trt <-as.factor(factorInfo$trt)

factorInfo$time <- as.factor(factorInfo$time)

# rownames have already been assigned
proSet <- Biobase::ExpressionSet(as.matrix(data))

 # log2 transformation if distribution isn't normal
 ex <- Biobase::exprs(proSet)
 qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                           na.rm = TRUE
 ))
 LogC <- (qx[5] > 100) ||
   (qx[6] - qx[1] > 50 && qx[2] > 0)
 if (LogC) {
   ex[which(ex <= 0)] <- NaN
   Biobase::exprs(proSet) <- log2(ex)
 }


proSet$sample <- factorInfo$sample
proSet$trt <- factorInfo$trt
proSet$time <- factorInfo$time
 #browser()
design <- stats::model.matrix(formula, proSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
#print(cont.matrix)
fit <- limma::lmFit(proSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)
#browser()
fit <- limma::eBayes(fit)
deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
return(list(deg, nDiffExpressed))
}
```

# Compare all time points and conditions

A loop through the data frame calculates all the conditions and adds them to a new data frame.


```{r}
##created for loop to calculate all diff ex stats
t0dat<-data.frame()
ddat<-data.frame()
for(d in c('DMSO','Mirdametinib','Selumetinib','Vorinostat')){
  for(tv in c('0.5','3','24')){
    for(s in c('JH.2.002','MN.2')){
      ##get matrix of values of sample at condition
      tmat<-longProt|>
        subset(sample==s)|>
        subset(drug==d)|>
        subset(time==tv)|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
      if(nrow(tmat)==0)##there is no data for this
        next
      ##matrix of values at DMSO 0
      cmat<-longProt|>
        subset(sample==s)|>
        subset(drug=='DMSO')|>
        subset(time=='0')|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
  
      if(d!='DMSO'){    
        #matrix of values at DMSO same time
        dmat<-longProt|>
          subset(sample==s)|>
          subset(drug=='DMSO')|>
          subset(time==tv)|>
          select(gene,condition,logRatio)|>
          distinct()|>
          tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
        ## get result compared to DMSO
        res<-getNdiffExpressedGenes(cbind(dmat,tmat), formula = ~ trt  + 0, analysisContrast =paste0('trt',d,'-trtDMSO')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
      }
           ##first get restult compare to t0
        res<-getNdiffExpressedGenes(cbind(cmat,tmat), formula = ~ time  + 0, analysisContrast =paste0('time',tv,'-time0')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        t0dat<-rbind(t0dat,tres)
      
    }
  }
}

t0counts = t0dat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(t0counts)
ggplot(t0counts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)



dcounts = ddat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(dcounts)
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)

#DMSOresult <- getNdiffExpressedGenes(cbind(DMSO0, DMSO0.5), formula = ~ time + sample + 0, analysisContrast ='time0.5-time0') 
#Mirdaresult <-getNdiffExpressedGenes(cbind(DMSO0, Mirda0.5), formula = ~ time + sample + 0,analysisContrast ='time0.5-time0')
#Seluresult <-getNdiffExpressedGenes(cbind(DMSO0, Selu0.5),formula = ~ time + sample + 0, analysisContrast ='time0.5-time0')
#Traresult <-getNdiffExpressedGenes(cbind(DMSO0, Tramet0.5), formula = ~ time + sample + 0,analysisContrast ='time0.5-time0') 
#Vorinoresult <-getNdiffExpressedGenes(cbind(DMSO0, Vorino0.5), formula = ~ time + sample + 0,analysisContrast ='time0.5-time0') 

#halfhour <- c("DMSO" = DMSOresult[[2]], "Mirdametinib" = Mirdaresult[[2]], "Selumetinib" = Seluresult[[2]], "Trametinib" = Traresult[[2]], "Vorinostat" = Vorinoresult[[2]])
```

### compare 3 hour to 0 timepoint, all drugs and DMSO


```{r}
#DMSOresult <- getNdiffExpressedGenes(cbind(DMSO0, DMSO3), formula = ~ time + sample + 0,analysisContrast ='time3-time0') #1
#Mirdaresult <-getNdiffExpressedGenes(cbind(DMSO0, Mirda3), formula = ~ time + sample + 0,analysisContrast ='time3-time0')  #309
#Vorinoresult <-getNdiffExpressedGenes(cbind(DMSO0, Vorino3), formula = ~ time + sample + 0, analysisContrast ='time3-time0')  #0

#threeHrs <- c("DMSO" = DMSOresult[[2]], "Mirdametinib" = Mirdaresult[[2]], "Vorinostat"= Vorinoresult[[2]])

```


### compare 24 hour to 0 timepoint, all drugs and DMSO


```{r}
#DMSOresult <- getNdiffExpressedGenes(cbind(DMSO0, DMSO24), formula = ~ time + sample + 0, analysisContrast ='time24-time0') #381
#Mirdaresult <-getNdiffExpressedGenes(cbind(DMSO0, Mirda24), formula = ~ time + sample + 0, analysisContrast ='time24-time0')  #939
#Vorinoresult <-getNdiffExpressedGenes(cbind(DMSO0, Vorino24), formula = ~ time + sample + 0, analysisContrast ='time24-time0') #219 
#twentyFourHrs <-c('DMSO' =DMSOresult[[2]], 'Mirdametinib'=Mirdaresult[[2]], "Vorinostat" =Vorinoresult[[2]])

```


```{r}

#plotData <- data.frame('diffexprCounts' = c(halfhour, threeHrs, twentyFourHrs), time = c(rep(0.5 ,5), rep(3,3), rep(24,3)), drug = c("DMSO", "Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat", rep(c("DMSO", "Mirdametinib", "Vorinostat"), 2)))

#ggplot2::ggplot(plotData, aes(x=factor(time, level=c('0.5', '3', '24')), fill=drug, y=diffexprCounts))+ 
#   geom_bar(position='dodge', stat='identity')+
#   geom_text(aes(label=diffexprCounts), position=position_dodge(width=0.9), vjust=-0.25)+ 
#   ggtitle('Differentially expressed phospho-sites compared to DMSO at 0 timepoint')+
#   xlab("Timepoint, in hours")+
#   ylab("# of Differentially Expressed Proteins")+
#   theme_bw()
# 
# 
# ggplot2::ggplot(plotData, aes(x=drug, fill=drug, y=diffexprCounts))+ 
#   geom_col() + 
#   facet_grid(~time) + 
#   geom_text(aes(label=diffexprCounts), position=position_dodge(width=0.9), vjust=-0.25) +
#   ggtitle('Differentially expressed phospho-sites compared to DMSO at 0 timepoint')+
#   #xlab("Timepoint, in hours")+
#   ylab("# of Differentially Expressed Proteins")+
#   theme_bw()+
#   theme(axis.text.x = element_text(angle = 90)) 
```

## Compare all drugs individually at timepoint 1/2 hour to DMSO at 1/2 hour

```{r}
# Mirdaresult <- getNdiffExpressedGenes(cbind(Mirda0.5, DMSO0.5), formula = ~ trt + sample + 0, analysisContrast ='trtMirdametinib-trtDMSO') #53
# Seluresult <- getNdiffExpressedGenes(cbind(Selu0.5, DMSO0.5), formula = ~ trt + sample + 0, analysisContrast ='trtSelumetinib-trtDMSO') #5 
# Trametresult <- getNdiffExpressedGenes(cbind(Tramet0.5, DMSO0.5), formula = ~ trt + sample + 0, analysisContrast ='trtTrametinib-trtDMSO') #15 
# Vorinoresult <- getNdiffExpressedGenes(cbind(Vorino0.5, DMSO0.5), formula = ~ trt + sample + 0, analysisContrast ='trtVorinostat-trtDMSO') #0
# 
# halfhour <- c("Mirdametinib" = Mirdaresult[[2]], "Selumetinib" = Seluresult[[2]], "Trametinib" = Traresult[[2]], "Vorinostat" = Vorinoresult[[2]])
```


## and at 3 hours
```{r}
# Mirdaresult <- getNdiffExpressedGenes(cbind(Mirda3, DMSO3), formula = ~ trt + sample + 0, analysisContrast ='trtMirdametinib-trtDMSO') #53
# 
# Vorinoresult <- getNdiffExpressedGenes(cbind(Vorino3, DMSO3), formula = ~ trt + sample + 0, analysisContrast ='trtVorinostat-trtDMSO') #0
# 
# threeHrs <- c("Mirdametinib" = Mirdaresult[[2]], "Vorinostat"= Vorinoresult[[2]])
```

### and at 24 hours
```{r}
# Mirdaresult <- getNdiffExpressedGenes(cbind(Mirda24, DMSO24), formula = ~ trt + sample + 0, analysisContrast ='trtMirdametinib-trtDMSO') #53
# 
# Vorinoresult <- getNdiffExpressedGenes(cbind(Vorino24, DMSO24), formula = ~ trt + sample + 0, analysisContrast ='trtVorinostat-trtDMSO') #0
# 
# twentyfourHrs <- c("Mirdametinib" = Mirdaresult[[2]], "Vorinostat"= Vorinoresult[[2]])

```

```{r}
# plotData <- data.frame('diffexprCounts' = c(halfhour, threeHrs, twentyfourHrs), time = c(rep(0.5 ,4), rep(3,2), rep(24,2)), drug = c( "Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat", rep(c( "Mirdametinib", "Vorinostat"), 2)))
# 
# ggplot2::ggplot(plotData, aes(x=drug, fill=drug, y=diffexprCounts))+ 
#   geom_col() + 
#   facet_grid(~time) + 
#   geom_text(aes(label=diffexprCounts), position=position_dodge(width=0.9), vjust=-0.25) +
#   
#   ggtitle('Differentially expressed proteins comparing drug to DMSO at each timepoint')+
#   #xlab("Timepoint, in hours")+
#   ylab("# of Differentially Expressed Proteins")+
#   theme_bw()+
#   theme(axis.text.x = element_text(angle = 90))
# 
# ggplot2::ggplot(plotData, aes(x=factor(time, level=c('0.5', '3', '24')), fill=drug, y=diffexprCounts))+ 
#   geom_bar(position='dodge', stat='identity')+
#   geom_text(aes(label=diffexprCounts), position=position_dodge(width=0.9), vjust=-0.25)+ 
#   ggtitle('Differentially expressed proteins comparing drug to DMSO at each timepoint')+
#   xlab("Timepoint, in hours")+
#   ylab("# of Differentially Expressed Proteins")+
#   theme_bw()

```
