---
title: "AUC Exploration"
author: "Sara Gosline"
date: "2026-02-28"
output: html_document
---

Here we explore the AUC and AUC shift values from the drug sensitivity data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../RNAseq/plottingParameters.R")
source("correlation_functions.R")


```

## Single drug properties

First we collect the baseline drug sensitivity measurements for single agents
using the combination data. 

```{r single drug data}

##averge ACUS
sing_exp <- sing_exp |>
  group_by(common_name, improve_sample_id,other_id,drug,treat_time) |>
  summarize(auc = mean(auc)) |>
  ungroup() |>
  subset(!is.na(auc)) |>
  mutate(treat_time = as.numeric(as.character(treat_time)))

```

We can revisit the distribution of AUC scores to see how varied they are
across samples. 
```{r AUC distribution}

ggplot(sing_exp, aes(x=common_name, fill = as.factor(treat_time)))+ 
  geom_bar(position = 'dodge') + ggtitle('Number of drugs per sample')


ggplot(sing_exp, aes(x = drug, fill = as.factor(treat_time)))+ 
  geom_bar(position = 'dodge') + ggtitle('Number of samples per drug')+
   theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) 

###how is the AUC distributed across samples
subset(sing_exp, !is.na(drug)) |>
  ggplot(aes(x = auc,group = drug, 
                  col = drug)) +
           scale_color_manual(values = drugCol) +
  geom_density() +
   theme_bw() + facet_grid(~treat_time)

ggplot(sing_exp, aes(x=drug, y=auc, col = drug)) +
    geom_boxplot()+
         geom_jitter() + 
         scale_color_manual(values = drugCol) +
  facet_grid(~treat_time)+ 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(x='Drug',y='AUC')

ggsave('aucs_by_drug.pdf')

###how is the AUC distributed across samples
subset(sing_exp, !is.na(drug)) |>
  ggplot(aes(x = auc,group = drug, 
                  fill = drug)) +
           scale_fill_manual(values = drugCol) +
    geom_histogram(position = 'dodge') +#position='dodge') +
     theme_bw()+ facet_grid(~treat_time)

```


now that we have the gene, sample, and drug names and metadata we can collect the 
drug response data



## Combination drug data

For the combination data we are looking at effects of secondary drugs, or 
the AUC shift, which is the _difference_ between the AUC of a second drug
after the addition of a first. Keep in mind it is NOT symmetric and
varies based on the impact of the first drug. 

First we can ask what drugs seem to have a positive effect on tumor reduction 
regardless of initial drug treatment. 
```{r plot data}
##now we can plot the AUC shift values
#names(drugCol)<- tolower(names(drugCol))

####how is the AUC shift distributed across samples
subset(exp, !is.na(drug)) |>
ggplot(aes(x=delta_auc,group = drug, 
                col = drug)) +
         scale_color_manual(values=drugCol)+
  geom_density()+facet_grid(treat_time  ~ .) +
   theme_bw()

subset(exp, !is.na(drug)) |>
  ggplot(aes(x=delta_auc,group = drug, 
                fill = drug)) +
         scale_fill_manual(values=drugCol) +
  geom_histogram(position='dodge', binwidth=.1) + facet_grid(treat_time  ~ .) +
   theme_bw()


```


Here we can see that olaparib almost seems to reduce single drug efficacy,
while trabectedin improves it at 48 hrs

### AUC shift by sample
We can further break this down by sample to see
which samples show the largest shift in AUC with a second drug. 
```{r AUC shift by sample}

subset(exp, !is.na(drug))|> ##get rid of drugs that are not in the final set
  ggplot(aes(x=delta_auc,group = drug, 
                col = drug)) +
         scale_color_manual(values=classCol)+
  geom_density()+facet_grid(treat_time ~ common_name) +
   theme_bw() 

subset(exp, !is.na(drug))|> ##get rid of drugs that are not in the final set
  ggplot(aes(x=delta_auc,group = drug, 
                fill = drug)) +
         scale_fill_manual(values=drugCol)+
  geom_histogram(position='dodge')+facet_grid(treat_time ~ common_name) +
   theme_bw()


```



This shows differing behavior by sample. we can also see
if the AUC shift is different based on initial drug, which 
is divided up along the columns below. 

### AUC shift by drug

```{r AUC shift by drug}

# subset(exp, !is.na(drug))|>
# ggplot(aes(x=delta_auc,group = drug, 
#                 col = drug)) +
#          scale_color_manual(values=drugCol)+
#   geom_density()+facet_grid(treat_time  ~ initial_drug) +
#     theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) +
#    theme_bw()

subset(exp, !is.na(drug))|>
ggplot(aes(x=drug, y=delta_auc, col = drug)) +
    geom_boxplot()+
         geom_jitter() + 
         scale_color_manual(values = drugCol) +
  facet_grid(~treat_time)+ 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(x='Drug',y='AUC shift')

ggsave('auc_shift_by_drug.pdf')

subset(exp, !is.na(drug)) |>
  subset(treat_time ==48) |>
ggplot(aes(x=initial_drug,y=drug)) +
         scale_fill_viridis_c()+
  geom_tile(aes(fill=delta_auc))+facet_grid(common_name ~ .) +
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) +
   theme_minimal()

subset(exp, !is.na(drug)) |>
  subset(treat_time == 120) |>
ggplot(aes(x = initial_drug,y = drug)) +
         scale_fill_viridis_c() +
  geom_tile(aes(fill = delta_auc)) + facet_grid(common_name ~ .) +
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) +
   theme_minimal()
```


### How do mirda + vorin look?

THen we can look direclty at the combination of mirda and vorinostat
```{r dauc mirda vorin}
subset(exp, !is.na(drug))|>
  subset(initial_drug%in%c('mirdametinib','vorinostat'))|>
    subset(drug%in%c('mirdametinib','vorinostat'))|>
ggplot(aes(x=initial_drug,y=drug)) +
         #scale__manual(values=drugCol)+
  geom_tile(aes(fill=delta_auc))+facet_grid(treat_time  ~ common_name) +
   theme_bw()

subset(exp, !is.na(drug)) |> subset(delta_auc>0)|>group_by(initial_drug,drug)|>summarize(number=n())

```


We have the files, now combine them in a way that can be used for modeling

### Compare mean AUC shift
Do the mean values agree with what we saw with the paper calculation?

```{r dauc}
md <- exp|>
  subset(!is.na(drug))|>
           group_by(drug,treat_time)|>
  summarize(meanDAUC = mean(delta_auc),  count = n_distinct(delta_auc))

ggplot(md,aes(x=drug,y=meanDAUC,fill=count)) +
  geom_bar(stat='identity')+facet_grid(~treat_time) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```
