---
title: "Linear Regression tests"
author: "Sara Gosline"
date: "2026-02-10"
output: html_document
---


# Regression analysis of drug response

Here we propose to use an elastic net regression model to predict drug response of treated samples.


## Get data from single agent drugs and baseline gene expression
We first collect the transcriptomic and drug response data from matched samples. It is important to note
that each model is built for a single drug, and there is not a lot of variability across drugs, the 
model is much more sensitive to the sample itself. 

```{r setup, include=FALSE}

library(ggplot2)
library(tidyr)
library(dplyr)

library(synapser)
synLogin()
source('../RNAseq/plottingParameters.R')
source('correlation_functions.R')

names(classCol) <- tolower(names(classCol))
if(!exists('full_tmat'))
  full_tmat <- get_harmonized_gex()

tmat <- full_tmat[,intersect(colnames(full_tmat), sing_exp$improve_sample_id)]
```

View AUC counts and see if we can filter for those drugs for which we have the
most data. 

```{r aucs}
ggplot(sing_exp, aes(x=common_name, y=auc, col = drug)) +
         geom_jitter() + 
         scale_color_manual(values = classCol) +
  facet_grid(~treat_time)+ 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 

ggplot(sing_exp, aes(x=drug, y=auc, col = drug)) +
    geom_boxplot()+
         geom_jitter() + 
         scale_color_manual(values = classCol) +
  facet_grid(~treat_time)+ 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 


ggplot(sing_exp, aes(x = drug, fill = as.factor(treat_time)))+ 
  geom_bar(position = 'dodge') + ggtitle('Number of samples per drug')+
   theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) 
ggsave('sample_counts.pdf')

scounts <- sing_exp|>
  group_by(treat_time,drug) |> 
  summarize(numSamp= n()) |> arrange(numSamp)

print(scounts)

tcounts <- scounts |> subset(numSamp > 5)

```

Obviously we have more single agent data for the drugs we analyze in combination
so maybe we should just use that. 

However we can still try to identify features that predict AUC.

## Build elastic net regression using leave one out cross validation

Our sample size is very small, so we restrict ourselves to those drugs with at
least 4 gene expression + drug response measurements (so we train on 3 and evaluate
on 1). 

```{r single drug regression}

library(glmnet)
#drug = 'mirdametinib'

alpha = 0.2
sig = 0.1

rres <- do.call(rbind, lapply(c(48,120), function(ttime){ #build model for each treatment time
  
  ##build model for each drug in our top counts list
  res <- do.call(rbind, lapply(unique(tcounts$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                  # sing_exp$expr_time == etime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  #df$delta_auc = as.numeric(tvec$delta_auc)

##here we do the LOO for each sample in the drug
  samp_preds <- do.call(rbind,lapply(rownames(df), function(samp){

    st <- setdiff(rownames(df), samp)
    
    dtrain <- as.matrix(df[st,])
  
    model <- glmnet(x = dtrain, y = tvec[st, 'auc'], alpha = alpha)
    coefs <-  setdiff(unique(rownames(which(coef(model) != 0, arr.ind =T))),'(Intercept)')
    preds <- predict(model, newx = df[samp,], s = sig) 
    return(c(preds = preds, features = paste(coefs, collapse=','), training_set = length(st)))
  }
))
  tvec$preds = unlist(samp_preds[,1])
  tvec$features = unlist(samp_preds[,2])
  tvec$training_set = unlist(samp_preds[,3])
  return(tvec)
  }))
 return(res) 
}))

rres$error = (as.numeric(rres$preds)-rres$auc)^2
```

### Model performance
The results show minimal error at 48 hours, with greater error at 120 hrs.

```{r pred results}


##get ranking to make plotting prettier
rs <-  rres |> group_by(drug, .add=TRUE, .drop=FALSE) |> 
  summarize(mse = mean(error)) |>
   arrange(mse)

#join ranking
rres <- rres |> left_join(rs)

##plot
ggplot(rres, aes(x=reorder(drug, mse), y = error, col = drug)) + 
  geom_boxplot() +
  geom_jitter() + 
  facet_grid(~treat_time)+
    scale_color_manual(values=drugCol) + 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 

#subset(rres,drug != 'iag933') 
rres |>
ggplot(aes(x=reorder(drug, mse), y = error, col = drug)) + 
  geom_boxplot() +
  geom_jitter() + 
  facet_grid(~treat_time)+
    scale_color_manual(values=drugCol) + 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(x='Drug', y='Cross validation error')

ggsave('fig3a_cv_error.pdf')
```

### Model features
Now we can investigate the features used to predict by building a model on all
the data.
```{r full model}

ttime = 48
early_models <- lapply(unique(tcounts$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  model <- glmnet(x = df, y = tvec$auc, alpha = alpha)
  return(model)
})

names(early_models) <- unique(tcounts$drug)

ttime = 120
late_models <- lapply(unique(tcounts$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  model <- glmnet(x = df, y = tvec$auc, alpha = alpha)
  return(model)
})

names(late_models) <- unique(tcounts$drug)

##now for each model we can visualize the genes
early_feats <- do.call(rbind,lapply(names(early_models), function(nmod){
  mod <- early_models[[nmod]]
  genes = setdiff(unique(rownames(which(coef(mod) != 0, arr.ind=T))),'(Intercept)')
  return(c(drug = nmod, features = paste(genes, collapse=',')))
})) |>
  as.data.frame() |>
  tidyr::separate_longer_delim(features, delim = ',')

late_feats <- do.call(rbind,lapply(names(late_models), function(nmod){
  mod <- late_models[[nmod]]
  genes = setdiff(unique(rownames(which(coef(mod) != 0, arr.ind=T))),'(Intercept)')
  return(c(drug = nmod, features = paste(genes, collapse=',')))
})) |>
  as.data.frame() |>
  tidyr::separate_longer_delim(features, delim = ',') 

feats <- rbind(mutate(late_feats, treat_time = 120), mutate(early_feats, treat_time = 48))

##we can also predict comvbination effects
feats |> group_by(drug, treat_time) |> summarize(numFeatures = n())
```
We can ask if there are specific pathways that are useful to predict drug response


```{r pathway enrichment}

library(leapR)
#ddata <- sing_exp |> tidyr::pivot_wider(names_from= drug, values_from=auc) |> tibble::column_to_rownames('improve_sample_id')
cdata <- feats |> 
  mutate(model_feature=TRUE) |> 
  tidyr::pivot_wider(names_from=drug, values_from=model_feature, values_fill=FALSE) 

#odata <- data.frame(features = setdiff(rownames(tmat), feats$features), |> tibble::column_to_rownames('features')
#exp_set <- SummarizedExperiment(assays=SimpleList(tmat), rowData=cdata, colData = ddata)


```


## Prediction of secondary effects from single agent treatment

Now we want to use single agent models to predict combo


### Combo gex data

```{r combo gex data}

ttmat <- full_tmat[,intersect(colnames(full_tmat), exp$improve_sample_id)]

```

### Combo model evaluation


Now we can try to use these values to predict AUC shift. Keep in mind that
the model predicts AUC, and the AUC shift is between -1 and 1,with 1 meaning
MORE drug sensitivity. 


```{r applying single drug models}

exp <- exp |> subset(!is.na(drug))



early_cpreds <- do.call(rbind, lapply(names(early_models), function(d){
  mod <- early_models[[d]]
  #feats <- intersect(rownames(mod$beta),rownames(stmat))
  pval <- predict(mod, newx = t(ttmat), s = sig)#[feats,])
  
  return(data.frame(drug = d, improve_sample_id = as.numeric(colnames(ttmat)), pred = pval[,1], pred_time = 48))
}) )

late_cpreds <- do.call(rbind, lapply(names(late_models), function(d){
  mod <- late_models[[d]]
  #feats <- intersect(rownames(mod$beta),rownames(stmat))
  pval <- predict(mod, newx = t(ttmat), s = sig)#[feats,])
  
  return(data.frame(drug = d, improve_sample_id = as.numeric(colnames(ttmat)), pred = pval[,1], pred_time = 120))
}) )


dpreds <- rbind(late_cpreds, early_cpreds) |>  left_join(samples) |>
  tidyr::separate(other_id, into=c('samp','initial_drug','time','treat','type'),sep='_') |>
  dplyr::select(drug,improve_sample_id, samp, initial_drug,time,treat, pred, pred_time ) |>
  distinct() |>
  mutate(initial_drug = tolower(initial_drug))

```

We can now plot the results by drug, the time at which they were predicted to have
an effect, and the expression measurement used to predict. 


```{r plot prediction results}
ggplot(dpreds, aes(x = pred, fill = drug)) + 
  geom_histogram(position = 'dodge') +
  scale_fill_manual(values=classCol) + facet_grid(~pred_time)


ggplot(dpreds, aes(x = drug, y = pred, col = initial_drug, shape = time)) + 
  geom_jitter() +
  scale_color_manual(values=classCol) +
  facet_grid( ~ pred_time) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(y='Predicted effect')
ggsave('fig2d_predictedEffect.pdf')

```

The predictions at 8hr and 24hr seem nearly identical.  Can we compare them?

```{r compare 24h and 8hr}

dpreds |> dplyr::select(-improve_sample_id)|>
  pivot_wider(names_from=time, values_from = pred) |>
  mutate(pred_diff = abs(`24hr`-`8hr`)) |>
  ggplot(aes(x = drug, y = pred_diff, col = drug)) + 
  geom_jitter() +
  scale_color_manual(values=classCol) +
  facet_grid( ~ pred_time) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(x='Drug')

```

So there is a difference in prediction! Seems to refect variability in experession result
with MEK inhibitors showing more variability.

### Top combos predicted
```{r sample combos}

dpreds$combination <- apply(dpreds, 1, function(x) paste(sort(c(x[['drug']],x[['initial_drug']])), collapse='+'))

dpreds$oppPred = 1-dpreds$pred


epreds <- subset(dpreds, pred_time == 48)
cmeans <- epreds |> group_by(combination, comboMean = mean(oppPred))

subset(epreds, initial_drug != drug) |>
  left_join(cmeans) |>
  subset(time = '8hr') |>
  ggplot(aes(x = samp, y = reorder(combination,comboMean), fill = oppPred)) + 
  geom_tile() +#facet_grid( ~ pred_time) + 
  labs(y='Combination') + scale_fill_continuous(high='turquoise4', low ='orchid4')


subset(epreds, initial_drug != drug) |>
  subset(initial_drug %in% exp$initial_drug & drug %in% exp$drug)|>
  left_join(cmeans) |>
  subset(time = '8hr') |>
  ggplot(aes(x = samp, y = reorder(combination,comboMean), fill = oppPred)) + 
  geom_tile() +#facet_grid( ~ pred_time) + 
  labs(y='Combination') + scale_fill_continuous(high='turquoise4', low ='orchid4')

ccounts <- epreds |> group_by(combination, pred_time, time) |> 
  summarize(meanAUC = mean(pred), numSamps = n()) |>
  tidyr::separate(combination, remove = FALSE, into=c('drug1','drug2'), sep =',')

pcounts <- epreds |> group_by(initial_drug, drug, time, pred_time) |>
  summarize(meanAUC = median(pred), numSamps = n())

mpcounts <-pcounts |> group_by(drug) |> summarize(mma =mean(meanAUC))

pcounts|> 
  left_join(mpcounts) |># subset(drug1 != drug2) |>
  #subset(numSamps > 2) |>
#  subset(pred_time == 120) |>
  ggplot(aes(x=initial_drug, y=reorder(drug,mma))) +
         scale_fill_viridis_c()+
  geom_tile(aes(fill=meanAUC))+facet_grid( ~ time) +
   theme_minimal()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 
```
These just all look the same, probably won't plot these.

###Best predictions for mirda, RMC38, and vorinostat and sn38
These four drugs seemed to have the greatest effect at the single agent level,

```{r selected preds}
names(drugCol) = tolower(names(drugCol))
dpreds |>
  subset(initial_drug %in% c('sn-38','mirdametinib','vorinostat','doxorubicin','tno155','capmatinib'))|>
  ggplot(aes(x=initial_drug, y = pred, col = drug)) + 
  geom_boxplot() + 
  #geom_jitter(aes(shape = samp)) +
  scale_color_manual(values = drugCol) 

```

### Comparing predictions to delta_AUC

```{r normalized predictions}


dpreds <- dpreds |> 
  mutate(oppPred = 1-pred) #|>
#  group_by(drug, .drop = FALSE) |> mutate(predRank = 1-rank(pred)/length(pred)) |>
#  ungroup() |>
#  group_by(improve_sample_id, .drop=FALSE) |> mutate(sampleRank = 1-rank(pred)/length(pred))


##rank d_auc by same metr
#exp <- exp |>
#  group_by(initial_drug, .drop = FALSE) |> mutate(dAUCRank = rank(delta_auc)/length(delta_auc))

```



```{r compare with experiment}

dat_with_pred <- dpreds |>
  dplyr::select(improve_sample_id, pred_time, drug, oppPred,pred) |>
  right_join(exp) |>
 # mutate(pd_auc = 1 - as.numeric(pred)) |> #this gets us in the right direction
 # mutate(rankErr = (predRank - dAUCRank)^2) |>
  mutate(error = abs(oppPred - delta_auc))
#|> ##still not quite right
#  subset(!is.na(pred))

dat_with_pred |>
  subset(pred_time == treat_time) |>
  ggplot(aes(x = drug, y = error,col = drug)) + geom_jitter() + 
    scale_color_manual(values = classCol) + 
    geom_boxplot() +
  facet_grid(expr_time ~ treat_time)+
  theme_bw() +
   theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 

ggsave('fig4c_secondaryDrugError.pdf')

dat_with_pred |>
  subset(treat_time == pred_time) |>
ggplot( aes(x = common_name, y = error, col = drug)) + 
  geom_jitter() + 
    scale_color_manual(values=classCol) + 
  facet_grid(~pred_time)+
  theme_bw()

ggsave('suppfig_secondaryDrugErrorBySample.pdf')

dat_with_pred |>
  subset(treat_time == pred_time) |>
ggplot( aes(x = delta_auc, y = log(pred), col = drug)) + 
  geom_point() + 
    scale_color_manual(values=classCol) + 
  facet_grid(expr_time~pred_time)+
  theme_bw()

```

