---
title: "Evaluate second drug pathways"
author: "Sara Gosline"
date: "2026-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("correlation_functions.R")
source("../RNAseq/plottingParameters.R")

```

## Data

We have reformatted the data into three files:
1. sample file describing the samples
2. gene file with gene metadata
3. transcriptomics data with gene expression data
4. combination data with delta AUC values
5. drug data with single AUC values

Most data is now read in by `correlation_functions.R`

```{r data, message=FALSE}

##get the transcript data, this is slow
full_tmat <- get_harmonized_gex()

```

now that we have the gene, sample, and drug names and metadata we can collect the 
drug response data


## Single drug properties

First we collect the baseline gene expression measurements. 
__TODO__: this is old data, we need the updated values with DMSO treated
samples.

```{r single drug data}

##averge ACUS
sing_exp <- sing_exp |>
  group_by(common_name, improve_sample_id,other_id,drug,treat_time) |>
  summarize(auc = mean(auc)) |>
  ungroup() |>
  subset(!is.na(auc)) |>
  mutate(treat_time = as.numeric(as.character(treat_time)))

```

We can revisit the distribution of AUC scores to see how varied they are
across samples. 
```{r AUC distribution}
drugnames <-  unique(ddrugs$chem_name)
drugCol <- colorRampPalette(RColorBrewer::brewer.pal(length(unique(drugnames)),"Paired"))(length(drugnames)) # need color ramp for >12 colors
names(drugCol) <- drugnames

###how is the AUC distributed across samples
subset(sing_exp, !is.na(drug)) |>
  ggplot(aes(x = auc,group = drug, 
                  col = drug)) +
           scale_color_manual(values = drugCol) +
  geom_density() +
   theme_bw()


###how is the AUC distributed across samples
subset(sing_exp, !is.na(drug)) |>
  ggplot(aes(x = auc,group = drug, 
                  fill = drug)) +
           scale_fill_manual(values = drugCol) +
    geom_histogram() +#position='dodge') +
     theme_bw()

```

## Combination drug data

For the combination data we are looking at effects of secondary drugs, or the
delta AUC. We first plot the values then look at specifics

First we can ask what drugs seem to have a positive effect on tumor reduction 
regardless of initial drug treatment. 
```{r plot data}
##now we can plot the delta AUC values
#names(drugCol)<- tolower(names(drugCol))
#drugnames <-  unique(exp$drug)
drugCol <- classCol <- colorRampPalette(RColorBrewer::brewer.pal(length(unique(drugnames)),"Paired"))(length(drugnames)) # need color ramp for >12 colors
names(drugCol)<-drugnames

####how is the delta AUC distributed across samples
subset(exp, !is.na(drug)) |>
ggplot(aes(x=delta_auc,group = drug, 
                col = drug)) +
         scale_color_manual(values=drugCol)+
  geom_density()+facet_grid(treat_time  ~ .) +
   theme_bw()

subset(exp, !is.na(drug)) |>
  ggplot(aes(x=delta_auc,group = drug, 
                fill = drug)) +
         scale_fill_manual(values=drugCol) +
  geom_histogram(position='dodge', binwidth=.1) + facet_grid(treat_time  ~ .) +
   theme_bw()


```

Here we can see that olaparib almost seems to reduce single drug efficacy,
while trabectedin improves it at 48 hrs

### Delta AUC by sample
We can further break this down by sample
```{r delta auc by sample}

subset(exp, !is.na(drug))|> ##get rid of drugs that are not in the final set
  ggplot(aes(x=delta_auc,group = drug, 
                col = drug)) +
         scale_color_manual(values=drugCol)+
  geom_density()+facet_grid(treat_time ~ common_name) +
   theme_bw()

subset(exp, !is.na(drug))|> ##get rid of drugs that are not in the final set
  ggplot(aes(x=delta_auc,group = drug, 
                fill = drug)) +
         scale_fill_manual(values=drugCol)+
  geom_histogram(position='dodge')+facet_grid(treat_time ~ common_name) +
   theme_bw()


```


this shows differing behavior by sample. we can also see
if the delta AUC is different based on initial drug, which 
is divided up along the columns below. 

### Delta auc by drug

```{r delta auc by drug}

subset(exp, !is.na(drug))|>
ggplot(aes(x=delta_auc,group = drug, 
                col = drug)) +
         scale_color_manual(values=drugCol)+
  geom_density()+facet_grid(treat_time  ~ initial_drug) +
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) +
   theme_bw()

subset(exp, !is.na(drug)) |>
  subset(treat_time ==48) |>
ggplot(aes(x=initial_drug,y=drug)) +
         scale_fill_viridis_c()+
  geom_tile(aes(fill=delta_auc))+facet_grid(common_name ~ .) +
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) +
   theme_minimal()

subset(exp, !is.na(drug)) |>
  subset(treat_time == 120) |>
ggplot(aes(x = initial_drug,y = drug)) +
         scale_fill_viridis_c() +
  geom_tile(aes(fill = delta_auc)) + facet_grid(common_name ~ .) +
    theme(axis.text.x = element_text(angle = 90, vjust=1, hjust=1)) +
   theme_minimal()
```

### How do mirda + vorin look?

THen we can look direclty at the combination of mirda and vorinostat
```{r dauc mirda vorin}
subset(exp, !is.na(drug))|>
  subset(initial_drug%in%c('mirdametinib','vorinostat'))|>
    subset(drug%in%c('mirdametinib','vorinostat'))|>
ggplot(aes(x=initial_drug,y=drug)) +
         #scale__manual(values=drugCol)+
  geom_tile(aes(fill=delta_auc))+facet_grid(treat_time  ~ common_name) +
   theme_bw()

subset(exp, !is.na(drug)) |> subset(delta_auc>0)|>group_by(initial_drug,drug)|>summarize(number=n())

```

We have the files, now combine them in a way that can be used for modeling
### Compare mean delta auc
Do the mean values agree with what we saw with the paper calculation?

```{r dauc}
md <- exp|>
  subset(!is.na(drug))|>
           group_by(drug,treat_time)|>
  summarize(meanDAUC = mean(delta_auc),  count = n_distinct(delta_auc))

ggplot(md,aes(x=drug,y=meanDAUC,fill=count)) +
  geom_bar(stat='identity')+facet_grid(~treat_time) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


# Evaluate correlations

Here we want to look at correlations between baseline gene expression and drug
response.

### Correlation of AUC and gene expression

Let's first look at the untreated gene expression data and see if we can 
find out features that predict AUC

```{r untreated gene expression}


tmat <-   full_tmat[,intersect(colnames(full_tmat), sing_exp$improve_sample_id)]

print('computing gene/aUC correlation')

#first focus on only durgs with 10 data points
#gvals <- md|>subset(count>10)

sing_exp <- sing_exp |> mutate(expr_time='0hr')

cor_res <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_cor(sing_exp,tmat,drug,'0hr',48))
    })) #end drug iter


cor_res |>
  ggplot(aes(x=corval,col=drug)) + geom_density() +
  scale_color_manual(values=drugCol) + 
  theme_bw()

```

We only have 8 samples here, but across drugs there is an enrichment of genes
that are negatively correlated with AUC, which means they are high when the drug
is MORE effective

### Correlation with known targets

We have a list of known drug targets, we can see if higher expression of those targets
is correlated with AUC

```{r cor targets}
drug.target.info <- read.csv("https://ndownloader.figshare.com/files/20237712") |>
  mutate(name = tolower(name))

sel.targs <- drug.target.info[drug.target.info$name %in% sing_exp$drug,c('name','moa','target')] |>
  unique() |>
  rbind(c('mirdametinib','MEK inhibitor','MAP2K1, MAP2K2')) |>#MEK
  rbind(c('tno155','SHP2 inhibitor','PTPN11')) |>
  rbind(c('trabectedin','','')) |>
  tidyr::separate_longer_delim(target,delim = ', ')


corplots <- lapply(unique(sing_exp$drug), function(d){
  corv <- subset(cor_res, drug == d)
  targs = subset(sel.targs, name == d)
  
  corv$target <- 'Not a target'
  corv$target[which(corv$gene %in%targs$gene)] <- 'Drug target'
  ggplot(corv, aes(x=corval, col=target)) + geom_density() + ggtitle(paste0(d, ' target correlation'))
  
})
#subset(cor_res,gene%in%sel.targs$target)|>View()

```

### Other promising correlated features



Now that we have the data we can look at its properties. 

### Treated gene expression data

We first read in the transcriptomics data that matches the combination experiments
```{r treated exp, echo=FALSE, warning=FALSE, message=FALSE}
##read in transcritomics, focus on gene name and average counts acros
##gene isoforms

tmat <- full_tmat[,intersect(colnames(full_tmat), exp$improve_sample_id)]


```

Significance code doesn't work yet
```{r cor sig, eval=FALSE}
print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(gvals$drug), function(drug){
      rbind(compute_sig(exp,tmat,drug,'8hr','48'),
            compute_sig(exp,tmat,drug,'24hr','48'),
            compute_sig(exp,tmat,drug,'8hr','120'),
            compute_sig(exp,tmat,drug,'24hr','120'))
    })) #end drug iter


```

Now we can summarize and calculate how many genes are significantly correlated 
by drug, time point. 

```{r number of genes}

cor_res <- cor_res |>
  mutate(dir=ifelse(corval<0,'neg','pos'))

##here is the count of correlates, based on 0.7 for the moment
cor_counts <- cor_res |>
  subset(abs(corval)>0.7) |> 
  group_by(drug,ttime,etime)|>
  summarize(gene_cor = n_distinct(gene))|>
  arrange(desc(gene_cor))

##let's look at these
ggplot(cor_counts,aes(x=drug,y=gene_cor,fill=etime)) +
         geom_bar(stat='identity',position='dodge') +
  facet_grid(ttime~.)

dir_counts <- cor_res |>
  subset(abs(corval) > 0.7) |> 
  group_by(drug,dir,ttime,etime) |>
  summarize(gene_cor = n_distinct(gene)) |>
  mutate(cor_genes = ifelse(dir == 'pos',gene_cor,-1*gene_cor)) |>
  arrange(desc(cor_genes))

##let's look at these
ggplot(dir_counts,aes(x = drug,y = cor_genes,fill = etime)) +
         geom_bar(stat='identity',position='dodge') +
  facet_grid(ttime~.)

```

## Pathway enrichment

Now we can look at which pathways are correlated with drug efficacy

```{r pathway, warning=FALSE, message=FALSE}
library(leapR)
library(SummarizedExperiment)
###move all data into expression set
#first format exp data

cor_mat <- cor_res |>
  dplyr::select(-dir)|>
  tidyr::unite(drug,ttime,etime,sep='_',col='condition') |>
  pivot_wider(names_from=condition,values_from=corval)|>
  tibble::column_to_rownames('gene')

df <- exp |> 
  subset(!is.na(drug)) |>
  tidyr::unite(drug, treat_time, 
               col = 'experiment', sep = '_') |>
  pivot_wider(names_from = experiment,
              values_from = auc) |>
  tibble::column_to_rownames('improve_sample_id')

eset <- SummarizedExperiment(assays = list(cor = cor_mat))
data("ncipid")
data('krbpaths')


cor_enrich <- do.call(rbind,lapply(colnames(cor_mat), function(e){
  print(e)
  suppressMessages(res <- leapR(geneset = krbpaths,
                                  assay_name = 'cor',
          enrichment_method = 'enrichment_in_order',
          primary_columns = e, 
          eset = eset))
    res |>
      mutate(condition = e) |>
      tibble::rownames_to_column('pathway')
})) |>
  tidyr::separate(condition,into=c('drug','treat_time','expr_time'), sep='_')

path_stats <- subset(cor_enrich,BH_pvalue < 0.01) |>
  group_by(drug,treat_time,expr_time) |>
  summarize(numSig=n_distinct(pathway)) 

ggplot(path_stats,aes(x = drug,y = numSig,fill = treat_time)) + 
  geom_bar(position = 'dodge',stat='identity') + facet_grid(expr_time~.)


path_counts <- subset(cor_enrich,BH_pvalue < 0.01) |> 
  group_by(pathway) |>
  summarize(n_drugs=n_distinct(drug))|>arrange(desc(n_drugs))

upaths <- subset(path_counts,n_drugs==1) ## unique pathways

subset(cor_enrich, BH_pvalue < 0.01) |>
  subset(pathway%in%upaths$pathway)
```
Not sure what to do with this - lots of pathways correlated, but they're all the same.

Let's try regression instead. 

## Try to predict AUC from gene expression


Here we try to use a predictive model. Here we use xgboost regression with LOO
cross validation to predict delta AUC in an unseen sample for each secondary drug. 

```{r regression}
library(xgboost)

#drug = 'mirdametinib'
etime = '8hr'
ttime = 120

params <- list(
  objective = "reg:squarederror", # Standard regression
  eta = 0.1,                     # Learning rate
  max_depth = 6,                 # Tree depth
  eval_metric = "rmse"           # Evaluation metric
)
  
rres <- do.call(rbind, lapply(c('8hr','24hr'), function(etime) {
  do.call(rbind, lapply(c(48,120),function(ttime) { 
    do.call(rbind, lapply(unique(exp$drug), function(drug) {
  
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- exp[exp$treat_time == ttime &
                   exp$expr_time == etime &
                   exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  df = data.frame(apply(tmat[, as.character(rownames(tvec))],1,log2))
  #df$delta_auc = as.numeric(tvec$delta_auc)



##here we do the LOO for each sample in the drug
  samp_preds <- lapply(rownames(df), function(samp){

    st <- setdiff(rownames(df), samp)
    dtrain <- xgb.DMatrix(as.matrix(df[st,]), label = tvec[st,'auc'])
  
  
    model <- xgboost(x = as.matrix(df[st,]), 
                     y = tvec[st,'auc'],
     objective = "reg:squarederror", # Standard regression
      learning_rate = 0.1,                     # Learning rate
      max_depth = 8,                 # Tree depth
      eval_metric = "rmse",
  #  data = dtrain,
     nrounds = 1000)
  #importance_matrix <- xgb.importance(model = model)
  #pheatmap::pheatmap(t(df)[importance_matrix$Feature,
  #                             as.character(rownames(tvec))],
  #                    annotation_col = tvec, cluster_cols = FALSE)

  preds <- predict(model, newdata = df[samp,], validate_features = TRUE)
  return(preds)
  }
)
  
  tvec$preds= unlist(samp_preds)
  return(tvec)
  })
)}))}))

```

once the regression runs we can evaluate which predictions are possible. 

```{r prediction plots}

rres <- rres|>mutate(error = (auc-preds)^2)

##plot by drug
ggplot(rres, aes(x=error,fill=drug)) + geom_histogram() + scale_fill_manual(values=drugCol) + facet_grid(treat_time~expr_time)

##plot by sample

```
It seems that the 8hr time point is the best at predicting 48 hr drug combination
effect of a single agent, though that varies for specific drugs


```{r compute gene wise correlations, warning=FALSE, message=FALSE}

print('computing gene/aUC correlation')

#first focus on only durgs with 10 data points
gvals <- md |> subset(count > 10)

exp <- exp |>
dplyr::rename(auc = 'delta_auc') |>
  mutate(treat_time = as.numeric(as.character(treat_time)))

cor_res <- do.call(rbind,lapply(unique(gvals$drug), function(drug){
      rbind(compute_cor(exp,tmat,drug,'8hr',48),
            compute_cor(exp,tmat,drug,'24hr',48),
            compute_cor(exp,tmat,drug,'8hr',120),
            compute_cor(exp,tmat,drug,'24hr',120))
    })) #end drug iter



```

```{r results}

##top combo is mirda + vorin, what predicts those? 


```

```{r lightgbm eval, eval=FALSE}
library(tidymodels)
library(bonsai)
library(lightgbm)


stable_folds <- vfold_cv(df, v = 5, repeats = 10)

lgbm_spec <- boost_tree(
  trees = 100,
  tree_depth = tune(),  # Tunable parameter
  learn_rate = tune()   # Tunable parameter
) %>%
   set_engine("lightgbm", 
             num_threads = 1,          # Prevent parallel conflicts
             min_data_in_leaf = 1,    # REQUIRED for 18 rows
             min_data_in_bin = 1      # REQUIRED for 18 rows
  )%>% 
  set_mode("regression")

# 2. Create LOOCV folds
# Warning: This creates n resamples (n = number of rows)
loocv_folds <- loo_cv(df)

# 3. Create a workflow
lgbm_wflow <- workflow() %>%
  add_model(lgbm_spec) %>%
  add_formula(delta_auc ~ .)

# 4. Define a small tuning grid
# LOOCV fits n * grid_size models. Keep size low to avoid crashes.
lgbm_grid <- grid_regular(
  tree_depth(range = c(3, 7)),
  learn_rate(range = c(-2, -1)),
  levels = 3
)

# 5. Tune the model
# Increase R's expression limit if you still hit stack overflow
options(expressions = 50000)

tuning_results <- lgbm_wflow %>%
  tune_grid(
    resamples = stable_folds,
    grid = lgbm_grid,
    metrics = metric_set(rmse, rsq)
  )

# 6. Select best parameters
best_params <- select_best(tuning_results, metric = "rmse")
final_wflow <- finalize_workflow(lgbm_wflow, best_params)
```
