---
title: "Linear Regression tests"
author: "Sara Gosline"
date: "2026-02-10"
output: html_document
---


# Regression analysis of drug response

Here we propose to use an elastic net regression model to predict drug response of treated samples.

We first collect the transcriptomic and drug response data from matched samples. It is important to note
that each model is built for a single drug, and there is not a lot of variability across drugs, the 
model is much more sensitive to the sample itself. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

tmat <-   full_tmat[,intersect(colnames(full_tmat), sing_exp$improve_sample_id)]

ggplot(sing_exp, aes(x=common_name, y=auc, col = drug, shape = as.factor(treat_time))) +
         geom_jitter() + 
         scale_color_manual(values = drugCol)

```


However we can still try to identify features that predict AUC.


del of drug responses
```{r single drug regression}

library(glmnet)
#drug = 'mirdametinib'
etime = '0hr'
ttime = 48
  
rres <- do.call(rbind, lapply(unique(sing_exp$drug), function(drug) {
  
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                   sing_exp$expr_time == etime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  #df$delta_auc = as.numeric(tvec$delta_auc)

##here we do the LOO for each sample in the drug
  samp_preds <- do.call(rbind,lapply(rownames(df), function(samp){

    st <- setdiff(rownames(df), samp)
    
    dtrain <- as.matrix(df[st,])
  
    model <- glmnet(x = dtrain, y = tvec[st, 'auc'], alpha = 0.5)
    coefs <-  setdiff(unique(rownames(which(coef(model) != 0, arr.ind=T))),'(Intercept)')
    preds <- predict(model, newx = df[samp,], s=0.001)
    return(c(preds = preds, features = paste(coefs, collapse=','), training_set=length(st)))
  }
))
  tvec$preds = unlist(samp_preds[,1])
  tvec$features = unlist(samp_preds[,2])
  tvec$training_set = unlist(samp_preds[,3])
  return(tvec)
  })
)

rres$error = (as.numeric(rres$preds)-rres$auc)^2
```

We use leave one out cross validation so we train on one set of samples and evaluate on the other. 

```{r pred results}


rs <-  rres |> group_by(drug, .add=TRUE, .drop=FALSE) |> 
  summarize(mse = mean(error)) |>
   arrange(mse)

rres <- rres|>left_join(rs)

ggplot(rres, aes(x=reorder(drug, mse), y = error, col = drug)) + 
  geom_boxplot() +
  geom_jitter() + 
    scale_color_manual(values=drugCol) + 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 


```

Now we can investigate the features used to predict by building a model on all
the data.
```{r full model}

full_models <- lapply(unique(sing_exp$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                   sing_exp$expr_time == etime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  model <- glmnet(x = df, y = tvec$auc, alpha = 0.5)
  return(model)
})

names(full_models) <- unique(sing_exp$drug)

##now for each model we can visualize the genes
feats <- do.call(rbind,lapply(names(full_models), function(nmod){
  mod <- full_models[[nmod]]
  genes = setdiff(unique(rownames(which(coef(mod) != 0, arr.ind=T))),'(Intercept)')
  return(c(drug = nmod, features = paste(genes, collapse=',')))
})) |>
  as.data.frame() |>
  tidyr::separate_longer_delim(features, delim = ',')

##we can also predict comvbination effects
feats |> group_by(drug) |> summarize(numFeatures = n())
```
We can ask if there are specific pathways that are useful to predict drug response


```{r pathway enrichment}

library(leapR)
ddata <- sing_exp |> tidyr::pivot_wider(names_from= drug, values_from=auc) |> tibble::column_to_rownames('improve_sample_id')
cdata <- feats |> mutate(model_feature=TRUE) |> tidyr::pivot_wider(names_from=drug, values_from=model_feature, values_fill=FALSE) 

odata <- data.frame(features = setdiff(rownames(tmat), feats$features), |> tibble::column_to_rownames('features')
exp_set <- SummarizedExperiment(assays=SimpleList(tmat), rowData=cdata, colData = ddata)


```


## Combo prediction data

Now we want to use single agent models to predict combo


### Combo gex data

```{r combo gex data}

tmat <- full_tmat[,intersect(colnames(full_tmat), exp$improve_sample_id)]

```

### Combo model evaluation


Now we can try to use these values to predict delta AUC. These values
are on a different scale, as tehy are meant to represent AUC, so let's scale
them linearly between -1 and 1. 


```{r applying single drug models}

dpreds <- do.call(rbind, lapply(names(full_models), function(d){
  mod <- full_models[[d]]
  #feats <- intersect(rownames(mod$beta),rownames(stmat))
  pval <- predict(mod, newx = t(tmat), s = 0.001)#[feats,])
  
  return(data.frame(drug = d, improve_sample_id = as.numeric(colnames(tmat)), pred = pval[,1]))
}) )

dpreds <- dpreds |>  left_join(samples) |>
  tidyr::separate(other_id, into=c('samp','initial_drug','time','treat','type'),sep='_') |>
  dplyr::select(drug,improve_sample_id, samp, initial_drug,time,treat, pred ) |>
  distinct() |>
  mutate(initial_drug = tolower(initial_drug))

ggplot(dpreds, aes(x = pred, fill = drug)) + 
  geom_histogram(position = 'dodge') +
  scale_fill_manual(values=drugCol)


ggplot(dpreds, aes(x = drug, y = pred, col = drug)) + 
  geom_jitter() +
  scale_color_manual(values=drugCol) +facet_grid(samp~.)

```

Some values are just all over the map. linear model wants to push values
between 0 and 1 (AUC) but realistically gets moved around. 

Since we are interested in what are the *best* predictions for each drug,  we 
will scale the values between 0 and 1 by drug.

```{r normalized predictions}


dpreds <- dpreds |> 
  mutate(oppPred = 1-pred) |>
  group_by(drug, .drop = FALSE) |> mutate(predRank = 1-rank(pred)/length(pred)) |>
  ungroup() |>
  group_by(improve_sample_id, .drop=FALSE) |> mutate(sampleRank = 1-rank(pred)/length(pred))


##rank d_auc by same metr

exp <- exp |>
  group_by(initial_drug, .drop = FALSE) |> mutate(dAUCRank = rank(delta_auc)/length(delta_auc))

```



```{r compare with experiment}

dat_with_pred <- dpreds |>
  dplyr::select(improve_sample_id, drug, oppPred,pred, predRank,sampleRank) |>
  right_join(exp) |>
 # mutate(pd_auc = 1 - as.numeric(pred)) |> #this gets us in the right direction
  mutate(rankErr = (predRank - dAUCRank)^2) |>
  mutate(err = (oppPred - delta_auc)^2)
#|> ##still not quite right
#  subset(!is.na(pred))

dat_with_pred |>
  ggplot(aes(x = expr_time, y = err,col = drug)) + geom_jitter() + 
    scale_color_manual(values = drugCol) + 
    geom_boxplot() +
  facet_grid(~treat_time)

dat_with_pred |>
  ggplot(aes(x = common_name, y = sampleRank, col = drug)) + geom_jitter() + 
    scale_color_manual(values=drugCol) + 
    geom_boxplot() +
  facet_grid(~treat_time)


dat_with_pred |>
  ggplot(aes(x = initial_drug, y = predRank, col = drug)) + geom_jitter() + 
    scale_color_manual(values = drugCol) + 
  geom_boxplot() +
  #facet_grid(~treat_time) +
  theme_bw()


ggplot(dat_with_pred, aes(x = common_name, y = err, col = drug)) + 
  geom_jitter() + 
    scale_color_manual(values=drugCol) + 
  theme_bw()


ggplot(dat_with_pred, aes(x = delta_auc, y = pred, col = drug)) + 
  geom_jitter() + 
    scale_color_manual(values=drugCol) + 
  geom_smooth(method='lm', aes(group=drug))+
  theme_bw()

subset(dat_with_pred, predRank > 0.75) |>
ggplot(aes(x = delta_auc, y = pred, col = drug)) + 
  geom_jitter() + 
    scale_color_manual(values=drugCol) + 
    geom_smooth(method='lm', aes(group=drug))+
  theme_bw()
```

