---
title: "4_Target_pathways"
author: "`r config::get(file = 'config.yml')$document$author`"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
    fig_caption: yes
    highlight: haddock
    number_sections: yes

knit: (
  function(inputFile, encoding) {
    config=config::get(file = 'config.yml') 
    
    pSubTitle <- "4_Target_pathways.pdf"
    
    base_output_dir <- paste0(config$Path,"/","Report")
    
    DIR=file.path(base_output_dir)
    
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      output_file = pSubTitle,
      output_dir =DIR )}
    )
---



```{r,include=F}
rm(list=ls())
```

```{r,include=F}
library(dplyr)
library(ggplot2)
library(synapser)
library(dplyr)
library(readr)
library(tidyr)

```


# Define your own config
```{r,setup, include=FALSE}
getwd()
config <- config::get(file = 'config.yml')
# 
Working_path=config$Path
base_output_dir <- paste0(Working_path,"/","Report")
email=config$Email
Authtoken=config$Token_file
knitr::opts_chunk$set(echo = F,warnings =F, message=F,include=F)
options(knitr.table.format = 'markdown',encoding = 'UTF-8',warnings =F, message=F ) 
# Check and create output directory if it doesn't exist
output_dir <- file.path(Working_path, "Data_preparation")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
Reference_path=config$Reference_path

dir(Reference_path)

getwd()
# load function
source(file.path("/Volumes/lyu.yang/MAC_1/Github/MPNST-PDX-MT/RNAseq","Rfunction.R" ))
```


```{r}
if (! file.exists( file.path(Reference_path,"Human_GSEA","Protein_coding_genelist.rdata"))){
   syID="syn63828384"
  synapser::synGet(syID,downloadLocation = Reference_path,ifcollision="keep.local")

}

```

# Download raw coutns, TPM and annoation data from synapse
```{r}
# Batch1 data dowloading from synapse
Rawdata_path=file.path(Working_path,"Rawdata_update")
Processed_path=file.path(Working_path,"Processed_data")
dir.create(Processed_path)
Rawdata_path_b1=file.path(Rawdata_path,"b1")
# login synapse
Token=read.table(Authtoken)%>%as.character()
synapser::synLogin(email=email,authToken=Token)
# download files from synapse
Rawcounts_syID_b1="syn64608365"
RawTPM_syID_b1="syn64608366"
Annotation_syID_b1="syn64608368"

for (i in c(Rawcounts_syID_b1,RawTPM_syID_b1,Annotation_syID_b1)){
   synapser::synGet(i,downloadLocation = Rawdata_path_b1,ifcollision="keep.local")

  
}


# Download Batch2 data from synapse

Rawdata_path=file.path(Working_path,"Rawdata_update")
Rawdata_path_b2=file.path(Rawdata_path,"b2")
# login synapse
Token=read.table(Authtoken)%>%as.character()
synapser::synLogin(email=email,authToken=Token)
# download files from synapse
Rawcounts_syID_b2="syn64608369"
RawTPM_syID_b2="syn64608370"
#Annotation_syID_b2="syn64608372" # DMSO_batch_version1
Annotation_syID_b2="syn66302373"

for (i in c(Rawcounts_syID_b2,RawTPM_syID_b2,Annotation_syID_b2)){
   synapser::synGet(i,downloadLocation = Rawdata_path_b2,ifcollision="keep.local")

  
}

Annotation_b2=read.table(file.path(Rawdata_path_b2,"01172025_Batch2_Samplemap_swap_DMSO.csv"),sep=",",header = T)%>%dplyr::mutate(Drug=ifelse(Full_name=="Trametinib","TRAM",Drug))

Drug_table_syID="syn64608501"

synapser::synGet(Drug_table_syID,downloadLocation = Rawdata_path,ifcollision="keep.local")

# human GMT files download
GMT_folder_id="syn63828408"
GSEA_path=file.path(Reference_path,"Human_GSEA")

download_synapse_folder(folder_id=GMT_folder_id, destination_path=GSEA_path)

```

```{r,include=F}
load(file.path(Reference_path,"Human_GSEA","Protein_coding_genelist.rdata"))

set.seed(12345)
dir(Rawdata_path_b1)
Raw_counts_b1=read.table(file.path(Rawdata_path_b1,"salmon.merged.gene_counts_corrected_id.csv"),sep=",",header = T,row.names = 1)
TPM_counts_b1=read.table(file.path(Rawdata_path_b1,"salmon.merged.gene_tpm_corrected_id.csv"),sep=",",header = T,row.names = 1)
Annotation_b1=read.table(file.path(Rawdata_path_b1,"01172025_Batch1_Samplemap.csv"),sep=",",header = T)
Annotation_b1$Drug=factor(Annotation_b1$Drug,levels = c("DMSO","MIR"))
Annotation_b1$Timepoint=factor(Annotation_b1$Timepoint,levels=c("0h","8h","24h"))
TPM_b1_2=TPM_counts_b1%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()
Counts_b1_2=Raw_counts_b1%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()%>%mutate_all(.,as.integer)

Raw_counts_b2=read.table(file.path(Rawdata_path_b2,"salmon.merged.gene_counts_corrected_id.csv"),sep=",",header = T)
TPM_counts_b2=read.table(file.path(Rawdata_path_b2,"salmon.merged.gene_tpm.tsv_corrected_ID.csv"),sep=",",header = T)
Annotation_b2$Drug=factor(Annotation_b2$Drug)
Annotation_b2$Timepoint=factor(Annotation_b2$Timepoint,levels=c("0h","8h","24h"))
TPM_b2_2=TPM_counts_b2%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()
Counts_b2_2=Raw_counts_b2%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()%>%mutate_all(.,as.integer)

```



```{r}
Drug_anno=readxl::read_excel(file.path(Rawdata_path,"Drug_annotation.xlsx"))%>%arrange(Drug_class)
Drug_name = Drug_anno$Drug %>% unique()%>%na.omit()
Drug_name=factor(Drug_name,levels = c("DMSO" ,"PAL",  "RIB" , "DEC" , "VOR" , "TRAM"  ,"MIR"  ,"SEL"  ,"OLA",  "TNO",  "RMC" , "CAP"  ,"TRAB", "DOX"  ,"IFO"  ,"IRI"))

length(Drug_name)
# Ensure you have 14 unique Drug names (1 for black, 13 other colors)
if (length(Drug_name) != 16) {
  stop("The number of unique drugs should be 16, including the one colored black.")
}

# Use RColorBrewer to get 13 colors from the Paired palette
paired_colors =RColorBrewer:: brewer.pal(n = 12, name = "Paired")
extra_color =RColorBrewer:: brewer.pal(n = 8, name = "Set3")[1:3]  # Take one color from another palette (e.g., Set3)


# Combine black and the 13 colors
drug_colors = c("black", paired_colors, extra_color)

# Ensure the color vector matches the order of Drug_name
names(drug_colors) = sort(Drug_name)

Timepoint_colors= RColorBrewer::brewer.pal(5,"Set1")[2:4]
names(Timepoint_colors)=c("0h","8h","24h")



```


```{r}
# RES and GSEA data download
load(file.path(Processed_path,"All_drug_RES_FGSEA.rdata"))
```

```{r}
if (!file.exists(file.path(GSEA_path,"Cancer_Pathways.csv"))) {
# Load required package
library(tibble)

# Create the cancer pathways data frame
cancer_pathways_df <- tribble(
  ~Pathway,                      ~Keywords,                     ~Key_Genes,                                  ~Key_Role,
  "PI3K-AKT-mTOR",               "PI3K AKT MTOR",               "PIK3CA, PIK3CB, AKT1, MTOR, PTEN",         "Regulates cell growth, survival, and metabolism",
  "RAS-RAF-MEK-ERK (MAPK)",      "RAS MAPK ERK",                "KRAS, NRAS, BRAF, MAP2K1 (MEK1), MAPK1",   "Controls cell proliferation and differentiation",
  "WNT/β-Catenin",               "WNT β-Catenin BETA-CATENIN",               "CTNNB1, APC, GSK3B, AXIN1",                "Regulates stem cell renewal and cell fate",
  "Notch Signaling",             "NOTCH DLL JAG",               "NOTCH1, NOTCH2, JAG1, DLL1, HES1",         "Controls cell differentiation and survival",
  "Hedgehog (Hh)",               "HEDGEHOG SHH PTCH",           "SHH, PTCH1, SMO, GLI1",                    "Essential for development, implicated in cancers",
  "Hippo",                       "HIPPO YAP TAZ",               "YAP1, TAZ, TEAD1, MST1, LATS1",            "Controls organ size, proliferation, and apoptosis",
  "JAK-STAT",                    "JAK STAT",                    "JAK1, JAK2, STAT3, SOCS1",                 "Mediates immune response and inflammation",
  "TP53 Tumor Suppressor",       "TP53 P53",                    "TP53, MDM2, CDKN1A, ATM, CHEK2",           "Regulates DNA damage response and apoptosis",
  "TGF-β Signaling",             "TGF SMAD",                     "TGFB1, TGFBR1, SMAD2, SMAD3, ZEB1",        "Controls cell proliferation and immune suppression",
  "NF-κB",                       "NFKB RELA",                    "NFKB1, RELA, IKBKG, TNF",                  "Regulates inflammation and cell survival",
  "MYC Pathway",                 "MYC MAX",                      "MYC, MAX, MXD1",                           "Regulates cell proliferation and metabolism",
  "DNA Damage Response (DDR)",   "DNA_REPAIR BRCA ATM",          "BRCA1, BRCA2, ATM, CHEK2",                 "Repairs DNA damage to prevent mutations",
  "Oxidative Stress/Nrf2",       "NRF2 KEAP1",                   "KEAP1, NFE2L2, CUL3, SQSTM1",              "Regulates cellular response to oxidative stress",
  "Autophagy",                   "AUTOPHAGY BECN1",              "BECN1, ATG5, ATG7, ULK1",                  "Maintains cell survival under stress conditions",
  "Cell Cycle (CDK4/6-RB1)",     "CDK RB1 E2F",                  "CDK4, CDK6, RB1, CCND1, E2F1",             "Regulates cell cycle progression and division",
  "RTK-RAS Pathway",             "RTK RAS EGFR",                 "EGFR, KRAS, BRAF, SHC1, SOS1",             "Growth signaling and differentiation",
  "Hypoxia/HIF-1α",              "HIF1A VEGFA",                  "HIF1A, VEGFA, EGLN1",                      "Adaptation to low oxygen, promotes angiogenesis",
  "Metabolic Pathways",          "CANCER METABOLISM AMPK GLUT1", "PKM2, GLUT1, LDHA, AMPK",                  "Cancer cells reprogram metabolism to sustain growth"
)

# Save to CSV file
write.csv(cancer_pathways_df, file.path(GSEA_path,"Cancer_Pathways.csv"), row.names = FALSE)

}


if (!file.exists(file.path(GSEA_path,"Cancer_Chemokine_Cytokine_Pathways.csv"))) {

# Create the cancer chemokine & cytokine pathways data frame
cancer_chemokine_cytokine_df <- tribble(
  ~Pathway,                      ~Keywords,                           ~Key_Genes,                                   ~Key_Role,
  "Chemokine Signaling Pathway", "CHEMOKINE CCR CXCR CCL CXCL",      "CXCR4, CXCR3, CCR5, CCL2, CCL5, CXCL8",      "Regulates immune cell migration, inflammation, and metastasis",
  "CXCR4/CXCL12 Axis",          "CXCR4 CXCL12 SDF-1",                "CXCR4, CXCL12",                             "Promotes tumor invasion, metastasis, and immune evasion",
  "CCR5/CCL5 Axis",             "CCR5 CCL5 RANTES",                  "CCR5, CCL5",                                "Enhances tumor cell migration and metastasis",
  "CCR2/CCL2 Axis",             "CCR2 CCL2 MCP-1",                   "CCR2, CCL2",                                "Recruits tumor-associated macrophages (TAMs)",
  "CXCR3/CXCL9-11 Axis",        "CXCR3 CXCL9 CXCL10 CXCL11",         "CXCR3, CXCL9, CXCL10, CXCL11",              "Influences immune surveillance and T-cell recruitment",
  "CXCR2/CXCL1-8 Axis",         "CXCR2 CXCL1 CXCL2 CXCL8 IL-8 IL8",      "CXCR2, CXCL1, CXCL2, CXCL8 (IL-8)",         "Promotes angiogenesis and tumor-associated neutrophils",
  "IL-6/STAT3 Pathway",         "IL6 STAT3 JAK",                     "IL6, IL6R, JAK1, STAT3, SOCS3",             "Drives chronic inflammation and promotes tumor growth",
  "TNF-α Signaling Pathway",    "TNF TNFR NFKB NF-κB",                    "TNF, TNFRSF1A, RELA, IKK",                  "Regulates cell survival, apoptosis, and inflammation",
  "IL-10 Anti-Inflammatory Pathway", "IL10 IL10R STAT3",             "IL10, IL10RA, STAT3, SOCS1",                "Suppresses immune response and promotes tumor escape",
  "TGF-β Signaling Pathway",    "TGF SMAD",                          "TGFB1, TGFBR1, SMAD2, SMAD3, ZEB1",         "Controls cell proliferation and immune suppression",
  "Type I Interferon Pathway",  "IFNA IFNB JAK STAT",                "IFNA1, IFNB1, JAK1, STAT1, IRF9",           "Activates anti-tumor immune response and inflammation",
  "Type II Interferon Pathway", "IFNG JAK STAT",                     "IFNG, JAK1, STAT1, IRF1",                   "Enhances immune cytotoxicity against tumors",
  "GM-CSF Pathway",             "GM-CSF CSF2 CSF3",                  "CSF2 (GM-CSF), CSF3, CSF2RA, JAK2",         "Regulates macrophage and neutrophil activity"
)

# Save as CSV
write.csv(cancer_chemokine_cytokine_df, file.path(GSEA_path,"Cancer_Chemokine_Cytokine_Pathways.csv"), row.names = FALSE)

print("Cancer_Chemokine_Cytokine_Pathways.csv has been saved successfully.")
}

```


```{r}
FGSEA_sig=get_sig_FGSEA (Result_list)
#%>%mutate(Drug=factor(FGSEA_sig$Drug,levels =Drug_anno$Drug ))%>%mutate(Timepoint=factor(FGSEA_sig$Timepoint,levels =c("8h","24h" ))) 
FGSEA_sig$Drug=factor(FGSEA_sig$Drug,levels =Drug_name)
FGSEA_sig$Timepoint=factor(FGSEA_sig$Timepoint,levels =c("8h","24h" ))
FGSEA_sig_org=FGSEA_sig[,c("Drug","Timepoint","Group")]%>%arrange(Drug,Timepoint)%>%unique()
FGSEA_sig$Group=factor(FGSEA_sig$Group,levels=FGSEA_sig_org$Group)
  
Group_anno=FGSEA_sig[,c("Group","Drug","Timepoint")]
```



\newpage

## Cell cycle related pathway  
```{r}
keywords <- "CELL_CYCLE"

add_keywords=c("")

# Print extracted keywords
keyword_pattern <-paste(c(keywords,add_keywords), collapse = "|")%>%gsub("\\|$","",.)
pathway_list_1=FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))
# Filter FGSEA results using grepl (case-insensitive matching)
filtered_fgsea <- FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))%>%arrange(Group)
#filtered_fgsea$Group
missing_samples=setdiff(unique(FGSEA_sig$Group),filtered_fgsea$Group)
missing_samples_anno=data.frame(Group=missing_samples)%>%left_join(., Group_anno)%>%mutate(pathway=filtered_fgsea$pathway[1],pval=NA,padj=NA,log2err=NA,ES=NA,NES=NA,size=NA,leadingEdge=NA)%>%.[,colnames(filtered_fgsea)]
filtered_fgsea2=rbind(filtered_fgsea,missing_samples_anno)%>%arrange(Group)

Plot_all=plot_pathway_heatmap_2(filtered_fgsea2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug",timepoint_col = "Timepoint",Drug_colors_set = drug_colors,Timepoint_color_set = Timepoint_colors)


```

```{r,fig.width=30,fig.height=28,include=T}
Plot_all
```

\newpage

```{r,include=T}
Select_pathway="FISCHER_G2_M_CELL_CYCLE"

Cell_cycle_UP=FGSEA_sig%>%dplyr::filter(pathway==Select_pathway)%>%dplyr::filter(NES>0)%>%arrange(Drug )
knitr::kable(Cell_cycle_UP[,c("pathway","NES","pval","Drug","Timepoint")],caption = "Group with Cell cycle upregulated")%>%kableExtra::kable_styling(latex_options = "scale_down")
```


 


# Mirda (batch1)

```{r,include=T}
Drug_name="MIR"
Select_pathway="FISCHER_G2_M_CELL_CYCLE"
anno_2=Annotation_b1[c("specimenID","Drug","Timepoint")]%>%arrange(Timepoint,Drug)%>%matrix.please()%>%as.data.frame()
TPM_b1_2_arrange=TPM_b1_2[,rownames(anno_2)]

number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])
Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name&pathway==Select_pathway)%>%arrange(NES) 

Gene_list=Test$leadingEdge%>%unlist()%>%as.character()
fig_title=paste0(Drug_name,"_",Select_pathway)
p2=TPM_heatmap(TPM_b1_2_arrange,Gene_list,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

```
```{r}
knitr::knit_exit()
```


```{r}
# Generare Heatmap by using significant genes from GSEA output
generate_TPM_heatmap <- function(Drug_name, Pathway_name=NULL,Gene_list=NULL) {
  # Get the experiment date for the specified drug
  Date_exp <- Annotation_b2[c("specimenID", "Drug", "Timepoint", "Date")] %>%
    dplyr::filter(Drug == Drug_name) %>%
    .$Date %>%
    unique()

  # Filter annotations for the drug and DMSO at the same experiment date
  anno_2 <- Annotation_b2[c("specimenID", "Drug", "Timepoint", "Date")] %>%
    arrange(Timepoint, Drug) %>%
    dplyr::filter(Drug %in% c(Drug_name, "DMSO")) %>%
    dplyr::filter(Date == Date_exp) %>%
    .[, c("specimenID", "Drug", "Timepoint")] %>%
    matrix.please() %>%
    as.data.frame()

  # Subset TPM data based on the arranged annotation
  TPM_b2_2_arrange <- TPM_b2_2[, rownames(anno_2)]

  # Count the number of samples at 8h timepoint
  number <- anno_2 %>%
    dplyr::filter(Timepoint == "8h") %>%
    nrow()

  # Set colors for annotation
  Color_list <- list(Drug = drug_colors[unique(anno_2$Drug)])

# Extract leading-edge genes for the pathway
if (! is.null(Pathway_name)){
  # Filter FGSEA results for the specified drug and pathway
  Test <- FGSEA_sig %>%
    dplyr::filter(Drug == Drug_name) %>%
    dplyr::filter(pathway == Pathway_name)   
  Pathway_pickup <- Test %>%
    dplyr::filter(pathway == Pathway_name) %>%
    .$leadingEdge %>%
    unlist() %>%
    as.character()
  }
  
  if (! is.null(Gene_list)){
    Pathway_pickup=Gene_list[,1]
    Pathway_name=colnames(Gene_list)
  }

   # Generate the title for the figure
  fig_title <- paste0(Drug_name, "_", Pathway_name)

  # Generate the heatmap
  p2 <- TPM_heatmap(TPM_b2_2_arrange, Pathway_pickup, 
                     Annotation_col = anno_2, 
                     Annotation_color = Color_list, 
                     gap_set = c(3, 3 + number, 3 + number), 
                     Title = fig_title)
  
  return(p2)
}

 

```

```{r}
Drug_check_list=Cell_cycle_UP$Drug%>%unique()
Plotlist=list()
for (i in Drug_check_list) {
  Drug_name=i
  Pahway_name="FISCHER_G2_M_CELL_CYCLE"
  Plotlist[[i]]=generate_TPM_heatmap(Drug_name,Pahway_name)%>%.[[4]]
  
}


```


```{r,fig.width=24,fig.height=40,include=T}
cowplot::plot_grid(plotlist = Plotlist,nrow=1)
```


```{r}
CDK_genes=data.frame(CDK_genes=c("CDK4","CDK6","CCND1"))
Drug_name="PAL"
P1=generate_TPM_heatmap(Drug_name,Pathway_name=NULL,Gene_list =CDK_genes )
Drug_name="RIB"
P2=generate_TPM_heatmap(Drug_name,Pathway_name=NULL,Gene_list =CDK_genes )
```

```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = list(P1$gtable,P2$gtable),nrow=1)
```



## MAPK inhibitor
```{r}
Drug_name=c("SEL","TRA")
MAPK_genes=data.frame(MAPK_sig=c("PHLDA1","SPRY2", "SPRY4", "ETV4", "ETV5", "DUSP4", "DUSP6", "CCND1", "EPHA2", "EPHA4") )
#Selected_path="WP_HOSTPATHOGEN_INTERACTION_OF_HUMAN_CORONAVIRUSES_MAPK_SIGNALING"

Plot_list=list()
#Plot_list2=list()
for (i in Drug_name) {
  Plot_list[[i]]=generate_TPM_heatmap(i,Pathway_name=NULL,Gene_list =MAPK_genes )%>%.$gtable
 # Plot_list2[[i]]=generate_TPM_heatmap(i,Pathway_name=Selected_path,Gene_list =NULL )%>%.$gtable
  
}

```


```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = Plot_list,nrow=1)
```

```{r}
knitr::knit_exit()
```


```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = Plot_list2,nrow=1)
```



```{r}
Drug_name="PAL"
Select_pathway="CHICAS_RB1_TARGETS_LOW_SERUM"
Date_exp=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%dplyr::filter(Drug==Drug_name)%>%.$Date%>%unique()
anno_2=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%arrange(Timepoint,Drug)%>%dplyr::filter(Drug%in%c(Drug_name,"DMSO"))%>%
                      dplyr::filter(Date==Date_exp)%>%.[,c("specimenID","Drug","Timepoint")]%>%matrix.please()%>%as.data.frame()

TPM_b2_2_arrange=TPM_b2_2[,rownames(anno_2)]
number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()

Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])

NES_neg=FGSEA_sig%>%dplyr::filter(Drug==Drug_name&NES<0)%>%arrange(NES)
Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name)%>%arrange(NES)%>%dplyr::filter(pathway==Select_pathway)

Pathway_pickup=Test%>%dplyr::filter(pathway==Select_pathway)%>%.$leadingEdge%>%unlist()%>%as.character()
fig_title=paste0(Drug_name,"_",Select_pathway)

p2=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

CDK_genes=data.frame(CDK_genes=c("CDK4","CDK6","CCND1"))

Test=generate_TPM_heatmap(Drug_name,Pathway_name=NULL,Gene_list =CDK_genes )

fig_title=paste0(Drug_name,"_","CDK genes")
p3=TPM_heatmap(TPM_b2_2_arrange,CDK_genes,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

PCA_Plot_3_shape(t(TPM_b2_2_arrange), Annotation_b2, "specimenID", "Drug", "Timepoint")+scale_color_manual(values=drug_colors)
```
```{r}
p2
```


# Mirda

```{r}
Drug_name="MIR"
anno_2=Annotation_b1[c("specimenID","Drug","Timepoint")]%>%arrange(Timepoint,Drug)%>%matrix.please()%>%as.data.frame()
TPM_b1_2_arrange=TPM_b1_2[,rownames(anno_2)]
number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()

Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])

MAPK_genes=c("PHLDA1","SPRY2", "SPRY4", "ETV4", "ETV5", "DUSP4", "DUSP6", "CCND1", "EPHA2", "EPHA4") 

p1=TPM_heatmap(TPM_b1_2_arrange,MAPK_genes,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="Mirda signature")

Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name&Timepoint=="8h")%>%arrange(NES)%>%dplyr::filter(grepl("MAPK|ERK|CYCLE",pathway))

Select_pathway="FISCHER_G2_M_CELL_CYCLE"
fig_title=paste0(Drug_name,"_",Select_pathway)
FISCHER_G1_S_CELL_CYCLE=Test%>%dplyr::filter(pathway==Select_pathway)%>%.$leadingEdge%>%unlist()%>%as.character()


p2=TPM_heatmap(TPM_b1_2_arrange,FISCHER_G1_S_CELL_CYCLE,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

```

```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = list(p1$gtable,p2$gtable),nrow=1)
```

# Selumetinib

```{r}
Drug_name="SEL"
Date_exp=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%dplyr::filter(Drug==Drug_name)%>%.$Date%>%unique()
anno_2=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%arrange(Timepoint,Drug)%>%dplyr::filter(Drug%in%c(Drug_name,"DMSO"))%>%
                      dplyr::filter(Date==Date_exp)%>%.[,c("specimenID","Drug","Timepoint")]%>%matrix.please()%>%as.data.frame()
TPM_b2_2_arrange=TPM_b2_2[,rownames(anno_2)]
number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()

Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])

p1=TPM_heatmap(TPM_b2_2_arrange,MAPK_genes,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="Mirda target genes")

Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name)%>%arrange(NES)%>%dplyr::filter(grepl("MAPK|ERK|CYCLE",pathway))%>%arrange(pathway)

Pahway_name="GOBP_ERK1_AND_ERK2_CASCADE"

Pathway_pickup=Test%>%dplyr::filter(pathway=="GOBP_ERK1_AND_ERK2_CASCADE")%>%.$leadingEdge%>%unlist()%>%as.character()

p2=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=Pahway_name)

Pahway_name="WP_HOSTPATHOGEN_INTERACTION_OF_HUMAN_CORONAVIRUSES_MAPK_SIGNALING"

Pathway_pickup=Test%>%dplyr::filter(pathway=="GOBP_ERK1_AND_ERK2_CASCADE")%>%.$leadingEdge%>%unlist()%>%as.character()

p3=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=Pahway_name)

Pahway_name="FISCHER_G2_M_CELL_CYCLE"
Pathway_pickup=Test%>%dplyr::filter(pathway==Pahway_name)%>%.$leadingEdge%>%unlist()%>%as.character()


p4=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=Pahway_name)



```
```{r,include=T,fig.width=12,fig.height=12}
cowplot::plot_grid(plotlist = list(p1$gtable,p2$gtable,p3$gtable,p4$gtable),nrow=2)
```

```{r}
knitr::knit_exit()
```


```{r}

Test

Test$leadingEdge
```


```{r}
Test=FGSEA_sig%>%dplyr::filter(Drug=="RIB"&Timepoint=="8h")%>%arrange(NES)%>%dplyr::filter(grepl("MAPK|ERK|CYCLE|RB1",pathway))

Test%>%dplyr::filter(pathway=="REACTOME_PERK_REGULATES_GENE_EXPRESSION")%>%.$leadingEdge
Test$leadingEdge
```



# Cytokine and chemokine pathways extraction


```{r}
# Read the CSV file containing cancer-related chemokine & cytokine pathways
cancer_chemo_pathways_df <- read_csv(file.path(GSEA_path, "Cancer_Chemokine_Cytokine_Pathways.csv"))
```

```{r,include=T}
knitr::kable(cancer_chemo_pathways_df,caption = "Cytokine and chemokine related pathways")%>%kableExtra::kable_styling(latex_options = c("scale_down"))
```


```{r}
# Extract unique keywords from the "Keywords" column

keywords <- unique(unlist(stringr::str_split(cancer_chemo_pathways_df$Keywords, " ")))
add_keywords=c("INTERLEUKIN","IL1","IL2")
# Print extracted keywords
keyword_pattern <-paste(c(keywords,add_keywords), collapse = "|")%>%gsub("\\|$","",.)
pathway_list_1=FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))
# Filter FGSEA results using grepl (case-insensitive matching)
Plot_list=list()
for (i in c("WP|HALLMARK|KEGG","GOBP")) {
  filtered_fgsea <- FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))%>%dplyr::filter(grepl(i,pathway))%>%arrange(Group)
    filtered_fgsea$Group
   missing_samples=setdiff(unique(FGSEA_sig$Group),filtered_fgsea$Group)
   missing_samples_anno=data.frame(Group=missing_samples)%>%left_join(., Group_anno)%>%mutate(pathway=filtered_fgsea$pathway[1],pval=NA,padj=NA,log2err=NA,ES=NA,NES=NA,size=NA,leadingEdge=NA)%>%.[,colnames(filtered_fgsea)]
   filtered_fgsea2=rbind(filtered_fgsea,missing_samples_anno)%>%arrange(Group)

   Plot_list[[i]]=plot_pathway_heatmap_2(filtered_fgsea2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug",timepoint_col = "Timepoint",Drug_colors_set = drug_colors,Timepoint_color_set = Timepoint_colors)

}


```

\newpage

```{r,fig.width=30,fig.height=24,include=T}
 Plot_list[[1]]
```

\newpage

```{r,fig.width=30,fig.height=18,include=T}
 Plot_list[[2]]
```



# Cancer-related pathways
```{r}
# Read the CSV file containing cancer-related chemokine & cytokine pathways
cancer_pathways_df <- read_csv(file.path(GSEA_path, "Cancer_Pathways.csv"))
```

\newpage
```{r,include=T}
knitr::kable(cancer_pathways_df,caption = "Cancer related pathways")%>%kableExtra::kable_styling()
```

```{r}
# Extract unique keywords from the "Keywords" column
keywords <- unique(unlist(stringr::str_split(cancer_pathways_df$Keywords, " ")))

add_keywords=c("")
# Print extracted keywords
keyword_pattern <-paste(c(keywords,add_keywords), collapse = "|")%>%gsub("\\|$","",.)
FGSEA_sig_rm=FGSEA_sig%>%dplyr::filter(!pathway%in%pathway_list_1$pathway)
# Filter FGSEA results using grepl (case-insensitive matching)
Plot_list=list()
for (i in c("WP","HALLMARK","KEGG","GOBP")) {
  filtered_fgsea <- FGSEA_sig_rm %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))%>%dplyr::filter(grepl(i,pathway))%>%arrange(Group)
    filtered_fgsea$Group
   missing_samples=setdiff(unique(FGSEA_sig$Group),filtered_fgsea$Group)
   missing_samples_anno=data.frame(Group=missing_samples)%>%left_join(., Group_anno)%>%mutate(pathway=filtered_fgsea$pathway[1],pval=NA,padj=NA,log2err=NA,ES=NA,NES=NA,size=NA,leadingEdge=NA)%>%.[,colnames(filtered_fgsea)]
   filtered_fgsea2=rbind(filtered_fgsea,missing_samples_anno)%>%arrange(Group)
   Plot_list[[i]]=plot_pathway_heatmap_2(filtered_fgsea2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug",timepoint_col = "Timepoint",Drug_colors_set = drug_colors,Timepoint_color_set = Timepoint_colors)

}
```
\newpage

```{r,fig.width=36,fig.height=28,include=T}
 Plot_list[[1]]
```

\newpage
```{r,fig.width=24,fig.height=18,include=T}
 Plot_list[[2]]
```

\newpage
```{r,fig.width=24,fig.height=18,include=T}
 Plot_list[[3]]
```

\newpage
```{r,fig.width=48,fig.height=32,include=T}
 Plot_list[[4]]
```

```{r}
knitr::knit_exit()
```



```{r}
Drug_name = RNAseq_anno$Drug %>% unique()
Drug_name=factor(Drug_name,levels = Drug_name)
# Use RColorBrewer to get 13 colors from the Paired palette
paired_colors =RColorBrewer:: brewer.pal(n = 12, name = "Paired")
extra_color =RColorBrewer:: brewer.pal(n = 8, name = "Set3")[1]  # Take one color from another palette (e.g., Set3)

# Combine black and the 13 colors
drug_colors = c("black", paired_colors, extra_color)[1:length(Drug_name)]
names(drug_colors)=Drug_name
Colour_frame=data.frame(Drug=names(drug_colors),Colour=drug_colors)

RNAseq_anno=RNAseq_anno%>%left_join(.,Colour_frame)
RNAseq_anno$Timepoint=factor(RNAseq_anno$Timepoint,levels = c("0h","8h","24h"))
RNAseq_anno$Drug=factor(RNAseq_anno$Drug,levels=Drug_name)

Sample_ID_Extract=RNAseq_anno[match(RNAseq_anno$Sample_name,colnames(TPM_2)),]

if (isTRUE(unique(Sample_ID_Extract$Sample_name==colnames(TPM_2))) ){
  print("it is correct")
  TPM_cor=TPM_2
   colnames(TPM_cor)=Sample_ID_Extract$Sample_ID
}

RNAseq_anno_org%>%group_by(Drug,Timepoint)%>%summarise(n=n())%>%arrange(Drug)
```


```{r}
Drug_select=RNAseq_anno%>%dplyr::filter(Drug_class==Drug_class_name)
Batch=Drug_select$Time%>%unique()
DMSO_group=RNAseq_anno%>%dplyr::filter(Drug_class=="Control")%>%dplyr::filter(Time==Batch)
Drug_DMSO_data=rbind(Drug_select,DMSO_group)%>%arrange(Timepoint,Drug)
TPM_cor_drug=TPM_cor[,c(DMSO_group$Sample_ID,Drug_DMSO_data$Sample_ID)]%>%.[,match(Drug_DMSO_data$Sample_ID,colnames(.))]
Sample_anno=Drug_DMSO_data
colnames(TPM_cor_drug)
rownames(Sample_anno)=Sample_anno$Sample_ID
Sample_anno_2=Sample_anno[c("Drug","Timepoint")]
Summary=Sample_anno_2%>%group_by(Drug,Timepoint)%>%summarise(n=n())%>%arrange(Timepoint,Drug)
Summary2=Sample_anno_2%>%group_by(Timepoint)%>%summarise(n=n())%>%arrange(Timepoint)
CUM1=cumsum(Summary$n)
CUM2=cumsum(Summary2$n)
Cut_col=c(CUM1,CUM2)%>%sort()

```


# Use DMSO group1 as control

```{r}

load(file.path(Process_path,Drug_class_name,"Chemo_Deseq2_OUT_vs_paired_DMSO" ))

cgwtools::lsdata(file.path(Process_path,Drug_class_name,"Chemo_Deseq2_OUT_vs_paired_DMSO" ))


Test=Drug_reslist[[1]]$RES$`8h`$DOX_8h_vs_DMSO_8h%>%as.data.frame()
Test%>%.["IL1B",]

Test2=Drug_reslist[[1]]$RES$`24h`$DOX_24h_vs_DMSO_24h%>%as.data.frame()
Test2%>%.["IL1B",]
```



```{r}
File_name1="Chemo_GSEA_VS_paried_DMSO.rdata"
if (file.exists(file.path(Process_path,File_name1))) {
 load(file.path(Process_path,File_name1 ))
 #fileexist=cgwtools::lsdata(file.path(Process_path,File_name1))
} else{
 GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="all.v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("c2|c5|h",.)]
 FGSEA_list <- list()

 # Loop through each drug in the Drug_reslist
for (i in names(Drug_reslist)) {
  Data <- Drug_reslist[[i]]
  Timepoint_list <- Data$RES  # Access the time points for this drug

  # Loop through each time point
  for (j in names(Timepoint_list)) {
    R1 <- Timepoint_list[[j]]

    # Loop through each replicate within the time point
    for (h in names(R1)) {
      data <- R1[[h]]

      # Filter significant genes and prepare ranked list
      res_q005 <- data %>%
        as.data.frame() %>%
        dplyr::filter(padj < 0.05)
    
      # Create a ranked vector for FGSEA using log2FoldChange
      res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
      ranks <- tibble::deframe(res_q005_rank)  # Convert to named vector

      # Store FGSEA results in a list
      out_list <- list()

      # Run FGSEA for each GMT file
      for (x in seq_along(GMT_files)) {
        pathways <- fgsea::gmtPathways(GMT_files[x])  # Load pathway data
        set.seed(12345)  # Set seed for reproducibility
        out_list[[x]] <- fgsea::fgseaMultilevel(
          pathways, ranks,
          minSize = 10,
          maxSize = 500,
          nPermSimple = 10000
        )
      }

      # Combine FGSEA results for all GMT files for this replicate
      #FGSEA_list[[paste(i, j, h, sep = "_")]] <- do.call("rbind", out_list)
      
      FGSEA_list[[h]] <- do.call("rbind", out_list)
    }
   }
}
  save(FGSEA_list,file=file.path(Process_path,File_name1 ))
 
}
```

```{r}
names(FGSEA_list)
```


```{r}
cgwtools::lsdata(file.path(Process_path,Drug_class_name,"Deseq2_OUT_vs_DMSO_d1.rdata" ))
#Sample_Drug_list= colnames(TPM_2)%>%.[grepl("TRAB|DOX|IFO|D1",.)]



```



```{r}
# Define DEGs from RES data (padj<0.05& |FC|>1.2)

DEGs_list <- list()
for (i in names(Drug_reslist)) {
  Data <- Drug_reslist[[i]]
  Timepoint_list <- Data$RES  # Access the time points for this drug

  # Loop through each time point
  for (j in names(Timepoint_list)) {
    R1 <- Timepoint_list[[j]]

    # Loop through each replicate within the time point
    for (h in names(R1)) {
      data <- R1[[h]]

      # Filter significant genes and prepare ranked list
      DEGs <- data %>%
        as.data.frame() %>%
        dplyr::filter(padj < 0.05)%>%dplyr::filter(abs(log2FoldChange)>0.27)%>%mutate(gene_symbol=rownames(.))
    
      
      # Store FGSEA results in a list
   
      DEGs_list[[h]]=DEGs
    }
  }  
}

names(DEGs_list)
DEGs_list[[1]]
All_DEGs=do.call("rbind",DEGs_list)%>%as.data.frame()%>%.$gene_symbol%>%unique()
```

# Drug 1_Hallmark

## pathway enrichemant
```{r}
Result_list=list()
Plotlist=list()
Drug_name_list=names(Drug_reslist)
Pathway_key="Hallmark|HALLMARK"

for (I in Drug_name_list) {
 Drug_name=I
 Drug_fgsea=names(FGSEA_list)%>%.[grepl(Drug_name,.)]
 names(FGSEA_list)
 Select_data=list()
 for(i in Drug_fgsea){
  Select_data[[i]]=FGSEA_list[[i]]  }

 all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
 NES_sig_pathway_impute=list()
 for(i in names(Select_data)){
  Nes_data=Select_data[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  }
# samples=names(NES_sig_pathway_impute)%>%.[grepl("DMSO",.)]
 S_data=NES_sig_pathway_impute
 names(NES_sig_pathway_impute)=factor(names(NES_sig_pathway_impute),levels = rev(names(NES_sig_pathway_impute)))
 result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
 result$group=factor(result$group,levels = names(NES_sig_pathway_impute))
#remotes::install_github("YuLab-SMU/ggbreak")
 Plotlist[[I]]=Enrich_dotplot_update(result) 
 Result_list[[I]]=result
 
}

```



```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[1]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[2]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[3]]
```


```{r}
Filter_result=list()
for (i in names(Result_list)){
   data=Result_list[[i]]%>%dplyr::filter(grepl("DMSO_8h|DMSO_24h",group))
   Unique_name=data%>%dplyr::filter(padj<0.05)%>%distinct(pathway) 
   Filter_result[[i]]=data%>%dplyr::filter(pathway%in%Unique_name$pathway)%>%arrange(pathway)%>%as.data.frame()
  
}
 

Filter_result_all=do.call("rbind",Filter_result)%>%as.data.frame()%>%tidyr::separate(.,col=group,into=c("Drug","timepoint"),sep="_",remove=F)%>%arrange(Drug,desc(timepoint))

Filter_result_all$Group=factor(Filter_result_all$Group,levels=unique(Filter_result_all$Group))
```




```{r,fig.width=12,fig.height=16,include=T}
plot_pathway_heatmap(Filter_result_all, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

```



```{r,fig.width=18,fig.height=48,include=T}

Hallmark_pathways=names(GSEA_pathway_list)%>%.[grepl("HALLMARK",.)]
Pathway_name="HALLMARK_INTERFERON_GAMMA_RESPONSE"
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```


# Drug Wiki

## pathway enrichemant
```{r}
Result_list=list()
Plotlist=list()
Drug_name_list=names(Drug_reslist)
Pathway_key="WP_"

for (I in Drug_name_list) {
 Drug_name=I
 Drug_fgsea=names(FGSEA_list)%>%.[grepl(Drug_name,.)]
 names(FGSEA_list)
 Select_data=list()
 for(i in Drug_fgsea){
  Select_data[[i]]=FGSEA_list[[i]]  }

 all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
 NES_sig_pathway_impute=list()
 for(i in names(Select_data)){
  Nes_data=Select_data[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  }
# samples=names(NES_sig_pathway_impute)%>%.[grepl("DMSO",.)]
 S_data=NES_sig_pathway_impute
 names(NES_sig_pathway_impute)=factor(names(NES_sig_pathway_impute),levels = rev(names(NES_sig_pathway_impute)))
 result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
 result$group=factor(result$group,levels = names(NES_sig_pathway_impute))
#remotes::install_github("YuLab-SMU/ggbreak")
 Plotlist[[I]]=Enrich_dotplot_update(result) 
 Result_list[[I]]=result
 
}

```

```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[1]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[2]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[3]]
```


```{r}
Filter_result=list()
for (i in names(Result_list)){
   data=Result_list[[i]]%>%dplyr::filter(grepl("DMSO_8h|DMSO_24h",group))
   Unique_name=data%>%dplyr::filter(padj<0.05)%>%distinct(pathway) 
   Filter_result[[i]]=data%>%dplyr::filter(pathway%in%Unique_name$pathway)%>%arrange(pathway)%>%as.data.frame()
  
}
 

Filter_result_all=do.call("rbind",Filter_result)%>%as.data.frame()%>%tidyr::separate(.,col=group,into=c("Drug","timepoint"),sep="_",remove=F)%>%arrange(Drug,desc(timepoint))

Filter_result_all$Group=factor(Filter_result_all$Group,levels=unique(Filter_result_all$Group))
```


## Signal pahtway

```{r}
unique(Filter_result_all$pathway)
```


```{r,fig.width=12,fig.height=18,include=T}
Data1=Filter_result_all%>%dplyr::filter(grepl("SIGNALING|INTERFERON|IFNG|INTERLEUKIN",pathway))
plot_pathway_heatmap(Data1, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```
\newpage
```{r,fig.width=12,fig.height=16,include=T}
Hallmark_pathways=names(GSEA_pathway_list)%>%.[grepl("WP_TNF",.)]
Pathway_name="WP_TNFALPHA_SIGNALING_PATHWAY"
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```

\newpage

```{r,fig.width=12,fig.height=6,include=T}
Hallmark_pathways=names(GSEA_pathway_list)%>%.[grepl("PI3K",.)]
Pathway_name="WP_TYPE_I_INTERFERON_INDUCTION_AND_SIGNALING_DURING_SARSCOV2_INFECTION"
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```
\newpage

```{r,fig.width=12,fig.height=30,include=T,}
Pathway_name="WP_PI3KAKT_SIGNALING_PATHWAY" 
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```


```{r,fig.width=12,fig.height=4}
INF_IL_gene_TPM=TPM_cor_drug%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(grepl("^IL|IFN",gene_symbol))%>%dplyr::filter(gene_symbol%in%All_DEGs)
genes=INF_IL_gene_TPM$gene_symbol
Pathway_name="Interleukin and Interferon genes"

#Col_anno=Colour_frame %>%dplyr::filter(Sample_ID==colnames(TPM_cor_drug))


```

\newpage
```{r,fig.width=12,fig.height=8,include=T}
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )

```


```{r,fig.width=12,fig.height=14,include=T}
Data2=Filter_result_all%>%dplyr::filter(grepl("CYCLE|GROWTH|MIGRATION|G2|G1|_S_|APOPTOSIS|ARREST|PATHWAY",pathway))%>%dplyr::filter(!pathway%in%Data1$pathway)
plot_pathway_heatmap(Data2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```



```{r,fig.width=12,fig.height=6,include=T}
Data3=Filter_result_all%>%dplyr::filter(grepl("GLYCOLYSIS|OXIDATIVE|TCA|METABOLISM|PYRUVATE|PENTOSE|FATTY|OXIDATION|MITOCHONDRIA",pathway))%>%dplyr::filter(!pathway%in%Data1$pathway)%>%dplyr::filter(!pathway%in%Data2$pathway)
plot_pathway_heatmap(Data3, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```


```{r,fig.width=12,fig.height=6,include=T}
Data4=Filter_result_all%>%dplyr::filter(grepl("CANCER|CARCINOMA",pathway))%>%dplyr::filter(!pathway%in%Data1$pathway)%>%dplyr::filter(!pathway%in%Data2$pathway)%>%dplyr::filter(!pathway%in%Data3$pathway)
plot_pathway_heatmap(Data4, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```





```{r}
knitr::knit_exit()
```


# Drug wikipathway

## pathway enrichemant
```{r}
Plotlist=list()
Pathway_key="WP_"
Drug_name_list=names(Drug_reslist)
for (I in Drug_name_list) {
 Drug_name=I
 Drug_fgsea=names(FGSEA_list)%>%.[grepl(Drug_name,.)]
 names(FGSEA_list)
 Select_data=list()
 for(i in Drug_fgsea){
  Select_data[[i]]=FGSEA_list[[i]]
  
 }

 all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
 NES_sig_pathway_impute=list()
 for(i in names(Select_data)){
  Nes_data=Select_data[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}
# samples=names(NES_sig_pathway_impute)%>%.[grepl("DMSO",.)]
 S_data=NES_sig_pathway_impute
 names(NES_sig_pathway_impute)=factor(names(NES_sig_pathway_impute),levels = rev(names(NES_sig_pathway_impute)))

 
 result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
 result$group=factor(result$group,levels = names(NES_sig_pathway_impute))
#remotes::install_github("YuLab-SMU/ggbreak")
 Plotlist[[I]]=Enrich_dotplot_update(result) 

 
}

```

```{r,fig.width=16,fig.height=18,include=T}
Plotlist[[1]]
```



\newpage
```{r,fig.width=16,fig.height=18,include=T}
Plotlist[[2]]
```


```{r,fig.width=16,fig.height=18,include=T}
Plotlist[[3]]
```


```{r}
knitr::knit_exit()
```


```{r,fig.width=12,fig.height=18,include=T}
Pathway_key="REACTOME"
result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
result$group=factor(names(NES_sig_pathway_impute),levels = names(NES_sig_pathway_impute))
P3=Enrich_dotplot(result)

P3
```


```{r,include=F}

out=list()

for (i in names(Drug_reslist)){
  Data=Drug_reslist[[i]]
  Sample_list=Data$Counts%>%colnames(.)
  Genes=rownames(Data$RES[[1]][[1]])
  TPM_pickup=TPM_counts%>%dplyr::filter(gene_name%in%Genes)
  TPM_pickup_2=TPM_pickup%>%.[,colnames(.) %in%Sample_list]
  rownames(TPM_pickup_2)=TPM_pickup$gene_name
  Annotion_file=Data$Anno%>%arrange(Timepoint,Drug)%>%dplyr::filter(Sample_name%in%colnames(TPM_pickup_2))
  Annotion_file_2=Annotion_file%>%.[,c("Timepoint","Drug")]
  TPM_pickup_3=TPM_pickup_2%>%as.data.frame()%>%.[Annotion_file$Sample_name] 
  rownames(TPM_pickup_3)=rownames(TPM_pickup_2)
  Color_Set1=RColorBrewer::brewer.pal(5,"Set1")[1:2]
  names(Color_Set1)=unique(Annotion_file$Drug)
  Color_Set2=RColorBrewer::brewer.pal(5,"Set1")[2:4]
  names(Color_Set2)=unique(Annotion_file$Timepoint)
 
  Color_list=list(Drug=Color_Set1,Timepoint=Color_Set2)
  number= Annotion_file_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
 out[[i]]=TPM_heatmap(TPM_pickup_3,MAPK_genes,Annotation_col=Annotion_file_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="MAPK genes")%>%.$gtable

}

```

```{r,fig.width=12,fig.height=8,include=T}
cowplot::plot_grid(plotlist=out)
```

# Wiki, Hallmark, KEGG, Reactome Pathways in Mirda 8 and 24h ("Activated": p<0.05 and NES>0, "Inhibited":p<0.05 and NES<0,



```{r}
names(FGSEA_list)

Select_data=list(SEL_8h_vs_DMSO_8h=FGSEA_list[["SEL_8h_vs_DMSO_8h"]],
           SEL_24h_vs_DMSO_24h=FGSEA_list[["SEL_24h_vs_DMSO_24h"]],
         TRA_8h_vs_DMSO_8h=FGSEA_list[["TRA_8h_vs_DMSO_8h"]],
        TRA_24h_vs_DMSO_24h=FGSEA_list[["TRA_24h_vs_DMSO_24h"]])

all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}
```

```{r}
Data

```


# SEL
## MAPK pathway
```{r}
SEL_data=NES_sig_pathway_impute[c("SEL_8h_vs_DMSO_8h" ,"SEL_24h_vs_DMSO_24h")]
Pathway_key="MAPK"
result <- Common_sig_pathway_NES(SEL_data, Pathway_key = Pathway_key)

P2=Enrich_dotplot(result)
```


```{r,fig.width=12,fig.height=5,include=T}
P2
```

```{r,fig.width=24,fig.height=12,include=T}
library(pathview)
library(KEGGREST)
kegg_list <- KEGGREST::keggList("pathway", "hsa")
KEGG_sig=kegg_list%>%.[grep("MAPK|Mapk|mapk",.)]
```

# SEL 8h vs DMOS 8h
```{r,fig.width=24,fig.height=12,include=T}
Drug_name="SEL"
Data=Drug_reslist[[Drug_name]]
Sample_list=Data$Counts%>%colnames(.)
Time_point="8h"
Sample_name="SEL_8h_vs_DMSO_8h"
Res_1=Data$RES[[Time_point]][[Sample_name]]%>%as.data.frame()%>%mutate(Group=Sample_name)%>%mutate(gene_symbol=rownames(.))

test_data_1 <-Res_1%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.27) # Example: Use one of your T-test results
# Run the pathview analysis
Pathway_ID=names(KEGG_sig)
Pathway_name=as.character( KEGG_sig)%>%strsplit(.,"-")%>%unlist()%>%.[1]%>%gsub(" $","",.)
pathview_out=file.path(".","Pathview")
Outname=paste0(Sample_name,"_",Pathway_name,".png")%>%gsub(" ","_",.)
```

```{r}
#https://bioconductor.org/packages/devel/bioc/vignettes/SBGNview/inst/doc/SBGNview.Vignette.html#53_Visualize_both_gene_data_and_compound_data
```


```{r,fig.width=24,fig.height=12,include=T}
pathview_result <- run_pathview_gene_symbol(test_data_1, pathway_name=Pathway_name, pathway_id=Pathway_ID, output_dir=pathview_out,outname=Outname)

```


```{r, out.width='100%', fig.align='center',include=T}
png_file=file.path(pathview_out,Outname)
knitr::include_graphics(png_file)
```

# SEL 24h vs DMOS 24h
```{r,fig.width=24,fig.height=12,include=T}
Drug_name="SEL"
Data=Drug_reslist[[Drug_name]]
Sample_list=Data$Counts%>%colnames(.)
Time_point="24h"
Sample_name="SEL_24h_vs_DMSO_24h"
Res_2=Data$RES[[Time_point]][[Sample_name]]%>%as.data.frame()%>%mutate(Group=Sample_name)%>%mutate(gene_symbol=rownames(.))

test_data_2<-Res_2%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.27)
```


```{r,fig.width=24,fig.height=12,include=T}
# Run the pathview analysis
Pathway_ID=names(KEGG_sig)
Pathway_name=as.character( KEGG_sig)%>%strsplit(.,"-")%>%unlist()%>%.[1]%>%gsub(" $","",.)
pathview_out=file.path(".","Pathview")
Outname=paste0(Sample_name,"_",Pathway_name,".png")%>%gsub(" ","_",.)
pathview_result <- run_pathview_gene_symbol(test_data_2, pathway_name=Pathway_name, pathway_id=Pathway_ID, output_dir=pathview_out,outname=Outname)

```

```{r, out.width='100%', fig.align='center',include=T}
png_file=file.path(pathview_out,Outname)
knitr::include_graphics(png_file)
```

## Combine
```{r}
Outname=paste0(Drug_name,"_",Pathway_name,".png")%>%gsub(" ","_",.)
```

```{r}
run_pathview_group_gene_symbol(
  test_data1 = test_data_1,
  test_data2 = test_data_2,
  pathway_name=Pathway_name, pathway_id=Pathway_ID, output_dir=pathview_out,outname=Outname)

 

```
```{r, out.width='100%', fig.align='center',include=T}
png_file=file.path(pathview_out,Outname)
knitr::include_graphics(png_file)
```


```{r}
knitr::knit_exit()
```



```{r}

SEL_data_pathway=lapply(SEL_data, function(x) x%>%dplyr::filter(grepl("AKT|ERK|MAPK_",pathway)))

NES_sig_pathway_pikcup=list()
all_sig_pathway_name=lapply(SEL_data_pathway,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
for(i in names(SEL_data)){
  Nes_data=SEL_data[[i]]
  NES_sig_pathway_pikcup[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

NES_sig_pathway_pikcup
```



```{r}
knitr::knit_exit()
```



```{r}
if (!grepl("Pathwaylist_sig_all", fileexist)){
 FGSEA_list=list()
 FGSEA_q005_list=list()
 for (i in 1:length(Res_list)){
  data=Res_list[[i]]
  res_q005=data%>%dplyr::filter(padj<0.05)
  name2=names(Res_list)[[i]]
  res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
  ranks <- tibble::deframe(res_q005_rank)
  out_list=list()
  for (x in 1:length(GMT_files)){
 
     data=fgsea::gmtPathways(GMT_files[x])
    set.seed(12345)
    out_list[[x]] <- fgsea::fgseaMultilevel(data, ranks,
                           minSize=10,
                           maxSize=500,nPermSimple = 10000)
  }
                           
  FGSEA_list[[name2]]=do.call("rbind",out_list)
  FGSEA_q005_list[[name2]]=do.call("rbind",out_list)%>%as.data.frame()%>%dplyr::filter(padj<0.05)

}

all_sig_pathway_name=lapply(FGSEA_q005_list,function(x) x$pathway )%>%unlist()%>%unique()


NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

 Pathwaylist_sig_all=do.call("rbind",NES_sig_pathway_impute)
 Pathwaylist_sig_all$group=factor(Pathwaylist_sig_all$group,c(names(NES_sig_pathway_impute)))
 
 if (!file.exists(file.path(Rawdata_path,File_name1 ))) {
    save(Pathwaylist_sig_all,FGSEA_q005_list,file=file.path(File_name1,File_name1 ))
   }
  
}

```




# Use DMSO group2 as control

```{r}
load(file.path(Process_path,"Deseq2_OUT_vs_DMSO_d2.rdata"))
```


```{r,include=F}


out=list()

for (i in names(Drug_reslist)){
  Data=Drug_reslist[[i]]
  Sample_list=Data$Counts%>%colnames(.)
  Genes=rownames(Data$RES[[1]][[1]])
  TPM_pickup=TPM_counts%>%dplyr::filter(gene_name%in%Genes)
  TPM_pickup_2=TPM_pickup%>%.[,colnames(.) %in%Sample_list]
  rownames(TPM_pickup_2)=TPM_pickup$gene_name
  Annotion_file=Data$Anno%>%arrange(Timepoint,Drug)%>%dplyr::filter(Sample_name%in%colnames(TPM_pickup_2))
  Annotion_file_2=Annotion_file%>%.[,c("Timepoint","Drug")]
  TPM_pickup_3=TPM_pickup_2%>%as.data.frame()%>%.[Annotion_file$Sample_name] 
  rownames(TPM_pickup_3)=rownames(TPM_pickup_2)
  Color_Set1=RColorBrewer::brewer.pal(5,"Set1")[1:2]
  names(Color_Set1)=unique(Annotion_file$Drug)
  Color_Set2=RColorBrewer::brewer.pal(5,"Set1")[2:4]
  names(Color_Set2)=unique(Annotion_file$Timepoint)
 
  Color_list=list(Drug=Color_Set1,Timepoint=Color_Set2)
  number= Annotion_file_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
 out[[i]]=TPM_heatmap(TPM_pickup_3,MAPK_genes,Annotation_col=Annotion_file_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="MAPK genes")%>%.$gtable

}

```

```{r,fig.width=12,fig.height=8,include=T}
cowplot::plot_grid(plotlist=out)
```


```{r}
knitr::knit_exit()
```

```{r}
TPM_heatmap(TPM_pickup_2,MAPK_genes,Annotation_col=Annotion_file,Annotation_color=Color_list,Title="MAPK genes")
```

```{r,include=F,fig.width=8,fig.height=5}
set.seed(12345)
Sample_anno_2$Drug=factor(Sample_anno_2$Drug,levels = c("DMSO","Mirda"))
VST_rm=t(deseq2VST)%>%as.data.frame()%>%.[rownames(.)!="sample.8hM2", ]%>%.[!grepl("0", rownames(.) ),]
PCA_2=PCA_Plot_3(VST_rm,Sample_anno_2,VAR="Sample_name","Drug","Timepoint")
```


```{r,include=T,fig.width=5,fig.height=5}

cowplot::plot_grid(PCA_2,labels=c("PCA plot"))
```

```{r,include=T}
# Sample_anno=data.frame(Sample_name=colnames(Raw_counts_2_flipped))%>%mutate(Drug=ifelse(grepl("D",Sample_name),"DMSO", ifelse(grepl("M",Sample_name),"Mirda","DMSO") ))%>%mutate(Timepoint=ifelse(grepl("0h",Sample_name),"0hour", ifelse(grepl("8h",Sample_name),"8hour","24hour") ))%>%mutate(Group=paste0(Drug,"_",Timepoint))%>%mutate(Timepoint=factor(.$Timepoint,levels = c("0hour","8hour","24hour")))%>%arrange(Timepoint,Drug)
# #Timepoint=factor(Sample_anno$Timepoint,levels = c("0hour","8hour","24hour"))
# 
# 
# knitr::kable(Sample_anno,caption="Flipped Sample annotation")%>%kableExtra::kable_styling()
```

# DEGs in Mirda 8h vs  DMSo 8h (3396 "padj<0.05,|FC|>1.2"); DEGs in Mirda 24h vs  DMSo 24h (3309 "padj<0.05,|FC|>1.2")
```{r}
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt, colData = Sample_anno,design =~Group)
dds= DESeq2::DESeq(deseq2Data)

Mirda_8h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_8hour","DMSO_8hour"))%>%as.data.frame()
Mirda_8h_vs_DMSO_sig=Mirda_8h_vs_DMSO%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_8h_vs_DMSO_sig)
Mirda_24h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_24hour","DMSO_24hour"))%>%as.data.frame()
Mirda_24h_vs_DMSO_sig=Mirda_24h_vs_DMSO%>%dplyr::filter(padj<0.05)%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_24h_vs_DMSO_sig)
```

# Use DEGs for pathway analysis (3396 at 8h and 3309 at 24h)

```{r,include=F,fig.width=8,fig.height=6}

Mirda_8h_vs_DMSO_sig_up=Mirda_8h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange>0)
Mirda_8h_vs_DMSO_sig_DN=Mirda_8h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange<0)
Mirda_24h_vs_DMSO_sig_up=Mirda_24h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange>0)
Mirda_24h_vs_DMSO_sig_DN=Mirda_24h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange<0)

set1 <-  rownames(Mirda_8h_vs_DMSO_sig_up)
set2 <-  rownames(Mirda_24h_vs_DMSO_sig_up)
x = list(set1, set2)
library("ggVennDiagram")
G1=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_up","Mirda_24h_up"),hjust=-0.5,set_color = "red",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))



set3 <-  rownames(Mirda_8h_vs_DMSO_sig_DN)
set4 <-  rownames(Mirda_24h_vs_DMSO_sig_DN)
x = list(set3, set4)
library("ggVennDiagram")
G2=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_DN","Mirda_24h_DN"),hjust=-0.5,set_color = "green",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))


```

```{r,fig.width=12,fig.height=6,include=T}
cowplot::plot_grid(G1,G2,labels=c("Upregulated_genes","Down_regulated_genes"))
```


```{r}
Res_list=list(Mirda_8h=Mirda_8h_vs_DMSO_sig,Mirda_24h=Mirda_24h_vs_DMSO_sig)
GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="all.v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("c2|c5|h",.)]
```

```{r}
if (file.exists(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))) {
 load(file.path(Rawdata_path,"2_Flipped_GSEA_data.rdata" ))
 #fileexist=cgwtools::lsdata(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))
} else{
  fileexist=c("NA")
  
}


```

# Wiki, Hallmark, KEGG, Reactome Pathways in Mirda 8 and 24h ("Activated": p<0.05 and NES>0, "Inhibited":p<0.05 and NES<0,
**"P_Activated":possible activated 0.07>p>0.05,NES>0; "P_Inhibited" possible inhibtied 0.07>p>0.05,NES<0; "no_change": P>0.07)**

```{r}
if (!grepl("Pathwaylist_sig_all", fileexist)){

 FGSEA_list=list()
 FGSEA_q005_list=list()
 for (i in 1:length(Res_list)){
  data=Res_list[[i]]
  res_q005=data%>%dplyr::filter(padj<0.05)
  name2=names(Res_list)[[i]]
  res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
  ranks <- tibble::deframe(res_q005_rank)
  out_list=list()
  for (x in 1:length(GMT_files)){
 
     data=fgsea::gmtPathways(GMT_files[x])
    set.seed(12345)
    out_list[[x]] <- fgsea::fgseaMultilevel(data, ranks,
                           minSize=10,
                           maxSize=500,nPermSimple = 10000)
  }
                           
  FGSEA_list[[name2]]=do.call("rbind",out_list)
  FGSEA_q005_list[[name2]]=do.call("rbind",out_list)%>%as.data.frame()%>%dplyr::filter(padj<0.05)

}

all_sig_pathway_name=lapply(FGSEA_q005_list,function(x) x$pathway )%>%unlist()%>%unique()


NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

 Pathwaylist_sig_all=do.call("rbind",NES_sig_pathway_impute)
 Pathwaylist_sig_all$group=factor(Pathwaylist_sig_all$group,c(names(NES_sig_pathway_impute)))
 
 if (!file.exists(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))) {
 save(Pathwaylist_sig_all,FGSEA_q005_list,file=file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))}
  



 }




```

```{r}
Pathwaylist_sig_all_pickup=Pathwaylist_sig_all%>%dplyr::filter(grepl("HALLMARK|WP_|KEGG|REACTOME",pathway ))

Pathwaylist_sig_all%>%dplyr::filter(grepl("MAPK",pathway))
unique(Pathwaylist_sig_all_pickup$group )


```

```{r}
Pathwaylist_sig_8_up=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES>0&group=="Mirda_8h"&padj<0.05)%>%arrange(desc(NES))
Pathwaylist_sig_8_DN=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES<0&group=="Mirda_8h"&padj<0.05)


Pathwaylist_sig_24_up=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES>0&group=="Mirda_24h"&padj<0.05)%>%arrange(desc(NES))
Pathwaylist_sig_24_DN=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES<0&group=="Mirda_24h"&padj<0.05)
```

```{r}
set1 <-  Pathwaylist_sig_8_up$pathway
set2 <-  Pathwaylist_sig_24_up$pathway
x = list(set1, set2)
library("ggVennDiagram")
G3=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_pathway_up","Mirda_24h_pathway_up"),hjust=-0.5,set_color = "red",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))



set3 <-  Pathwaylist_sig_8_DN$pathway
set4 <-  Pathwaylist_sig_24_DN$pathway
x = list(set3, set4)
library("ggVennDiagram")
G4=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_pathway_DN","Mirda_24h_pathway_DN"),hjust=-0.5,set_color = "green",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))

```


```{r,fig.width=16,fig.height=7,include=T}
cowplot::plot_grid(G3,G4,labels=c("upregulated pathways","Downregulated pathways"))
```


```{r}
set1 <-  Pathwaylist_sig_8_up$pathway
set2 <-  Pathwaylist_sig_24_up$pathway
Up_pathways=unique(set1,set2)
```


```{r}
set3 <-  Pathwaylist_sig_8_DN$pathway
set4 <-  Pathwaylist_sig_24_DN$pathway

DN_pathways=unique(set3,set4)
```


```{r}
## Pathway network
human_input_8h=Mirda_8h_vs_DMSO_sig%>%mutate(Gene.symbol=rownames(.)) %>%.[,c("Gene.symbol","log2FoldChange","padj")]%>%setNames(c("Gene.symbol","logFC","adj.P.Val"))%>%na.omit() 
 
human_input_8h_up=human_input_8h%>%dplyr::filter(logFC>0)
human_input_8h_DN=human_input_8h%>%dplyr::filter(logFC<0)

human_input_24h=Mirda_24h_vs_DMSO_sig%>%mutate(Gene.symbol=rownames(.)) %>%.[,c("Gene.symbol","log2FoldChange","padj")]%>%setNames(c("Gene.symbol","logFC","adj.P.Val"))%>%na.omit() 
 
human_input_24h_up=human_input_24h%>%dplyr::filter(logFC>0)
human_input_24h_DN=human_input_24h%>%dplyr::filter(logFC<0)

```

```{r}




```


```{r}
Term_gene_graph_2=function (result_df, num_terms = 10, layout = "stress", use_description = FALSE, 
      node_size = "num_genes", node_colors = c("#E5D7BF", "green", "red", "grey")) {
  
  # for single plot
  
  if (!is.numeric(num_terms) & !is.null(num_terms)) {
    stop("`num_terms` must either be numeric or NULL!")
  }
  if (!is.logical(use_description)) {
    stop("`use_description` must either be TRUE or FALSE!")
  }
  ID_column <- ifelse(use_description, "Term_Description", 
                      "ID")
  val_node_size <- c("num_genes", "p_val")
  if (!node_size %in% val_node_size) {
    stop("`node_size` should be one of ", paste(dQuote(val_node_size), 
                                                collapse = ", "))
  }
  if (!is.data.frame(result_df)) {
    stop("`result_df` should be a data frame")
  }
  necessary_cols <- c(ID_column, "lowest_p", "Up_regulated", 
                      "Down_regulated", "no_change","group")  # Include "no_change"
  if (!all(necessary_cols %in% colnames(result_df))) {
    stop(paste(c("All of", paste(necessary_cols, collapse = ", "), 
                 "must be present in `results_df`!"), collapse = " "))
  }
  if (length(node_colors) != 4) {  # Corrected length check
    stop("`node_colors` must contain exactly 4 colors")  # Updated error message
  }
  if (!is.null(num_terms)) {
    if (nrow(result_df) < num_terms) {
      num_terms <- NULL
    }
  }
  result_df<- result_df[order(result_df$Term_Description, decreasing = FALSE),  ]
  if (!is.null(num_terms)) {
        result_df <- result_df[1:num_terms, ]
    }
    graph_df <- data.frame()
  for (i in base::seq_len(nrow(result_df))) {
        up_genes <- unlist(strsplit(result_df$Up_regulated[i], 
            ", "))
        down_genes <- unlist(strsplit(result_df$Down_regulated[i], 
            ", "))
        
        no_change= unlist(strsplit(result_df$no_change[i], 
            ", "))
        
        for (gene in c(up_genes, down_genes,no_change)) {
            graph_df <- rbind(graph_df, data.frame(Term = result_df[i, 
                ID_column], Gene = gene))
        }
    }
  up_genes <- unlist(lapply(result_df$Up_regulated, function(x) strsplit(x, ", ")))
  DN_genes=  unlist(lapply(result_df$Down_regulated, function(x) strsplit(x, ", ")))
  g <- igraph::graph_from_data_frame(graph_df, directed = FALSE)
  cond_term <- names(igraph::V(g)) %in% result_df[, ID_column]
  cond_up_gene <- names(igraph::V(g)) %in% up_genes
  cond_down_gene=  names(igraph::V(g)) %in% DN_genes
  node_type <- ifelse(cond_term, "term", 
                      ifelse(cond_up_gene, "up", 
                             ifelse(cond_down_gene, "down", "no_change")))
  node_type <- factor(node_type, levels = c("term", "up", "down", "no_change"))
  node_type <- factor(node_type, levels = c("term", "up", "down","no_change"))
    igraph::V(g)$type <- node_type
  type_descriptions <- c(term = "enriched term", up = "up-regulated gene", 
                         down = "down-regulated gene", no_change = "no_change gene")  # Include "no_change" description
   type_descriptions <- type_descriptions[levels(node_type)]
   names(node_colors) <- c("term", "up", "down", "no_change")  # Added "no_change"
   node_colors <- node_colors[levels(node_type)]
    if (node_size == "num_genes") {
        sizes <- igraph::degree(g)
        sizes <- ifelse(igraph::V(g)$type == "term", sizes, 2)
        size_label <- "# genes"
    }
    else {
        idx <- match(names(igraph::V(g)), result_df[, ID_column])
        sizes <- -log10(result_df$lowest_p[idx])
        sizes[is.na(sizes)] <- 2
        size_label <- "-log10(p)"
    }
     
    
    igraph::V(g)$size <- sizes
    igraph::V(g)$label.cex <- 0.5
    igraph::V(g)$frame.color <- "gray"
    
    p <- ggraph::ggraph(g, layout = layout)
    
    xy <- graphlayouts::layout_with_stress(g)
    p <- ggraph::ggraph(g, layout = xy)
    
    p <- p + ggraph::geom_edge_link(alpha = 0.8, colour = "darkgrey")
    p <- p + ggraph::geom_node_point(ggplot2::aes(color = .data$type, 
        size = .data$size))
    p <- p + ggplot2::scale_size(range = c(5, 10), breaks = round(seq(round(min(igraph::V(g)$size)), 
        round(max(igraph::V(g)$size)), length.out = 4)), name = size_label)
    p <- p + ggplot2::theme_void()
    p <- p + suppressWarnings(ggraph::geom_node_text(ggplot2::aes(label = .data$name), 
        nudge_y = 0.2, repel = TRUE, max.overlaps = 20))
    p <- p + ggplot2::scale_color_manual(values = node_colors, 
        name = NULL, labels = type_descriptions)
    if (is.null(num_terms)) {
        p <- p + ggplot2::ggtitle("Term-Gene Graph")
    }
    else {
        p <- p + ggplot2::ggtitle("Term-Gene Graph", subtitle = paste(c("Top", 
            num_terms, "terms"), collapse = " "))
    }
    p <- p + ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5), 
        plot.subtitle = ggplot2::element_text(hjust = 0.5))
    return(p)
}

```


```{r,fig.width=24,fig.height=8,include=T}
# NES plot
Both_up=intersect(set1,set2)
Both_DN=intersect(set3,set4)

Select_path=Pathwaylist_sig_all%>%dplyr::filter( pathway%in%c(Both_up,Both_DN))
Output_both=Enrich_dotplot(Select_path)
Output_both
```

```{r,fig.height=6,fig.width=12,include=T}
Pathway_pickup=c(Both_up,Both_DN)
Pathway_list=GSEA_pathway_list[Pathway_pickup]
Pathway_singscore= singcore_GSEA(Flip_TPM_2_arrange,Pathway_list)%>%t()%>%as.data.frame()
TITLE="Significant Pathway"
Keypahtway_singscore=paste0(names(Pathway_list),"_singscore")

GSEA_heatmap(Pathway_singscore,Flip_TPM_2_anno_2,Color_list,Title = TITLE)
```

## 24h UP and Down

```{r,fig.width=18,fig.height=10,include=T}
Pathway_get=c(setdiff(set4,set3), setdiff(set2,set1))%>%.[!grepl("REACTOME",.)]%>%.[grepl("COMPLEMENT|SIGNALING|INTEFERIN|APOPTOSIS",.)]

Select_path=Pathwaylist_sig_all%>%dplyr::filter( pathway%in%Pathway_get)
Output=Enrich_dotplot(Select_path)
Output
```
```{r,fig.height=6,fig.width=12,include=T}
Pathway_pickup=Pathway_get
Pathway_list=GSEA_pathway_list[Pathway_pickup]
Pathway_singscore= singcore_GSEA(Flip_TPM_2_arrange,Pathway_list)%>%t()%>%as.data.frame()
TITLE="Significant Pathway"
Keypahtway_singscore=paste0(names(Pathway_list),"_singscore")

GSEA_heatmap(Pathway_singscore,Flip_TPM_2_anno_2,Color_list,Title = TITLE)
```


```{r}
knitr::knit_exit()
```


```{r}

# Raw_counts_add_control=Raw_counts_2%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)
# Sample_anno_2=data.frame(Sample_name=colnames(Raw_counts_add_control))%>%left_join(.,Sample_anno)%>%mutate(Drug=ifelse(is.na(Drug),"Mirda",Drug))%>%mutate(Timepoint=ifelse(is.na(Timepoint),"0hour",Timepoint))%>%mutate(Group=paste0(Drug,"_",Timepoint))
# rownames(Sample_anno_2)=Sample_anno_2$Sample_name
# Sample_anno_2$Timepoint=factor(Sample_anno_2$Timepoint,levels = c("0hour","8hour","24hour"))
# library(SummarizedExperiment)
# library(DESeq2)
#   keep.exprs <- selectGenes_BY_group(Raw_counts_add_control, min.count=10, N=0.90,Sample_anno_2)
# keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
# myFilt <- Raw_counts_add_control[rownames(keep.exprs_2),]
# 
# dim(myFilt)

library(SummarizedExperiment)
library(DESeq2)
keep.exprs <- selectGenes_BY_group(Raw_counts_2_flipped, min.count=10, N=0.90,Sample_anno)
keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
myFilt <- Raw_counts_2[rownames(keep.exprs_2),]

dim(myFilt)
```


```{r}
# Flipping TPM data
Flip_TPM_2=TPM_counts[,c(2:17)]%>%dplyr::filter(gene_name%in%rownames(myFilt) )%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()

colnames(Flip_TPM_2)=gsub("D","DC",colnames(Flip_TPM_2))%>%gsub("M","D",.)%>%gsub("DC","M",.)


Flip_TPM_2_anno=data.frame(Sample_name=colnames(Flip_TPM_2))%>%left_join(., Sample_anno)%>%arrange(Timepoint,Drug)

Flip_TPM_2_arrange=Flip_TPM_2[,Flip_TPM_2_anno$Sample_name]

Flip_TPM_2_anno_2=Flip_TPM_2_anno["Group"]
rownames(Flip_TPM_2_anno_2)=Flip_TPM_2_anno$Sample_name
unique(rownames(Flip_TPM_2_anno_2))
Color_Set=RColorBrewer::brewer.pal(5,"Set1")
names(Color_Set)=unique(Flip_TPM_2_anno_2$Group)
Color_list=list(Group=Color_Set)
```


```{r}
# Time course analyses with LRT
#The LRT test can be especially helpful when performing time course analyses. We can use the LRT to explore whether there are any significant differences in treatment effect between any of the timepoints.**

#For have an experiment looking at the effect of treatment over time on mice of two different genotypes. We could use a design formula for our ‘full model’ that would include the major sources of variation in our data: genotype, treatment, time, and our main condition of interest, which is the difference in the effect of treatment over time (treatment:time).**
# #https://rstudio-pubs-static.s3.amazonaws.com/329027_593046fb6d7a427da6b2c538caf601e1.html
# deseq2Data_add_control=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt , colData = Sample_anno_2,design = ~Drug+Timepoint+Drug:Timepoint)
# deseq2Data_add_control$Timepoint=relevel(deseq2Data_add_control$Timepoint,"0hour")
# dds_LRT= DESeq2::DESeq(deseq2Data_add_control,test="LRT", reduced =~Drug+Timepoint)
# DESeq2::resultsNames(dds_LRT)

deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =Raw_counts_2_flipped, colData = Sample_anno,design = ~Timepoint)
deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
```

```{r}
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt, colData = Sample_anno,design =~Group)
dds= DESeq2::DESeq(deseq2Data)
# DEGs: FC 1.2 and padj>0.05
Mirda_8h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_8hour","DMSO_8hour"))%>%as.data.frame()
Mirda_8h_vs_DMSO_sig=Mirda_8h_vs_DMSO%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_8h_vs_DMSO_sig)
Mirda_24h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_24hour","DMSO_24hour"))%>%as.data.frame()
Mirda_24h_vs_DMSO_sig=Mirda_24h_vs_DMSO%>%dplyr::filter(padj<0.05)%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_24h_vs_DMSO_sig)
```

```{r}

```




# Save data
```{r}
if (!file.exists(file.path(Rawdata_path,"Flipped_Deseq2_output.rdata"))){
save(myFilt,Sample_anno_2,deseq2VST,dds_LRT,file=file.path(Rawdata_path,"Flipped_Deseq2_output.rdata"))
}
```

```{r}
knitr::knit_exit()
```


