---
title: "5_Drug_PCA_abnormal_samples"
author: "`r config::get(file = 'config.yml')$document$author`"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
    fig_caption: yes
    highlight: haddock
    number_sections: yes

knit: (
  function(inputFile, encoding) {
    config=config::get(file = 'config.yml') 
    
    pSubTitle <- "5_Drug_PCA_abnormal_samples"
    
    base_output_dir <- paste0(config$Path,"/","Report")
    
    DIR=file.path(base_output_dir)
    
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      output_file = pSubTitle,
      output_dir =DIR )}
    )
---



```{r,include=F}
rm(list=ls())
```

```{r,include=F}
library(dplyr)
library(ggplot2)
library(synapser)
library(dplyr)
library(readr)
library(tidyr)

```

```{r}
dir()
source(file.path(getwd(),"Data_input.R"))
```




```{r}
# RES and GSEA data download
load(file.path(Processed_path,"All_drug_RES_FGSEA.rdata"))
```

```{r}
Drug_names=Drug_anno$Full_name%>%unique()%>%na.omit()

Gene_sig_path=file.path(Reference_path,"Gene_signature")
dir(Gene_sig_path)
# RB loss of function signature
RB_sig_genes=read.csv(file.path(Gene_sig_path,"RB_LOF_sig.csv"),header = T)%>%.[,1]
```


```{r}
GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="all.v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("h|c2|c4|c5|c6|c7|c8",.)]

pathways_genes <- fgsea::gmtPathways(GMT_files[1])

for (i in 2:length(GMT_files)){
  pathways2 <- fgsea::gmtPathways(GMT_files[i])
  pathways_genes =append(pathways_genes ,pathways2)
}
                            
if(!file.exists(file.path(Processed_path,"pathways_genes.rdata"))){
  All_pathway_singscore=singcore_GSEA(TPM_all,pathways_genes)   

  save(All_pathway_singscore,file=file.path(Processed_path,"pathways_genes.rdata"))
} else {
  load(file.path(Processed_path,"pathways_genes.rdata"))
  
}


# combine the data from all pathways singscore that has same row number with col number of TPM_all



```


```{r}
Annotation_all_rm_0h=Annotation_all%>%dplyr::filter(Timepoint!="0h")

Singscore_ttest=perform_drug_vs_DMSO_tests(singscore_data = All_pathway_singscore, annotation_table = Annotation_all_rm_0h)
singscore_ttest_p005 <- Singscore_ttest %>% dplyr::filter(p_value < 0.05)%>%arrange(p_value)
Dataset=c("HALLMARK","WIKI","KEGG","REACTOME")

singscore_ttest_p005_filter=singscore_ttest_p005%>%dplyr::filter(grepl("HALLMARK|REACTOME|KEGG|WP_",pathway))
singscore_ttest_p005_filter$pathway

names(Singscore_ttest)
```

# Mirda in Batch1
```{r}
Drug_name="MIR"

Input=singscore_ttest_p005_filter%>%dplyr::filter(grepl(Drug_name,pathway))

#Input=Singscore_ttest%>%dplyr::filter(Group%in%c("PAL_8h", "PAL_24h"))

Drug_Samples=Annotation_all%>%dplyr::filter(Drug==Drug_name)
DMSO_Samples=Annotation_all%>%dplyr::filter(Drug=="DMSO"&Date==Drug_Samples$Date[1])
TPM_filter=TPM_all[,colnames(TPM_all)%in%c(Drug_Samples$specimenID,DMSO_Samples$specimenID)]

#%>%.$specimenID
#combined_sig <- extract_and_merge_significant_pathways(Input)
 
#colnames(combined_sig)
#Test=prepare_dotplot_data(combined_sig) %>%mutate(Timepoint=factor(Timepoint,levels = c("8h","24h")))

#P1=plot_singscore_dot(Test)


# use these keywords to filter the pathway
#keyword_pattern <- paste(palbociclib_pathway_keywords, collapse = "|") %>% gsub("\\|$", "", .)
# Filter FGSEA results using grepl (case-insensitive matching)
#filtered_singscore <- Input %>% dplyr::filter(grepl(keyword_pattern, pathway, ignore.case = TRUE)) %>% arrange(Group)
```

```{r,fig.width=22,fig.height=16,include=T}
PCA_Plot_3_shape(t(TPM_filter), Annotation_all, "specimenID", "Drug", "Timepoint") + 
  scale_color_manual(values = drug_colors) +
  scale_shape_manual(values = c(16, 17, 18)) +
  labs(title = paste0("PCA plot for ", Drug_name), x = "PC1", y = "PC2") +
  theme_bw() + 
   theme(legend.position = "top")+geom_text(aes(label = specimenID),hjust = 1, vjust = 0, size = 3, angle = 45) 

```

```{r}
Plist <- list()
Drugs <- Annotation_all |>
  filter(Drug != "MIR", Drug != "DMSO")
unique(Drugs$Drug_class)
for (i in unique(Drugs$Drug_class)) {
  Drug_choose <- Drugs |>
    filter(Drug_class == i) |>
    pull(Drug) |>
    na.omit()
  
  
  Plist2 <- list()
  for (j in Drug_choose) {
    Drug_name <- j
    
    # Get sample IDs for the drug
    Drug_Samples <- Annotation_all |>
      filter(Drug == Drug_name)
    
  
    
    specimen_ID <- Drug_Samples$specimenID
    
    # Get matched DMSO samples (same date)
    DMSO_Samples <- Annotation_all |>
      filter(Drug == "DMSO", Date == Drug_Samples$Date[1],Timepoint!="0h")
    
    # Filter TPM data to relevant columns
    TPM_data <- TPM_all[, colnames(TPM_all) %in% c(specimen_ID, DMSO_Samples$specimenID)]
    
    # Prepare annotations for PCA
    Annotation_data <- Annotation_all |>
      filter(specimenID %in% colnames(TPM_data))
    
    # Generate PCA plot
    Plist2[[j]] <-  PCA_Plot_3_shape(t(TPM_data), Annotation_all, "specimenID", "Drug", "Timepoint") + 
  
  #scale_color_manual(values = drug_colors) +
  scale_shape_manual(values = c(16, 17, 18)) +
  scale_color_manual(values = drug_colors) +
  labs(title = paste0(i,":", Drug_name), x = "PC1", y = "PC2") +
  theme_bw() +
  theme(legend.position = "top")+geom_text(aes(label = specimenID),hjust = 1, vjust = -1, size = 6, angle = 45) 
  }
  
  # Combine individual drug plots into one for each drug class
  Plist[[i]] <- cowplot::plot_grid(plotlist = Plist2, nrow = 1)
}
```
# Drugs in batch2
## cMET inhitor,DNMTi,PARPi

```{r,fig.width=24,fig.height=16,include=T}
list_data=list(Plist[[1]],Plist[[2]],Plist[[4]])
cowplot::plot_grid(plotlist = list_data, nrow = 1)
```


## Chemo
```{r,fig.width=24,fig.height=16,include=T}
Plist[[3]]
```


# CDK4/6 inhibitors
```{r,fig.width=18,fig.height=12,include=T}
Plist[[5]]
```

# SHP2 inhibitor
```{r,fig.width=18,fig.height=12,include=T}
Plist[[6]]
```

# MEKi
```{r,fig.width=18,fig.height=12,include=T}
Plist[[7]]
```

```{r,fig.width=8,fig.height=6,include=T}
Plist[[8]]
```



```{r}
knitr::knit_exit()
```


```{r}
Drug_name="PAL"
#Input=Singscore_ttest%>%dplyr::filter(Group%in%c("PAL_8h", "PAL_24h"))

Drug_Samples=Annotation_all%>%dplyr::filter(Drug==Drug_name)
DMSO_Samples=Annotation_all%>%dplyr::filter(Drug=="DMSO"&Date==Drug_Samples$Date[1])
TPM_filter=TPM_all[,colnames(TPM_all)%in%c(Drug_Samples$specimenID,DMSO_Samples$specimenID)]

#%>%.$specimenID
#combined_sig <- extract_and_merge_significant_pathways(Input)
 
#colnames(combined_sig)
#Test=prepare_dotplot_data(combined_sig) %>%mutate(Timepoint=factor(Timepoint,levels = c("8h","24h")))

#P1=plot_singscore_dot(Test)


palbociclib_pathway_keywords <- c(
  "CELL_CYCLE",
  "G1_S_TRANSITION",
  "G2M_CHECKPOINT",
  "RB_PATHWAY",
  "E2F_TARGETS",
  "MYC_TARGETS",
  "DNA_REPLICATION",
  "DNA_REPAIR",
  "MITOTIC_SPINDLE",
  "PROLIFERATION",
  "PI3K_AKT_MTOR_SIGNALING",
  "MTORC1_SIGNALING",
  "SASP",
  "SENESCENCE",
  "P53_PATHWAY"
)
# use these keywords to filter the pathway
keyword_pattern <- paste(palbociclib_pathway_keywords, collapse = "|") %>% gsub("\\|$", "", .)
# Filter FGSEA results using grepl (case-insensitive matching)
filtered_singscore <- Input %>% dplyr::filter(grepl(keyword_pattern, pathway, ignore.case = TRUE)) %>% arrange(Group)
```


```{r,fig.width=22,fig.height=16,include=T}
PCA_Plot_3_shape(t(TPM_filter), Annotation_all, "specimenID", "Drug", "Timepoint") + 
  scale_color_manual(values = drug_colors) +
  scale_shape_manual(values = c(16, 17, 18)) +
  labs(title = paste0("PCA plot for ", Drug_name), x = "PC1", y = "PC2") +
  theme_bw() +
  theme(legend.position = "top")+geom_text(aes(label = specimenID),hjust = 1, vjust = 0, size = 3, angle = 45) 

```

 
```{r,include=T,fig.width=42,fig.height=16}
names(Singscore_ttest)

Input=Singscore_ttest[c("IFO_8h", "IFO_24h")]

combined_sig <- extract_and_merge_significant_pathways(Input)

colnames(combined_sig)
Test=prepare_dotplot_data(combined_sig) %>%mutate(Timepoint=factor(Timepoint,levels = c("8h","24h")))

P2=plot_singscore_dot(Test)
```

```{r,fig.width=22,fig.height=16,include=T}
P2
```
 
```{r,include=F,fig.width=42,fig.height=16}
names(Singscore_ttest)

Input=Singscore_ttest[c("RIB_8h", "RIB_24h")]

combined_sig <- extract_and_merge_significant_pathways(Input)
Test=prepare_dotplot_data(combined_sig) %>%mutate(Timepoint=factor(Timepoint,levels = c("8h","24h")))

P3=plot_singscore_dot(Test)
```

# Ttest data
```{r,fig.width=22,fig.height=16,include=T}
P3
```

```{r}
knitr::knit_exit()
```



# Irinotecan signautre: PMID: 28454366

```{r}
MIR_sig_genes=c("PHLDA1","SPRY2", "SPRY4", "ETV4", "ETV5", "DUSP4", "DUSP6", "CCND1", "EPHA2", "EPHA4") 
IRI_sig_genes=c(
  "LIMS2", "TUBA1B", "KLHDC2", "PDGFC", "TGFB1I1", 
  "ARHGAP24", "PPAP2A", "DNASE2B", "KIRREL", 
  "EDA2R", "VWA5B1", "DAPK1", "FBXL21"
)
Pathway_list=list(RB_LOF=RB_sig_genes,MIR_sig=MIR_sig_genes,IRI_sig=IRI_sig_genes)
Ouput=generate_singscore_plots(TPM_all, Pathway_list, Annotation_all, drug_colors)

```

```{r,include=T,fig.width=12,fig.height=16}
Ouput[[1]]
```

```{r,include=T,fig.width=12,fig.height=16}
Ouput[[2]]
```

```{r,include=T,fig.width=12,fig.height=16}
Ouput[[3]]
```

```{r}
knitr::knit_exit()
```


```{r}
Pathway_list=list(RB_LOF=RB_sig_genes)
names(Pathway_list)
library(singscore)
Pathway_singscore= singcore_GSEA(TPM_all,Pathway_list)%>%t()%>%as.data.frame()
Pathway_singscore_anno=t(Pathway_singscore)%>%as.data.frame()%>%mutate(specimenID=rownames(.))%>%left_join(.,Annotation_all)
#%>%dplyr::filter(Drug%in%c("DMSO","RIB","PAL","TRAB","DOX"))


# write a function to plot box plot for pathway_singscore anno, group by Date,Drug and Timepoint,input is singscore data and sample_annotation
P2=ggplot(Pathway_singscore_anno,aes(x=Group,y=RB_LOF,fill=Drug,color=Timepoint))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color=Timepoint),size=2,alpha=0.5)+
  scale_fill_manual(values = drug_colors)+
  scale_color_manual(values =Timepoint_colors)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title="RB_LOF signature",x="Drug",y="Singscore")+
  facet_wrap(~Date,nrow=3)

 

```

```{r,include=T,fig.width=12,fig.height=16}
P2
```

```{r}


Pathway_list=list(MIR_sig=MIR_sig_genes)

library(singscore)
Pathway_singscore= singcore_GSEA(TPM_all,Pathway_list)%>%t()%>%as.data.frame()
Pathway_singscore_anno=t(Pathway_singscore)%>%as.data.frame()%>%mutate(specimenID=rownames(.))%>%left_join(.,Annotation_all)
#%>%dplyr::filter(Drug%in%c("DMSO","RIB","PAL","TRAB","DOX"))


# write a function to plot box plot for pathway_singscore anno, group by Date,Drug and Timepoint,input is singscore data and sample_annotation
P3=ggplot(Pathway_singscore_anno,aes(x=Group,y=MIR_sig,fill=Drug))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color=Timepoint),size=2,alpha=0.5)+
  scale_fill_manual(values = drug_colors)+
 # scale_color_manual(values =Timepoint_colors)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title="MIR_sig",x="Drug",y="Singscore")+
  facet_wrap(~Date,nrow=3)
```


```{r,include=T,fig.width=12,fig.height=16}
P3
```

```{r}
knitr::knit_exit()
```


```{r}
FGSEA_sig=get_sig_FGSEA (Result_list)
#%>%mutate(Drug=factor(FGSEA_sig$Drug,levels =Drug_anno$Drug ))%>%mutate(Timepoint=factor(FGSEA_sig$Timepoint,levels =c("8h","24h" ))) 
FGSEA_sig$Drug=factor(FGSEA_sig$Drug,levels =Drug_name)
FGSEA_sig$Timepoint=factor(FGSEA_sig$Timepoint,levels =c("8h","24h" ))
FGSEA_sig_org=FGSEA_sig[,c("Drug","Timepoint","Group")]%>%arrange(Drug,Timepoint)%>%unique()
FGSEA_sig$Group=factor(FGSEA_sig$Group,levels=FGSEA_sig_org$Group)
  
Group_anno=FGSEA_sig[,c("Group","Drug","Timepoint")]
```



\newpage

## Cell cycle related pathway  
```{r}
keywords <- "CELL_CYCLE"

add_keywords=c("")

# Print extracted keywords
keyword_pattern <-paste(c(keywords,add_keywords), collapse = "|")%>%gsub("\\|$","",.)
pathway_list_1=FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))
# Filter FGSEA results using grepl (case-insensitive matching)
filtered_fgsea <- FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))%>%arrange(Group)
#filtered_fgsea$Group
missing_samples=setdiff(unique(FGSEA_sig$Group),filtered_fgsea$Group)
missing_samples_anno=data.frame(Group=missing_samples)%>%left_join(., Group_anno)%>%mutate(pathway=filtered_fgsea$pathway[1],pval=NA,padj=NA,log2err=NA,ES=NA,NES=NA,size=NA,leadingEdge=NA)%>%.[,colnames(filtered_fgsea)]
filtered_fgsea2=rbind(filtered_fgsea,missing_samples_anno)%>%arrange(Group)

Plot_all=plot_pathway_heatmap_2(filtered_fgsea2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug",timepoint_col = "Timepoint",Drug_colors_set = drug_colors,Timepoint_color_set = Timepoint_colors)


```

```{r,fig.width=30,fig.height=28,include=T}
Plot_all
```

\newpage

```{r,include=T}
Select_pathway="FISCHER_G2_M_CELL_CYCLE"

Cell_cycle_UP=FGSEA_sig%>%dplyr::filter(pathway==Select_pathway)%>%dplyr::filter(NES>0)%>%arrange(Drug )
knitr::kable(Cell_cycle_UP[,c("pathway","NES","pval","Drug","Timepoint")],caption = "Group with Cell cycle upregulated")%>%kableExtra::kable_styling(latex_options = "scale_down")
```


 


# Mirda (batch1)

```{r,include=T}
Drug_name="MIR"
Select_pathway="FISCHER_G2_M_CELL_CYCLE"
anno_2=Annotation_b1[c("specimenID","Drug","Timepoint")]%>%arrange(Timepoint,Drug)%>%matrix.please()%>%as.data.frame()
TPM_b1_2_arrange=TPM_b1_2[,rownames(anno_2)]

number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])
Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name&pathway==Select_pathway)%>%arrange(NES) 

Gene_list=Test$leadingEdge%>%unlist()%>%as.character()
fig_title=paste0(Drug_name,"_",Select_pathway)
p2=TPM_heatmap(TPM_b1_2_arrange,Gene_list,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

```

```{r}
# Generare Heatmap by using significant genes from GSEA output
generate_TPM_heatmap <- function(Drug_name, Pathway_name=NULL,Gene_list=NULL) {
  # Get the experiment date for the specified drug
  Date_exp <- Annotation_b2[c("specimenID", "Drug", "Timepoint", "Date")] %>%
    dplyr::filter(Drug == Drug_name) %>%
    .$Date %>%
    unique()

  # Filter annotations for the drug and DMSO at the same experiment date
  anno_2 <- Annotation_b2[c("specimenID", "Drug", "Timepoint", "Date")] %>%
    arrange(Timepoint, Drug) %>%
    dplyr::filter(Drug %in% c(Drug_name, "DMSO")) %>%
    dplyr::filter(Date == Date_exp) %>%
    .[, c("specimenID", "Drug", "Timepoint")] %>%
    matrix.please() %>%
    as.data.frame()

  # Subset TPM data based on the arranged annotation
  TPM_b2_2_arrange <- TPM_b2_2[, rownames(anno_2)]

  # Count the number of samples at 8h timepoint
  number <- anno_2 %>%
    dplyr::filter(Timepoint == "8h") %>%
    nrow()

  # Set colors for annotation
  Color_list <- list(Drug = drug_colors[unique(anno_2$Drug)])

# Extract leading-edge genes for the pathway
if (! is.null(Pathway_name)){
  # Filter FGSEA results for the specified drug and pathway
  Test <- FGSEA_sig %>%
    dplyr::filter(Drug == Drug_name) %>%
    dplyr::filter(pathway == Pathway_name)   
  Pathway_pickup <- Test %>%
    dplyr::filter(pathway == Pathway_name) %>%
    .$leadingEdge %>%
    unlist() %>%
    as.character()
  }
  
  if (! is.null(Gene_list)){
    Pathway_pickup=Gene_list[,1]
    Pathway_name=colnames(Gene_list)
  }

   # Generate the title for the figure
  fig_title <- paste0(Drug_name, "_", Pathway_name)

  # Generate the heatmap
  p2 <- TPM_heatmap(TPM_b2_2_arrange, Pathway_pickup, 
                     Annotation_col = anno_2, 
                     Annotation_color = Color_list, 
                     gap_set = c(3, 3 + number, 3 + number), 
                     Title = fig_title)
  
  return(p2)
}

 

```

```{r}
Drug_check_list=Cell_cycle_UP$Drug%>%unique()
Plotlist=list()
for (i in Drug_check_list) {
  Drug_name=i
  Pahway_name="FISCHER_G2_M_CELL_CYCLE"
  Plotlist[[i]]=generate_TPM_heatmap(Drug_name,Pahway_name)%>%.[[4]]
  
}


```


```{r,fig.width=24,fig.height=40,include=T}
cowplot::plot_grid(plotlist = Plotlist,nrow=1)
```


```{r}
CDK_genes=data.frame(CDK_genes=c("CDK4","CDK6","CCND1"))
Drug_name="PAL"
P1=generate_TPM_heatmap(Drug_name,Pathway_name=NULL,Gene_list =CDK_genes )
Drug_name="RIB"
P2=generate_TPM_heatmap(Drug_name,Pathway_name=NULL,Gene_list =CDK_genes )
```

```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = list(P1$gtable,P2$gtable),nrow=1)
```



## MAPK inhibitor
```{r}
Drug_name=c("SEL","TRA")
MAPK_genes=data.frame(MAPK_sig=c("PHLDA1","SPRY2", "SPRY4", "ETV4", "ETV5", "DUSP4", "DUSP6", "CCND1", "EPHA2", "EPHA4") )
#Selected_path="WP_HOSTPATHOGEN_INTERACTION_OF_HUMAN_CORONAVIRUSES_MAPK_SIGNALING"

Plot_list=list()
#Plot_list2=list()
for (i in Drug_name) {
  Plot_list[[i]]=generate_TPM_heatmap(i,Pathway_name=NULL,Gene_list =MAPK_genes )%>%.$gtable
 # Plot_list2[[i]]=generate_TPM_heatmap(i,Pathway_name=Selected_path,Gene_list =NULL )%>%.$gtable
  
}

```


```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = Plot_list,nrow=1)
```

```{r}
knitr::knit_exit()
```


```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = Plot_list2,nrow=1)
```



```{r}
Drug_name="PAL"
Select_pathway="CHICAS_RB1_TARGETS_LOW_SERUM"
Date_exp=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%dplyr::filter(Drug==Drug_name)%>%.$Date%>%unique()
anno_2=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%arrange(Timepoint,Drug)%>%dplyr::filter(Drug%in%c(Drug_name,"DMSO"))%>%
                      dplyr::filter(Date==Date_exp)%>%.[,c("specimenID","Drug","Timepoint")]%>%matrix.please()%>%as.data.frame()

TPM_b2_2_arrange=TPM_b2_2[,rownames(anno_2)]
number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()

Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])

NES_neg=FGSEA_sig%>%dplyr::filter(Drug==Drug_name&NES<0)%>%arrange(NES)
Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name)%>%arrange(NES)%>%dplyr::filter(pathway==Select_pathway)

Pathway_pickup=Test%>%dplyr::filter(pathway==Select_pathway)%>%.$leadingEdge%>%unlist()%>%as.character()
fig_title=paste0(Drug_name,"_",Select_pathway)

p2=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

CDK_genes=data.frame(CDK_genes=c("CDK4","CDK6","CCND1"))

Test=generate_TPM_heatmap(Drug_name,Pathway_name=NULL,Gene_list =CDK_genes )

fig_title=paste0(Drug_name,"_","CDK genes")
p3=TPM_heatmap(TPM_b2_2_arrange,CDK_genes,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

PCA_Plot_3_shape(t(TPM_b2_2_arrange), Annotation_b2, "specimenID", "Drug", "Timepoint")+scale_color_manual(values=drug_colors)
```
```{r}
p2
```


# Mirda

```{r}
Drug_name="MIR"
anno_2=Annotation_b1[c("specimenID","Drug","Timepoint")]%>%arrange(Timepoint,Drug)%>%matrix.please()%>%as.data.frame()
TPM_b1_2_arrange=TPM_b1_2[,rownames(anno_2)]
number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()

Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])

MAPK_genes=c("PHLDA1","SPRY2", "SPRY4", "ETV4", "ETV5", "DUSP4", "DUSP6", "CCND1", "EPHA2", "EPHA4") 

p1=TPM_heatmap(TPM_b1_2_arrange,MAPK_genes,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="Mirda signature")

Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name&Timepoint=="8h")%>%arrange(NES)%>%dplyr::filter(grepl("MAPK|ERK|CYCLE",pathway))

Select_pathway="FISCHER_G2_M_CELL_CYCLE"
fig_title=paste0(Drug_name,"_",Select_pathway)
FISCHER_G1_S_CELL_CYCLE=Test%>%dplyr::filter(pathway==Select_pathway)%>%.$leadingEdge%>%unlist()%>%as.character()


p2=TPM_heatmap(TPM_b1_2_arrange,FISCHER_G1_S_CELL_CYCLE,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=fig_title)

```

```{r,include=T,fig.width=12,fig.height=6}
cowplot::plot_grid(plotlist = list(p1$gtable,p2$gtable),nrow=1)
```

# Selumetinib

```{r}
Drug_name="SEL"
Date_exp=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%dplyr::filter(Drug==Drug_name)%>%.$Date%>%unique()
anno_2=Annotation_b2[c("specimenID","Drug","Timepoint","Date")]%>%arrange(Timepoint,Drug)%>%dplyr::filter(Drug%in%c(Drug_name,"DMSO"))%>%
                      dplyr::filter(Date==Date_exp)%>%.[,c("specimenID","Drug","Timepoint")]%>%matrix.please()%>%as.data.frame()
TPM_b2_2_arrange=TPM_b2_2[,rownames(anno_2)]
number= anno_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()

Color_list=list(Drug= drug_colors[unique(anno_2$Drug)])

p1=TPM_heatmap(TPM_b2_2_arrange,MAPK_genes,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="Mirda target genes")

Test=FGSEA_sig%>%dplyr::filter(Drug==Drug_name)%>%arrange(NES)%>%dplyr::filter(grepl("MAPK|ERK|CYCLE",pathway))%>%arrange(pathway)

Pahway_name="GOBP_ERK1_AND_ERK2_CASCADE"

Pathway_pickup=Test%>%dplyr::filter(pathway=="GOBP_ERK1_AND_ERK2_CASCADE")%>%.$leadingEdge%>%unlist()%>%as.character()

p2=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=Pahway_name)

Pahway_name="WP_HOSTPATHOGEN_INTERACTION_OF_HUMAN_CORONAVIRUSES_MAPK_SIGNALING"

Pathway_pickup=Test%>%dplyr::filter(pathway=="GOBP_ERK1_AND_ERK2_CASCADE")%>%.$leadingEdge%>%unlist()%>%as.character()

p3=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=Pahway_name)

Pahway_name="FISCHER_G2_M_CELL_CYCLE"
Pathway_pickup=Test%>%dplyr::filter(pathway==Pahway_name)%>%.$leadingEdge%>%unlist()%>%as.character()


p4=TPM_heatmap(TPM_b2_2_arrange,Pathway_pickup,Annotation_col=anno_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title=Pahway_name)



```
```{r,include=T,fig.width=12,fig.height=12}
cowplot::plot_grid(plotlist = list(p1$gtable,p2$gtable,p3$gtable,p4$gtable),nrow=2)
```

```{r}
knitr::knit_exit()
```


```{r}

Test

Test$leadingEdge
```


```{r}
Test=FGSEA_sig%>%dplyr::filter(Drug=="RIB"&Timepoint=="8h")%>%arrange(NES)%>%dplyr::filter(grepl("MAPK|ERK|CYCLE|RB1",pathway))

Test%>%dplyr::filter(pathway=="REACTOME_PERK_REGULATES_GENE_EXPRESSION")%>%.$leadingEdge
Test$leadingEdge
```



# Cytokine and chemokine pathways extraction


```{r}
# Read the CSV file containing cancer-related chemokine & cytokine pathways
cancer_chemo_pathways_df <- read_csv(file.path(GSEA_path, "Cancer_Chemokine_Cytokine_Pathways.csv"))
```

```{r,include=T}
knitr::kable(cancer_chemo_pathways_df,caption = "Cytokine and chemokine related pathways")%>%kableExtra::kable_styling(latex_options = c("scale_down"))
```


```{r}
# Extract unique keywords from the "Keywords" column

keywords <- unique(unlist(stringr::str_split(cancer_chemo_pathways_df$Keywords, " ")))
add_keywords=c("INTERLEUKIN","IL1","IL2")
# Print extracted keywords
keyword_pattern <-paste(c(keywords,add_keywords), collapse = "|")%>%gsub("\\|$","",.)
pathway_list_1=FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))
# Filter FGSEA results using grepl (case-insensitive matching)
Plot_list=list()
for (i in c("WP|HALLMARK|KEGG","GOBP")) {
  filtered_fgsea <- FGSEA_sig %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))%>%dplyr::filter(grepl(i,pathway))%>%arrange(Group)
    filtered_fgsea$Group
   missing_samples=setdiff(unique(FGSEA_sig$Group),filtered_fgsea$Group)
   missing_samples_anno=data.frame(Group=missing_samples)%>%left_join(., Group_anno)%>%mutate(pathway=filtered_fgsea$pathway[1],pval=NA,padj=NA,log2err=NA,ES=NA,NES=NA,size=NA,leadingEdge=NA)%>%.[,colnames(filtered_fgsea)]
   filtered_fgsea2=rbind(filtered_fgsea,missing_samples_anno)%>%arrange(Group)

   Plot_list[[i]]=plot_pathway_heatmap_2(filtered_fgsea2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug",timepoint_col = "Timepoint",Drug_colors_set = drug_colors,Timepoint_color_set = Timepoint_colors)

}


```

\newpage

```{r,fig.width=30,fig.height=24,include=T}
 Plot_list[[1]]
```

\newpage

```{r,fig.width=30,fig.height=18,include=T}
 Plot_list[[2]]
```



# Cancer-related pathways
```{r}
# Read the CSV file containing cancer-related chemokine & cytokine pathways
cancer_pathways_df <- read_csv(file.path(GSEA_path, "Cancer_Pathways.csv"))
```

\newpage
```{r,include=T}
knitr::kable(cancer_pathways_df,caption = "Cancer related pathways")%>%kableExtra::kable_styling()
```

```{r}
# Extract unique keywords from the "Keywords" column
keywords <- unique(unlist(stringr::str_split(cancer_pathways_df$Keywords, " ")))

add_keywords=c("")
# Print extracted keywords
keyword_pattern <-paste(c(keywords,add_keywords), collapse = "|")%>%gsub("\\|$","",.)
FGSEA_sig_rm=FGSEA_sig%>%dplyr::filter(!pathway%in%pathway_list_1$pathway)
# Filter FGSEA results using grepl (case-insensitive matching)
Plot_list=list()
for (i in c("WP","HALLMARK","KEGG","GOBP")) {
  filtered_fgsea <- FGSEA_sig_rm %>%  filter(grepl(keyword_pattern, pathway, ignore.case = TRUE))%>%dplyr::filter(grepl(i,pathway))%>%arrange(Group)
    filtered_fgsea$Group
   missing_samples=setdiff(unique(FGSEA_sig$Group),filtered_fgsea$Group)
   missing_samples_anno=data.frame(Group=missing_samples)%>%left_join(., Group_anno)%>%mutate(pathway=filtered_fgsea$pathway[1],pval=NA,padj=NA,log2err=NA,ES=NA,NES=NA,size=NA,leadingEdge=NA)%>%.[,colnames(filtered_fgsea)]
   filtered_fgsea2=rbind(filtered_fgsea,missing_samples_anno)%>%arrange(Group)
   Plot_list[[i]]=plot_pathway_heatmap_2(filtered_fgsea2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug",timepoint_col = "Timepoint",Drug_colors_set = drug_colors,Timepoint_color_set = Timepoint_colors)

}
```
\newpage

```{r,fig.width=36,fig.height=28,include=T}
 Plot_list[[1]]
```

\newpage
```{r,fig.width=24,fig.height=18,include=T}
 Plot_list[[2]]
```

\newpage
```{r,fig.width=24,fig.height=18,include=T}
 Plot_list[[3]]
```

\newpage
```{r,fig.width=48,fig.height=32,include=T}
 Plot_list[[4]]
```

```{r}
knitr::knit_exit()
```



```{r}
Drug_name = RNAseq_anno$Drug %>% unique()
Drug_name=factor(Drug_name,levels = Drug_name)
# Use RColorBrewer to get 13 colors from the Paired palette
paired_colors =RColorBrewer:: brewer.pal(n = 12, name = "Paired")
extra_color =RColorBrewer:: brewer.pal(n = 8, name = "Set3")[1]  # Take one color from another palette (e.g., Set3)

# Combine black and the 13 colors
drug_colors = c("black", paired_colors, extra_color)[1:length(Drug_name)]
names(drug_colors)=Drug_name
Colour_frame=data.frame(Drug=names(drug_colors),Colour=drug_colors)

RNAseq_anno=RNAseq_anno%>%left_join(.,Colour_frame)
RNAseq_anno$Timepoint=factor(RNAseq_anno$Timepoint,levels = c("0h","8h","24h"))
RNAseq_anno$Drug=factor(RNAseq_anno$Drug,levels=Drug_name)

Sample_ID_Extract=RNAseq_anno[match(RNAseq_anno$Sample_name,colnames(TPM_2)),]

if (isTRUE(unique(Sample_ID_Extract$Sample_name==colnames(TPM_2))) ){
  print("it is correct")
  TPM_cor=TPM_2
   colnames(TPM_cor)=Sample_ID_Extract$Sample_ID
}

RNAseq_anno_org%>%group_by(Drug,Timepoint)%>%summarise(n=n())%>%arrange(Drug)
```


```{r}
Drug_select=RNAseq_anno%>%dplyr::filter(Drug_class==Drug_class_name)
Batch=Drug_select$Time%>%unique()
DMSO_group=RNAseq_anno%>%dplyr::filter(Drug_class=="Control")%>%dplyr::filter(Time==Batch)
Drug_DMSO_data=rbind(Drug_select,DMSO_group)%>%arrange(Timepoint,Drug)
TPM_cor_drug=TPM_cor[,c(DMSO_group$Sample_ID,Drug_DMSO_data$Sample_ID)]%>%.[,match(Drug_DMSO_data$Sample_ID,colnames(.))]
Sample_anno=Drug_DMSO_data
colnames(TPM_cor_drug)
rownames(Sample_anno)=Sample_anno$Sample_ID
Sample_anno_2=Sample_anno[c("Drug","Timepoint")]
Summary=Sample_anno_2%>%group_by(Drug,Timepoint)%>%summarise(n=n())%>%arrange(Timepoint,Drug)
Summary2=Sample_anno_2%>%group_by(Timepoint)%>%summarise(n=n())%>%arrange(Timepoint)
CUM1=cumsum(Summary$n)
CUM2=cumsum(Summary2$n)
Cut_col=c(CUM1,CUM2)%>%sort()

```


# Use DMSO group1 as control

```{r}

load(file.path(Process_path,Drug_class_name,"Chemo_Deseq2_OUT_vs_paired_DMSO" ))

cgwtools::lsdata(file.path(Process_path,Drug_class_name,"Chemo_Deseq2_OUT_vs_paired_DMSO" ))


Test=Drug_reslist[[1]]$RES$`8h`$DOX_8h_vs_DMSO_8h%>%as.data.frame()
Test%>%.["IL1B",]

Test2=Drug_reslist[[1]]$RES$`24h`$DOX_24h_vs_DMSO_24h%>%as.data.frame()
Test2%>%.["IL1B",]
```



```{r}
File_name1="Chemo_GSEA_VS_paried_DMSO.rdata"
if (file.exists(file.path(Process_path,File_name1))) {
 load(file.path(Process_path,File_name1 ))
 #fileexist=cgwtools::lsdata(file.path(Process_path,File_name1))
} else{
 GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="all.v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("c2|c5|h",.)]
 FGSEA_list <- list()

 # Loop through each drug in the Drug_reslist
for (i in names(Drug_reslist)) {
  Data <- Drug_reslist[[i]]
  Timepoint_list <- Data$RES  # Access the time points for this drug

  # Loop through each time point
  for (j in names(Timepoint_list)) {
    R1 <- Timepoint_list[[j]]

    # Loop through each replicate within the time point
    for (h in names(R1)) {
      data <- R1[[h]]

      # Filter significant genes and prepare ranked list
      res_q005 <- data %>%
        as.data.frame() %>%
        dplyr::filter(padj < 0.05)
    
      # Create a ranked vector for FGSEA using log2FoldChange
      res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
      ranks <- tibble::deframe(res_q005_rank)  # Convert to named vector

      # Store FGSEA results in a list
      out_list <- list()

      # Run FGSEA for each GMT file
      for (x in seq_along(GMT_files)) {
        pathways <- fgsea::gmtPathways(GMT_files[x])  # Load pathway data
        set.seed(12345)  # Set seed for reproducibility
        out_list[[x]] <- fgsea::fgseaMultilevel(
          pathways, ranks,
          minSize = 10,
          maxSize = 500,
          nPermSimple = 10000
        )
      }

      # Combine FGSEA results for all GMT files for this replicate
      #FGSEA_list[[paste(i, j, h, sep = "_")]] <- do.call("rbind", out_list)
      
      FGSEA_list[[h]] <- do.call("rbind", out_list)
    }
   }
}
  save(FGSEA_list,file=file.path(Process_path,File_name1 ))
 
}
```

```{r}
names(FGSEA_list)
```


```{r}
cgwtools::lsdata(file.path(Process_path,Drug_class_name,"Deseq2_OUT_vs_DMSO_d1.rdata" ))
#Sample_Drug_list= colnames(TPM_2)%>%.[grepl("TRAB|DOX|IFO|D1",.)]



```



```{r}
# Define DEGs from RES data (padj<0.05& |FC|>1.2)

DEGs_list <- list()
for (i in names(Drug_reslist)) {
  Data <- Drug_reslist[[i]]
  Timepoint_list <- Data$RES  # Access the time points for this drug

  # Loop through each time point
  for (j in names(Timepoint_list)) {
    R1 <- Timepoint_list[[j]]

    # Loop through each replicate within the time point
    for (h in names(R1)) {
      data <- R1[[h]]

      # Filter significant genes and prepare ranked list
      DEGs <- data %>%
        as.data.frame() %>%
        dplyr::filter(padj < 0.05)%>%dplyr::filter(abs(log2FoldChange)>0.27)%>%mutate(gene_symbol=rownames(.))
    
      
      # Store FGSEA results in a list
   
      DEGs_list[[h]]=DEGs
    }
  }  
}

names(DEGs_list)
DEGs_list[[1]]
All_DEGs=do.call("rbind",DEGs_list)%>%as.data.frame()%>%.$gene_symbol%>%unique()
```

# Drug 1_Hallmark

## pathway enrichemant
```{r}
Result_list=list()
Plotlist=list()
Drug_name_list=names(Drug_reslist)
Pathway_key="Hallmark|HALLMARK"

for (I in Drug_name_list) {
 Drug_name=I
 Drug_fgsea=names(FGSEA_list)%>%.[grepl(Drug_name,.)]
 names(FGSEA_list)
 Select_data=list()
 for(i in Drug_fgsea){
  Select_data[[i]]=FGSEA_list[[i]]  }

 all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
 NES_sig_pathway_impute=list()
 for(i in names(Select_data)){
  Nes_data=Select_data[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  }
# samples=names(NES_sig_pathway_impute)%>%.[grepl("DMSO",.)]
 S_data=NES_sig_pathway_impute
 names(NES_sig_pathway_impute)=factor(names(NES_sig_pathway_impute),levels = rev(names(NES_sig_pathway_impute)))
 result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
 result$group=factor(result$group,levels = names(NES_sig_pathway_impute))
#remotes::install_github("YuLab-SMU/ggbreak")
 Plotlist[[I]]=Enrich_dotplot_update(result) 
 Result_list[[I]]=result
 
}

```



```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[1]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[2]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[3]]
```


```{r}
Filter_result=list()
for (i in names(Result_list)){
   data=Result_list[[i]]%>%dplyr::filter(grepl("DMSO_8h|DMSO_24h",group))
   Unique_name=data%>%dplyr::filter(padj<0.05)%>%distinct(pathway) 
   Filter_result[[i]]=data%>%dplyr::filter(pathway%in%Unique_name$pathway)%>%arrange(pathway)%>%as.data.frame()
  
}
 

Filter_result_all=do.call("rbind",Filter_result)%>%as.data.frame()%>%tidyr::separate(.,col=group,into=c("Drug","timepoint"),sep="_",remove=F)%>%arrange(Drug,desc(timepoint))

Filter_result_all$Group=factor(Filter_result_all$Group,levels=unique(Filter_result_all$Group))
```




```{r,fig.width=12,fig.height=16,include=T}
plot_pathway_heatmap(Filter_result_all, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

```



```{r,fig.width=18,fig.height=48,include=T}

Hallmark_pathways=names(GSEA_pathway_list)%>%.[grepl("HALLMARK",.)]
Pathway_name="HALLMARK_INTERFERON_GAMMA_RESPONSE"
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```


# Drug Wiki

## pathway enrichemant
```{r}
Result_list=list()
Plotlist=list()
Drug_name_list=names(Drug_reslist)
Pathway_key="WP_"

for (I in Drug_name_list) {
 Drug_name=I
 Drug_fgsea=names(FGSEA_list)%>%.[grepl(Drug_name,.)]
 names(FGSEA_list)
 Select_data=list()
 for(i in Drug_fgsea){
  Select_data[[i]]=FGSEA_list[[i]]  }

 all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
 NES_sig_pathway_impute=list()
 for(i in names(Select_data)){
  Nes_data=Select_data[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  }
# samples=names(NES_sig_pathway_impute)%>%.[grepl("DMSO",.)]
 S_data=NES_sig_pathway_impute
 names(NES_sig_pathway_impute)=factor(names(NES_sig_pathway_impute),levels = rev(names(NES_sig_pathway_impute)))
 result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
 result$group=factor(result$group,levels = names(NES_sig_pathway_impute))
#remotes::install_github("YuLab-SMU/ggbreak")
 Plotlist[[I]]=Enrich_dotplot_update(result) 
 Result_list[[I]]=result
 
}

```

```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[1]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[2]]
```

\newpage
```{r,fig.width=16,fig.height=18,include=F}
Plotlist[[3]]
```


```{r}
Filter_result=list()
for (i in names(Result_list)){
   data=Result_list[[i]]%>%dplyr::filter(grepl("DMSO_8h|DMSO_24h",group))
   Unique_name=data%>%dplyr::filter(padj<0.05)%>%distinct(pathway) 
   Filter_result[[i]]=data%>%dplyr::filter(pathway%in%Unique_name$pathway)%>%arrange(pathway)%>%as.data.frame()
  
}
 

Filter_result_all=do.call("rbind",Filter_result)%>%as.data.frame()%>%tidyr::separate(.,col=group,into=c("Drug","timepoint"),sep="_",remove=F)%>%arrange(Drug,desc(timepoint))

Filter_result_all$Group=factor(Filter_result_all$Group,levels=unique(Filter_result_all$Group))
```


## Signal pahtway

```{r}
unique(Filter_result_all$pathway)
```


```{r,fig.width=12,fig.height=18,include=T}
Data1=Filter_result_all%>%dplyr::filter(grepl("SIGNALING|INTERFERON|IFNG|INTERLEUKIN",pathway))
plot_pathway_heatmap(Data1, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```
\newpage
```{r,fig.width=12,fig.height=16,include=T}
Hallmark_pathways=names(GSEA_pathway_list)%>%.[grepl("WP_TNF",.)]
Pathway_name="WP_TNFALPHA_SIGNALING_PATHWAY"
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```

\newpage

```{r,fig.width=12,fig.height=6,include=T}
Hallmark_pathways=names(GSEA_pathway_list)%>%.[grepl("PI3K",.)]
Pathway_name="WP_TYPE_I_INTERFERON_INDUCTION_AND_SIGNALING_DURING_SARSCOV2_INFECTION"
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```
\newpage

```{r,fig.width=12,fig.height=30,include=T,}
Pathway_name="WP_PI3KAKT_SIGNALING_PATHWAY" 
genes=intersect(GSEA_pathway_list[[Pathway_name]],All_DEGs)
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )
```


```{r,fig.width=12,fig.height=4}
INF_IL_gene_TPM=TPM_cor_drug%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(grepl("^IL|IFN",gene_symbol))%>%dplyr::filter(gene_symbol%in%All_DEGs)
genes=INF_IL_gene_TPM$gene_symbol
Pathway_name="Interleukin and Interferon genes"

#Col_anno=Colour_frame %>%dplyr::filter(Sample_ID==colnames(TPM_cor_drug))


```

\newpage
```{r,fig.width=12,fig.height=8,include=T}
pheatmap_pathway_4(TPM_cor_drug,genes,Pathway_name,Annotation =Sample_anno_2,Cut_col )

```


```{r,fig.width=12,fig.height=14,include=T}
Data2=Filter_result_all%>%dplyr::filter(grepl("CYCLE|GROWTH|MIGRATION|G2|G1|_S_|APOPTOSIS|ARREST|PATHWAY",pathway))%>%dplyr::filter(!pathway%in%Data1$pathway)
plot_pathway_heatmap(Data2, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```



```{r,fig.width=12,fig.height=6,include=T}
Data3=Filter_result_all%>%dplyr::filter(grepl("GLYCOLYSIS|OXIDATIVE|TCA|METABOLISM|PYRUVATE|PENTOSE|FATTY|OXIDATION|MITOCHONDRIA",pathway))%>%dplyr::filter(!pathway%in%Data1$pathway)%>%dplyr::filter(!pathway%in%Data2$pathway)
plot_pathway_heatmap(Data3, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```


```{r,fig.width=12,fig.height=6,include=T}
Data4=Filter_result_all%>%dplyr::filter(grepl("CANCER|CARCINOMA",pathway))%>%dplyr::filter(!pathway%in%Data1$pathway)%>%dplyr::filter(!pathway%in%Data2$pathway)%>%dplyr::filter(!pathway%in%Data3$pathway)
plot_pathway_heatmap(Data4, pathway_col = "pathway", padj_col = "padj", NES_col = "NES", group_col = "Group",drug_col = "Drug")

 
```





```{r}
knitr::knit_exit()
```


# Drug wikipathway

## pathway enrichemant
```{r}
Plotlist=list()
Pathway_key="WP_"
Drug_name_list=names(Drug_reslist)
for (I in Drug_name_list) {
 Drug_name=I
 Drug_fgsea=names(FGSEA_list)%>%.[grepl(Drug_name,.)]
 names(FGSEA_list)
 Select_data=list()
 for(i in Drug_fgsea){
  Select_data[[i]]=FGSEA_list[[i]]
  
 }

 all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
 NES_sig_pathway_impute=list()
 for(i in names(Select_data)){
  Nes_data=Select_data[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}
# samples=names(NES_sig_pathway_impute)%>%.[grepl("DMSO",.)]
 S_data=NES_sig_pathway_impute
 names(NES_sig_pathway_impute)=factor(names(NES_sig_pathway_impute),levels = rev(names(NES_sig_pathway_impute)))

 
 result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
 result$group=factor(result$group,levels = names(NES_sig_pathway_impute))
#remotes::install_github("YuLab-SMU/ggbreak")
 Plotlist[[I]]=Enrich_dotplot_update(result) 

 
}

```

```{r,fig.width=16,fig.height=18,include=T}
Plotlist[[1]]
```



\newpage
```{r,fig.width=16,fig.height=18,include=T}
Plotlist[[2]]
```


```{r,fig.width=16,fig.height=18,include=T}
Plotlist[[3]]
```


```{r}
knitr::knit_exit()
```


```{r,fig.width=12,fig.height=18,include=T}
Pathway_key="REACTOME"
result <- Common_sig_pathway_NES(S_data, Pathway_key = Pathway_key)
result$group=factor(names(NES_sig_pathway_impute),levels = names(NES_sig_pathway_impute))
P3=Enrich_dotplot(result)

P3
```


```{r,include=F}

out=list()

for (i in names(Drug_reslist)){
  Data=Drug_reslist[[i]]
  Sample_list=Data$Counts%>%colnames(.)
  Genes=rownames(Data$RES[[1]][[1]])
  TPM_pickup=TPM_counts%>%dplyr::filter(gene_name%in%Genes)
  TPM_pickup_2=TPM_pickup%>%.[,colnames(.) %in%Sample_list]
  rownames(TPM_pickup_2)=TPM_pickup$gene_name
  Annotion_file=Data$Anno%>%arrange(Timepoint,Drug)%>%dplyr::filter(Sample_name%in%colnames(TPM_pickup_2))
  Annotion_file_2=Annotion_file%>%.[,c("Timepoint","Drug")]
  TPM_pickup_3=TPM_pickup_2%>%as.data.frame()%>%.[Annotion_file$Sample_name] 
  rownames(TPM_pickup_3)=rownames(TPM_pickup_2)
  Color_Set1=RColorBrewer::brewer.pal(5,"Set1")[1:2]
  names(Color_Set1)=unique(Annotion_file$Drug)
  Color_Set2=RColorBrewer::brewer.pal(5,"Set1")[2:4]
  names(Color_Set2)=unique(Annotion_file$Timepoint)
 
  Color_list=list(Drug=Color_Set1,Timepoint=Color_Set2)
  number= Annotion_file_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
 out[[i]]=TPM_heatmap(TPM_pickup_3,MAPK_genes,Annotation_col=Annotion_file_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="MAPK genes")%>%.$gtable

}

```

```{r,fig.width=12,fig.height=8,include=T}
cowplot::plot_grid(plotlist=out)
```

# Wiki, Hallmark, KEGG, Reactome Pathways in Mirda 8 and 24h ("Activated": p<0.05 and NES>0, "Inhibited":p<0.05 and NES<0,



```{r}
names(FGSEA_list)

Select_data=list(SEL_8h_vs_DMSO_8h=FGSEA_list[["SEL_8h_vs_DMSO_8h"]],
           SEL_24h_vs_DMSO_24h=FGSEA_list[["SEL_24h_vs_DMSO_24h"]],
         TRA_8h_vs_DMSO_8h=FGSEA_list[["TRA_8h_vs_DMSO_8h"]],
        TRA_24h_vs_DMSO_24h=FGSEA_list[["TRA_24h_vs_DMSO_24h"]])

all_sig_pathway_name=lapply(Select_data,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}
```

```{r}
Data

```


# SEL
## MAPK pathway
```{r}
SEL_data=NES_sig_pathway_impute[c("SEL_8h_vs_DMSO_8h" ,"SEL_24h_vs_DMSO_24h")]
Pathway_key="MAPK"
result <- Common_sig_pathway_NES(SEL_data, Pathway_key = Pathway_key)

P2=Enrich_dotplot(result)
```


```{r,fig.width=12,fig.height=5,include=T}
P2
```

```{r,fig.width=24,fig.height=12,include=T}
library(pathview)
library(KEGGREST)
kegg_list <- KEGGREST::keggList("pathway", "hsa")
KEGG_sig=kegg_list%>%.[grep("MAPK|Mapk|mapk",.)]
```

# SEL 8h vs DMOS 8h
```{r,fig.width=24,fig.height=12,include=T}
Drug_name="SEL"
Data=Drug_reslist[[Drug_name]]
Sample_list=Data$Counts%>%colnames(.)
Time_point="8h"
Sample_name="SEL_8h_vs_DMSO_8h"
Res_1=Data$RES[[Time_point]][[Sample_name]]%>%as.data.frame()%>%mutate(Group=Sample_name)%>%mutate(gene_symbol=rownames(.))

test_data_1 <-Res_1%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.27) # Example: Use one of your T-test results
# Run the pathview analysis
Pathway_ID=names(KEGG_sig)
Pathway_name=as.character( KEGG_sig)%>%strsplit(.,"-")%>%unlist()%>%.[1]%>%gsub(" $","",.)
pathview_out=file.path(".","Pathview")
Outname=paste0(Sample_name,"_",Pathway_name,".png")%>%gsub(" ","_",.)
```

```{r}
#https://bioconductor.org/packages/devel/bioc/vignettes/SBGNview/inst/doc/SBGNview.Vignette.html#53_Visualize_both_gene_data_and_compound_data
```


```{r,fig.width=24,fig.height=12,include=T}
pathview_result <- run_pathview_gene_symbol(test_data_1, pathway_name=Pathway_name, pathway_id=Pathway_ID, output_dir=pathview_out,outname=Outname)

```


```{r, out.width='100%', fig.align='center',include=T}
png_file=file.path(pathview_out,Outname)
knitr::include_graphics(png_file)
```

# SEL 24h vs DMOS 24h
```{r,fig.width=24,fig.height=12,include=T}
Drug_name="SEL"
Data=Drug_reslist[[Drug_name]]
Sample_list=Data$Counts%>%colnames(.)
Time_point="24h"
Sample_name="SEL_24h_vs_DMSO_24h"
Res_2=Data$RES[[Time_point]][[Sample_name]]%>%as.data.frame()%>%mutate(Group=Sample_name)%>%mutate(gene_symbol=rownames(.))

test_data_2<-Res_2%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.27)
```


```{r,fig.width=24,fig.height=12,include=T}
# Run the pathview analysis
Pathway_ID=names(KEGG_sig)
Pathway_name=as.character( KEGG_sig)%>%strsplit(.,"-")%>%unlist()%>%.[1]%>%gsub(" $","",.)
pathview_out=file.path(".","Pathview")
Outname=paste0(Sample_name,"_",Pathway_name,".png")%>%gsub(" ","_",.)
pathview_result <- run_pathview_gene_symbol(test_data_2, pathway_name=Pathway_name, pathway_id=Pathway_ID, output_dir=pathview_out,outname=Outname)

```

```{r, out.width='100%', fig.align='center',include=T}
png_file=file.path(pathview_out,Outname)
knitr::include_graphics(png_file)
```

## Combine
```{r}
Outname=paste0(Drug_name,"_",Pathway_name,".png")%>%gsub(" ","_",.)
```

```{r}
run_pathview_group_gene_symbol(
  test_data1 = test_data_1,
  test_data2 = test_data_2,
  pathway_name=Pathway_name, pathway_id=Pathway_ID, output_dir=pathview_out,outname=Outname)

 

```
```{r, out.width='100%', fig.align='center',include=T}
png_file=file.path(pathview_out,Outname)
knitr::include_graphics(png_file)
```


```{r}
knitr::knit_exit()
```



```{r}

SEL_data_pathway=lapply(SEL_data, function(x) x%>%dplyr::filter(grepl("AKT|ERK|MAPK_",pathway)))

NES_sig_pathway_pikcup=list()
all_sig_pathway_name=lapply(SEL_data_pathway,function(x) x%>%as.data.frame%>%dplyr::filter(padj<0.05)%>%.$pathway )%>%unlist()%>%unique()
for(i in names(SEL_data)){
  Nes_data=SEL_data[[i]]
  NES_sig_pathway_pikcup[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

NES_sig_pathway_pikcup
```



```{r}
knitr::knit_exit()
```



```{r}
if (!grepl("Pathwaylist_sig_all", fileexist)){
 FGSEA_list=list()
 FGSEA_q005_list=list()
 for (i in 1:length(Res_list)){
  data=Res_list[[i]]
  res_q005=data%>%dplyr::filter(padj<0.05)
  name2=names(Res_list)[[i]]
  res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
  ranks <- tibble::deframe(res_q005_rank)
  out_list=list()
  for (x in 1:length(GMT_files)){
 
     data=fgsea::gmtPathways(GMT_files[x])
    set.seed(12345)
    out_list[[x]] <- fgsea::fgseaMultilevel(data, ranks,
                           minSize=10,
                           maxSize=500,nPermSimple = 10000)
  }
                           
  FGSEA_list[[name2]]=do.call("rbind",out_list)
  FGSEA_q005_list[[name2]]=do.call("rbind",out_list)%>%as.data.frame()%>%dplyr::filter(padj<0.05)

}

all_sig_pathway_name=lapply(FGSEA_q005_list,function(x) x$pathway )%>%unlist()%>%unique()


NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

 Pathwaylist_sig_all=do.call("rbind",NES_sig_pathway_impute)
 Pathwaylist_sig_all$group=factor(Pathwaylist_sig_all$group,c(names(NES_sig_pathway_impute)))
 
 if (!file.exists(file.path(Rawdata_path,File_name1 ))) {
    save(Pathwaylist_sig_all,FGSEA_q005_list,file=file.path(File_name1,File_name1 ))
   }
  
}

```




# Use DMSO group2 as control

```{r}
load(file.path(Process_path,"Deseq2_OUT_vs_DMSO_d2.rdata"))
```


```{r,include=F}


out=list()

for (i in names(Drug_reslist)){
  Data=Drug_reslist[[i]]
  Sample_list=Data$Counts%>%colnames(.)
  Genes=rownames(Data$RES[[1]][[1]])
  TPM_pickup=TPM_counts%>%dplyr::filter(gene_name%in%Genes)
  TPM_pickup_2=TPM_pickup%>%.[,colnames(.) %in%Sample_list]
  rownames(TPM_pickup_2)=TPM_pickup$gene_name
  Annotion_file=Data$Anno%>%arrange(Timepoint,Drug)%>%dplyr::filter(Sample_name%in%colnames(TPM_pickup_2))
  Annotion_file_2=Annotion_file%>%.[,c("Timepoint","Drug")]
  TPM_pickup_3=TPM_pickup_2%>%as.data.frame()%>%.[Annotion_file$Sample_name] 
  rownames(TPM_pickup_3)=rownames(TPM_pickup_2)
  Color_Set1=RColorBrewer::brewer.pal(5,"Set1")[1:2]
  names(Color_Set1)=unique(Annotion_file$Drug)
  Color_Set2=RColorBrewer::brewer.pal(5,"Set1")[2:4]
  names(Color_Set2)=unique(Annotion_file$Timepoint)
 
  Color_list=list(Drug=Color_Set1,Timepoint=Color_Set2)
  number= Annotion_file_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
 out[[i]]=TPM_heatmap(TPM_pickup_3,MAPK_genes,Annotation_col=Annotion_file_2,Annotation_color=Color_list,gap_set = c(3,3+number,3+number) ,Title="MAPK genes")%>%.$gtable

}

```

```{r,fig.width=12,fig.height=8,include=T}
cowplot::plot_grid(plotlist=out)
```


```{r}
knitr::knit_exit()
```

```{r}
TPM_heatmap(TPM_pickup_2,MAPK_genes,Annotation_col=Annotion_file,Annotation_color=Color_list,Title="MAPK genes")
```

```{r,include=F,fig.width=8,fig.height=5}
set.seed(12345)
Sample_anno_2$Drug=factor(Sample_anno_2$Drug,levels = c("DMSO","Mirda"))
VST_rm=t(deseq2VST)%>%as.data.frame()%>%.[rownames(.)!="sample.8hM2", ]%>%.[!grepl("0", rownames(.) ),]
PCA_2=PCA_Plot_3(VST_rm,Sample_anno_2,VAR="Sample_name","Drug","Timepoint")
```


```{r,include=T,fig.width=5,fig.height=5}

cowplot::plot_grid(PCA_2,labels=c("PCA plot"))
```

```{r,include=T}
# Sample_anno=data.frame(Sample_name=colnames(Raw_counts_2_flipped))%>%mutate(Drug=ifelse(grepl("D",Sample_name),"DMSO", ifelse(grepl("M",Sample_name),"Mirda","DMSO") ))%>%mutate(Timepoint=ifelse(grepl("0h",Sample_name),"0hour", ifelse(grepl("8h",Sample_name),"8hour","24hour") ))%>%mutate(Group=paste0(Drug,"_",Timepoint))%>%mutate(Timepoint=factor(.$Timepoint,levels = c("0hour","8hour","24hour")))%>%arrange(Timepoint,Drug)
# #Timepoint=factor(Sample_anno$Timepoint,levels = c("0hour","8hour","24hour"))
# 
# 
# knitr::kable(Sample_anno,caption="Flipped Sample annotation")%>%kableExtra::kable_styling()
```

# DEGs in Mirda 8h vs  DMSo 8h (3396 "padj<0.05,|FC|>1.2"); DEGs in Mirda 24h vs  DMSo 24h (3309 "padj<0.05,|FC|>1.2")
```{r}
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt, colData = Sample_anno,design =~Group)
dds= DESeq2::DESeq(deseq2Data)

Mirda_8h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_8hour","DMSO_8hour"))%>%as.data.frame()
Mirda_8h_vs_DMSO_sig=Mirda_8h_vs_DMSO%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_8h_vs_DMSO_sig)
Mirda_24h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_24hour","DMSO_24hour"))%>%as.data.frame()
Mirda_24h_vs_DMSO_sig=Mirda_24h_vs_DMSO%>%dplyr::filter(padj<0.05)%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_24h_vs_DMSO_sig)
```

# Use DEGs for pathway analysis (3396 at 8h and 3309 at 24h)

```{r,include=F,fig.width=8,fig.height=6}

Mirda_8h_vs_DMSO_sig_up=Mirda_8h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange>0)
Mirda_8h_vs_DMSO_sig_DN=Mirda_8h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange<0)
Mirda_24h_vs_DMSO_sig_up=Mirda_24h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange>0)
Mirda_24h_vs_DMSO_sig_DN=Mirda_24h_vs_DMSO_sig%>%dplyr::filter(log2FoldChange<0)

set1 <-  rownames(Mirda_8h_vs_DMSO_sig_up)
set2 <-  rownames(Mirda_24h_vs_DMSO_sig_up)
x = list(set1, set2)
library("ggVennDiagram")
G1=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_up","Mirda_24h_up"),hjust=-0.5,set_color = "red",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))



set3 <-  rownames(Mirda_8h_vs_DMSO_sig_DN)
set4 <-  rownames(Mirda_24h_vs_DMSO_sig_DN)
x = list(set3, set4)
library("ggVennDiagram")
G2=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_DN","Mirda_24h_DN"),hjust=-0.5,set_color = "green",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))


```

```{r,fig.width=12,fig.height=6,include=T}
cowplot::plot_grid(G1,G2,labels=c("Upregulated_genes","Down_regulated_genes"))
```


```{r}
Res_list=list(Mirda_8h=Mirda_8h_vs_DMSO_sig,Mirda_24h=Mirda_24h_vs_DMSO_sig)
GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="all.v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("c2|c5|h",.)]
```

```{r}
if (file.exists(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))) {
 load(file.path(Rawdata_path,"2_Flipped_GSEA_data.rdata" ))
 #fileexist=cgwtools::lsdata(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))
} else{
  fileexist=c("NA")
  
}


```

# Wiki, Hallmark, KEGG, Reactome Pathways in Mirda 8 and 24h ("Activated": p<0.05 and NES>0, "Inhibited":p<0.05 and NES<0,
**"P_Activated":possible activated 0.07>p>0.05,NES>0; "P_Inhibited" possible inhibtied 0.07>p>0.05,NES<0; "no_change": P>0.07)**

```{r}
if (!grepl("Pathwaylist_sig_all", fileexist)){

 FGSEA_list=list()
 FGSEA_q005_list=list()
 for (i in 1:length(Res_list)){
  data=Res_list[[i]]
  res_q005=data%>%dplyr::filter(padj<0.05)
  name2=names(Res_list)[[i]]
  res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
  ranks <- tibble::deframe(res_q005_rank)
  out_list=list()
  for (x in 1:length(GMT_files)){
 
     data=fgsea::gmtPathways(GMT_files[x])
    set.seed(12345)
    out_list[[x]] <- fgsea::fgseaMultilevel(data, ranks,
                           minSize=10,
                           maxSize=500,nPermSimple = 10000)
  }
                           
  FGSEA_list[[name2]]=do.call("rbind",out_list)
  FGSEA_q005_list[[name2]]=do.call("rbind",out_list)%>%as.data.frame()%>%dplyr::filter(padj<0.05)

}

all_sig_pathway_name=lapply(FGSEA_q005_list,function(x) x$pathway )%>%unlist()%>%unique()


NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

 Pathwaylist_sig_all=do.call("rbind",NES_sig_pathway_impute)
 Pathwaylist_sig_all$group=factor(Pathwaylist_sig_all$group,c(names(NES_sig_pathway_impute)))
 
 if (!file.exists(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))) {
 save(Pathwaylist_sig_all,FGSEA_q005_list,file=file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))}
  



 }




```

```{r}
Pathwaylist_sig_all_pickup=Pathwaylist_sig_all%>%dplyr::filter(grepl("HALLMARK|WP_|KEGG|REACTOME",pathway ))

Pathwaylist_sig_all%>%dplyr::filter(grepl("MAPK",pathway))
unique(Pathwaylist_sig_all_pickup$group )


```

```{r}
Pathwaylist_sig_8_up=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES>0&group=="Mirda_8h"&padj<0.05)%>%arrange(desc(NES))
Pathwaylist_sig_8_DN=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES<0&group=="Mirda_8h"&padj<0.05)


Pathwaylist_sig_24_up=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES>0&group=="Mirda_24h"&padj<0.05)%>%arrange(desc(NES))
Pathwaylist_sig_24_DN=Pathwaylist_sig_all_pickup%>%dplyr::filter(NES<0&group=="Mirda_24h"&padj<0.05)
```

```{r}
set1 <-  Pathwaylist_sig_8_up$pathway
set2 <-  Pathwaylist_sig_24_up$pathway
x = list(set1, set2)
library("ggVennDiagram")
G3=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_pathway_up","Mirda_24h_pathway_up"),hjust=-0.5,set_color = "red",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))



set3 <-  Pathwaylist_sig_8_DN$pathway
set4 <-  Pathwaylist_sig_24_DN$pathway
x = list(set3, set4)
library("ggVennDiagram")
G4=ggVennDiagram:: ggVennDiagram(x, label_alpha = 0,category.names =c("Mirda_8h_pathway_DN","Mirda_24h_pathway_DN"),hjust=-0.5,set_color = "green",label_color = "firebrick")+ scale_x_continuous(expand = expansion(mult = .2))

```


```{r,fig.width=16,fig.height=7,include=T}
cowplot::plot_grid(G3,G4,labels=c("upregulated pathways","Downregulated pathways"))
```


```{r}
set1 <-  Pathwaylist_sig_8_up$pathway
set2 <-  Pathwaylist_sig_24_up$pathway
Up_pathways=unique(set1,set2)
```


```{r}
set3 <-  Pathwaylist_sig_8_DN$pathway
set4 <-  Pathwaylist_sig_24_DN$pathway

DN_pathways=unique(set3,set4)
```


```{r}
## Pathway network
human_input_8h=Mirda_8h_vs_DMSO_sig%>%mutate(Gene.symbol=rownames(.)) %>%.[,c("Gene.symbol","log2FoldChange","padj")]%>%setNames(c("Gene.symbol","logFC","adj.P.Val"))%>%na.omit() 
 
human_input_8h_up=human_input_8h%>%dplyr::filter(logFC>0)
human_input_8h_DN=human_input_8h%>%dplyr::filter(logFC<0)

human_input_24h=Mirda_24h_vs_DMSO_sig%>%mutate(Gene.symbol=rownames(.)) %>%.[,c("Gene.symbol","log2FoldChange","padj")]%>%setNames(c("Gene.symbol","logFC","adj.P.Val"))%>%na.omit() 
 
human_input_24h_up=human_input_24h%>%dplyr::filter(logFC>0)
human_input_24h_DN=human_input_24h%>%dplyr::filter(logFC<0)

```

```{r}




```


```{r}
Term_gene_graph_2=function (result_df, num_terms = 10, layout = "stress", use_description = FALSE, 
      node_size = "num_genes", node_colors = c("#E5D7BF", "green", "red", "grey")) {
  
  # for single plot
  
  if (!is.numeric(num_terms) & !is.null(num_terms)) {
    stop("`num_terms` must either be numeric or NULL!")
  }
  if (!is.logical(use_description)) {
    stop("`use_description` must either be TRUE or FALSE!")
  }
  ID_column <- ifelse(use_description, "Term_Description", 
                      "ID")
  val_node_size <- c("num_genes", "p_val")
  if (!node_size %in% val_node_size) {
    stop("`node_size` should be one of ", paste(dQuote(val_node_size), 
                                                collapse = ", "))
  }
  if (!is.data.frame(result_df)) {
    stop("`result_df` should be a data frame")
  }
  necessary_cols <- c(ID_column, "lowest_p", "Up_regulated", 
                      "Down_regulated", "no_change","group")  # Include "no_change"
  if (!all(necessary_cols %in% colnames(result_df))) {
    stop(paste(c("All of", paste(necessary_cols, collapse = ", "), 
                 "must be present in `results_df`!"), collapse = " "))
  }
  if (length(node_colors) != 4) {  # Corrected length check
    stop("`node_colors` must contain exactly 4 colors")  # Updated error message
  }
  if (!is.null(num_terms)) {
    if (nrow(result_df) < num_terms) {
      num_terms <- NULL
    }
  }
  result_df<- result_df[order(result_df$Term_Description, decreasing = FALSE),  ]
  if (!is.null(num_terms)) {
        result_df <- result_df[1:num_terms, ]
    }
    graph_df <- data.frame()
  for (i in base::seq_len(nrow(result_df))) {
        up_genes <- unlist(strsplit(result_df$Up_regulated[i], 
            ", "))
        down_genes <- unlist(strsplit(result_df$Down_regulated[i], 
            ", "))
        
        no_change= unlist(strsplit(result_df$no_change[i], 
            ", "))
        
        for (gene in c(up_genes, down_genes,no_change)) {
            graph_df <- rbind(graph_df, data.frame(Term = result_df[i, 
                ID_column], Gene = gene))
        }
    }
  up_genes <- unlist(lapply(result_df$Up_regulated, function(x) strsplit(x, ", ")))
  DN_genes=  unlist(lapply(result_df$Down_regulated, function(x) strsplit(x, ", ")))
  g <- igraph::graph_from_data_frame(graph_df, directed = FALSE)
  cond_term <- names(igraph::V(g)) %in% result_df[, ID_column]
  cond_up_gene <- names(igraph::V(g)) %in% up_genes
  cond_down_gene=  names(igraph::V(g)) %in% DN_genes
  node_type <- ifelse(cond_term, "term", 
                      ifelse(cond_up_gene, "up", 
                             ifelse(cond_down_gene, "down", "no_change")))
  node_type <- factor(node_type, levels = c("term", "up", "down", "no_change"))
  node_type <- factor(node_type, levels = c("term", "up", "down","no_change"))
    igraph::V(g)$type <- node_type
  type_descriptions <- c(term = "enriched term", up = "up-regulated gene", 
                         down = "down-regulated gene", no_change = "no_change gene")  # Include "no_change" description
   type_descriptions <- type_descriptions[levels(node_type)]
   names(node_colors) <- c("term", "up", "down", "no_change")  # Added "no_change"
   node_colors <- node_colors[levels(node_type)]
    if (node_size == "num_genes") {
        sizes <- igraph::degree(g)
        sizes <- ifelse(igraph::V(g)$type == "term", sizes, 2)
        size_label <- "# genes"
    }
    else {
        idx <- match(names(igraph::V(g)), result_df[, ID_column])
        sizes <- -log10(result_df$lowest_p[idx])
        sizes[is.na(sizes)] <- 2
        size_label <- "-log10(p)"
    }
     
    
    igraph::V(g)$size <- sizes
    igraph::V(g)$label.cex <- 0.5
    igraph::V(g)$frame.color <- "gray"
    
    p <- ggraph::ggraph(g, layout = layout)
    
    xy <- graphlayouts::layout_with_stress(g)
    p <- ggraph::ggraph(g, layout = xy)
    
    p <- p + ggraph::geom_edge_link(alpha = 0.8, colour = "darkgrey")
    p <- p + ggraph::geom_node_point(ggplot2::aes(color = .data$type, 
        size = .data$size))
    p <- p + ggplot2::scale_size(range = c(5, 10), breaks = round(seq(round(min(igraph::V(g)$size)), 
        round(max(igraph::V(g)$size)), length.out = 4)), name = size_label)
    p <- p + ggplot2::theme_void()
    p <- p + suppressWarnings(ggraph::geom_node_text(ggplot2::aes(label = .data$name), 
        nudge_y = 0.2, repel = TRUE, max.overlaps = 20))
    p <- p + ggplot2::scale_color_manual(values = node_colors, 
        name = NULL, labels = type_descriptions)
    if (is.null(num_terms)) {
        p <- p + ggplot2::ggtitle("Term-Gene Graph")
    }
    else {
        p <- p + ggplot2::ggtitle("Term-Gene Graph", subtitle = paste(c("Top", 
            num_terms, "terms"), collapse = " "))
    }
    p <- p + ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5), 
        plot.subtitle = ggplot2::element_text(hjust = 0.5))
    return(p)
}

```


```{r,fig.width=24,fig.height=8,include=T}
# NES plot
Both_up=intersect(set1,set2)
Both_DN=intersect(set3,set4)

Select_path=Pathwaylist_sig_all%>%dplyr::filter( pathway%in%c(Both_up,Both_DN))
Output_both=Enrich_dotplot(Select_path)
Output_both
```

```{r,fig.height=6,fig.width=12,include=T}
Pathway_pickup=c(Both_up,Both_DN)
Pathway_list=GSEA_pathway_list[Pathway_pickup]
Pathway_singscore= singcore_GSEA(Flip_TPM_2_arrange,Pathway_list)%>%t()%>%as.data.frame()
TITLE="Significant Pathway"
Keypahtway_singscore=paste0(names(Pathway_list),"_singscore")

GSEA_heatmap(Pathway_singscore,Flip_TPM_2_anno_2,Color_list,Title = TITLE)
```

## 24h UP and Down

```{r,fig.width=18,fig.height=10,include=T}
Pathway_get=c(setdiff(set4,set3), setdiff(set2,set1))%>%.[!grepl("REACTOME",.)]%>%.[grepl("COMPLEMENT|SIGNALING|INTEFERIN|APOPTOSIS",.)]

Select_path=Pathwaylist_sig_all%>%dplyr::filter( pathway%in%Pathway_get)
Output=Enrich_dotplot(Select_path)
Output
```
```{r,fig.height=6,fig.width=12,include=T}
Pathway_pickup=Pathway_get
Pathway_list=GSEA_pathway_list[Pathway_pickup]
Pathway_singscore= singcore_GSEA(Flip_TPM_2_arrange,Pathway_list)%>%t()%>%as.data.frame()
TITLE="Significant Pathway"
Keypahtway_singscore=paste0(names(Pathway_list),"_singscore")

GSEA_heatmap(Pathway_singscore,Flip_TPM_2_anno_2,Color_list,Title = TITLE)
```


```{r}
knitr::knit_exit()
```


```{r}

# Raw_counts_add_control=Raw_counts_2%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)
# Sample_anno_2=data.frame(Sample_name=colnames(Raw_counts_add_control))%>%left_join(.,Sample_anno)%>%mutate(Drug=ifelse(is.na(Drug),"Mirda",Drug))%>%mutate(Timepoint=ifelse(is.na(Timepoint),"0hour",Timepoint))%>%mutate(Group=paste0(Drug,"_",Timepoint))
# rownames(Sample_anno_2)=Sample_anno_2$Sample_name
# Sample_anno_2$Timepoint=factor(Sample_anno_2$Timepoint,levels = c("0hour","8hour","24hour"))
# library(SummarizedExperiment)
# library(DESeq2)
#   keep.exprs <- selectGenes_BY_group(Raw_counts_add_control, min.count=10, N=0.90,Sample_anno_2)
# keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
# myFilt <- Raw_counts_add_control[rownames(keep.exprs_2),]
# 
# dim(myFilt)

library(SummarizedExperiment)
library(DESeq2)
keep.exprs <- selectGenes_BY_group(Raw_counts_2_flipped, min.count=10, N=0.90,Sample_anno)
keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
myFilt <- Raw_counts_2[rownames(keep.exprs_2),]

dim(myFilt)
```


```{r}
# Flipping TPM data
Flip_TPM_2=TPM_counts[,c(2:17)]%>%dplyr::filter(gene_name%in%rownames(myFilt) )%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()

colnames(Flip_TPM_2)=gsub("D","DC",colnames(Flip_TPM_2))%>%gsub("M","D",.)%>%gsub("DC","M",.)


Flip_TPM_2_anno=data.frame(Sample_name=colnames(Flip_TPM_2))%>%left_join(., Sample_anno)%>%arrange(Timepoint,Drug)

Flip_TPM_2_arrange=Flip_TPM_2[,Flip_TPM_2_anno$Sample_name]

Flip_TPM_2_anno_2=Flip_TPM_2_anno["Group"]
rownames(Flip_TPM_2_anno_2)=Flip_TPM_2_anno$Sample_name
unique(rownames(Flip_TPM_2_anno_2))
Color_Set=RColorBrewer::brewer.pal(5,"Set1")
names(Color_Set)=unique(Flip_TPM_2_anno_2$Group)
Color_list=list(Group=Color_Set)
```


```{r}
# Time course analyses with LRT
#The LRT test can be especially helpful when performing time course analyses. We can use the LRT to explore whether there are any significant differences in treatment effect between any of the timepoints.**

#For have an experiment looking at the effect of treatment over time on mice of two different genotypes. We could use a design formula for our ‘full model’ that would include the major sources of variation in our data: genotype, treatment, time, and our main condition of interest, which is the difference in the effect of treatment over time (treatment:time).**
# #https://rstudio-pubs-static.s3.amazonaws.com/329027_593046fb6d7a427da6b2c538caf601e1.html
# deseq2Data_add_control=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt , colData = Sample_anno_2,design = ~Drug+Timepoint+Drug:Timepoint)
# deseq2Data_add_control$Timepoint=relevel(deseq2Data_add_control$Timepoint,"0hour")
# dds_LRT= DESeq2::DESeq(deseq2Data_add_control,test="LRT", reduced =~Drug+Timepoint)
# DESeq2::resultsNames(dds_LRT)

deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =Raw_counts_2_flipped, colData = Sample_anno,design = ~Timepoint)
deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
```

```{r}
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt, colData = Sample_anno,design =~Group)
dds= DESeq2::DESeq(deseq2Data)
# DEGs: FC 1.2 and padj>0.05
Mirda_8h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_8hour","DMSO_8hour"))%>%as.data.frame()
Mirda_8h_vs_DMSO_sig=Mirda_8h_vs_DMSO%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_8h_vs_DMSO_sig)
Mirda_24h_vs_DMSO=DESeq2::results(dds,contrast = c("Group" ,"Mirda_24hour","DMSO_24hour"))%>%as.data.frame()
Mirda_24h_vs_DMSO_sig=Mirda_24h_vs_DMSO%>%dplyr::filter(padj<0.05)%>%dplyr::filter(padj<0.05&abs(log2FoldChange)>0.26)
nrow(Mirda_24h_vs_DMSO_sig)
```

```{r}

```




# Save data
```{r}
if (!file.exists(file.path(Rawdata_path,"Flipped_Deseq2_output.rdata"))){
save(myFilt,Sample_anno_2,deseq2VST,dds_LRT,file=file.path(Rawdata_path,"Flipped_Deseq2_output.rdata"))
}
```

```{r}
knitr::knit_exit()
```


