---
title: "7_DEGs_mining_DMSO_samples"
author: "`r config::get(file = 'config.yml')$document$author`"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    latex_engine: pdflatex
    fig_caption: yes
    highlight: haddock
    number_sections: yes

knit: (
  function(inputFile, encoding) {
    config=config::get(file = 'config.yml') 
    
    pSubTitle <- paste0(config$PT_ID, "7_DEGs_mining_DMSO_samples.pdf")
    
    base_output_dir <- paste0(config$Path,"/","Report")
    
    DIR=file.path(base_output_dir)
    
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      output_file = pSubTitle,
      output_dir =DIR )}
    )
---



```{r,include=F}
rm(list=ls())
```

```{r,include=F}
library(dplyr)
library(ggplot2)
library(synapser)
library(dplyr)
library(readr)
library(tidyr)
library(DESeq2)
library(grid)
library("tidyverse")
#library(SummarizedExperiment)

# install config
#BiocManager::install("SummarizedExperiment")


```




```{r}
source(file.path(getwd(),"Data_input.R"))

```



```{r}
# v3.9
#BiocManager::install("SummarizedExperiment") solve the error "names' slot must be set to null at all time"
#install summaryexperiment version 1.3
#BiocManager::install("SummarizedExperiment",version = "1.3")
```


```{r}
dir(Reference_path)
GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("c2|c5|h",.)]
# Define annotation for pheatmap
All_anno=rbind(Annotation_b1, Annotation_b2[, colnames(Annotation_b1)])
All_TPM=cbind(TPM_b1_2, TPM_b2_2)
new_anno <- All_anno %>%
    mutate(Sample_name = specimenID) %>%dplyr::filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

rownames(new_anno) <- new_anno$specimenID
new_anno$Group <- factor(new_anno$Group, levels = unique(new_anno$Group))
tpm=All_TPM[,new_anno$specimenID]

group_anno <- data.frame(Timepoint=new_anno$Timepoint,Drug=new_anno$Drug)
rownames(group_anno) <- rownames(new_anno)
Timepoint_colors <- c(
  "0h" = "#E69F00",  # Orange
  "8h" = "#56B4E9",  # Light blue
  "24h" = "#009E73"  # Teal
)
# list colors in breww Dark2
names(Timepoint_colors) <- levels(new_anno$Timepoint)

anno_colors <- list( Timepoint = Timepoint_colors,Drug=drug_colors)


neg_colors <- colorRampPalette(c("blue", "white"))(50)  # For -1 to 0
pos_colors <- colorRampPalette(c("white", "#990000"))(50)
custom_colors <- c(neg_colors, pos_colors)

# Define symmetric breaks
breaks_list <- seq(-1, 1, length.out = 101)

```
 

# All samples in 12192024 batch
```{r}
Outlier_samples=c("")
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == "12192024") %>%dplyr::filter(!specimenID %in% Outlier_samples)
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
Cor_method="pearson"
pheatlist=list()
PCA_list=list()
for (i in Drug_list) {
  drug_pickup <- Annotation_file_pickup %>% filter(Drug == i)
  Paired_DMSO <- Annotation_file_pickup %>% filter(Drug == "DMSO" & Date == unique(drug_pickup$Date))
  new_anno <- rbind(drug_pickup, Paired_DMSO) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

  rownames(new_anno) <- new_anno$specimenID
  new_anno$Group <- factor(new_anno$Group, levels = unique(new_anno$Group))
 # new_counts <-Counts_b2_2[, rownames(new_anno)]
  tpm=TPM_b2_2[,rownames(new_anno)]
 
  PCA_list[[i]]=PCA_Plot_3_shape(t(tpm),  Annotation_b2, "specimenID", "Drug", "Timepoint")+scale_color_manual(values=drug_colors)+ 
     scale_color_manual(values = drug_colors) +
     scale_shape_manual(values = c(16, 17, 18)) +
    labs(title = paste0("PCA plot for ", Drug_name), x = "PC1", y = "PC2") +
    theme_bw() + 
     theme(legend.position = "top")+geom_text(aes(label = specimenID),hjust = 1, vjust = -1, size = 3, angle = 45) 
 
  cor_mat <- cor(tpm, method = Cor_method)
  group_anno <- data.frame(Group = new_anno$Group)
  #rownames(group_anno) <- rownames(new_anno)
  #group_levels <- levels(new_anno$Group)
  #group_colors <- c("#999999", "#000000", "#D55E00","#E69F00")
  #names(group_colors) <- group_levels
  Group_anno_2= data.frame(Timepoint=new_anno$Timepoint,Drug=new_anno$Drug)
  rownames(Group_anno_2) <- rownames(new_anno)
  anno_color2=anno_colors
  anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno$Drug))]
  
# list colors in breww Dark2
  names(Timepoint_colors) <- levels(new_anno$Timepoint)
  pheatlist[[i]]=pheatmap::pheatmap(cor_mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = TRUE,
         main = i,
         annotation_col = Group_anno_2,
          annotation_row = Group_anno_2,
          annotation_colors = anno_color2,
        color = pos_colors )[[4]]
    #     color = colorRampPalette(c("blue", "white", "red"))(100))
         
}


```


```{r,fig.width=12,fig.height=16,include=T}
cowplot::plot_grid(plotlist = pheatlist, ncol = 2, nrow = 4)
```

```{r,fig.width=12,fig.height=18,include=T}
cowplot::plot_grid(plotlist =PCA_list, ncol = 2, nrow = 4)
```

# Remove "b2_12" and "b2_52" in DMSO group

```{r}
Outlier_samples=c("b2_12", "b2_52")
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == "12192024") %>%dplyr::filter(!specimenID %in% Outlier_samples)
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
Cor_method="pearson"
pheatlist=list()
PCA_list=list()
for (i in Drug_list) {
  drug_pickup <- Annotation_file_pickup %>% filter(Drug == i)
  Paired_DMSO <- Annotation_file_pickup %>% filter(Drug == "DMSO" & Date == unique(drug_pickup$Date))
  new_anno <- rbind(drug_pickup, Paired_DMSO) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

  rownames(new_anno) <- new_anno$specimenID
  new_anno$Group <- factor(new_anno$Group, levels = unique(new_anno$Group))
 # new_counts <-Counts_b2_2[, rownames(new_anno)]
  tpm=TPM_b2_2[,rownames(new_anno)]
  cor_mat <- cor(tpm, method = Cor_method)
  group_anno <- data.frame(Group = new_anno$Group)
  #rownames(group_anno) <- rownames(new_anno)
  #group_levels <- levels(new_anno$Group)
  #group_colors <- c("#999999", "#000000", "#D55E00","#E69F00")
  #names(group_colors) <- group_levels
  Group_anno_2= data.frame(Timepoint=new_anno$Timepoint,Drug=new_anno$Drug)
  rownames(Group_anno_2) <- rownames(new_anno)
  anno_color2=anno_colors
  anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno$Drug))]
  PCA_list[[i]]=PCA_Plot_3_shape(t(tpm),  Annotation_b2, "specimenID", "Drug", "Timepoint")+scale_color_manual(values=drug_colors)+ 
     scale_color_manual(values = drug_colors) +
     scale_shape_manual(values = c(16, 17, 18)) +
    labs(title = paste0("PCA plot for ", Drug_name), x = "PC1", y = "PC2") +
    theme_bw() + 
     theme(legend.position = "top")+geom_text(aes(label = specimenID),hjust = 1, vjust = -1, size = 3, angle = 45) 
# list colors in breww Dark2
  names(Timepoint_colors) <- levels(new_anno$Timepoint)
  pheatlist[[i]]=pheatmap::pheatmap(cor_mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = TRUE,
         main = i,
         annotation_col = Group_anno_2,
          annotation_row = Group_anno_2,
          annotation_colors = anno_color2,
        color = pos_colors )[[4]]
    #     color = colorRampPalette(c("blue", "white", "red"))(100))
         
}


```

```{r,fig.width=12,fig.height=16,include=T}
cowplot::plot_grid(plotlist = pheatlist, ncol = 2, nrow = 4)
```

```{r,fig.width=12,fig.height=16,include=T}
cowplot::plot_grid(plotlist = PCA_list, ncol = 2, nrow = 4)
```



# Remove only "b2_52" in DMSO group

```{r}
Outlier_samples=c("b2_52")
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == "12192024") %>%dplyr::filter(!specimenID %in% Outlier_samples)
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
Cor_method="pearson"
pheatlist=list()
PCA_list=list()
for (i in Drug_list) {
  drug_pickup <- Annotation_file_pickup %>% filter(Drug == i)
  Paired_DMSO <- Annotation_file_pickup %>% filter(Drug == "DMSO" & Date == unique(drug_pickup$Date))
  new_anno <- rbind(drug_pickup, Paired_DMSO) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

  rownames(new_anno) <- new_anno$specimenID
  new_anno$Group <- factor(new_anno$Group, levels = unique(new_anno$Group))
 # new_counts <-Counts_b2_2[, rownames(new_anno)]
  tpm=TPM_b2_2[,rownames(new_anno)]
  cor_mat <- cor(tpm, method = Cor_method)
  group_anno <- data.frame(Group = new_anno$Group)
  #rownames(group_anno) <- rownames(new_anno)
  #group_levels <- levels(new_anno$Group)
  #group_colors <- c("#999999", "#000000", "#D55E00","#E69F00")
  #names(group_colors) <- group_levels
  Group_anno_2= data.frame(Timepoint=new_anno$Timepoint,Drug=new_anno$Drug)
  rownames(Group_anno_2) <- rownames(new_anno)
  anno_color2=anno_colors
  anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno$Drug))]
  PCA_list[[i]]=PCA_Plot_3_shape(t(tpm),  Annotation_b2, "specimenID", "Drug", "Timepoint")+scale_color_manual(values=drug_colors)+ 
     scale_color_manual(values = drug_colors) +
     scale_shape_manual(values = c(16, 17, 18)) +
    labs(title = paste0("PCA plot for ", Drug_name), x = "PC1", y = "PC2") +
    theme_bw() + 
     theme(legend.position = "top")+geom_text(aes(label = specimenID),hjust = 1, vjust = -1, size = 3, angle = 45) 
# list colors in breww Dark2
  names(Timepoint_colors) <- levels(new_anno$Timepoint)
  pheatlist[[i]]=pheatmap::pheatmap(cor_mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = TRUE,
         main = i,
         annotation_col = Group_anno_2,
          annotation_row = Group_anno_2,
          annotation_colors = anno_color2,
        color = pos_colors )[[4]]
    #     color = colorRampPalette(c("blue", "white", "red"))(100))
         
}


```

```{r,fig.width=12,fig.height=16,include=T}
cowplot::plot_grid(plotlist = pheatlist, ncol = 2, nrow = 4)
```

```{r,fig.width=12,fig.height=16,include=T}
cowplot::plot_grid(plotlist = PCA_list, ncol = 2, nrow = 4)
```


```{r}
knitr::knit_exit()
```


# Minus samples b2_12,b2_52 in 12192024 batch
```{r,include=F}
#file.remove(file.path(Processed_path,"All_drug_RES_FGSEA.rdata"))
if (file.exists(file.path(Processed_path,"12172024_rm_DMSO_DEGs_GSEA.rdata"))) {
  load(file.path(Processed_path,"12172024_rm_DMSO_DEGs_GSEA.rdata"))
  
} else {

  
Rawcounts <- Counts_b2_2 %>%
  select(-c("b2_16", "b2_52"))
Result_list <- list()

for (i in Drug_list) {
  drug_pickup <- Annotation_file_pickup %>% filter(Drug == i)
  Paired_DMSO <- Annotation_file_pickup %>% filter(Drug == "DMSO" & Date == unique(drug_pickup$Date))
  new_anno <- rbind(drug_pickup, Paired_DMSO) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()

  rownames(new_anno) <- new_anno$specimenID
  new_anno$Group <- factor(new_anno$Group, levels = unique(new_anno$Group))
  colnames(Rawcounts) 
  new_counts <- Counts_b2_2%>%as.data.frame()%>%.[, rownames(new_anno)]

  if (all(colnames(new_counts) == rownames(new_anno))) {
    message("Sample matching correct")
  }

  keep.exprs <- selectGenes_BY_group(new_counts, min.count = 10, N = 0.90, new_anno)
  keep.exprs_2 <- keep.exprs%>%as.data.frame() %>% mutate(Sum = rowSums(.)) %>% filter(Sum > 1)
  myFilt <- Counts_b2_2[rownames(keep.exprs_2), ]
  myFilt <- myFilt[, colnames(myFilt) %in% rownames(new_anno)] %>% as.data.frame()
  new_anno <- new_anno[colnames(myFilt), ]
  all(colnames(myFilt) == rownames(new_anno))
  # find NA in myfilt
  if (any(is.na(myFilt))) {
    stop("There are NA values in the filtered count matrix.")
  }
  if (any(is.na(new_anno))) {
    stop("There are NA values in the annotation data frame.")
  }

  #stopifnot(all(colnames(myFilt) == rownames(new_anno)))
  setClassUnion("ExpData", c("matrix", "SummarizedExperiment"))
  deseq2Data <- DESeq2::DESeqDataSetFromMatrix(countData = myFilt, colData = new_anno, design = ~Group)
  dds <- DESeq2::DESeq(deseq2Data)
  res_list <- list()
  FGSEA_list <- list()
  Timepoint <- c("8h", "24h")

  for (t in Timepoint) {
    drug_0h <- paste0(i, "_0h")
    DMSO_t <- paste0("DMSO_", t)
    group1 <- paste0(i, "_", t)

    DESeq2::resultsNames(dds) # Check the names of the results 
    Res_out <- DESeq2::results(dds, contrast = c("Group", group1, DMSO_t))
    DEGs <- as.data.frame(Res_out) %>%dplyr:: filter(padj < 0.05 & abs(log2FoldChange) > 0.27)

    name2 <- paste0(i, "_", t, "_vs_DMSO_", t)
    name_deg <- paste0(i, "_", t, "_vs_DMSO_", t, "_DEGs")
    res_list[[t]] <- list()
    res_list[[t]][[name2]] <- Res_out
    res_list[[t]][[name_deg]] <- DEGs

    data <- as.data.frame(Res_out)
    res_q005 <- filter(data, padj < 0.05)
    res_q005_rank <- arrange(res_q005, log2FoldChange) %>%
      select(log2FoldChange) %>%
      tibble::rownames_to_column() %>%
      tibble::deframe()

    out_list <- list()
    for (x in seq_along(GMT_files)) {
      pathways <- fgsea::gmtPathways(GMT_files[x])
      set.seed(12345)
      out_list[[x]] <- fgsea::fgseaMultilevel(pathways, res_q005_rank,
                                              minSize = 10, maxSize = 500, nPermSimple = 10000)
    }

    FGSEA_out <- do.call("rbind", out_list)
    FGSEA_list[[t]] <- list()
    FGSEA_list[[t]][[name2]] <- FGSEA_out
  }

  Result_list[[i]] <- list(Res = res_list, FGSEA = FGSEA_list)
}

Result_list[[1]]$Res[[2]][[2]] 

names(Result_list) 

save(Result_list, file = file.path(Processed_path, "12172024_rm_DMSO_DEGs_GSEA.rdata"))
}
```




```{r}
# Function to extract DEG counts from Result_list

# Extract DEG counts.
DEG_summary <- get_DEG_counts(Result_list,logFC=0.27)

# Reshape data for plotting
DEG_summary_long <- DEG_summary %>%
  pivot_longer(cols = c("Upregulated", "Downregulated"), names_to = "Direction", values_to = "Count")

DEG_summary_long$Timepoint=factor(DEG_summary_long$Timepoint,levels=c("8h","24h"))
# Plot the number of upregulated and downregulated genes
P1=ggplot(DEG_summary_long, aes(x = Timepoint, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Drug, scales = "free_y") +
  theme_minimal() +ylim(c(0,4000))+
  labs(
    title = "Number of Upregulated and Downregulated Genes,|log2FC|>0.27, padj<0.05",
    x = "Timepoint",
  #  y = paste0("padj<0.05","_","log2FC>0.27"),
    fill = "Regulation Direction"
  ) +
  theme(
     axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10, face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  )

```

```{r,fig.width=12,fig.height=16,include=T}


P1
```


```{r}
knitr::knit_exit()
```



# first experiment on date 10172023 (Drug and DMSO are seperated)


```{r,fig.width=12,fig.height=16,include=T}
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == "10172023") 
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
new_anno2 <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

# label b2_12 and b2_52 WITH RED IN PHEATMAP

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)


```


# Mir,TRAB and 12192024 DMSO group (TRAB is in the same batch in 12192024 DMSO)

```{r,fig.width=12,fig.height=16,include=T}
Annotation_file_pickup <- new_anno 
Drug_list <- c("MIR","TRAB")
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
Drug_data <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%dplyr::filter(Drug%in%c(Drug_list)) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

DMSO_data <- new_anno %>% dplyr::filter(Drug %in% c("DMSO"))  %>%
    filter(Timepoint != "0h"&Date!="12272024") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

new_anno2=rbind(Drug_data, DMSO_data) 
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug, Exp_date=as.character(new_anno2$Date))
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

# label b2_12 and b2_52 WITH RED IN PHEATMAP

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)


```

# Mir,TRAB and 12272024 DMSO group (TRAB is NOT in the  batch in 12272024 DMSO)

```{r,fig.width=12,fig.height=16,include=T}
Annotation_file_pickup <- new_anno 
Drug_list <- c("MIR","TRAB")
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
Drug_data <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%dplyr::filter(Drug%in%c(Drug_list)) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

DMSO_data <- new_anno %>% dplyr::filter(Drug %in% c("DMSO"))  %>%
    filter(Timepoint != "0h"&Date!="12192024") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)

new_anno2=rbind(Drug_data, DMSO_data) 
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug, Exp_date=as.character(new_anno2$Date))
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

# label b2_12 and b2_52 WITH RED IN PHEATMAP

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)


```


# DMSO in 12192024 and 12272024 (batch effect in DMSO groups from different experiments)
```{r,fig.width=12,fig.height=10,include=F}
Exp_date=c("12192024","12272024")
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date %in% Exp_date) 
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
new_anno2 <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h"&Drug=="DMSO") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug,Exp_date=as.character(new_anno2$Date))
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
anno_color2$Exp_date <- c("12192024" = "#E69F00", "12272024" = "#56B4E9")  # Custom colors for Exp_date
cor_mat <- cor(tpm, method = "pearson")

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)
```


```{r,fig.width=12,fig.height=10,include=T}
# You can confirm it with: names(P$gtable$grobs)
col_labels_grob <- P$gtable$grobs[[5]]

# Extract original column names (ensure the order matches)
col_labels <- col_labels_grob$label

# Create a color vector: red if label matches, else black
label_colors <- ifelse(col_labels %in% c("b2_12", "b2_52"), "red", "black")

# Modify the text color
col_labels_grob$gp <- gpar(col = label_colors)

# Assign back the modified grob
P$gtable$grobs[[5]] <- col_labels_grob

### ===== Modify Row Name Colors =====
# Access by name, not index (more reliable)
row_labels_grob <- P$gtable$grobs[[6]]

# Safely extract row labels
if (!is.null(row_labels_grob$label)) {
  row_labels <- row_labels_grob$label

  # Assign red to selected row names
  row_colors <- ifelse(row_labels %in% c("b2_12", "b2_52"), "red", "black")

  # Apply text colors
  row_labels_grob$gp <- gpar(col = row_colors)

  # Put modified grob back
  P$gtable$grobs[[6]] <- row_labels_grob
}

# Draw updated plot
grid.newpage()
grid.draw(P$gtable)

```

# experiment on date 12192024 

```{r,fig.width=12,fig.height=16,include=F}
Exp_date="12192024"
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == Exp_date) 
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
new_anno2 <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors,

)
```


```{r,fig.width=12,fig.height=16,include=T}
# You can confirm it with: names(P$gtable$grobs)
col_labels_grob <- P$gtable$grobs[[5]]

# Extract original column names (ensure the order matches)
col_labels <- col_labels_grob$label

# Create a color vector: red if label matches, else black
label_colors <- ifelse(col_labels %in% c("b2_12", "b2_52"), "red", "black")

# Modify the text color
col_labels_grob$gp <- gpar(col = label_colors)

# Assign back the modified grob
P$gtable$grobs[[5]] <- col_labels_grob

### ===== Modify Row Name Colors =====
# Access by name, not index (more reliable)
row_labels_grob <- P$gtable$grobs[[6]]

# Safely extract row labels
if (!is.null(row_labels_grob$label)) {
  row_labels <- row_labels_grob$label

  # Assign red to selected row names
  row_colors <- ifelse(row_labels %in% c("b2_12", "b2_52"), "red", "black")

  # Apply text colors
  row_labels_grob$gp <- gpar(col = row_colors)

  # Put modified grob back
  P$gtable$grobs[[6]] <- row_labels_grob
}

# Draw updated plot
grid.newpage()
grid.draw(P$gtable)


```

```{r,fig.width=12,fig.height=12,include=T}
Exp_date="12192024"
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == Exp_date) 
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]

for (i in Drug_list) {
  new_anno2 <- Annotation_file_pickup%>%dplyr::filter(Drug%in%c("DMSO",i)) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors,
  silent = TRUE
)

# You can confirm it with: names(P$gtable$grobs)
col_labels_grob <- P$gtable$grobs[[5]]

# Extract original column names (ensure the order matches)
col_labels <- col_labels_grob$label

# Create a color vector: red if label matches, else black
label_colors <- ifelse(col_labels %in% c("b2_12", "b2_52"), "red", "black")

# Modify the text color
col_labels_grob$gp <- gpar(col = label_colors)

# Assign back the modified grob
P$gtable$grobs[[5]] <- col_labels_grob

### ===== Modify Row Name Colors =====
# Access by name, not index (more reliable)
row_labels_grob <- P$gtable$grobs[[6]]

# Safely extract row labels
if (!is.null(row_labels_grob$label)) {
  row_labels <- row_labels_grob$label

  # Assign red to selected row names
  row_colors <- ifelse(row_labels %in% c("b2_12", "b2_52"), "red", "black")

  # Apply text colors
  row_labels_grob$gp <- gpar(col = row_colors)

  # Put modified grob back
  P$gtable$grobs[[6]] <- row_labels_grob
}

# Draw updated plot
grid.newpage()
grid.draw(P$gtable)

}
```

# experiment on date 12272024 (rm RIB and NOR)

```{r,fig.width=12,fig.height=16,include=T}
Exp_date="12272024"
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == Exp_date) 
Drug_list <- unique(Annotation_file_pickup$Drug)
Drug_list <- Drug_list[!grepl("DMSO", Drug_list)]
new_anno2 <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

P=pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)



```

```{r,fig.width=12,fig.height=12,include=T}
Drug_list=Drug_list[!Drug_list %in% c("RIB", "VOR")]
for (i in Drug_list) {
  new_anno2 <- Annotation_file_pickup%>%dplyr::filter(Drug%in%c("DMSO",i)) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h")  %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)
 
}
```

# RIB and VOR
```{r,fig.width=12,fig.height=8,include=T}
Annotation_file_pickup <- All_anno %>%
  dplyr::filter(Date == Exp_date) 
 Drug_list=c("RIB", "VOR")
for (i in Drug_list) {
  new_anno2 <- Annotation_file_pickup%>%dplyr::filter(Drug%in%c("DMSO",i)) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h")  %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")

pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)
 
}
```

# RIB and VOR, USing DMSO in 12192024 as control
```{r,fig.width=12,fig.height=8,include=T}

Drug_list=c("RIB", "VOR")
for (i in Drug_list) {
  Drug_data=Annotation_file_pickup%>%dplyr::filter(Drug%in%c(i)) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h")  %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
  
  DMSO_data=new_anno%>%dplyr::filter(Drug%in%c("DMSO")) %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h"&Date=="12192024")  %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
  new_anno2 <- rbind(Drug_data, DMSO_data) 
  rownames(new_anno2) <- new_anno2$specimenID

  tpm=All_TPM[,new_anno2$specimenID]

  Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug,Exp_date=as.character(new_anno2$Date))
  rownames(Group_anno_2) <- rownames(new_anno2)
  anno_color2=anno_colors
  anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
  
  cor_mat <- cor(tpm, method = "pearson")

 pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)
 
}
```

```{r}
knitr::knit_exit()
```


```{r,fig.width=12,fig.height=16,include=T}
new_anno2 <- Annotation_file_pickup %>%
    mutate(Sample_name = specimenID) %>%
    filter(Timepoint != "0h") %>%
    mutate(Group = paste0(Drug, "_", Timepoint))%>%select(specimenID, Group, Sample_name, Timepoint, Drug, Date)%>%as.data.frame()%>%arrange(Drug,Timepoint)
rownames(new_anno2) <- new_anno2$specimenID

tpm=All_TPM[,new_anno2$specimenID]

Group_anno_2= data.frame(Timepoint=new_anno2$Timepoint,Drug=new_anno2$Drug)
rownames(Group_anno_2) <- rownames(new_anno2)
anno_color2=anno_colors
anno_color2$Drug <- anno_color2$Drug[as.character(unique(new_anno2$Drug))]
cor_mat <- cor(tpm, method = "pearson")



pheatmap::pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = Group_anno_2,
  annotation_row = Group_anno_2,
  annotation_colors = anno_color2,
  color = pos_colors 
)



```


```{r}
# Choose DMSO samples from cor_mat
DMSO_samples <- new_anno %>% filter(Drug == "DMSO") %>% pull(specimenID)
# Extract the correlation values for DMSO samples
cor_DMSO <- cor_mat[DMSO_samples, DMSO_samples]

pheatmap::pheatmap(
  cor_DMSO,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = TRUE,
  main = "Sample Correlation Heatmap (TPM)",
  annotation_col = group_anno,
  annotation_row = group_anno,
  annotation_colors = anno_colors,
  color = pos_colors 
)
```



```{r}
knitr::knit_exit()
```


```{r}
knitr::knit_exit()
```


