---
title: "Spheroid Proteomics Differential Expression Analysis"
author:
  - name: "Yannick Mahlich"
    affiliation: "Pacific Northwest National Laboratory"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide'}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
```

Next the import the proteomics for the global proteomics abundance from spheroids.
The data file is located on synapse and is stored as Excel sheet.

```{r import-proteomics-data, results='hide'}
synapseFileID = "syn72245489"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsData <- read_xlsx(fileEntity$path, sheet = 1)
tmp <- synapser::synLogout()
```

Finally we retrieve the metadata for the experiments from synapse.
The experimental setup pools proteins from several wells into one MS/MS run to
attain better proteomics coverage.
The information which wells were pooled together is contained in the metadata
table, specifically it contains the description of which chemical agent was used
on the spheroids in the drug response experiment and at which timepoint the
sample was extracted.
Sheet 2 of the Excel document retrieved from synapes contains the mapping that
describes which DIA run is associated to which drug and sampling time point.

```{r import-meta-data, results='hide'}
synapseFileID = "syn72245597"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsMeta <- read_xlsx(fileEntity$path, sheet = 2)
tmp <- synapser::synLogout()
```

## Data preparation

Before running the actual differential expression analysis we need to first
clean the data. This consits of several steps:

- Matching DIA run identifiers to chemical agent & sample time descriptor
- Missingness removal (i.e. exclude genes from consideration that weren't
  detected in a certain fraction of MS/MS runs)
- remove additional identifier columns (move into separate DataFrame)
- log2 transform the abundance / intensity data
- impute missing values (media center based on other readouts for the protein)

### Column renaming and general cleanup

Steps of the data cleanup process are:

- creating a "look up vector" for the renaming 
- renaming the columns with skipping of columns not in the look up vector via
  `any_of()`
- move additional identifier columns into seperate DF, and keep one we can use
  for a potential later enrichment analysis
```{r data-cleanup}
# creating a "look up vector" based on the metadata table
colNameVec <- setNames(proteomicsMeta$sample, proteomicsMeta$condition)

# using the look up vector to rename column
proteomicsDataClean <- proteomicsData %>% 
  dplyr::rename(dplyr::any_of(colNameVec))

# moving "unnecessary columns" into seperate DataFrame

proteomicsMapping <- proteomicsDataClean[0:4]
proteomicsDataClean <- proteomicsDataClean %>%
  column_to_rownames(., var = "Protein.Names") %>%
  within(., rm('Protein.Group', 'Genes', 'First.Protein.Description'))


head(proteomicsDataClean)
```
### Missingness removal
Next we perform a missingness removal i.e. we remove proteins from consideration
that aren't detected in at least xx% of the MS/MS DIA readouts

*This is currently a place holder. Needs confirmation!*

```{r missingness-removal, eval=FALSE}
proteomicsDataClean <- proteomicsDataClean[rowSums(is.na(proteomicsDataClean)) <= 3,]
head(proteomicsDataClean)
```

### log2 transform data and impute missing data
Following the removal of proteins with high missingness across the different
experiments, we will convert all columns to numeric such that we can
subsequently log2 transform each value.

```{r log-transform}

proteomicsDataClean <- proteomicsDataClean %>%
  mutate(across(everything(), as.numeric)) %>% # turn everything numeric
  log2(.) # log transform all cells

head(proteomicsDataClean)
```

### Imputation of NAs - mean protein abundance
The next step in the data preparation is to peform imputation for missing
protein abundances i.e. proteins that were not detected by the DIA workflow.
To this end we first calculate the mean protein abundance per protein across all
experiments.
Then we iterate over all columns and use the `coalesce()` function to set a any
cell that is NA to the mean of the row calculated prior.
Finally we remove the temporary column that had the mean values stored.

```{r impute-mean, eval=FALSE}
proteomicsDataClean <- proteomicsDataClean %>%
  mutate(imputeVal = rowMeans(., na.rm = TRUE)) %>% # calculate mean abundance per protein 
  mutate(., across(!imputeVal, ~coalesce(.x, imputeVal))) %>% # iterate over all columns and set N/As to mean
  within(., rm('imputeVal')) # remove 'imputeVal' column

head(proteomicsDataClean)
```

### Alternative imputation via minimum protein abundance
Alternatively to the imputation above we can also calculate the minimum of the
protein abundances across all DIA measurements.
Note that this method is currently not used in the notebook.
To use minimum based imputation set cell `eval=TRUE` (and add `eval=FALSE`
above).
```{r impute-minimum, eval=FALSE}
proteomicsDataClean <- proteomicsDataClean %>%
  rowwise %>%
  mutate(imputeVal = min(c_across(), na.rm = TRUE)) %>% # calculate median abundance per protein 
  mutate(., across(!imputeVal, ~coalesce(.x, imputeVal))) %>% # iterate over all columns and set N/As to median
  within(., rm('imputeVal')) # remove 'imputeVal' column

head(proteomicsDataClean)
```

### Imputation via minimum protein abundance of sample
The last possible way to impute that we implement is to calculate the minimum
protein abundance within a sample (e.g. one replicate of one experiment).
For now this is the implementation of imputation that we will use.

The implementation makes use of the `colMins` function from the `matrixStats`
package to calculate a vector that records the minimal protein intensity values
per column (sample). This vector/list is then used with `tidyr::replace_na()`
to back fill all N/A values in a given column with the calculated column
minimum.
```{r impute-sample-minimum}
sampleMins <- colMins(
  as.matrix(proteomicsDataClean),
  na.rm = TRUE
  )

proteomicsDataClean <- proteomicsDataClean %>%
  replace_na(., as.list(sampleMins))

head(proteomicsDataClean)
```

### Averaging protein abundances across replicates
Since we want to do differential expression analysis and limma needs multiple 
replicates per condition the averaging below won't be needed. For now we will
keep the code, but not execute it.
```{r dereplication, eval=FALSE}
experiments <- gsub("_r[123]$", "", colnames(proteomicsDataClean)) %>% unique()
for (experiment in experiments) {
  proteomicsDataClean <- proteomicsDataClean %>%
    rownames_to_column(., 'proteins') %>%
    rowwise %>%
    mutate(., "{experiment}" := mean(c_across(starts_with(experiment)))) %>%
    select(., !starts_with(paste0(experiment, "_"))) %>%
    column_to_rownames(., 'proteins')
}
head(proteomicsDataClean)
```

## Differential Expression analysis
With all the data imported we can focus on the actual differential expression
analysis.

To that end we will first define some helper functions that will make use of the
limma package to calculate said differential expression results.

### Function definitions
```{r limma-diffex}

getDiffExprTable <- function(data, x, y) {
  if (!is.data.frame(data)){
    stop("Argument 'data' must be data.frame", call. = FALSE)
  }
  if (!is.vector(x)){
    stop("Argument 'x' must be vector", call. = FALSE)
  }
  if (!is.vector(y)){
    stop("Argument 'y' must be vector", call. = FALSE)
  }
  
  fac <- factor(rep(c(2, 1), c(length(y), length(x))))
  design <- stats::model.matrix(~fac)
  fit <- limma::lmFit(data[, c(y, x)], design)
  fit <- limma::eBayes(fit)
  resultTable <- limma::topTable(fit, coef = 2, number = Inf)
  return(resultTable)
}
```

```{r diffExp}
x <- c('24hr_DMSO_r1', '24hr_DMSO_r2', '24hr_DMSO_r3')
y <- c('24hr_Mirda_r1', '24hr_Mirda_r2', '24hr_Mirda_r3')
diffExpResults <- getDiffExprTable(proteomicsDataClean, x, y)
diffExpResults
```
