---
title: "Spheroid Phospho Proteomics Analysis"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
---


## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide'}
library(synapser)
library(readxl)
library(readr)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
```

Next is the import of the phospho proteomics abundances from spheroids.
The data file is located on synapse and is stored as tsv.

```{r import-phospho-data, results='hide'}
synapseFileID = "syn72351272"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsData <- readr::read_tsv(
  fileEntity$path,
  na = c("0", NA) # the file stored on synapse has '0's instead of NA or ''
  )
tmp <- synapser::synLogout()
```

```{r import-meta-data, results='hide'}
synapseFileID = "syn72245597"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsMeta <- read_xlsx(fileEntity$path, sheet = 3)
tmp <- synapser::synLogout()
```


## Defining helper functions

*TODO: Need to offload the helper functions into a seperate R file!*

Next we will be defining some helper functions. Since we will be storing all the
experimental results in a `SummarizedExperiment` object, those functions will
work with said object structure.

Specifically we define the following functions:

- `log_transform(...)` which will take a `SummarizedExperiment` object, retrieve
  the defined assay and log transform all intensity values and return a
  `SummarizedExperiment` object containing the log transformed values
- `impute_proteomics(...)` which will take a `SummarizedExeperiment` and impute
  any missing intensity values in the defined assay. The function will return an
  updated `SummarizedExepriment` object.

```{r}
log_transform <- function(experiment,
                          assay = "proteomics"){
  
  proteomics_data <- SummarizedExperiment::assays(experiment)[[assay]] %>%
    mutate(across(everything(), as.numeric)) %>% # turn everything numeric
    log2(.) # log transform all cells
  
  SummarizedExperiment::assays(
    experiment, 
    withDimnames = FALSE)[[assay]] <- proteomics_data 
  return(experiment)
}

impute_proteomics <- function(experiment,
                              assay = 'proteomics',
                              imputation_method = 'SampMin'){
  
  # retrieve assay from SummarizedExperiment and make sure that all values
  # are numeric such that we can impute
  proteomics_data <- SummarizedExperiment::assays(experiment)[[assay]] %>%
    mutate(across(everything(), as.numeric)) # turn every cell numeric
  
  if(imputation_method == "SampMin") {
    
    # first calculates the minimum protein intensity per sample (column) using
    # `matrixStats::colMins()`
    sample_mins <- matrixStats::colMins(
      as.matrix(proteomics_data),
      na.rm = TRUE
      )
    # then NAs in each sample (column) to the sample minimum (stored as list)
    proteomics_data <- proteomics_data %>%
      replace_na(., as.list(sample_mins))
    
  } else if (imputation_method == "means") {
    
    # calculates the mean abundance of a given feature (protein) across all
    # samples (columns) via `base::rowMeans()` stores it temporarily into a new
    # column `imputeVal` and then iterates over all columns (excluding the newly
    # generated column) to set N/As to the imputation value using 
    # `dplyr::coalesce`. Finally the temporary column is removed from the data.
    proteomics_data <- proteomics_data %>%
      mutate(imputeVal = rowMeans(., na.rm = TRUE)) %>%
      mutate(., across(!imputeVal, ~coalesce(.x, imputeVal))) %>%
      within(., rm('imputeVal'))
    
  } else if (imputation_method == "median") {
    
    # analogous to `imputation_method == "median"`. Uses `dplyr::rowwise` with
    # `dplyr::c_across()` in the `median()` call to achieve the same thing,
    # except calculating the medians of a given feature across all samples. 
    proteomics_data <- proteomics_data %>%
      rowwise %>%
      mutate(imputeVal = median(c_across(), na.rm = TRUE)) %>%
      mutate(., across(!imputeVal, ~coalesce(.x, imputeVal))) %>%
      within(., rm('imputeVal'))
    
  } else {
    stop()
  }
  
  # finally we store the imputed data back into the `SummarizedExperiment`
  # object and return it. 
  SummarizedExperiment::assays(
    experiment, 
    withDimnames = FALSE)[[assay]] <- proteomics_data 
  return(experiment)
  
} 

calc_diff_ex <- function(experiment,
                         group_1,
                         group_2,
                         assay = "proteomics"){
  
  
  
    # retrieving the samples (column ids) of the groups we want to use for the
    # differential expression analysis
    samples_1 <- which(
      SummarizedExperiment::colData(experiment)[["group"]] == group_1
      )
    samples_2 <- which(
      SummarizedExperiment::colData(experiment)[["group"]] == group_2
      ) 
    
    # next we set up the design matrix for the fit function used by limma to
    # determine the differentially expressed features. The design matrix uses
    # a factor based on the samples
    fac <- factor(rep(c(2, 1), c(length(samples_2), length(samples_1))))
    design <- stats::model.matrix(~fac)

    # next we extract the assay of interest from the `SummarizedExperiment`
    # object and create the differential expression result table. Note that we
    # switch the order of groups here for `limma`
    dat <- SummarizedExperiment::assays(experiment)[[assay]]
    
    result_table <- dat[, c(samples_2, samples_1)] %>%
      limma::lmFit(., design) %>%
      limma::eBayes(.) %>%
      limma::topTable(., coef = 2, number = Inf, sort.by = "none")
  
    # The assumption is that the input `SummarizedExperiment` had only one assay
    # associated with it, hence we will just augment the object and return it
    # containing the results from limma.
  
    # comparison_name <- paste(group_1, "_vs_", group_2, sep = "")
    # colnames_res <- paste(paste(comparison_name, 
    #                             colnames(result_table), "limma", sep = "."))
    # 
    # colnames(result_table) <- colnames_res
    result_table <- result_table %>%
      rownames_to_column(., var="Protein.With.Phospho.Site") %>%
      as.tibble(.)

    # This makes sure that the differential expression results for each feature
    # are properly aligned
    SummarizedExperiment::rowData(experiment) <- left_join(
      as_tibble(SummarizedExperiment::rowData(experiment)),
      result_table,
      by = 'Protein.With.Phospho.Site'
      )
  
    return(experiment)
}

prep_data <- function(proteomics_data, proteomics_meta, missingness_cutoff = 0.5) {
    
  proteomicsMetaClean <- proteomics_meta %>%
    separate_wider_regex(
      .,
      col = "condition",
      c("group" = ".*", "_", "replicate" = ".*"),
      cols_remove = FALSE
      ) %>% 
    separate_wider_delim(
      .,
      col = "group",
      delim = "_",
      names = c("timepoint", "agent"),
      too_few = "align_start",
      cols_remove = FALSE
      )
  
  proteomicsDataClean <- proteomics_data %>%
    unite(., "Protein.With.Phospho.Site", Protein, Residue, Site, sep = "_") %>%
    column_to_rownames(., var = "Protein.With.Phospho.Site") %>%
    within(., rm("Protein.Names", "Gene.Names", "Sequence"))
  
  # TODO: add check to make sure missgness_cutoff is [0, 1]
  proteomicsDataClean %<>%
    .[rowSums(!is.na(.))/ncol(.) >= missingness_cutoff,] %>%
    rownames_to_column(., var = "Protein.With.Phospho.Site")
  
  proteomicsMapping <- proteomics_data[0:5] %>%
    separate_wider_delim(
      .,
      col = "Gene.Names",
      names = c("Gene.Names", NA),
      delim = ";",
      cols_remove=TRUE,
      too_few = "align_start",
      too_many = "drop"
      ) %>%
    unite(., "Protein.With.Phospho.Site", Protein, Residue, Site,
          sep = "_", remove = FALSE) %>%
    semi_join(., proteomicsDataClean, by = "Protein.With.Phospho.Site")

  proteomicsDataClean %<>% column_to_rownames(., var = "Protein.With.Phospho.Site")
    
  experiment <- SummarizedExperiment(
    assays = list(proteomics = proteomicsDataClean),
    rowData = proteomicsMapping,
    colData = proteomicsMetaClean
  )
  
  return(experiment)
}
```

## Data preparation

Before running the actual differential expression analysis we need to first
clean the data. This consits of several steps:

- Matching DIA run identifiers to chemical agent & sample time descriptor
- Missingness removal (i.e. exclude genes from consideration that weren't
  detected in a certain fraction of MS/MS runs)
- remove additional identifier columns (move into separate DataFrame)
- log2 transform the abundance / intensity data
- impute missing values (media center based on other readouts for the protein)

- `prep_data` will take of most of the data prep including the missingness
  removal since we need to make sure that the identifier table and data table 
  that goes into the `SummarizedExperiment` match up.
- `log_transform` & `impute_proteomics` take care of their namesake.

```{r}
experiment <- prep_data(proteomicsData, proteomicsMeta, 0) %>%
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method = "SampMin")
```


## Differential Expression analysis

### PCA (No missingness removal)
With all the data imported we can focus on the actual differential expression
analysis.
First we will perform a PCA analysis of the underlying proteomics data to check
if there is meaningful signal separation of the individual treatment agents,
as well as timepoints.

*Note that no phospho sites have been removed due to missingness removal*
*i.e. all N/As are being imputed!*

```{r}
principal_components <- prcomp(t(assay(experiment))) %$%
  cbind(x, colData(experiment))

plot <- (
  ggplot(
    principal_components,
    aes(
      x = PC1,
      y = PC2,
      col = agent,
      shape = timepoint,
      )
    )
  + geom_point(size = 2)
  + theme_minimal()
  + scale_color_manual(values = c("DMSO" = "#e66101", "Mirda" = "#5e3c99", "NA" = "grey"))
  + labs(
    title = "PCA of spheroid phospho proteomics data (NO missingness removal)",
  )
)

plot
```

### PCA (50% missingness removal)
Now we repeat the PCA plot but with 50% missingness removal, i.e. only protein
phosphorylations that were detected in >= 80% of the samples are being
considered for the PCA generation.

```{r}
experiment <- prep_data(proteomicsData, proteomicsMeta, 0.50) %>%
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method = "SampMin")

principal_components <- prcomp(t(assay(experiment))) %$%
  cbind(x, colData(experiment))

plot <- (
  ggplot(
    principal_components,
    aes(
      x = PC1,
      y = PC2,
      col = agent,
      shape = timepoint,
      )
    )
  + geom_point(size = 2)
  + theme_minimal()
  + scale_color_manual(values = c("DMSO" = "#e66101", "Mirda" = "#5e3c99", "NA" = "grey"))
  + labs(
    title = "PCA of spheroid phospho proteomics data (50% missingness removal)",
  )
)

plot
```
As can be seen, the data after missingness removal separates much better than
the full dataset, albeit mostly by timepoint and not so much by drug.

*For the remainder of the analysis we will use the dataset after missingness
removal*

### Number of significant differentally expressed phospho sites 
Next we will perform the actual differential expression analysis.
We will define a significantly differentially experessed phospho site as

- having an adjusted p-Value < 0.05
- having a logFC >= 1 or logFC < -1

We will be investigating specifically the expression differential between DMSO
and Mirdametenib treated samples both at the 24h as well as 48h time point.
The codeblock below performs the differential expression analyses using the
previously defined function `calc_diff_ex`.
The we select the proteins that show a significant differential in their
expression between the two groups and prepare a `data.frame` for plotting the
results.
```{r differential-expression-dmso-mirda}
dmso_mirda_24h <- calc_diff_ex(experiment, group_1 = "24hr_DMSO", group_2 = "24hr_Mirda")
dmso_mirda_48h <- calc_diff_ex(experiment, group_1 = "48hr_DMSO", group_2 = "48hr_Mirda")

dfPlot <- subset(as_tibble(rowData(dmso_mirda_24h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val'), adj.P.Val < 0.05) %>% mutate(expression_change = case_when(
  logFC >= 1 ~ 'up',
  logFC < -1 ~ 'down'
))
dfPlot$timepoint = "24hr"

dfPlot2 <- subset(as_tibble(rowData(dmso_mirda_48h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val'), adj.P.Val < 0.05) %>% mutate(expression_change = case_when(
  logFC >= 1 ~ 'up',
  logFC < -1 ~ 'down'
))
dfPlot2$timepoint = '48hr'

dfPlot <- rbind(dfPlot, dfPlot2) %>% subset(., !is.na(expression_change))
```

The code block below plots the number of significantly up and down regulated
proteins.
```{r plot-expression-differential}
cpalette = c('up' = '#74add1', 'down' = '#f46d43')

plot <- (
  ggplot(
    data = dfPlot,
    aes(timepoint)
  )
  + geom_bar(aes(fill=expression_change), position = 'dodge2')
  + scale_fill_manual(values = cpalette)
  + theme_minimal()
  + labs(
    title = 'Differentially expressed proteins in DMSO vs. Mirdametinib',
    fill = 'expression'
  )
)
plot
```

## Volcano plots of differentially expressed porteins
Next we generate volcano plots to investigate the significant proteins further.
For each time point one volcano plot will be generated.

### DMSO vs Mirdametenib at 24h
```{r volcano-plot-24h}

dfPlot <- subset(as.tibble(rowData(dmso_mirda_24h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val')) %>% 
  mutate(expression_change = case_when(
    (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
    (logFC < -1 & adj.P.Val < 0.05) ~ 'down',
    .default = 'not significant', 
  )) %>%
  mutate(labels = case_when(
    expression_change %in% c('up', 'down') ~ Protein.With.Phospho.Site,
    .default = NA
  ))

dfPlot$timepoint = "24hr"


plot <- (
  ggplot(
    data = dfPlot,
    aes(
      x = logFC,
      y = -log10(adj.P.Val),
      col = expression_change,
      # label = labels
      )
    )
    + geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed')
    + geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed')
    + geom_point(size = .75)
    # + geom_text_repel(max.overlaps = 10, size = 2)
    + scale_color_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + theme_minimal()
    + labs(
      title = 'Differentially expressed proteins in DMSO vs. Mirdametinib at 24h',
      col = 'expression',
      y = expression("-log"[10]*"p-Value")
    )
)
plot
```

### DMSO vs Mirdametenib at 48h
```{r volcano-plot-48h}

dfPlot <- subset(as.tibble(rowData(dmso_mirda_48h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val')) %>% 
  mutate(expression_change = case_when(
    (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
    (logFC < -1 & adj.P.Val < 0.05) ~ 'down',
    .default = 'not significant', 
  )) %>%
  mutate(labels = case_when(
    expression_change %in% c('up', 'down') ~ Protein.With.Phospho.Site,
    .default = NA
  ))

dfPlot$timepoint = "48hr"


plot <- (
  ggplot(
    data = dfPlot,
    aes(
      x = logFC,
      y = -log10(adj.P.Val),
      col = expression_change,
      # label = labels
      )
    )
    + geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed')
    + geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed')
    + geom_point(size = .75)
    # + geom_text_repel(max.overlaps = 10, size = 2)
    + scale_color_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + theme_minimal()
    + labs(
      title = 'Differentially expressed proteins in DMSO vs. Mirdametinib at 48h',
      col = 'expression',
      y = expression("-log"[10]*"p-Value")
    )
)
plot
```

