---
title: "Ex1 Global proteomics"
output: html_document
---

The global proteomics pipeline uses the R package [PlexedPiper](https://github.com/vladpetyuk/PlexedPiper). It also requires a connection to the DMS to access data packages.


```{r, echo=F}
library(knitr)
knitr::opts_chunk$set(message=F, warning=F)
t0 <- Sys.time(); print(t0)
```


```{r setup}
library(plyr)
library(MSnID)
library(PlexedPiper)
library(dplyr)
library(PNNL.DMS.utils)
library(Biostrings)
library(tidyverse)

# only human library: 5792
# human hg38, uniprot mouse and rat: 5878
data_package_num <- 5878
data_folder <- "data/global_data_w_mouse_rat"
crosstab_prefix <- "Chr8_UMN1"

dir.create(data_folder)

```


# Study design

Study design information in PlexedPiper is encoded in three tables: fractions, samples, and references. 
These tables can be made using metadata and should be stored on the DMS before processing.

```{r read_study_design}
fractions <- read.table("study_design/global_study_design/fractions.txt", sep = "\t", header = T)
samples <- read.table("study_design/global_study_design/samples.txt", sep = "\t", header = T)
references <- read.table("study_design/global_study_design/references.txt", sep = "\t", header = T)

```

# Process MS-GF+ data

MS-GF+ data is processed in several steps. First, read MS-GF+ output from the DMS. (This step can take a while).

```{r read_msgf}
msgf_data_path <- file.path(data_folder, "msgfData_original.RData")
if (file.exists(msgf_data_path)) {
  load(msgf_data_path)
} else {
  msnid <- read_msgf_data_from_DMS(data_package_num)
  save(msnid, file=msgf_data_path)
}
show(msnid)

```


## Correct peak selection and peptide FDR
We use the target-decoy search strategy method described in [(Elias 2010)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2922680/). 
Filtering is done first at peptide level, then at protein level, both with max FDR of 1%.

```{r}
## peak selection & Peptide filter
msnid <- correct_peak_selection(msnid)   ## optional for phospho
msnid <- filter_msgf_data_peptide_level(msnid, 0.01)


```


## Remap accessions

This function remaps UniProt protein accessions to gene symbol.

```{r remap_accessions}
options(timeout = 120)
fst_dms_pth <- path_to_FASTA_used_by_DMS(data_package_num)
fst <- readAAStringSet(fst_dms_pth)

conv <- data.frame(NAME = names(fst)) %>%
  mutate(UNIPROT = str_extract(names(fst), "^(\\S)+"),
         SYMBOL = str_extract(names(fst), "GN=(\\S)+"),
         HUMAN = grepl("Homo sapiens", names(fst)),
         MOUSE = grepl("Mus musculus", names(fst)),
         RAT = grepl("Rattus norvegicus", names(fst))) %>%
  dplyr::select(-NAME) %>%
  mutate(UNIPROT = sub(".*\\|(.*)\\|.*", "\\1", UNIPROT),
         SYMBOL = sub("GN=", "", SYMBOL))
conv$SPECIES <- NA
conv[conv$HUMAN,]$SPECIES <- "Human"
conv[conv$MOUSE,]$SPECIES <- "Mouse"
conv[conv$RAT,]$SPECIES <- "Rat"

msnid_prot <- msnid

organisms <- c("Homo sapiens", "Mus musculus", "Rattus norvegicus")

# extract non-human data
nonHuman <- c("Mouse", "Rat")
msnid.list <- list()
for (i in nonHuman) {
  temp.msnid <- msnid
  temp.accessions <- temp.msnid$accession
  temp.accessions[!grepl(i, temp.accessions,
                           ignore.case = TRUE)] <- NA
  temp.msnid$accession <- temp.accessions
  show(temp.msnid)
  temp.msnid <- apply_filter(temp.msnid, "!is.na(accession)")
  show(temp.msnid)
  msnid.list[[i]] <- temp.msnid
}

# extract human data - we can use this approach because of different library format for human
temp.msnid <- msnid
temp.accessions <- temp.msnid$accession
for (i in nonHuman) {
  temp.accessions[grepl(i, temp.accessions, ignore.case = TRUE)] <- NA
}
temp.msnid$accession <- temp.accessions
show(temp.msnid)
temp.msnid <- apply_filter(temp.msnid, "!is.na(accession)")
show(temp.msnid)
msnid.list[["Human"]] <- temp.msnid

### extract or convert to UniProt IDs
## extract UniProt IDs from non-human
for (i in nonHuman) {
  temp.accessions <- data.frame(msnid.list[[i]]$accession)
  temp.accessions$new <- NA
  for (j in 1:nrow(temp.accessions)) {
   temp.accessions$new[j] <- strsplit(temp.accessions[j,1], "[|]")[[1]][2] 
  }
  msnid.list[[i]]$accession <- temp.accessions$new
  show(msnid.list[[i]])
  msnid.list[[i]] <- apply_filter(msnid.list[[i]], "!is.na(accession)")
  show(msnid.list[[i]])
}

## convert RefSeq to UniProt IDs for human
# make conversion table from fasta
conv_human <- data.frame(NAME = names(fst)) %>%
  mutate(REFSEQ = str_extract(names(fst), "^(\\S)+"),
         HUMAN = grepl("Homo sapiens", names(fst)))
conv_human <- conv_human[conv_human$HUMAN,]

# get conversion for RefSeq to UniProt accessions
library(httr)

isJobReady <- function(jobId) {
  pollingInterval = 5
  nTries = 20
  for (i in 1:nTries) {
    url <- paste("https://rest.uniprot.org/idmapping/status/", jobId, sep = "")
    r <- GET(url = url, accept_json())
    status <- content(r, as = "parsed")
    if (!is.null(status[["results"]]) || !is.null(status[["failedIds"]])) {
      return(TRUE)
    }
    if (!is.null(status[["messages"]])) {
      print(status[["messages"]])
      return (FALSE)
    }
    Sys.sleep(pollingInterval)
  }
  return(FALSE)
}

getResultsURL <- function(redirectURL) {
  if (grepl("/idmapping/results/", redirectURL, fixed = TRUE)) {
    url <- gsub("/idmapping/results/", "/idmapping/stream/", redirectURL)
  } else {
    url <- gsub("/results/", "/results/stream/", redirectURL)
  }
}

files = list(
  ids = paste(conv_human$REFSEQ, collapse=","),
  from = "RefSeq_Protein",
  to = "UniProtKB"
)
r <- POST(url = "https://rest.uniprot.org/idmapping/run", body = files, encode = "multipart", accept_json())
submission <- content(r, as = "parsed")

if (isJobReady(submission[["jobId"]])) {
  url <- paste("https://rest.uniprot.org/idmapping/details/", submission[["jobId"]], sep = "")
  r <- GET(url = url, accept_json())
  details <- content(r, as = "parsed")
  url <- getResultsURL(details[["redirectURL"]])
  # Using TSV format see: https://www.uniprot.org/help/api_queries#what-formats-are-available
  url3 <- paste(url, "?format=xlsx", sep="")
  r <- GET(url = url3, accept_json())
  download.file(url3, "temp_uniprot.xlsx")
  resultsTable <- readxl::read_excel("temp_uniprot.xlsx")
  #print(head(resultsTable))
}

uniprot.results <- distinct(resultsTable[,1:2]) # no rows lost but there are multiple uniprot IDs for some proteins
RefSeq <- unique(uniprot.results$From)
refseq.uniprot <- data.frame(RefSeq)
refseq.uniprot$UniProt <- NA
for (i in 1:nrow(refseq.uniprot)) {
  refseq.uniprot$UniProt[i] <- paste(uniprot.results[uniprot.results$From == RefSeq[i],]$Entry, collapse="|")
}

# switch RefSeq to UniProt IDs
temp.accessions <- data.frame(msnid.list[["Human"]]$accession)
colnames(temp.accessions) <- "RefSeq"
temp.accessions2 <- merge(temp.accessions, refseq.uniprot, all.x = TRUE)
sortAccessions <- match(temp.accessions$RefSeq, temp.accessions2$RefSeq)
temp.accessions2 <- temp.accessions2[sortAccessions,] # match original order
msnid.list[["Human"]]$accession <- temp.accessions2$UniProt
show(msnid.list[["Human"]])
msnid.list[["Human"]] <- apply_filter(msnid.list[["Human"]], "!is.na(accession)")
show(msnid.list[["Human"]])

# group by unique peptide sequence since peptides may have multiple protein entries
species <- names(msnid.list)
for (i in species) {
  temp.psms <- psms(msnid.list[[i]])
  message(i, " msnid before grouping by peptide")
  show(msnid.list[[i]])
  og.colnames <- colnames(temp.psms)
  #numCols <- colnames(dplyr::select_if(temp.psms, is.numeric))
  #nonNumCols <- colnames(temp.psms)[!(colnames(temp.psms) %in% numCols)]
  temp.psms.agg <- ddply(temp.psms, .(Dataset, FragMethod, spectrumFile, peptide, pepSeq, isDecoy), summarize,
                         Protein = ifelse(length(unique(Protein)) > 1, paste0(unique(Protein), collapse = ", "), unique(Protein)),
                         accession = ifelse(length(unique(accession)) > 1, paste0(unique(accession), collapse = "|"), unique(accession)),
                         ResultID = mean(ResultID, na.rm = TRUE),
                         Scan = mean(Scan, na.rm = TRUE),
                         SpecIndex = mean(SpecIndex, na.rm = TRUE),
                         Charge = mean(Charge, na.rm = TRUE),
                         PrecursorMZ = mean(PrecursorMZ, na.rm = TRUE),
                         DelM = mean(DelM, na.rm = TRUE),
                         DelM_PPM = mean(DelM_PPM, na.rm = TRUE),
                         MH = mean(MH, na.rm = TRUE),
                         NTT = mean(NTT, na.rm = TRUE),
                         DeNovoScore = mean(DeNovoScore, na.rm = TRUE),
                         MSGFScore = mean(MSGFScore, na.rm = TRUE),
                         MSGFDB_SpecEValue = mean(MSGFDB_SpecEValue, na.rm = TRUE),
                         Rank_MSGFDB_SpecEValue = mean(Rank_MSGFDB_SpecEValue, na.rm = TRUE),
                         EValue = mean(EValue, na.rm = TRUE),
                         QValue = mean(QValue, na.rm = TRUE),
                         PepQValue = mean(PepQValue, na.rm = TRUE),
                         IsotopeError = mean(IsotopeError, na.rm = TRUE),
                         calculatedMassToCharge = mean(calculatedMassToCharge, na.rm = TRUE),
                         chargeState = mean(chargeState, na.rm = TRUE),
                         experimentalMassToCharge = mean(experimentalMassToCharge, na.rm = TRUE),
                         spectrumID = mean(spectrumID, na.rm = TRUE),
                         msmsScore = mean(msmsScore, na.rm = TRUE),
                         absParentMassErrorPPM = mean(absParentMassErrorPPM, na.rm = TRUE))
  psms(msnid.list[[i]]) <- temp.psms.agg[,og.colnames]
  message(i, " msnid after grouping by peptide")
  show(msnid.list[[i]])
}

```


## Accession FDR


```{r protein_level_filter }
compute_num_peptides_per_1000aa_BG_human <- function(msnid, fst_dms_pth, refseq.uniprot) {
  mySequences <- readAAStringSet(fst_dms_pth)
  prot_lengths <- data.frame(accession = sub(" .*", "", names(mySequences)), 
    Length = width(mySequences), stringsAsFactors = FALSE)
  # accession should be the first column of refseq.uniprot
  colnames(refseq.uniprot)[1] <- "accession"
  length.acc.before <- nrow(prot_lengths)
  prot_lengths <- merge(prot_lengths, refseq.uniprot, all.x = TRUE)
  length.acc.after <- nrow(prot_lengths)
  if (length.acc.before != length.acc.after) {stop("merge changed length of prot_lengths")}
  prot_lengths$accession <- NULL
  prot_lengths <- prot_lengths[,c(colnames(refseq.uniprot)[2], "Length")]
  colnames(prot_lengths)[1] <- "accession"
  
  prot_lengths <- mutate(prot_lengths, accession = paste0("XXX_", 
    accession)) %>% rbind(prot_lengths, .)
  pepN1000 <- psms(msnid) %>% dplyr::select(accession, peptide, 
    isDecoy) %>% distinct() %>% group_by(accession) %>% 
    summarise(pepN = n()) %>% inner_join(prot_lengths, by = "accession") %>% 
    mutate(pep_per_1000 = 1000 * pepN/Length)
  pepN1000_vec <- pepN1000$pep_per_1000
  names(pepN1000_vec) <- pepN1000$accession
  msnid$peptides_per_1000aa <- pepN1000_vec[msnid$accession]
  return(msnid)
}

compute_num_peptides_per_1000aa_BG_nonHuman <- function(temp.msnid, fst_dms_pth) {
  mySequences <- readAAStringSet(fst_dms_pth)
  prot_lengths <- data.frame(accession = sub(" .*", "", names(mySequences)), 
    Length = width(mySequences), stringsAsFactors = FALSE)
  for (j in 1:nrow(prot_lengths)) {
   prot_lengths$accession[j] <- strsplit(prot_lengths$accession[j], "[|]")[[1]][2] 
  }
  
  prot_lengths <- mutate(prot_lengths, accession = paste0("XXX_", 
    accession)) %>% rbind(prot_lengths, .)
  pepN1000 <- psms(temp.msnid) %>% dplyr::select(accession, peptide, 
    isDecoy) %>% distinct() %>% group_by(accession) %>% 
    summarise(pepN = n()) %>% inner_join(prot_lengths, by = "accession") %>% 
    mutate(pep_per_1000 = 1000 * pepN/Length)
  pepN1000_vec <- pepN1000$pep_per_1000
  names(pepN1000_vec) <- pepN1000$accession
  temp.msnid$peptides_per_1000aa <- pepN1000_vec[temp.msnid$accession]
  return(temp.msnid)
}

for (i in species) {
  if (i == "Human") {
    msnid.list[[i]] <- compute_num_peptides_per_1000aa_BG_human(msnid.list[[i]], fst_dms_pth, refseq.uniprot)
  } else {
    msnid.list[[i]] <- compute_num_peptides_per_1000aa_BG_nonHuman(msnid.list[[i]], fst_dms_pth)
  }
  
  msnid.list[[i]] <- apply_filter(msnid.list[[i]], "!is.na(accession)")
  msnid.list[[i]] <- filter_msgf_data_protein_level(msnid.list[[i]], 0.01)
  message(i, " after filtering at protein level based on # of peptides per 1000 AAs")
  show(msnid.list[[i]]) 
}

```


## Parsimonious inference

To reduce number of protein identifications, we use a parsimonious inference algorithm described in 
[(Zhang et al. 2007)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2810678/).

```{r parsimonious_inference}
for (i in species) {
  msnid.list[[i]] <- infer_parsimonious_accessions(msnid.list[[i]], unique_only=FALSE)
  show(msnid.list[[i]])
}

```


## Remove Decoys

```{r}
for (i in species) {
  msnid.list[[i]] <- apply_filter(msnid.list[[i]], "!isDecoy")
  show(msnid.list[[i]])
}
save(msnid.list, file=file.path(data_folder, paste0("msgfData_filtered_", paste(species, collapse="-"), ".RData")))
```


## Process MASIC data

Output from the MASIC software is read from DMS, then filtered by interference score.


```{r read_masic}
masic_data_path <- file.path(data_folder, "masicData_original.RData")

options(timeout = 120)
if (file.exists(masic_data_path)) {
  load(masic_data_path)
} else {
  masic_data <- read_masic_data_from_DMS(data_package_num,
                                       interference_score = TRUE)
  save(masic_data, file=masic_data_path)
}

nrow(masic_data)
```


```{r filter_masic}
masic_data <- filter_masic_data(masic_data,
                                interference_score_threshold = 0.5,
                                s2n_threshold = 0)
save(masic_data, file=file.path(data_folder, "masicData_filtered.RData"))

nrow(masic_data)
```


## Create crosstab

The quantitative crosstab combines MS/MS identifications with reporter ion intensities. 
Abundances are taken relative to the reference channel and then log-transformed.

```{r create_global_gene_crosstab}
for (i in species) {
  temp.msnid <- msnid.list[[i]]
  psms(temp.msnid)$accession <- sub("^.*\\|(.*)\\|.*$", "\\1", psms(temp.msnid)$accession)
  
  psms(temp.msnid) <- psms(temp.msnid) %>%
    filter(!grepl("Contaminant", accession))
  
  aggregation_level <- c("accession")
  
  # remove non-samples (empty), but keep references
  samples <- samples[samples$ReporterAlias != "EMPTY", ]
  
  quant_cross_tab_prot <- create_crosstab(temp.msnid, 
                                          masic_data, 
                                          aggregation_level, 
                                          fractions, samples, references)
  dim(quant_cross_tab_prot)
  if (is.null(dim(quant_cross_tab_prot))) { # mouse only has 1 accession so it got converted into a vector
    quant_cross_tab_prot <- as.data.frame(quant_cross_tab_prot)
    colnames(quant_cross_tab_prot) <- unique(temp.msnid$accession)
    quant_cross_tab_prot <- t(quant_cross_tab_prot)
  }
  write.table(quant_cross_tab_prot, file=file.path(data_folder, paste(crosstab_prefix, paste0("crosstab_global_prot_original_", i, ".txt"), sep="_")),
              quote=F, sep="\t") 
}
# the only mouse uniprot accession remaining is A8DUK4 Beta-globin (gene symbol Hbb-bs)

```


