---
title: "Ex1 Global proteomics"
output: html_document
---

The global proteomics pipeline uses the R package [PlexedPiper](https://github.com/vladpetyuk/PlexedPiper). It also requires a connection to the DMS to access data packages.


```{r, echo=F}
library(knitr)
knitr::opts_chunk$set(message=F, warning=F)
t0 <- Sys.time(); print(t0)
```


```{r setup}
library(plyr)
library(MSnID)
library(PlexedPiper)
library(dplyr)
library(PNNL.DMS.utils)
library(Biostrings)
library(tidyverse)
library(tidyr)

# only HUMAN library: 5792
# HUMAN hg38, uniprot MOUSE and RAT: 5878
data_package_num <- 5878
data_folder <- "data/global_data_w_MOUSE_RAT_20240816"
crosstab_prefix <- "Chr8_UMN1"

dir.create(data_folder)

```

# Study design

Study design information in PlexedPiper is encoded in three tables: fractions, samples, and references. 
These tables can be made using metadata and should be stored on the DMS before processing.

```{r read_study_design}
fractions <- read.table("study_design/global_study_design/fractions.txt", sep = "\t", header = T)
samples <- read.table("study_design/global_study_design/samples.txt", sep = "\t", header = T)
references <- read.table("study_design/global_study_design/references.txt", sep = "\t", header = T)

```

# Process MS-GF+ data

MS-GF+ data is processed in several steps. First, read MS-GF+ output from the DMS. (This step can take a while).

```{r read_msgf}
msgf_data_path <- file.path(data_folder, "msgfData_original.RData")
if (file.exists(msgf_data_path)) {
  load(msgf_data_path)
} else {
  msnid <- read_msgf_data_from_DMS(data_package_num)
  save(msnid, file=msgf_data_path)
}
show(msnid)

```


## Correct peak selection and peptide FDR
We use the target-decoy search stRATegy method described in [(Elias 2010)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2922680/). 
Filtering is done first at peptide level, then at protein level, both with max FDR of 1%.

```{r select peak and filter peptides}
## peak selection & Peptide filter
msnid <- correct_peak_selection(msnid)   ## optional for phospho
msnid <- filter_msgf_data_peptide_level(msnid, 0.01)
show(msnid)

```

## Make sure each peptide is represented by 1 species
``` {r species selection}
## label accessions with species
message("before making sure peptides are species-specific")
show(msnid)
temp.psms <- psms(msnid)
temp.psms$species <- "Unknown"

# first tag human peptides
temp.psms[startsWith(temp.psms$accession, "NP_"),]$species <- "HUMAN"
humanPeptides <- unique(temp.psms[temp.psms$species == "HUMAN",]$peptide)

# remove any non-human entries for human peptides since we expect human to be most abundant
message(nrow(temp.psms), " rows before removing non-human peptides also found in humans")
temp.psms[temp.psms$peptide %in% humanPeptides & !startsWith(temp.psms$accession, "NP_"),]$species <- NA
temp.psms <- temp.psms[!is.na(temp.psms$species),]
message(nrow(temp.psms), " rows after removing non-human peptides also found in humans")

# next tag mouse peptides
temp.psms[endsWith(temp.psms$accession, "_MOUSE"),]$species <- "MOUSE"
mousePeptides <- unique(temp.psms[temp.psms$species == "MOUSE",]$peptide)

# then remove any rat entries for mouse peptides since we expect mouse to be second most abundant
#message(nrow(temp.psms), " rows before removing non-mouse peptides also found in mouse")
temp.psms[temp.psms$peptide %in% mousePeptides & temp.psms$species == "Unknown",]$species <- NA
temp.psms <- temp.psms[!is.na(temp.psms$species),]
message(nrow(temp.psms), " rows after removing non-mouse peptides also found in mouse")

# next tag rat peptides
temp.psms[endsWith(temp.psms$accession, "_RAT"),]$species <- "RAT"

# remove any peptides not assigned to human, mouse, or rat
temp.psms <- temp.psms[temp.psms$species != "Unknown",]
message(nrow(temp.psms), " rows after removing non-human/mouse/rat peptides")

psms(msnid) <- temp.psms
message("after making sure peptides are species-specific")
show(msnid)
```

## Accession FDR
```{r protein_level_filter }

# fst_dms_pth_2 <- remap_fasta_entry_names(fst_dms_pth, conv, "^.*\\|(.+?)(-\\d+)?\\|.*") 
# 
# ## peptide density and accession filter
# msnid <- compute_num_peptides_per_1000aa(msnid, fst_dms_pth_2)
# msnid <- filter_msgf_data_protein_level(msnid, 0.01)
# show(msnid)
options(timeout = 120)
fst_dms_pth <- path_to_FASTA_used_by_DMS(data_package_num)
msnid <- compute_num_peptides_per_1000aa(msnid, fst_dms_pth)

msnid <- filter_msgf_data_protein_level(msnid, 0.01)
message("after filtering at protein level based on # of peptides per 1000 AAs")
show(msnid) 

message(length(msnid$species[msnid$species=="HUMAN"]), " human entries")
message(length(msnid$species[msnid$species=="MOUSE"]), " mouse entries")
message(length(msnid$species[msnid$species=="RAT"]), " rat entries")

```


## Parsimonious inference

To reduce number of protein identifications, we use a parsimonious inference algorithm described in 
[(Zhang et al. 2007)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2810678/).

```{r parsimonious_inference}
msnid <- infer_parsimonious_accessions(msnid, unique_only=FALSE)
show(msnid)
message(length(msnid$species[msnid$species=="HUMAN"]), " human entries")
message(length(msnid$species[msnid$species=="MOUSE"]), " mouse entries")
message(length(msnid$species[msnid$species=="RAT"]), " rat entries")
```


## Remove Decoys

```{r}
msnid <- apply_filter(msnid, "!isDecoy")
show(msnid)
message(length(msnid$species[msnid$species=="HUMAN"]), " human entries")
message(length(msnid$species[msnid$species=="MOUSE"]), " mouse entries")
message(length(msnid$species[msnid$species=="RAT"]), " rat entries")
save(msnid, file=file.path(data_folder, paste0("msgfData_filtered_.RData")))
```


## Process MASIC data

Output from the MASIC software is read from DMS, then filtered by interference score.


```{r read_masic}
masic_data_path <- file.path(data_folder, "masicData_original.RData")

options(timeout = 120)
if (file.exists(masic_data_path)) {
  load(masic_data_path)
} else {
  masic_data <- read_masic_data_from_DMS(data_package_num,
                                       interference_score = TRUE)
  save(masic_data, file=masic_data_path)
}

nrow(masic_data)
```


```{r filter_masic}
masic_data <- filter_masic_data(masic_data,
                                interference_score_threshold = 0.5,
                                s2n_threshold = 0)
save(masic_data, file=file.path(data_folder, "masicData_filtered.RData"))

nrow(masic_data)
```


## Create crosstab

The quantitative crosstab combines MS/MS identifications with reporter ion intensities. 
Abundances are taken relative to the reference channel and then log-transformed.

```{r create_global_gene_crosstab}
# psms(msnid)$accession <- sub("^.*\\|(.*)\\|.*$", "\\1", psms(msnid)$accession)

psms(msnid) <- psms(msnid) %>%
  filter(!grepl("Contaminant", accession))

aggregation_level <- c("accession")

# remove non-samples (empty), but keep references
samples <- samples[samples$ReporterAlias != "EMPTY", ]

quant_cross_tab_prot <- create_crosstab(msnid, 
                                        masic_data, 
                                        aggregation_level, 
                                        fractions, samples, references)
dim(quant_cross_tab_prot)
if (is.null(dim(quant_cross_tab_prot))) { # MOUSE only has 1 accession so it got converted into a vector
  quant_cross_tab_prot <- as.data.frame(quant_cross_tab_prot)
  colnames(quant_cross_tab_prot) <- unique(msnid$accession)
  quant_cross_tab_prot <- t(quant_cross_tab_prot)
}
write.table(quant_cross_tab_prot, file=file.path(data_folder, paste(crosstab_prefix, paste0("crosstab_global_prot_original.txt"), sep="_")),
            quote=F, sep="\t") 
message(length(rownames(quant_cross_tab_prot)[startsWith(rownames(quant_cross_tab_prot), "NP_")]), " human protein entries")
message(length(rownames(quant_cross_tab_prot)[endsWith(rownames(quant_cross_tab_prot), "_MOUSE")]), " mouse protein entries")
message(length(rownames(quant_cross_tab_prot)[startsWith(rownames(quant_cross_tab_prot), "_RAT")]), " rat protein entries")

```

