---
title: "Differential Expression Analysis Without Contaminated Samples"
author: "Ruby Fore, Sara Gosline, and Belinda Garana"
date: "2024-08-13"
output:
  html_document: default
  pdf_document: default
---
```{r, echo=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggfortify)
library(UpSetR)
synapser::synLogin()
#corrected global proteomics: syn61695507
# corrected phospho proteomics: 'syn61695519'
# syn62015820: corrected global proteomics with mouse and rat in addition to human libraries
fileEntity <- synapser::synGet('syn61695519')
origProt <- read.table(fileEntity$path, header = T)

# general > 5 and known contaminant > 2 outliers: syn62034810
# only known contaminant > 1.75 outliers: syn62039783
# rel abundance > 5 in human, mouse, or rat: syn62049775
# rel abundance > 5 in human, mouse, or rat after grouping peptides: syn62071033
# rel abundance > 5 in human, mouse, or rat after grouping peptides & selecting species: syn62178788
outliers <- read.csv(synapser::synGet("syn62178788")$path)

origProt <- origProt[,colnames(origProt)[!(colnames(origProt) %in% outliers$Sample)]]
```

# Proteomics analysis
Here we evaluate the first evaluate the proteomics data derived from the PDX-MT experiments.

### Remove NAs
Proteins that are missing >50% of their data are removed from this analysis. However we retain the rest and impute them using the mean value for that protein.

```{r remove NAs}
# get the indices of NA's
na_indices <- which(is.na(origProt), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
correctedProteomics<-origProt
correctedProteomics[na_indices] <- rowMeans(origProt, na.rm=T)[na_indices[,1]]

#na_indices
sum(is.na(correctedProteomics))
dim(correctedProteomics)

```


### Collect sample metdata

Here I moved the data to long form to easily summarize, filter, and plot.

```{r remove outliers reshape}
library(dplyr)
#noOutliersCorrectedProteomics <- dplyr::select(correctedProteomics, -one_of(outliers))

samps<-colnames(correctedProteomics)
longProt<-correctedProteomics|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)
longProt$time <- as.numeric(longProt$time)

##collect sample metadata, annotate our outliers
meta<-longProt|>
  dplyr::select(condition,sample,drug,time)|>
  distinct()|>
  tibble::column_to_rownames('condition')
meta$time <- factor(meta$time, levels = c("0", "0.5", "3", "24"))
print(meta)
```

First we check to see how the proteins are distributed across the samples
```{r protein distribution}
##quick summary of distributions 
ggplot(longProt,aes(x=condition,y=logRatio,fill=drug))+geom_violin()+theme(axis.text.x = element_text(angle = 90))
```

Here we get a larger than expected dynamic range for some samples. This suggests that something is amiss. All of these values are the log ratio of what was in the sample compared to what was in the reference control.

How does this look without NA imputation? Are there specific proteins that are worse than others?

```{r protein outliers}

longOrig<-origProt|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)|>
  subset(!is.na(logRatio))
longOrig$time <- factor(longOrig$time, levels = c("0", "0.5", "3", "24"))

ggplot(longOrig,aes(x=condition,y=logRatio,fill=drug))+geom_violin()+theme(axis.text.x = element_text(angle = 90))

bigProts<-subset(longOrig,abs(logRatio)>3)|>
  dplyr::group_by(gene)|>
  summarize(count=n())|>
  arrange(desc(count))|>
  subset(count>2)|>
  arrange(desc(count))
print(bigProts)
longOrig|>
  subset(gene%in%bigProts$gene)|>
  ggplot(aes(x=gene,y=logRatio))+geom_violin()+theme(axis.text.x = element_text(angle = 90))

```
While something is amiss with many proteins, COL19A1 is the biggest outlier - there are no samples with a regular amount. Could this protein come from another species? 

Either way we can remove some samples that have too many of these problematic proteins:

```{r outlier list}
##what if we filter by logratio?
outliers<-subset(longProt,abs(logRatio)>5)|>
  group_by(condition)|>
  summarize(numOut=n())|>
  arrange(numOut)|>
  subset(numOut>1)
  
meta<-meta|>
  tibble::rownames_to_column('condition')|>
    mutate(isOutlier=condition%in%outliers$condition)|>
  tibble::column_to_rownames('condition')

print(outliers)
```
I decided that maybe the samples with large LFCs are outliers and removed those. This doesn't always seem to be the case, some outliers just look like regular samples. 

```{r plot outliers}
#PCA plots
pc<-prcomp(t(correctedProteomics))
autoplot(pc,data = meta,color='drug',shape='isOutlier')
autoplot(pc,data = meta,color='time',shape='isOutlier')+facet_grid(sample~drug)
```

Now we can see that these outliers are off and filter them out.

```{r remove outliers}
longProt<-longProt|>
  subset(!condition%in%outliers$condition)
meta<-meta|>
  subset(!isOutlier)

noOutliersCorrectedProteomics<-longProt|>
  dplyr::select(condition,gene,logRatio)|>
  tidyr::pivot_wider(names_from=condition,values_from=logRatio)|>
  tibble::column_to_rownames('gene')|>
  as.matrix()

pc2<-prcomp(t(noOutliersCorrectedProteomics))
autoplot(pc2,data = meta,color='drug',shape='sample')
autoplot(pc2,data = meta,color='drug',shape='sample')+facet_grid(~time)
ggsave('protPCA.png')

```
Now the result is much more divided by sample, which we expect.

## Differential expression

Now we want to evaluate if proteins are differentially expressed. We create a simple function for differental expression using LIMMA iterate over all the conditions.

```{r}
# function to get N genes, adj.P.value < 0.05
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  

factorInfo <- data.frame(stringr::str_split_fixed(names(data), "_", 5))

names(factorInfo) <- c('sample', 'trt', 'time','hr', 'rep')
factorInfo$sample <- as.factor(factorInfo$sample)
factorInfo$trt <-as.factor(factorInfo$trt)

factorInfo$time <- as.factor(factorInfo$time)

# rownames have already been assigned
proSet <- Biobase::ExpressionSet(as.matrix(data))

 # log2 transformation if distribution isn't normal
 ex <- Biobase::exprs(proSet)
 qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                           na.rm = TRUE
 ))
 LogC <- (qx[5] > 100) ||
   (qx[6] - qx[1] > 50 && qx[2] > 0)
 if (LogC) {
   ex[which(ex <= 0)] <- NaN
   Biobase::exprs(proSet) <- log2(ex)
 }


proSet$sample <- factorInfo$sample
proSet$trt <- factorInfo$trt
proSet$time <- factorInfo$time
 #browser()
design <- stats::model.matrix(formula, proSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
#print(cont.matrix)
fit <- limma::lmFit(proSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)
#browser()
fit <- try(limma::eBayes(fit), silent = TRUE) # added try functionality since removed contaminated samples
if (inherits(fit, "try-error")) {
  deg <- data.frame()
  nDiffExpressed <- NA
} else {
  deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

  nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
}

return(list(deg, nDiffExpressed))
}
```


We first check the results compared to time 0. A loop through the data frame calculates all the conditions and adds them to a new data frame, both comparing each value to time 0 and also each value to the DMSO at the same time point.

```{r}
##created for loop to calculate all diff ex stats
t0dat<-data.frame()
ddat<-data.frame()
for(d in c('DMSO','Mirdametinib','Selumetinib','Vorinostat')){
  for(tv in c('0.5','3','24')){
    for(s in c('JH.2.002','MN.2')){
      ##get matrix of values of sample at condition
      tmat<-longProt|>
        subset(sample==s)|>
        subset(drug==d)|>
        subset(time==tv)|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
      if(nrow(tmat)==0)##there is no data for this
        next
      ##matrix of values at DMSO 0
      cmat<-longProt|>
        subset(sample==s)|>
        subset(drug=='DMSO')|>
        subset(time=='0')|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
  
      if(d!='DMSO'){    
        #matrix of values at DMSO same time
        dmat<-longProt|>
          subset(sample==s)|>
          subset(drug=='DMSO')|>
          subset(time==tv)|>
          select(gene,condition,logRatio)|>
          distinct()|>
          tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
        ## get result compared to DMSO
        res<-getNdiffExpressedGenes(cbind(dmat,tmat), formula = ~ trt  + 0, analysisContrast =paste0('trt',d,'-trtDMSO')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
      }
           ##first get restult compare to t0
        res<-getNdiffExpressedGenes(cbind(cmat,tmat), formula = ~ time  + 0, analysisContrast =paste0('time',tv,'-time0')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        t0dat<-rbind(t0dat,tres)
      
    }
  }
}
ddat$time <- as.numeric(ddat$time)
t0counts = t0dat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
t0counts$time <- as.numeric(t0counts$time)
print(t0counts)
ggplot(t0counts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)+
  ggtitle("Number of differentially expressed proteins compared to time 0")+theme(axis.text.x = element_text(angle = 45))

```

DMSO seems to show a lot of changes, what happens when we only compare treated to DMSO? This tells us the biological differences.

```{r DMSO comparison}


dcounts = ddat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(dcounts)
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)+
  ggtitle("Number of differentially expressed proteins compared to DMSO")+theme(axis.text.x = element_text(angle = 45))
```

We dont get that many changes. Maybe the time point is too early? Vorinostat exhibits some changes. Let's do an upset plot to see if any proteins are shared across conditions.


```{r upset}

ddat|>
  subset(adj.P.Val<0.1)|>
  select(Gene,drug,time,samp)|>
  tidyr::unite(drug,time,samp,col='condition')|>
  mutate(present=1)|>
  tidyr::pivot_wider(names_from=condition,values_from=present,values_fill=0)|>
  tibble::column_to_rownames('Gene')|>
#  as.matrix()|>
    upset()

ddat|>
  subset(drug=='Vorinostat')|>
  subset(adj.P.Val<0.1)|>
  group_by(Gene)|>
  summarize(count=n())|>
  subset(count>1)

```

Indeed there are 9 proteins that are differentially expressed in both vorinostat conditions.

## Evaluate known target genes of various drugs

While we are using a large p-value threshold, maybe the target genes we expect are just above the p-value, or due to noise are not differentially expressed. Let's check to see  how known proteins behave.
First we can evaluate MEK, the known target of Mirdametinib and Selumetinib.
.
```{r}
genelist<-c('MAPK1','MAPK3','MAP2K1','MAP2K2','MAP2K4','MAP2K6')
longProt$SUB_SITE <- longProt$gene
longProt$gene <- strsplit(longProt$gene, "-")[[1]][1]

mtargs <- subset(longProt,gene%in%genelist)
if (nrow(mtargs) > 0) {
  mtargs$`Timepoint (hrs)` <- factor(mtargs$time, levels = c("0", "0.5", "3", "24"))
  
  mddat <- subset(ddat,Gene%in%genelist)
  mddat$`Timepoint (hrs)` <- factor(mddat$time, levels = c("0", "0.5", "3", "24"))
  
  # plot
  ggplot(mtargs,aes(x=`Timepoint (hrs)`,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)
  ggsave('mekProteins_geneSymbols.png')
  mddat|>
    mutate(sig=(adj.P.Val<0.1))|>
    ggplot(aes(x=`Timepoint (hrs)`,y=logFC,shape=sig,col=drug))+
    geom_point()+facet_grid(samp~Gene)
  ggsave('mekProteinsDiffEx_geneSymbols.png') 
}
```

Now we can look at those transcripts that we know change, particularly at the 8hr time point.

```{r mek trans targets}
genelist<-c('DUSP6','ETV5','SPRY2','SPRY4','EPHA4','ETV4','EPHA2','DUSP4','CCND1')
mtargs <- subset(longProt,gene%in%genelist)
if (nrow(mtargs) > 0) {
  mtargs$`Timepoint (hrs)` <- factor(mtargs$time, levels = c("0", "0.5", "3", "24"))
  
  mddat <- subset(ddat,Gene%in%genelist)
  mddat$`Timepoint (hrs)` <- factor(mddat$time, levels = c("0", "0.5", "3", "24"))
  
  # plot
  ggplot(mtargs,aes(x=`Timepoint (hrs)`,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)
  ggsave('mekTranscriptProteins_geneSymbols.png')
  mddat|>
    mutate(sig=(adj.P.Val<0.1))|>
    ggplot(aes(x=`Timepoint (hrs)`,y=logFC,shape=sig,col=drug))+
    geom_point()+facet_grid(samp~Gene)
  ggsave('mekTranscriptProteinsDiffEx_geneSymbols.png') 
}
```

Vorinostat is known to affect HDAC proteins. So we can plot those four here.

```{r hdac}
hdacs<-c('HDAC1','HDAC2','HDAC3','HDAC6')
mtargs <- subset(longProt,gene%in%hdacs)
if (nrow(mtargs) > 0) {
  mtargs$`Timepoint (hrs)` <- factor(mtargs$time, levels = c("0", "0.5", "3", "24"))
  
  mddat <- subset(ddat,Gene%in%hdacs)
  mddat$`Timepoint (hrs)` <- factor(mddat$time, levels = c("0", "0.5", "3", "24"))
  
  # plot
  ggplot(mtargs,aes(x=`Timepoint (hrs)`,y=logRatio,fill=drug))+geom_boxplot()+facet_grid(sample~gene)
  ggsave('hdacProteins_geneSymbols.png')
  mddat|>
    mutate(sig=(adj.P.Val<0.1))|>
    ggplot(aes(x=`Timepoint (hrs)`,y=logFC,shape=sig,col=drug))+
    geom_point()+facet_grid(samp~Gene)
  ggsave('hdacProteinsDiffEx_geneSymbols.png') 
}
```
