---
title: "Cross-model summary"
author: "Sara Gosline"
date: "2025-06-12"
output: html_document
---

The goal of this document is to compare the various model systems. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo= TRUE)
library(tidyverse)
library(synapser)

```

## Get all sample information

We are currently maintaining a table of all the information on our model systems. The full model catalog can be found [here](https://www.synapse.org/Synapse:syn53503360/tables/). 

We are removing some samples for which sequencing between samples does no align

```{r get sample info}
sl <- synLogin()
all_data <- synTableQuery("select * from syn53503360")$asDataFrame()

badsamps <- c('JH-2-009')#,'MN-1','WU-505','WU-536','WU-545','MN-3')

pcols <- c(`JH-2-002` =  'sienna',`JH-2-079c` = 'turquoise4',
           `MN-4`  =  'steelblue4', `WU-225`  =  'darkgoldenrod',
           `JH-2-031`  = 'pink3', `JH-2-103`  =  'chartreuse4',
           `MN-1` = 'darkred', `MN-2` = 'cornsilk3', `MN-3` = 'mediumorchid4', 
           `WU-355` = 'salmon', `WU-368` = 'orangered3')

all_data <- subset(all_data,!Sample %in% badsamps) |>
  mutate(drugData  =  !is.na(MicroTissueDrugFolder))
```



## RNA Seq comparisons

You can also embed plots, for example:

```{r seq, echo = FALSE, message = FALSE, warning = FALSE}

do_deseq_import <- function(file) {
  if (!require('DESeq2')) {
    BiocManager::install("DESeq2")
    library(DESeq2)
  }
  
  if (!require('tximportData')) {
    BiocManager::install("tximportData")
    library(tximportData)
  }
  
  if (!require('tximport')) {
    BiocManager::install("tximport")
    library(tximport)
  }
  
  dir <- system.file("extdata", package  =  "tximportData")
  tx2gene <- read.csv(file.path(dir, "tx2gene.gencode.v27.csv"))
  
  qnt.table <-
    data.frame(counts  =  tximport::tximport(file, type  =  'salmon',
                                           tx2gene  =  tx2gene)$counts) %>%
    tibble::rownames_to_column('GENEID')
  
  return(qnt.table)
}

rna <- all_data |>
  dplyr::select(Sample,ends_with('RNASeq'))

seqdata <- do.call(rbind,plyr::alply(rna,1,function(x){
  pat  =  x[[1]]
  pdx  =  x[[2]]
  tumor  =  x[[3]]
  
  pd <- NULL
  td <- NULL
  if (!is.na(pdx))
    pd <- do_deseq_import(synGet(pdx)$path) |>
        mutate(sample  =  'PDX', patient  =  pat, synId = pdx)
  if (!is.na(tumor))
    td <- do_deseq_import(synGet(tumor)$path) |>
        mutate(sample  =  'Tumor', patient  =  pat, synId = tumor)
  return(rbind(pd,td))
}))

```

Now that we have the data we can map and load into DESeq for column normalization

```{r map}

seqdata <- seqdata |>
  tidyr::separate_wider_delim(cols = c(GENEID),names = c('GENE','VERS'),delim = '.')


if (!require('EnsDb.Hsapiens.v86')) {
    BiocManager::install('EnsDb.Hsapiens.v86')
    library(EnsDb.Hsapiens.v86)
  }
database <- EnsDb.Hsapiens.v86
pmap <-  ensembldb::select(
    database,
    keys  =  list(
      GeneIdFilter(seqdata$GENE),
      TxBiotypeFilter("protein_coding")
    ),
    columns  =  c("GENENAME")
  ) %>%
  dplyr::rename(GENE  =  'GENEID') %>%
  right_join(seqdata) %>%
  subset(TXBIOTYPE  ==  'protein_coding') %>%
  dplyr::select(GENE, GENENAME) %>% distinct()#%>%
# tibble::column_to_rownames('GENEID')


  counts <- seqdata |>
      subset(GENE %in% pmap$GENE) %>% ## this is important as we filter for protein coding genes
      dplyr::select(GENE, counts, synId) %>%
    tidyr::pivot_wider(values_from  =  'counts', names_from  =  'synId') %>%
    tibble::column_to_rownames('GENE') %>%
    round()

  cdata <- all_data |> 
    dplyr::select(Sample,Age,Sex, PDX_RNASeq,Tumor_RNASeq, drugData) |> 
    tidyr::pivot_longer(c('PDX_RNASeq','Tumor_RNASeq'),names_to = 'source',values_to = 'synId') |>
    mutate(source = gsub('_RNASeq','',source)) |>
    subset(!is.na(synId)) |>
    tibble::column_to_rownames('synId')
  
  dds <- DESeq2::DESeqDataSetFromMatrix(countData  =  counts,
                                        colData  =  cdata,
                                        design  =  ~ Sex)# + Clinical.Status)
  
  keep <- rowSums(counts(dds)) >=  10
  dds <- dds[keep, ]
  dds <- DESeq2::DESeq(dds)
##now get all the SF files, read them in 
vsd <- DESeq2::vst(dds)

```

Now we can try to plot

```{R plot}
library(ggplot2)
library(ggrepel)
pcaData <- plotPCA(vsd, intgroup = c("Sample", "source", "drugData"), returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))



p <- ggplot(pcaData, aes(PC1, PC2, shape = source, color = Sample)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()+
  ggtitle('Samples by normalized expression') 

p+geom_label_repel(aes(label = Sample))+theme_minimal()+scale_color_manual(values = pcols)

 ggsave('rnaSeqPCA.pdf')
 
 ##then subset for those which have drug data
 p <- subset(pcaData, drugData)|>
   ggplot(aes(PC1, PC2, shape = source, color = Sample)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()+
  ggtitle('Samples by normalized expression') 

p+geom_label_repel(aes(label = Sample))+theme_minimal()+scale_color_manual(values = pcols)

 ggsave('rnaSeqDrugPCA.pdf')
 
p <- subset(pcaData, source == 'PDX')|>
   ggplot(aes(PC1, PC2, shape = drugData, color = Sample)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()+
  ggtitle('Samples by normalized expression')+scale_color_manual(values = pcols)

p+geom_label_repel(aes(label = Sample))+theme_minimal()

 ggsave('rnaSeqDrugPDX.pdf')
 
```

## Copy number comparisons
Probably don't need to do these, but at the same time maybe?

## Drug response comparisons

First let's get the older data
```{r get original data}


orig_scores <- readr::read_tsv(synGet('syn68384507')$path)

orig_scores$improve_drug_id[which(orig_scores$improve_drug_id == 'CUDC907C')]<-'CUDC907'

orig_scores |>
   subset(dose_response_metric  ==  'auc') |>
  ggplot(aes(x  =  as.factor(time),y  =  dose_response_value,fill  =  improve_sample_id)) +
    geom_bar(position  =  'dodge',stat  =  'identity') +
    facet_grid(study~improve_drug_id)+scale_fill_manual(values = pcols)

ggsave('origDrugAUC.pdf',width = 24,height = 3)
```

How does the drug response compare to the Musyc scores?

```{r drug preds}

single_scores <- readr::read_tsv(synGet('syn65941820')$path)

single_scores |>
   subset(dose_response_metric  ==  'auc') |>
  ggplot(aes(x  =  as.factor(time),y  =  dose_response_value,fill  =  improve_sample_id)) +
    geom_bar(position  =  'dodge',stat  =  'identity') +
    facet_grid(~improve_drug_id)+scale_fill_manual(values = pcols)

ggsave('singleDrugAUC.pdf',width = 20)

#single_scores |>
#    subset(dose_response_metric  =  =  'auc') |>
#   ggplot(aes(x  =  as.factor(time),y  =  dose_response_value,fill  =  improve_drug_id))+
#  geom_bar(stat  =  'identity',position  =  'dodge') + facet_grid(improve_sample_id~.)

combo_scores <- readr::read_csv(synGet('syn68258295')$path)

##combinations have different effects on different samples
ggplot(combo_scores,aes(x = as.factor(time),y  =  log_alpha12,
                        fill = sample)) + geom_boxplot() +
  scale_fill_manual(values = pcols)


p1 <- ggplot(combo_scores,aes(x = time,
                        y = log_alpha12,fill = sample)) + 
  facet_grid(drug1~drug2)+
  geom_bar(stat = 'identity',position = 'dodge') +
   scale_fill_manual(values = pcols)

p2 <- ggplot(combo_scores,aes(x = time,
                        y = log_alpha21,fill = sample)) +
  facet_grid(drug1~drug2) +
  geom_bar(stat = 'identity',position = 'dodge') +
   scale_fill_manual(values = pcols)

cowplot::plot_grid(p1,p2,ncol = 1)

ggsave('musycBarplots.pdf',width = 10,height = 10)


```


Here we see that the 2 day timepoint shows greater effect over the five day in general

```{r effect comparison}
a1<-ggplot(combo_scores,aes(x=as.factor(time),y=log_alpha21,fill=sample))+geom_boxplot()+scale_fill_manual(values=pcols)

a2<-ggplot(combo_scores,aes(x=as.factor(time),y=log_alpha12,fill=sample))+geom_boxplot()+scale_fill_manual(values=pcols)

cowplot::plot_grid(a1,a2)+ggtitle('2D measurment shows greater effects')


```

```{r scores by drug}
##music gives three measurements that provide a profile of drug sensitivity rather than just synergy
ggplot(combo_scores,
       aes(x = log_alpha21,y = log_alpha12,
           colour = beta,shape = sample,size = 4)) +
  geom_point() +
  scale_color_viridis_c() +
  facet_grid(drug1~drug2) + 
  theme_bw()

ggsave('musyc_scores.pdf',width = 12)

```


Now can we compare musyc results to DMEA results?

```{R comparison}

combineddata <- readr::read_csv(synGet('syn68327933')$path)

ggplot(combineddata,aes(x=Time,y=NES,fill=DrugTreatment))+geom_bar(position='dodge',stat='identity')+facet_grid(~Drug_set)

full_data<-combineddata|>left_join(combo_scores)

##first let's check which time points. 

p1<-  ggplot(full_data,aes(x=log_alpha12,y=NES,shape=Time,col=time))+
  geom_point()+facet_grid(DrugTreatment~Drug_set)

p2<-  ggplot(full_data,aes(x=log_alpha21,y=NES,shape=Time,col=time))+
  geom_point()+facet_grid(DrugTreatment~Drug_set)

cowplot::plot_grid(p1,p2)
```

this isn't great, let's try to evaluate one drug at a time

```{r dmea}

dmea <- readr::read_csv(synGet('syn65672258')$path)

ds <- c("HDAC inhibitor","MEK inhibitor","MET inhibitor","DNA alkylating agent",'CDK inhibitor' )
drugs <-c('TNO155','Mirdametinib','Doxorubicin','Vorinostat','Trabectidin',"Olaparib",'Trabectedin','Palbociclib')

drugs <- c(drugs,tolower(drugs))

preds <- dmea |>
  subset(DrugTreatment %in% drugs) |>
  subset(Drug_set %in% ds) |>
  ggplot(aes(x = Drug_set,col = MPNST,y = NES,size = -log10(FDR_q_value))) +
  geom_point() +
  facet_grid(Time~DrugTreatment) +
  scale_color_manual(values = pcols) +
  coord_flip() +
  theme_bw()

ggsave('selectPreds.pdf',preds,height=5)
###now we can do a side by side with our drug scores
actual <- combo_scores |>
  subset(drug1 %in% drugs) |>
  subset(drug2 %in% drugs) |>
ggplot(aes(x = log_alpha21,y = log_alpha12,
           color = sample,size=beta,shape=as.factor(time))) +
  geom_point() +
 scale_color_manual(values = pcols) +
  facet_grid(drug1~drug2) + 
  theme_bw()

ggsave('selectActuals.pdf',actual,height=5)
cowplot::plot_grid(preds,actual)

ggsave('predsVsActual.pdf',width=20, height=5)


```


This is too much, let'st ry to focus even more

```{r mirda}

subset(dmea,DrugTreatment == 'Mirdametinib') |>
  subset(Time == '24h') |>
  subset(FDR_q_value < 0.1) |>
  subset(NES < 0) |>
  ggplot(aes(x = reorder(Drug_set,NES),y = NES,fill=MPNST))+
  #facet_grid(Time~.)+
  geom_bar(position='dodge',stat='identity')+
  coord_flip()+
  scale_fill_manual(values=pcols)+
  ggtitle("Mirdametinib based predictions")

ggsave('mirdaPreds.pdf')

subset(dmea,DrugTreatment=='TNO155')|>
  subset(Time=='24h') |>
  subset(FDR_q_value<0.25)|>
  subset(NES<0)|>
  ggplot(aes(x=reorder(Drug_set,NES),y=NES,fill=MPNST))+
  #facet_grid(Time~.)+
  geom_bar(position='dodge',stat='identity')+
  coord_flip()+
  scale_fill_manual(values=pcols)+
  ggtitle("TNO155 based predictions")

ggsave('tnoPreds.pdf')

subset(dmea,DrugTreatment=='Doxorubicin')|>
  subset(FDR_q_value<0.25)|>
  subset(Time=='24h') |>
  subset(NES<0)|>
  ggplot(aes(x=reorder(Drug_set,NES),y=NES,fill=MPNST))+
  #facet_grid(Time~.)+
  geom_bar(position='dodge',stat='identity')+
  coord_flip()+
  scale_fill_manual(values=pcols)+
  ggtitle("Doxorubicin based predictions")

ggsave('doxPreds.pdf')
```

Now we can compare to actual values, suggest new ones. 
