---
title: "Spheroid Proteomics Differential Expression Analysis"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    toc: true
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```

Next the import the proteomics for the global proteomics abundance from spheroids.
The data file is located on synapse and is stored as Excel sheet.

```{r import-proteomics-data, results='hide'}
synapseFileID = "syn72245489"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsData <- read_xlsx(fileEntity$path, sheet = 1)
tmp <- synapser::synLogout()
```

Finally we retrieve the metadata for the experiments from synapse.
The experimental setup pools proteins from several wells into one MS/MS run to
attain better proteomics coverage.
The information which wells were pooled together is contained in the metadata
table, specifically it contains the description of which chemical agent was used
on the spheroids in the drug response experiment and at which timepoint the
sample was extracted.
Sheet 2 of the Excel document retrieved from synapes contains the mapping that
describes which DIA run is associated to which drug and sampling time point.

```{r import-meta-data, results='hide'}
synapseFileID = "syn72245597"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsMeta <- read_xlsx(fileEntity$path, sheet = 2)
tmp <- synapser::synLogout()
```


## Data preparation

Before running the actual differential expression analysis we need to first
clean the data. This consits of several steps:

- Matching DIA run identifiers to chemical agent & sample time descriptor
- Missingness removal (i.e. exclude genes from consideration that weren't
  detected in a certain fraction of MS/MS runs)
- remove additional identifier columns (move into separate DataFrame)
- log2 transform the abundance / intensity data
- impute missing values (media center based on other readouts for the protein)


```{r prep-data}
experiment <- prep_global_prot_data(proteomicsData, proteomicsMeta, 0) %>%
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method = "SampMin")
```

## Differential Expression analysis

### PCA
With all the data imported we can focus on the actual differential expression
analysis.
First we will perform a PCA analysis of the underlying proteomics data to check
if there is meaningful signal separation of the individual treatment agents,
as well as timepoints.

```{r PCA, message = FALSE}
principal_components <- prcomp(t(assay(experiment))) %$%
  cbind(x, colData(experiment))

plot <- (
  ggplot(
    principal_components,
    aes(
      x = PC1,
      y = PC2,
      col = agent,
      shape = timepoint,
      label = condition
      )
    )
  + geom_point(size = 2)
  # + geom_label_repel(max.overlaps = 10, size = 2.5, box.padding = 0.5)
  + geom_text_repel(max.overlaps = 10, size = 2.5)
  + theme_minimal()
  + theme(legend.position = "bottom")
  + scale_color_manual(values = c("DMSO" = "#e66101", "Mirda" = "#5e3c99", "NA" = "grey"))
  + labs(
    title = "Spheroid global proteomics; DMSO vs. Mirdametinib",
    subtitle = "full set of proteins; SampMin imputation"
  )
)

plot

ggsave("../output_figures/global_prot_PCA.png",
       plot = plot,
       dpi = 600,
       height = 6,
       width = 6,
       )
```

The PCA demonstrates that the proteomics data indeed separates nicely based on
both agent as well as timepoint.

### PCA after removal of features with >50% missingness

```{r}
experiment <- prep_global_prot_data(proteomicsData, proteomicsMeta, 0.5) %>%
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method = "SampMin")

principal_components <- prcomp(t(assay(experiment))) %$%
  cbind(x, colData(experiment))

plot <- (
  ggplot(
    principal_components,
    aes(
      x = PC1,
      y = PC2,
      col = agent,
      shape = timepoint,
      label = condition
      )
    )
  + geom_point(size = 2)
  # + geom_label_repel(max.overlaps = 10, size = 2.5, box.padding = 0.5)
  + geom_text_repel(max.overlaps = 10, size = 2.5)
  + theme_minimal()
  + theme(legend.position = "bottom")
  + scale_color_manual(values = c("DMSO" = "#e66101", "Mirda" = "#5e3c99", "NA" = "grey"))
  + labs(
    title = "Spheroid global proteomics; DMSO vs. Mirdametinib",
    subtitle = "proteins with >50% missingness removed; SampMin imputation"
  )
)

plot

ggsave("../output_figures/global_prot_PCA_50pct_missingness_removed.png",
       plot = plot,
       dpi = 600,
       height = 6,
       width = 6,
       )
```

### Number of significant differentally expressed proteins 
Next we will perform the actual differential expression analysis.
We will define a significantly differentially experessed protein as a protein that

- has an adjusted p-Value < 0.05
- has a logFC >= 1 or logFC < -1

We will be investigating specifically the expression differential between DMSO
and Mirdametenib treated samples both at the 24h as well as 48h time point.
The codeblock below performs the differential expression analyses using the
previously defined function `calc_diff_ex`.
The we select the proteins that show a significant differential in their
expression between the two groups and prepare a `data.frame` for plotting the
results.
```{r differential-expression-dmso-mirda}
dmso_mirda_24h <- calc_diff_ex(experiment, group_1 = "24hr_DMSO", group_2 = "24hr_Mirda")
dmso_mirda_48h <- calc_diff_ex(experiment, group_1 = "48hr_DMSO", group_2 = "48hr_Mirda")

dfPlot <- subset(as_tibble(rowData(dmso_mirda_24h)),
                 select = c('Genes', 'logFC', 'adj.P.Val'),
                 adj.P.Val < 0.05) %>%
          mutate(expression_change = case_when(
                 logFC >= 1 ~ 'up',
                 logFC <= -1 ~ 'down'
                 ))
dfPlot$timepoint = "24hr"

dfPlot2 <- subset(as_tibble(rowData(dmso_mirda_48h)),
                  select = c('Genes', 'logFC', 'adj.P.Val'),
                  adj.P.Val < 0.05) %>%
           mutate(expression_change = case_when(
                  logFC >= 1 ~ 'up',
                  logFC <= -1 ~ 'down'
                  ))
dfPlot2$timepoint = '48hr'

dfPlot <- rbind(dfPlot, dfPlot2) %>% subset(., !is.na(expression_change))
```

The code block below plots the number of significantly up and down regulated
proteins.
```{r plot-expression-differential, message=FALSE}
cpalette = c('up' = '#74add1', 'down' = '#f46d43')

plot <- (
  ggplot(
    data = dfPlot,
    aes(timepoint)
  )
  + geom_bar(aes(fill=expression_change), position = 'dodge2')
  + scale_fill_manual(values = cpalette)
  + theme_minimal()
  + labs(
    title = 'Spheroid global proteomics; DMSO vs Mirdametinib', 
    subtitle = "Counts of significant differentially expressed proteins ( pVal < 0.05; -1  >= logFC >= 1 )",
    fill = 'expression'
  )
)
plot

ggsave("../output_figures/global_prot_diff_exp_counts.png", dpi = 600)
```

## Volcano plots of differentially expressed porteins
Next we generate volcano plots to investigate the significant proteins further.
For each time point one volcano plot will be generated.

### DMSO vs Mirdametenib at 24h
```{r volcano-plot-24h, message=FALSE}

dfPlot <- subset(as.tibble(rowData(dmso_mirda_24h)), select = c('Genes', 'logFC', 'adj.P.Val')) %>% 
  mutate(expression_change = case_when(
    (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
    (logFC <= -1 & adj.P.Val < 0.05) ~ 'down',
    .default = 'not significant', 
  )) %>%
  mutate(labels = case_when(
    expression_change %in% c('up', 'down') ~ Genes,
    .default = NA
  ))

dfPlot$timepoint = "24hr"


plot <- (
  ggplot(
    data = dfPlot,
    aes(
      x = logFC,
      y = -log10(adj.P.Val),
      col = expression_change,
      # label = labels
      )
    )
    + geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed')
    + geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed')
    + geom_point(size = .75)
    # + geom_text_repel(max.overlaps = 10, size = 2)
    + scale_color_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + coord_cartesian(xlim = c(-15, 15), ylim = c(0, 3))
    + theme_minimal()
    + theme(legend.position = "bottom")
    + labs(
      title = 'Spheroid global proteomics',
      subtitle = "DMSO vs. Mirdametinib at 24h",
      col = 'expression',
      y = expression("-log"[10]*"p-Value")
    )
)
plot

ggsave(
  filename = "../output_figures/global_prot_diff_exp_volcano_24h.png",
  plot = plot,
  dpi = 600,
  height = 6,
  width = 6,
  scale = .8,
  )
```

### DMSO vs Mirdametenib at 48h
```{r volcano-plot-48h, message=FALSE}

dfPlot <- subset(as.tibble(rowData(dmso_mirda_48h)), select = c('Genes', 'logFC', 'adj.P.Val')) %>% 
  mutate(expression_change = case_when(
    (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
    (logFC <= -1 & adj.P.Val < 0.05) ~ 'down',
    .default = 'not significant', 
  )) %>%
  mutate(labels = case_when(
    expression_change %in% c('up', 'down') ~ Genes,
    .default = NA
  ))

dfPlot$timepoint = "48hr"


plot <- (
  ggplot(
    data = dfPlot,
    aes(
      x = logFC,
      y = -log10(adj.P.Val),
      col = expression_change,
      # label = labels
      )
    )
    + geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed')
    + geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed')
    + geom_point(size = .75)
    # + geom_text_repel(max.overlaps = 10, size = 2)
    + scale_color_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + coord_cartesian(xlim = c(-15, 15), ylim = c(0, 3))
    + theme_minimal()
    + theme(legend.position = "bottom")
    + labs(
      title = 'Spheroid global proteomics',
      subtitle = "DMSO vs. Mirdametinib at 48h",
      col = 'expression',
      y = expression("-log"[10]*"p-Value")
    )
)
plot

ggsave(
  filename = "../output_figures/global_prot_diff_exp_volcano_48h.png",
  plot = plot,
  dpi = 600,
  height = 6,
  width = 6,
  scale = .8,
  )
```

## Pathway Enrichment

```{r load-geneset, results='hide'}
# data("msigdb")
data("ncipid")
data("krbpaths")
```

Now we can ask which pathways are over-represented in both 24h and 48h time points. 
```{r leapR-pathway-enrichment, message=FALSE}

late_res <- leapR(geneset = krbpaths, eset = dmso_mirda_48h, enrichment_method = "enrichment_in_sets", assay_name = 'proteomics', id_column = "Genes",
      primary_column = "adj.P.Val", threshold = 0.05,
      greaterthan = FALSE)
plot <- plot_leapr_bar(late_res,'Top pathways at 48h')
ggsave("../output_figures/global_prot_pathway_enrichment_krbpaths_48h_top.png", plot = plot, dpi = 600)
plot

early_res <- leapR(geneset = krbpaths, eset = dmso_mirda_24h, enrichment_method = "enrichment_in_sets", assay_name = 'proteomics', id_column = "Genes",
      primary_column = "adj.P.Val", threshold = 0.05,
      greaterthan = FALSE)

plot <- plot_leapr_bar(early_res,'Top pathways at 24h')
ggsave("../output_figures/global_prot_pathway_enrichment_krbpaths_24h_top.png", plot = plot, dpi = 600)
plot

```

This shows significant pathways being altered at 48h, suggesting a more informative time point. We
can evaluate the expression of those proteins in that pathway as well to see how they
are expressed across time points.

```{r path heatmaps, message=FALSE}

#create a table of pathways and genes by ingroup name
sig_lates <- subset(late_res, BH_pvalue < 0.05) |>
  tibble::rownames_to_column('Pathway') |>
  dplyr::select(Pathway,ingroupnames)

##
plots <- lapply(unique(sig_lates$Pathway), function(path){
  genes <- stringr::str_split(subset(sig_lates, Pathway==path)[,'ingroupnames'],
                              pattern=', ') |>
    unlist()
  
  #get the data and metadata out of the expression object
  mat <- assay(dmso_mirda_48h)[rowData(dmso_mirda_48h)$Genes %in% genes,]
  rownames(mat) <-  intersect(rowData(dmso_mirda_48h)$Genes,genes)
  annot <- colData(dmso_mirda_24h)[,c('timepoint','agent','replicate')] |>
    as.data.frame()
  ###create heatmap with metadata
  plot <- pheatmap::pheatmap(log10(mat),annotation_col = annot, 
                     cluster_cols = FALSE, 
                     main=paste0(path,' proteins from 48h enrichment'),
                     labels_col = rep("",ncol(mat)))
  plot
  ggsave(filename = paste0("../output_figures/global_prot_pathway_enrichment_48h_heatmap_", path, ".png"), plot = plot, dpi = 600)
  
})

```


Some of these look kind of unimpressive. Maybe we should plot logFCs instead?

```{r leapR by Rank}

late_rank <- leapR(geneset = krbpaths, eset = dmso_mirda_48h, enrichment_method = "enrichment_in_order", assay_name = 'proteomics', id_column = "Genes",
      primary_columns = "logFC") |>
  suppressWarnings()

plot_leapr_bar(late_rank,'Top pathways at 48h')

early_rank <- leapR(geneset = krbpaths, eset = dmso_mirda_24h, enrichment_method = "enrichment_in_order", assay_name = 'proteomics', id_column = "Genes",
      primary_columns = "logFC") |>
  suppressWarnings()

plot_leapr_bar(early_rank,'Top pathways at 24h')

```

At some point we should dive into why this is the case - not a lot of agreement between
order-based and set-based enrichment. 
