---
title: "Spheroid Phospho Proteomics Analysis"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
 #pdf_document:
#    toc: true
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
---


## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide'}
library(synapser)
library(readxl)
library(readr)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
```


```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


Next is the import of the phospho proteomics abundances from spheroids.
The data file is located on synapse and is stored as tsv.

**Note this version of the notebook uses syn72351272v2 and syn72245597v3**

```{r import-phospho-data, results='hide'}
synapseFileID = "syn72351272"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsData <- readr::read_tsv(
  fileEntity$path,
  na = c("0", NA) # the file stored on synapse has '0's instead of NA or ''
  )
tmp <- synapser::synLogout()
```

```{r import-meta-data, results='hide'}
synapseFileID = "syn72245597"

tmp <- synapser::synLogin(silent = TRUE)
fileEntity <- synapser::synGet(synapseFileID)
proteomicsMeta <- read_xlsx(fileEntity$path, sheet = 3)
##try altnerate naming
proteomicsMeta = data.frame(proteomicsMeta,altSample=colnames(proteomicsData)[7:21])
tmp <- synapser::synLogout()
```


## Data preparation

Before running the actual differential expression analysis we need to first
clean the data. This consits of several steps:

- Matching DIA run identifiers to chemical agent & sample time descriptor
- Missingness removal (i.e. exclude genes from consideration that weren't
  detected in a certain fraction of MS/MS runs)
- remove additional identifier columns (move into separate DataFrame)
- log2 transform the abundance / intensity data
- impute missing values (media center based on other readouts for the protein)

- `prep_data` will take of most of the data prep including the missingness
  removal since we need to make sure that the identifier table and data table 
  that goes into the `SummarizedExperiment` match up.
- `log_transform` & `impute_proteomics` take care of their namesake.

```{r}
experiment <- prep_prot_data(proteomicsData, proteomicsMeta, type = 'phospho') %>%
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method = "SampMin")
```


## Differential Expression analysis

### PCA (No missingness removal)
With all the data imported we can focus on the actual differential expression
analysis.
First we will perform a PCA analysis of the underlying proteomics data to check
if there is meaningful signal separation of the individual treatment agents,
as well as timepoints.

*Note that no phospho sites have been removed due to missingness removal*
*i.e. all N/As are being imputed!*

```{r}
principal_components <- prcomp(t(assay(experiment))) %$%
  cbind(x, colData(experiment))

plot <- (
  ggplot(
    principal_components,
    aes(
      x = PC1,
      y = PC2,
      col = agent,
      shape = timepoint,
      label = condition
      )
    )
  + geom_point(size = 2)
  + geom_text_repel(max.overlaps = 10, size = 2)
  + theme_minimal()
  + theme(legend.position = "bottom")
  + scale_color_manual(values = c("DMSO" = "#e66101", "Mirda" = "#5e3c99", "NA" = "grey"))
  + labs(
    title = "Spheroid phospho proteomics; DMSO vs. Mirdametinib",
    subtitle = "no missingness filter; SampMin imputation"
  )
)

plot
ggsave("../output_figures/phospho_prot_PCA.png",
       plot = plot,
       dpi = 600,
       height = 6,
       width = 6,
       )
```

### PCA (50% missingness removal)
Now we repeat the PCA plot but with 50% missingness removal, i.e. only protein
phosphorylations that were detected in >= 80% of the samples are being
considered for the PCA generation.

```{r}
experiment <- prep_prot_data(proteomicsData, proteomicsMeta, type = 'phospho', 0.5) %>%
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method = "SampMin")

principal_components <- prcomp(t(assay(experiment))) %$%
  cbind(x, colData(experiment))

plot <- (
  ggplot(
    principal_components,
    aes(
      x = PC1,
      y = PC2,
      col = agent,
      shape = timepoint,
      label = condition
      )
    )
  + geom_point(size = 2)
  + geom_text_repel(max.overlaps = 10, size = 2)
  + theme_minimal()
  + theme(legend.position = "bottom")
  + scale_color_manual(values = c("DMSO" = "#e66101", "Mirda" = "#5e3c99", "NA" = "grey"))
  + labs(
    title = "Spheroid phospho proteomics; DMSO vs. Mirdametinib",
    subtitle = "proteins with >50% missingness removed; SampMin imputation"
  )
)

plot

ggsave("../output_figures/phospho_prot_PCA_50pct_missingness_removed.png",
       plot = plot,
       dpi = 600,
       height = 6,
       width = 6,
       )
```
As can be seen, the data after missingness removal separates much better than
the full dataset, albeit mostly by timepoint and not so much by drug.

One mayor point to mention here is that there seem to be two pairs of conditions
that indicate that some sample swapping might have occurred during the data
processing. Those samples are:

- 24hr_Mirda_r1 & 48hr_Mirda_r1
- 0hr_r1 & 48hr_DMSO_r3

If those two sample pairs would swap their respecive lables the PCA would truly
place all points together in space accoring to their timepoint.

### Evaluation of known targets

There are some genes that are down-regulated upon mek inhibition. what is the
expression of these?

```{r meki genes}
down_genes <- c("DUSP4", "DUSP6", "SPRY2", "SPRY4", "ETV4", "ETV5", "CCND1", "CDK5R1", "DDX21", "NGFR", "BCL3", "STK17B")

  plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=down_genes,
                                  timepoints = c('24hr','48hr'),
                                  ftype =  'Gene.Names',
                                  fmapping = 'Gene.Names') +
   ggtitle('Mirda targs')
```

### Investigation into the sample discrepancy

```{r sample-discrepancy, eval=FALSE}
experiment <- prep_prot_data(proteomicsData, proteomicsMeta, type='phospho',0)
summary(as_tibble(assay(experiment)))

```
```{r, eval=FALSE}
tmp <- as_tibble(assay(experiment)) %>% reframe(, n = colSums(is.na(across(everything())))) %>% cbind(., rownames(t(assay(experiment)))) %>% as_tibble()
tmp %<>% dplyr::rename(., count_NA = n) %>% dplyr::rename(., id = `rownames(t(assay(experiment)))`)
subset(tmp, id %in% c('cNF_spheroid_P_DIA_01', 'cNF_spheroid_P_DIA_13', 'cNF_spheroid_P_DIA_07', 'cNF_spheroid_P_DIA_12'))
```
```{r, eval=FALSE}
subset(tmp, !id %in% c('cNF_spheroid_P_DIA_01', 'cNF_spheroid_P_DIA_13', 'cNF_spheroid_P_DIA_07', 'cNF_spheroid_P_DIA_12'))
```

### Plot of known targetes

Mirdametinib has a few known targets, we can plot them here.

```{r mirda known targs, message=FALSE}
 direct_targs <- c('MAPK3-T202','MAPK3-Y204','MAPK1-Y187','MAPK1-T185')


 plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=direct_targs,
                                  timepoints = c('24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct mirda targs')

```
This is promising, let's look at other targets

```{r secondary targs, message=FALSE}

  second_targs <- c('RPS6KA1-T359', 'RPS6KA1-S363','RPS6KA1-T573',
                   'RPS6KA3-T365', "RPS6KA3-S369","RPS6KA5-T581", "RPS6KA5-S360")

  plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=second_targs,
                                  timepoints = c('24hr','48hr'),
                                  ftype =  'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Secondary mirda targs')
```

### Number of significant differentally expressed phospho sites 
Next we will perform the actual differential expression analysis.
We will define a significantly differentially experessed phospho site as

- having an adjusted p-Value < 0.05
- having a logFC >= 1 or logFC < -1

We will be investigating specifically the expression differential between DMSO
and Mirdametenib treated samples both at the 24h as well as 48h time point.
The codeblock below performs the differential expression analyses using the
previously defined function `calc_diff_ex`.
The we select the proteins that show a significant differential in their
expression between the two groups and prepare a `data.frame` for plotting the
results.
```{r differential-expression-dmso-mirda}
dmso_mirda_24h <- calc_diff_ex(experiment, group_1 = "24hr_DMSO", group_2 = "24hr_Mirda", align_by = "Protein.With.Phospho.Site")
dmso_mirda_48h <- calc_diff_ex(experiment, group_1 = "48hr_DMSO", group_2 = "48hr_Mirda", align_by = "Protein.With.Phospho.Site")

dfPlot <- subset(as_tibble(rowData(dmso_mirda_24h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val'), adj.P.Val < 0.05) %>% mutate(expression_change = case_when(
  logFC >= 1 ~ 'up',
  logFC < -1 ~ 'down'
))
dfPlot$timepoint = "24hr"

dfPlot2 <- subset(as_tibble(rowData(dmso_mirda_48h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val'), adj.P.Val < 0.05) %>% mutate(expression_change = case_when(
  logFC >= 1 ~ 'up',
  logFC < -1 ~ 'down'
))
dfPlot2$timepoint = '48hr'

dfPlot <- rbind(dfPlot, dfPlot2) %>% subset(., !is.na(expression_change))
```

The code block below plots the number of significantly up and down regulated
proteins.
```{r plot-expression-differential}
cpalette = c('up' = '#74add1', 'down' = '#f46d43')

plot <- (
  ggplot(
    data = dfPlot,
    aes(timepoint)
  )
  + geom_bar(aes(fill=expression_change), position = 'dodge2')
  + scale_fill_manual(values = cpalette)
  + theme_minimal()
  + labs(
    title = 'Spheroid phospho proteomics; DMSO vs Mirdametinib', 
    subtitle = "Counts of significant differentially expressed proteins ( pVal < 0.05; -1  >= logFC >= 1 )",
    fill = 'expression'
  )
)
plot
```

## Volcano plots of differentially expressed porteins
Next we generate volcano plots to investigate the significant proteins further.
For each time point one volcano plot will be generated.

### DMSO vs Mirdametenib at 24h
```{r volcano-plot-24h}

dfPlot <- subset(as.tibble(rowData(dmso_mirda_24h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val','quantile')) %>% 
  mutate(expression_change = case_when(
    (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
    (logFC < -1 & adj.P.Val < 0.05) ~ 'down',
    .default = 'not significant', 
  )) %>%
  mutate(labels = case_when(
    expression_change %in% c('up', 'down') ~ Protein.With.Phospho.Site,
    .default = NA
  ))

dfPlot$quantile <- factor(dfPlot$quantile, levels=c("10%", "20%", "50%", "100%"))
dfPlot$timepoint = "24hr"


plot <- (
  ggplot(
    data = dfPlot,
    aes(
      x = logFC,
      y = -log10(adj.P.Val),
      col = expression_change,
      shape = quantile,
      # label = labels
      )
    )
    + geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed')
    + geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed')
    + geom_point(size = .75)
    # + geom_text_repel(max.overlaps = 10, size = 2)
    + scale_color_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + coord_cartesian(xlim = c(-11, 11), ylim = c(0, 4))
    + theme_minimal()
    + theme(legend.position = "bottom", legend.box = "vertical")
    + labs(
      title = 'Spheroid phospho proteomics',
      subtitle = "DMSO vs. Mirdametinib at 24h",
      col = 'expression',
      y = expression("-log"[10]*"p-Value")
    )
)
plot

ggsave(
  filename = "../output_figures/phospho_prot_diff_exp_volcano_24h.png",
  plot = plot,
  dpi = 600,
  height = 6,
  width = 6,
  scale = .8,
  )
```

```{r sig-phospho-site-counts-24h}
plot <- (dfPlot |> 
 # subset(expression_change %in% c('up','down')) |> 
  ggplot(aes(x=quantile,fill=expression_change))
    + geom_bar(position='dodge')
    + scale_fill_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + scale_y_log10()
    + theme_minimal()
    + theme(legend.position = "bottom")
    + labs(    
      title = 'Spheroid phospho proteomics; DMSO vs Mirdametinib at 24h', 
      subtitle = "Significant phospho sites: pVal < 0.05; -1  >= logFC >= 1",
      fill = 'expression',
      )
  )
plot

ggsave(
  filename = "../output_figures/phospho_prot_sig_site_count_24h.png",
  plot = plot,
  dpi = 600,
  height = 6,
  width = 6,
  )
```

### DMSO vs Mirdametenib at 48h
```{r volcano-plot-48h}

dfPlot <- subset(as.tibble(rowData(dmso_mirda_48h)), select = c('Protein.With.Phospho.Site', 'logFC', 'adj.P.Val','quantile')) %>% 
  mutate(expression_change = case_when(
    (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
    (logFC < -1 & adj.P.Val < 0.05) ~ 'down',
    .default = 'not significant', 
  )) %>%
  mutate(labels = case_when(
    expression_change %in% c('up', 'down') ~ Protein.With.Phospho.Site,
    .default = NA
  ))

dfPlot$quantile <- factor(dfPlot$quantile, levels=c("10%", "20%", "50%", "100%"))
dfPlot$timepoint = "48hr"


plot <- (
  ggplot(
    data = dfPlot,
    aes(
      x = logFC,
      y = -log10(adj.P.Val),
      col = expression_change,
      shape = quantile
      # label = labels
      )
    )
    + geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed')
    + geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed')
    + geom_point(size = .75)
    # + geom_text_repel(max.overlaps = 10, size = 2)
    + scale_color_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + coord_cartesian(xlim = c(-11, 11), ylim = c(0, 4))
    + theme_minimal()
    + theme(legend.position = "bottom", legend.box = "vertical")
    + labs(
      title = 'Spheroid phospho proteomics',
      subtitle = "DMSO vs. Mirdametinib at 48h",
      col = 'expression',
      y = expression("-log"[10]*"p-Value")
    )
)
plot

ggsave(
  filename = "../output_figures/phospho_prot_diff_exp_volcano_48h.png",
  plot = plot,
  dpi = 600,
  height = 6,
  width = 6,
  scale = .8,
  )

```

```{r }
plot <- (dfPlot |> 
 # subset(expression_change %in% c('up','down')) |> 
  ggplot(aes(x=quantile,fill=expression_change))
    + geom_bar(position='dodge')
    + scale_fill_manual(values = c('up' = '#74add1', 'down' = '#f46d43', 'not significant' = 'grey'))
    + scale_y_log10()
    + theme_minimal()
    + theme(legend.position = "bottom")
    + labs(    
      title = 'Spheroid phospho proteomics; DMSO vs Mirdametinib at 48h', 
      subtitle = "Significant phospho sites: pVal < 0.05; -1  >= logFC >= 1",
      fill = 'expression',
      )
  )
plot

ggsave(
  filename = "../output_figures/phospho_prot_sig_site_count_48h.png",
  plot = plot,
  dpi = 600,
  height = 6,
  width = 6,
  )
```
### Pathway enrichment for Kinase substrate Pathways

First we load the gene set table we need for the pathway enrichment that will
be done by leapR.

```{r load-leapR-gmt}
data("kinasesubstrates")
```

Next we can use the already established differential expression data and run
that data through `leapR`.
We are using the `kinasesubstrates` gene sets here, to look for enriched
kinase substrate pathways.
The first set we will look at is enrichment at the **24h timepoint** between 
**DMSO** and **Mirdametinib**.

```{r kinasesubstrate-enrichment-24h, message=FALSE}
early_res <- leapR(geneset = kinasesubstrates,
                   eset = dmso_mirda_24h,
                   enrichment_method = "enrichment_in_sets",
                   assay_name = 'proteomics',
                   id_column = "Gene.With.Phospho.Site",
                   primary_column = "adj.P.Val",
                   threshold = 0.05,
                   greaterthan = FALSE)

plot <- plot_leapr_bar(early_res,'Top pathways at 24h')
ggsave("../output_figures/phospho_prot_pathway_enrichment_kinasesubstrates_24h_top.png", plot = plot, dpi = 600)
plot
```
Finally we do the same for the **48h timepoint**.

```{r kinasesubstrate-enrichment-48h, message=FALSE}
late_res <- leapR(geneset = kinasesubstrates,
                   eset = dmso_mirda_48h,
                   enrichment_method = "enrichment_in_sets",
                   assay_name = 'proteomics',
                   id_column = "Gene.With.Phospho.Site",
                   primary_column = "adj.P.Val",
                   threshold = 0.05,
                   greaterthan = FALSE)

plot <- plot_leapr_bar(late_res,'Top pathways at 48h')
ggsave("../output_figures/phospho_prot_pathway_enrichment_kinasesubstrates_48h_top.png", plot = plot, dpi = 600)
plot
```


## Rank enrichment

let's calculate order enrichment

```{r order}


late_rank <- leapR(geneset = kinasesubstrates, eset = dmso_mirda_48h, enrichment_method = "enrichment_in_order", assay_name = 'proteomics', id_column = "Gene.With.Phospho.Site",
      primary_columns = "logFC") |>
  suppressWarnings()

plot_leapr_bar(late_rank,'Top pathways at 48h')

early_rank <- leapR(geneset = kinasesubstrates, eset = dmso_mirda_24h, enrichment_method = "enrichment_in_order", assay_name = 'proteomics', id_column = "Gene.With.Phospho.Site",
      primary_columns = "logFC") |>
  suppressWarnings()

plot_leapr_bar(early_rank,'Top pathways at 24h')

```
## Combine with MN2

Make sure to run the MN2 notebook first!

```{r mn2}

###THESE TWO FILES WERE WRITTEN BY mn2_global_params.Rmd
mn2_phospho <- readr::read_csv('mn2_phospho_enrichment_results.csv')

edf <- early_rank |>
  tibble::rownames_to_column('feature') |>
  dplyr::select(feature, zscore, pval = 'BH_pvalue') |>
  mutate(agent = 'Mirda', timepoint = '24hr', sig=FALSE)

edf$sig[which(edf$pval < 0.05)] <- TRUE

ldf <-  late_rank |>
  tibble::rownames_to_column('feature') |>
  dplyr::select(feature, zscore, pval = 'BH_pvalue') |>
  mutate(agent = 'Mirda', timepoint = '48hr', sig=FALSE)

ldf$sig[which(ldf$pval < 0.05)] <- TRUE

jhu_full <- rbind(edf, ldf) |>
  mutate(sample = 'JHU-2-079c')

both_samps <- mn2_phospho |>
  mutate(sample = 'MN2') |>
  rbind(jhu_full) |>
  subset(agent=='Mirda')


tps <- both_samps|>
  subset(sig) |>
  group_by(feature) |>
  summarize(meanz = mean(zscore, na.rm=T)) |>
  arrange(rev(meanz))
  
forder <- unique(tps$feature)
torder <- c('8hr','24hr','48hr')
both_samps$timepoint <- factor(both_samps$timepoint, levels = torder)



maxz = max(abs(both_samps$zscore))
ggplot(both_samps, aes(y = feature, x = timepoint, color = zscore, size = -log10(pval))) +
   geom_point() +
   ggplot2::scale_y_discrete(limits = forder) +
  # ggplot2::scale_x_discrete(limits = torder) +
   scale_color_gradient2(low="#2266bb",high="#cc3311", mid="lightgrey",
                         limits=c(-maxz, maxz)) +
   theme_minimal(base_size=12)+
   geom_point(data = subset(both_samps, sig), col = "black", stroke = 1, shape = 21) +
    facet_grid(~sample)+
   theme(axis.title=element_blank(), 
        axis.text.x=element_text(angle=45, vjust=1, hjust=1))+
  ggtitle("Combined Phosphoproteomic pathways")

ggsave('two_sample_mirda_phospho.png',width=8)

```
