---
title: "Phosphoproteomic Timepoint Comparison"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
 # pdf_document:
#    toc: false
  html_notebook:
    toc: true
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(gridExtra)
library(ggalluvial)
library(sva)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


## Data collection and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting phospho-proteomics abundance data
synapseFileID = "syn74589783"
fileEntity <- synapser::synGet(synapseFileID)
phospho_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data_raw <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}
# cleaning up the column names for the phospho-proteomics data
names(phospho_dat) <- names(phospho_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\PhosphoRawData\\\\", "", .) %>% 
  gsub("_20Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)


```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data_raw %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data_raw %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data_raw %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)

meta_data <- list(
  "global" = meta_data_global,
  "phospho" = meta_data_phospho
)
```


## Differential expression analysis parameter sweep


First we want to evaluate the missingness options to see how they alter the
differential expression and pathways. 

```{r diff-ex-parameter-sweep, eval=FALSE}
missingness_cutoff = 0
samp_cutoff= 0
imputation_method = "none"

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")



logfc = 0.5
pval = 0.05

experiment <- prep_experiment_data(
  phospho_dat,
  data_type = 'phospho',
   meta_data_phospho,
  missingness_cutoff,
  samp_cutoff,
  imputation_method
  )

##one table for diffex diffex stats, one for enrichment
all_diff_ex <- tibble()

for (mc in c(0,0.25,0.5,0.75)){
  for( sc in c(0.1, 0.25, 0.5)){
    for( im in c('SampMin','means')){
      for(agent in agents){
        for(timepoint in timepoints){
          ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
          group_1 = paste0(timepoint, "_", ctrl)
          group_2 = paste0(timepoint, "_", agent)
          plate_sel <- unique(experiment[, experiment$group == group_2]$plate)
          experiment_subset <- exp_subset(experiment, agent, timepoint) |>#)experiment[, experiment$plate == plate_sel]|>
          filter_for_sample_cutoff(sc) |>
          filter_for_feature_missingness(mc) |>
          impute_proteomics(imputation_method = im) |>
          log_transform()
  
          align_by_group <- ifelse(test = (prot_type == 'global'), yes = "Protein.Group",
                       no = "Protein.With.Phospho.Site")
          diff_ex <- calc_diff_ex(experiment_subset, group_1, group_2,
                        align_by=align_by_group) 
          ##getting stats
          sig_diff <- as.tibble(rowData(diff_ex))|>
            dplyr::select(Feature = 'Gene.With.Phospho.Site','logFC','adj.P.Val') |>
            subset(abs(logFC) > logfc) |>
            subset(adj.P.Val<pval) 
          n_up = subset(sig_diff, logFC>0) |>nrow()
          n_down = subset(sig_diff, logFC<0) |>nrow()
          
          stat_list <- tibble(exp=list(diff_ex), #yes i'm embedding an object in a table for leapR i hate it
                            agent = agent, 
                            timepoint = timepoint, 
                            missing = mc, 
                            sample = sc, 
                            imputation = im,
                            up_features = n_up,
                            down_features = n_down)
         all_diff_ex <- rbind(all_diff_ex, stat_list)
        }
      }
    }
  }
}

```

Now that we have one master table of all diffex calls across parameters, we can
plot the number of features we get for each time point and sample


### Plot across parameters

First we get the diffex stats and plot
```{r diffex stats}


all_diff_ex |> 
  mutate(df = -1 * down_features) |>
  tidyr::pivot_longer(c(up_features, df), names_to='dir', values_to='features') |>
  ggplot(aes(x=timepoint, y=features, fill=agent, shape = imputation)) + 
  geom_bar(stat='identity', position = 'dodge') +
  facet_grid(missing~sample)

```

The 48 hr time point has the most number of chemicals represented,
and three drugs are clear outliers, so now we can focus on other parms

```{r params}

all_diff_ex |> 
  mutate(df = -1 * down_features) |>
  subset(!agent%in%c('CAP','DOXO','IAG')) |>
  tidyr::pivot_longer(c(up_features, df), names_to='dir', values_to='features') |>
  subset(timepoint == '48hr') |>
  ggplot(aes(x=missing, y=features, fill=agent, shape = imputation)) + 
  geom_bar(stat='identity', position = 'dodge') +
  facet_grid(~sample)


```
from this it looks like we should not filter for sample missingness at all...

### Pathway enrichment across parameters
Now we can see how many pathways are enriched across parameters.
I first create some helper funnctions so we can call them from mutate.

```{r load-geneset, results='hide'}
data("kinasesubstrates")

ppval <- 0.05

##create function to call leapR
get_leapr_order <- function(exp,
                          geneset=kinasesubstrates,
                          id_column="Gene.With.Phospho.Site"){
  res <- NULL
  try(res <- leapR(
      geneset = geneset,
      eset = unlist(exp),
      enrichment_method = "enrichment_in_order",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "logFC") |>
     subset(BH_pvalue < ppval))
  return(nrow(res))
}

get_leapr_sets <- function(exp,
                          geneset=kinasesubstrates,
                          id_column="Gene.With.Phospho.Site"){
  res <- NULL
  try(res <- leapR(
        geneset = geneset,
      eset = unlist(exp),
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "adj.P.Val",
        threshold = 0.05,
      greaterthan = FALSE) |>
     subset(BH_pvalue < ppval) 
  )
  return(nrow(res))
}

```

Now we cycle through the parameters again and calculate the differentially
enriched kinases. 

#### Phospho

```{r, warning=FALSE, echo=FALSE, message=FALSE}
pval <- 0.1
prot_type <- "phospho"
id_column <- "Gene.With.Phospho.Site"
geneset <- kinasesubstrates

dp <- all_diff_ex |>
  subset(missing!=0.75)

dp$numpathsOrder <- sapply(dp$exp,get_leapr_order)
dp$numPathsSet <- sapply(dp$exp,get_leapr_sets)

```


Now we have calculated the number of kinases and can plot across various parameters

```{r plot path enrichment}


```
