---
title: "Global Timepoint Comparison"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
 # pdf_document:
#    toc: false
  html_notebook:
    toc: true
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(gridExtra)
library(ggalluvial)
library(sva)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


## Data collection and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting the global proteomics abundance data
synapseFileID = "syn74552028"
fileEntity <- synapser::synGet(synapseFileID)
global_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )



# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data_raw <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}

# cleaning up the column names for the global proteomics data
names(global_dat) <- names(global_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\rawdata\\\\", "", .) %>% 
  gsub("_16Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)




```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data_raw %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data_raw %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data_raw %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)

meta_data <- list(
  "global" = meta_data_global,
  "phospho" = meta_data_phospho
)
```





## Differential expression analysis parameter sweep


First we want to evaluate the missingness options to see how they alter the
differential expression and pathways. 

```{r diff-ex-parameter-sweep, eval=FALSE}
missingness_cutoff = 0
samp_cutoff= 0
imputation_method = "none"

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")

prot_type = 'global'

logfc = 0.5
pval = 0.05

experiment <- prep_experiment_data(
  global_dat,
  data_type = 'global',
   meta_data_global,
  missingness_cutoff,
  samp_cutoff,
  imputation_method
  )

##one table for diffex diffex stats, one for enrichment
all_diff_ex <- tibble()

for (mc in c(0,0.1,0.25,0.5)){ ##fraction of features that are present
  for(sc in c(0.1, 0.25, 0.5)){ #fraction of samples that have a measuremetn
    for( im in c('SampMin','means')){
      for(agent in agents){
        for(timepoint in timepoints){
          ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
          group_1 = paste0(timepoint, "_", ctrl)
          group_2 = paste0(timepoint, "_", agent)
          plate_sel <- unique(experiment[, experiment$group == group_2]$plate)
          experiment_subset <- exp_subset(experiment, agent, timepoint) |>#)experiment[, experiment$plate == plate_sel]|>
          filter_for_sample_cutoff(sc) |>
          filter_for_feature_missingness(mc) |>
          impute_proteomics(imputation_method = im)
  
          align_by_group <- ifelse(test = (prot_type == 'global'), yes = "Protein.Group",
                       no = "Protein.With.Phospho.Site")
          diff_ex <- calc_diff_ex(experiment_subset, group_1, group_2,
                        align_by=align_by_group) 
          ##getting stats
          sig_diff <- as.tibble(rowData(diff_ex))|>
           # dplyr::select(Feature = 'Gene','logFC','adj.P.Val') |>
            subset(abs(logFC) > logfc) |>
            subset(adj.P.Val < pval) 
          n_up = subset(sig_diff, logFC > 0) |>nrow()
          n_down = subset(sig_diff, logFC < 0) |>nrow()
          
          stat_list <- tibble(exp = list(diff_ex), #yes i'm embedding an object in a table for leapR i hate it
                            agent = agent, 
                            timepoint = timepoint, 
                            missing = mc, 
                            sample = sc, 
                            imputation = im,
                            up_features = n_up,
                            down_features = n_down)
         all_diff_ex <- rbind(all_diff_ex, stat_list)
        }
      }
    }
  }
}

```


Now we can go ahead and plot the results


### Plot across parameters

First we get the diffex stats and plot
```{r diffex stats}


all_diff_ex |> 
  subset(imputation=='means') |>
  mutate(df = -1 * down_features) |>
  tidyr::pivot_longer(c(up_features, df), names_to='dir', values_to='features') |>
  ggplot(aes(x=timepoint, y=features, fill=agent, shape = imputation)) + 
  geom_bar(stat='identity') +
  facet_grid(missing~sample) + 
    ggtitle("Means imputation")
  
  all_diff_ex |> 
  subset(imputation=='SampMin') |>
  mutate(df = -1 * down_features) |>
  tidyr::pivot_longer(c(up_features, df), names_to='dir', values_to='features') |>
  ggplot(aes(x=timepoint, y=features, fill=agent, shape = imputation)) + 
  geom_bar(stat='identity') +
  facet_grid(missing~sample) + 
    ggtitle("sample minimum imputation")

```

Now we can refine our search based on what we see.



### Pathway enrichment across parameters
Now we can see how many pathways are enriched across parameters.

I first create three helper functions to evaluate differences in substrate 
distribution, up-regulation, and down-regulation. 

```{r load-geneset, results='hide'}
data("ncipid")

ppval <- 0.05

hoc <- msigdbr::msigdbr(collection = "H")

df_to_gmt_long <- function(df, gene_col, pathway_col, output_file) {
  # Split the genes into a list based on their pathway
  gene_lists <- split(df[[gene_col]], df[[pathway_col]])
  
  # Open the file and write each pathway as a tab-separated line
  con <- file(output_file, "w")
  for (pathway in names(gene_lists)) {
    genes <- unique(gene_lists[[pathway]])
    line <- c(pathway, "null", genes)
    writeLines(paste(line, collapse = "\t"), con)
  }
  close(con)
}

df_to_gmt_long(hoc, 'gene_symbol','gs_name','hallmarks.gmt')

hgs <- leapR::read_gene_sets('hallmarks.gmt')
hgs$matrix <- t(hgs$matrix) ##why are we always backward shere? 
##create function to call leapR
get_leapr_order <- function(exp,
                          geneset=ncipid,
                          id_column="Genes"){
  res <- data.frame()
  try(
    res <- leapR(
      geneset = geneset,
      eset = unlist(exp),
      enrichment_method = "enrichment_in_order",
      assay_name = 'proteomics',
      id_column = id_column,
     # method = 'ztest',
      primary_column = "logFC") |>
      tibble::rownames_to_column('feature')
      )
  return(res)
}


get_leapr_sig <- function(exp,
                          geneset = ncipid,
                          id_column ='Genes'){
    res <- data.frame()
  try(res <- leapR(
        geneset = geneset,
      eset = unlist(exp),
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "adj.P.Val",
        threshold = 0.05,
      greaterthan = FALSE) |>
      tibble::rownames_to_column('feature')
  )
  return(res)
}
get_leapr_sets_down <- function(exp,
                          geneset=ncipid,
                          id_column="Genes"){
  res <- data.frame()
  try(res <- leapR(
        geneset = geneset,
      eset = unlist(exp),
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "logFC",
        threshold = -1,
      greaterthan = FALSE) |>
      tibble::rownames_to_column('feature')
  )
  return(res)
}

get_leapr_sets_up <- function(exp,
                          geneset=ncipid,
                          id_column="Genes"){
  res <- data.frame()
  try(res <- leapR(
        geneset = geneset,
      eset = unlist(exp),
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "logFC",
        threshold = 1,
      greaterthan = TRUE)  |>
      tibble::rownames_to_column('feature')
      )
  return(res)
}

get_sig_feats <- function(res, ppval = 0.1){
    subset(res,BH_pvalue < ppval) |>
    nrow()
}

```

Now we cycle through the parameters again and calculate the differentially
enriched kinases. 

## Global pathway enrichment

Here we look at how many kinases are enriched using three approaches:
1. Enrichment in order/ks test
2. how many pathways have increased activity
3. how many pathways have decreased activity

```{r, warning=FALSE, echo=FALSE, message=FALSE, error=FALSE}

prot_type <- "global"
id_column <- "Genes"


##let's run all the enrichment
dp <- all_diff_ex |>
#  subset(missing != 0.75) |>
  subset(imputation == 'SampMin') |>
  rowwise() |>
  mutate(pathEnrichOrder = list(get_leapr_order(exp, geneset=hgs))) |>
  rowwise() |>
  mutate(pathEnrichPval = list(get_leapr_sig(exp, geneset=hgs)))

##then get all the counts we need

dp <- dp |>
  rowwise() |>
  mutate(numpathsOrder = get_sig_feats(pathEnrichOrder)) |>
  rowwise() |>
  mutate(pathEnrichPval = get_sig_feats(pathEnrichPval))


```


Now we have calculated the number of kinases and can plot across various parameters

```{r plot path enrichment}

dp |>
  ggplot(aes(x=timepoint, y = numpathsOrder, fill = agent))+
  geom_bar(stat='identity',position = 'dodge')+
  facet_grid(sample~missing) +
  ggtitle('Enriched pathways in order')


dp |>
  ggplot(aes(x=timepoint, y = pathEnrichPval, fill = agent))+
  geom_bar(stat='identity',position = 'dodge')+
  facet_grid(sample~missing) +
  ggtitle('Enriched pathways significant')

```

## Known global targets

For some genes there is a clear response


```{r knwon mirda targs}

down_genes <- c("DUSP4", "DUSP6", "SPRY2", "SPRY4", "ETV4", "ETV5", "CCND1", "CDK5R1", "DDX21", "NGFR", "BCL3", "STK17B")

  plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=down_genes,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype =  'Genes',
                                  fmapping = 'Genes') +
   ggtitle('Mirda targs')
  
```
Next we can plot hte pathways at each time point. It seems that no filtering is needed for these datsets.


```{r plot enriched pathways, warning=FALSE}

 rp <- subset(dp, missing == 0.5) |>
   subset(imputation == 'SampMin') |>
   subset(sample == 0.5)
  
full_tab <- do.call(rbind, lapply(unique(rp$agent), function(ag) {
  ap <- subset(rp, agent == ag)
  
  tab <- format_enrich_result(
    ap$pathEnrichOrder,  #list of enrichment results
    ap$timepoint, #list of timepoints
    ag,
    pval_thresh = 0.1)
  
  if(nrow(tab) > 0) {
  p <- plot_enrich_result(
   tab,
     ap$exp, #list of differential expression results
   ap$pathEnrichOrder,
    agent = ag,
    fmapping = 'Genes')
  print(p)
  }
  return(tab)
}))

full_tab <- full_tab|>
  rowwise() |>
  mutate(pathway = gsub("HALLMARK_","",feature)) |>
  mutate(pathway = tolower(gsub('_',' ', pathway)))

##ploting params
tps <- full_tab|>
  group_by(pathway) |>
  summarize(meanz = mean(zscore)) |>
  arrange(rev(meanz))
  
forder <- unique(tps$pathway)
torder <- c('8hr','24hr','48hr')
full_tab$timepoint <- factor(full_tab$timepoint, levels = torder)



maxz = max(abs(full_tab$zscore))
ggplot(full_tab, aes(y = pathway, x = agent, color = zscore, size = -log10(pval))) +
   geom_point() +
   ggplot2::scale_y_discrete(limits = forder) +
  # ggplot2::scale_x_discrete(limits = torder) +
   scale_color_gradient2(low="#2266bb",high="#cc3311", mid="lightgrey",
                         limits=c(-maxz, maxz)) +
   theme_minimal(base_size=12)+
  facet_grid(~timepoint)+
   geom_point(data = subset(full_tab, sig), col = "black", stroke = 1, shape = 21) +
   theme(axis.title=element_blank(), 
        axis.text.x=element_text(angle=45, vjust=1, hjust=1))+
  ggtitle("MN2 Global proteomic pathways")

readr::write_csv(full_tab, 'mn2_global_enrichment_results.csv')
ggsave('allPathEnrichment.pdf',width=12,height=8)


```
