---
title: "MN2 global proteomics exploration with Non-negative Matrix Factorization"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
 # pdf_document:
#    toc: false
  html_notebook:
    toc: true
---

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```

```{r imports, echo=FALSE, results='hide'}
library(synapser)
library(readxl)
library(readr)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(NMF)
library(purrr)
```


## Data collection and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting the global proteomics abundance data
synapseFileID = "syn74552028"
fileEntity <- synapser::synGet(synapseFileID)
global_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# getting phospho-proteomics abundance data
synapseFileID = "syn74589783"
fileEntity <- synapser::synGet(synapseFileID)
phospho_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data_raw <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}

# cleaning up the column names for the global proteomics data
names(global_dat) <- names(global_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\rawdata\\\\", "", .) %>% 
  gsub("_16Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)

# cleaning up the column names for the phospho-proteomics data
names(phospho_dat) <- names(phospho_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\PhosphoRawData\\\\", "", .) %>% 
  gsub("_20Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)


prot_data <- list(
  "global" = global_dat,
  "phospho" = phospho_dat
)

```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data_raw %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data_raw %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data_raw %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)

meta_data <- list(
  "global" = meta_data_global,
  "phospho" = meta_data_phospho
)
```

As final perpartion we define a few bespoke helper functions that are only
applicable to this dataset:

```{r bespoke-helper-functions}

prep_experiment_data <- function(
    prot_data,
    meta_data,
    missingness_cutoff=0,
    samp_cutoff = list(global = 0, phospho = 0),
    imputation_method=c("none", 'SampMin', 'means', 'median'), 
    batch_correct = FALSE
    ){
  
  imputation_method_sel <- match.arg(imputation_method)
  
  if(!is.na(missingness_cutoff) & !(missingness_cutoff >=0 | missingness_cutoff <= 1)){
    stop(
      base::paste(
        "missingness_cutoff needs to be '0 <= missingness_cutoff <=1'",
        "or 'NULL'. Supplied value:", missingness_cutoff),
      call. = FALSE
    )
  }
  
  experiments <- list()
  for(data_type in c('global', 'phospho')){
      
    experiment <- prep_prot_data(proteomics_data = prot_data[[data_type]],
                                 proteomics_meta = meta_data[[data_type]],
                                 missingness_cutoff = missingness_cutoff,
                                 samp_cutoff = samp_cutoff[[data_type]],
                                 type = data_type, 
                                 imputation_method=imputation_method_sel)
    
    if(imputation_method_sel == "none"){
      experiment %<>%
        log_transform(., assay = "proteomics")
    }
    
    if(!(imputation_method_sel == 'none')){
      experiment %<>% 
        impute_proteomics(., assay = "proteomics", imputation_method_sel) %>%
        log_transform(., assay = "proteomics")
      
    }
    
    if(batch_correct && imputation_method_sel != 'none'){
      #first we have to remove zero hour beacuse it is confounded by batch
      #experiment <- experiment[,-which(colData(experiment)$timepoint=='0hr')]
      prot_matrix <- assay(experiment)
      batch_var = colData(experiment)$plate
      cond_var = colData(experiment)$agent
      mod_matrix <- model.matrix(~ as.factor(agent), data = colData(experiment))
      corrected_prot <- sva::ComBat(
             dat = prot_matrix,
             batch = batch_var,
             mod = mod_matrix,
             par.prior = FALSE
              )
    assay(experiment,withDimnames = FALSE) <- corrected_prot  

      
    }
    experiments[[data_type]] <- experiment
  }
  
  experiments
}


```


## Data quality control

As a first step we are doing some quality control & assessment for the global
and phospho-proteomics data.
To that end we are generating PCA plots for each of the plates that have been
prepared and run through the Mass-spectrometry pipeline.
Each plate contains 60 wells, each containing spheroids treated with a chemical
agent where four wells each were pooled into one sample.

In general each plate contains:

- 4 chemical agents (3 samples each)
- 1 control (DMSO - 3 samples each)
- at one timepoint.

We generate `SummarizedExperiment` objects for the global & phospho-proteomics
data with *NO* missingness cut-off for feature removal and *Sample Minimum*
*imputation*. The imputation step is necessary such that principal components
can be extracted.

```{r prep-data-for-PCA, eval=TRUE}

missingness_cutoff = 0
imputation_method = "SampMin"

samp_cutoff = list(global = 0, phospho = 0)

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  samp_cutoff,
  imputation_method,
  # batch_correct = TRUE
)


```

## Non-negative Matrix Factorization

The `NMF` module contains functions to run NMF with a `matrix`, `data.frame` or 
`ExpressionSet` as input. We are taking the `matrix` that is contained within 
the `Assay` object inside of the `SummarizedExperiment` as input.



```{r}
# getting only the phospho proteomics data
experiments_phospho <- experiments$phospho

# for testing purposese retrieving only one plate
exp <- experiments_phospho[, experiments_phospho$plate == 1]

# retrieve the phosphosite expression matrix
# TODO: we probably want to exclude the rows that are entirely 0s (after plate subsetting)
exp_assay <- assay(exp)

# Here we are doing a rank sweep to find the optimal rank
# the second argument can be defined as a range (in this case 2-6) and will run
# NMF for each rank
# 
# nrun defines how often the NMF is rerun and a consensus is built to generate
# stable results. For demonstration this is currently set to 10 but really
# should be 50-100 (maybe even higher?)
nrun <- 10
estimate_rank <- nmf(exp_assay, 2:6, nrun=nrun)
plot(estimate_rank)

```
Using the rank survey (above) we can see that rank 4 should yield be "best" results.
Hence we will set the rank to 4 and rerun NMF to extract the metagenes.
```{r}
# running the actual NMF
# nrun runs NMF n times and chooses the "best"

rank <- 4
res <- nmf(exp_assay, rank=rank, nrun=nrun)
```

```{r}
# consensus(res) extracts the consensus matrix from the NMF results / model object
# it effectively lists the "distance" between the coefficients (samples) based on the metagenes
consensus_matrix <- consensus(res)

# getting additional annotations (agent descriptors) for and additional labeling of the heatmap
annotation <- colData(exp)[['agent']]

# plotting the consensus heatmap
consensusmap(res, annCol=annotation)
```
Next we can extract the meta genes and coefficients.

```{r}
# Extracting W (meta_genes) & H (coefficients)
meta_genes <- basis(res)
coefficients_matrix <- coef(res)
```


Finally we can extract the features that are part of each metagene

```{r}
# extract the features for the NMF result
features <- extractFeatures(res)

# the result from `extractFeatures` returns a list of meta genes where each
# metagene in itself is a list of indices corresponding to the rows in W.
# We can retrieve the actual feature names by mapping back to the rownames of W.
features_mapped <- features %>% map(~ map_vec(.x, ~rownames(meta_genes)[.x]))

# Return the now mapped / named feature lists (1 per meta_gene)
features_mapped
```

Now we can do the same thing dynamically for other plates with the functions
defined in `src/helper_funcs.R`.
In theory the rank analysis should be rerun for each plate individually, but for
demonstration purposes we will just leave it at 4.

```{r}

res <- nmf_per_plate(experiments_phospho, plate=2, rank, nrun)
features <- get_features(res)
features[1]
```



