---
title: "Targeted site comparison"
author: "Sara Gosline"
date: "2026-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(gridExtra)
library(ggalluvial)
library(sva)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


## Data collection and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting phospho-proteomics abundance data
synapseFileID = "syn74589783"
fileEntity <- synapser::synGet(synapseFileID)
phospho_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data_raw <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}
# cleaning up the column names for the phospho-proteomics data
names(phospho_dat) <- names(phospho_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\PhosphoRawData\\\\", "", .) %>% 
  gsub("_20Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)


```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data_raw %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data_raw %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data_raw %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)

meta_data <- list(
  "global" = meta_data_global,
  "phospho" = meta_data_phospho
)
```

Now load in the data

```{r load dadta}
missingness_cutoff = 0
samp_cutoff= 0
imputation_method = "none"

  
experiment <- prep_experiment_data(
  phospho_dat,
  data_type = 'phospho',
   meta_data_phospho,
  missingness_cutoff,
  samp_cutoff,
  imputation_method
  )

```

## Individual kinase analysis

Now we can look into the individual drug targets to see where we observe the 
expected results. Since we collected all the results in the table we should be 
individually.

### Mirdametinib MAP2K1 and MAP2K2
```{r mirda targs, eval = FALSE, warning=FALSE, error=FALSE}
##now we have the 
targs <- c('MAP2K1','MAP2K2')

getPvalueForFeature <- function(tab, protein){
  subset(tab, feature==protein)|>
    dplyr::select(pvalue)
}

 m2k <-dp |>
  subset(agent%in% c('Mirda')) |>
  rowwise() |>
  mutate(map2k1_order_pval = getPvalueForFeature(kinEnrichOrder,'MAP2K1')) |>
  rowwise() |>
  mutate(map2k1_up_pval = getPvalueForFeature(kinEnrichUp,'MAP2K1')) |>
  rowwise() |>
  mutate(map2k1_down_pval = getPvalueForFeature(kinEnrichDown,'MAP2K1')) 
 
 
 m2k2 <-dp |>
  subset(agent%in% c('Mirda')) |>
  rowwise() |>
  mutate(map2k1_order_pval = getPvalueForFeature(kinEnrichOrder,'MAP2K2')) |>
  rowwise() |>
  mutate(map2k1_up_pval = getPvalueForFeature(kinEnrichUp,'MAP2K2')) |>
  rowwise() |>
  mutate(map2k1_down_pval = getPvalueForFeature(kinEnrichDown,'MAP2K2')) 
 
```
This didnt' work very well, going to plot targets manually

Mirdametinib blocks the non-ATP-competitive allosteric site of MEK1/2, preventing the dual-phosphorylation of downstream ERK1/2. Literature validation for this signature, downstream targets, and functional impacts can be reviewed via [PMID: 39695081](https://pubmed.ncbi.nlm.nih.gov/39695081/) and [PMID: 41816962](https://pubmed.ncbi.nlm.nih.gov/41816962/).


| Protein Node | Gene Name-Phosphosite | Functional Significance | Source Evidence |
| :--- | :--- | :--- | :--- |
| **Core Pathway** | MAPK3-T202<br>MAPK3-Y204<br>MAPK1-T185<br>MAPK1-Y187 | Primary ERK1/2 activation loop (TEY motif) | [PMC11083651](https://pmc.ncbi.nlm.nih.gov/articles/PMC11083651/) |
| **Downstream Kinases** | RPS6KA1-T359<br>RPS6KA1-S363<br>RPS6KA1-T573<br>RPS6KA3-T365<br>RPS6KA3-S369<br>RPS6KA5-T581<br>RPS6KA5-S360 | RSK and MSK family activation loops | [PMC11083651](https://pmc.ncbi.nlm.nih.gov/articles/PMC11083651/) |
| **Transcription Factors** | FOSL1-S252<br>FOSL1-S265<br>FOS-S362<br>FOS-S374 | FRA1 and c-Fos stabilization dynamics | [PMC11083651](https://pmc.ncbi.nlm.nih.gov/articles/PMC11083651/) |
| **Cytoskeleton & Adapters** | STMN1-S25<br>GAB1-S552 | Microtubule and scaffold assembly | [PMC11083651](https://pmc.ncbi.nlm.nih.gov/articles/PMC11083651/) |
| **Feedback Loops** | MAP2K1-T292<br>MAP2K1-T386<br>SOS1-S1137<br>SOS1-S1167<br>SOS1-S1178<br>SOS1-S1193<br>EGFR-T669 | ERK-mediated inhibitory feedback loop sites | [PMC11655561](https://pmc.ncbi.nlm.nih.gov/articles/PMC11655561/) |


 
```{r plot results, echo = FALSE, message= FALSE}
 direct_targs <- c('MAPK3-T202','MAPK3-Y204','MAPK1-Y187','MAPK1-T185')


 plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=direct_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct mirda targs')
 
 plot_features_across_timepoints(experiment,
                                 agent =setdiff(experiment$agent,'DMSO'),
                                  features=direct_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct mirda targs')
 
 
 
```






```{r secondary mirdametinib targets, message=FALSE} 
 
  second_targs <- c('RPS6KA1-T359', 'RPS6KA1-S363','RPS6KA1-T573',
                   'RPS6KA3-T365', "RPS6KA3-S369","RPS6KA5-T581", "RPS6KA5-S360")

  plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=second_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype =  'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Secondary mirda targs')

```


```{r other affected targets, message=FALSE}  
 cyto_targs <-  c("FOSL1-S252", "FOSL1-S265", 
                  "FOS-S362", "FOS-S374",
                  "STMN1-S25", "GAB1-S552")

   plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=cyto_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype =  'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Other mirda targs')
   

```
   
```{r known feedback mechanisms, message=FALSE}
   feedback_targs <- c( "MAP2K1-T292", "MAP2K1-T386", 
                     "SOS1-S1137", "SOS1-S1167", 
                    "SOS1-S1178", "SOS1-S1193", "EGFR-T669")  
    plot_features_across_timepoints(experiment,
                                 agent = 'Mirda',
                                  features=feedback_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype =  'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Potential feedback mirda targs')
  
```




### Palbociclib CDK4 and CDK6 

Now we look at targets that would be affected by CDK4/6 inhibition

Palbociclib stops G1-phase cell cycle progression by selectively preventing CDK4/6 from phosphorylating Retinoblastoma protein (Rb). Quantitative proteomic, chemoproteomic, and phenotypic shift evidence can be examined in [PMID: 27571378](https://pubmed.ncbi.nlm.nih.gov/27571378/) and [PMC9273754](https://pmc.ncbi.nlm.nih.gov/articles/PMC9273754/).


| Protein Node | Gene Name-Phosphosite | Functional Significance | Source Evidence |
| :--- | :--- | :--- | :--- |
| **Primary Target** | RB1-S780<br>RB1-S795<br>RB1-S807<br>RB1-S811 | Canonical CDK4/6 inhibitory sites on Rb | [PMC6317929](https://pmc.ncbi.nlm.nih.gov/articles/PMC6317929/) / [PMID: 27571378](https://pubmed.ncbi.nlm.nih.gov/27571378/) |
| **Downstream E2F Targets** | CDK2-T160<br>CCNE1-S384<br>E2F1-S364 | Transcriptional suppression of G1/S machinery | [PMC8794533](https://pmc.ncbi.nlm.nih.gov/articles/PMC8794533/) |
| **Cell Cycle & Replication** | MCM2-S40<br>MCM2-S108<br>FOXM1-T600<br>FOXM1-S715 | Pre-replication complex and mitotic signaling loops | [PMC6115463](https://pmc.ncbi.nlm.nih.gov/articles/PMC6115463/) |
| **Feedback Loops** | CDKN1A-T145<br>CDKN1B-T157<br>CDKN1B-T187 | p21/p27 cell-localization stability shifts | [PMC6343717](https://pmc.ncbi.nlm.nih.gov/articles/PMC6343717/) |

---


```{r palbo kinases, message=FALSE}

direct_targs <- c("RB1-S780", "RB1-S795", "RB1-S807", "RB1-S811")

 plot_features_across_timepoints(experiment,
                                 agent = 'Palbo',
                                  features=direct_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct palbo targs')

```


```{r downstream targets, message=FALSE}
downstream_targs <- c("CDK2-T160", "CCNE1-S384", "E2F1-S364")

 plot_features_across_timepoints(experiment,
                                 agent = 'Palbo',
                                  features=downstream_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Secondary palbo targs')

```


why is this failing? 
```{r other targets, message=FALSE, eval = FALSE}
other_targs <- c("MCM2-S40","MCM2-S41", "MCM2-S108", "FOXM1-T600", "FOXM1-S715")##added S41 since it was cloes


 plot_features_across_timepoints(experiment,
                                 agent = 'Palbo',
                                  features=other_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Other palbo targs')



```

```{r feedback mechanisms, message=FALSE}
feedback_targs <- c("CDKN1A-T145", "CDKN1B-T157", "CDKN1B-T187")


 plot_features_across_timepoints(experiment,
                                 agent = 'Palbo',
                                  features=feedback_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Likely feedback palbo targs')

```


### Afatinib EGFR

Afatinib covalently binds to EGFR (ERBB1), HER2 (ERBB2), and HER4 (ERBB4), globally abolishing receptor trans-autophosphorylation and silencing its downstream docking partners. Direct chemical kinome analysis and resistance tracking data can be found via [PMID: 24435321](https://pubmed.ncbi.nlm.nih.gov/24435321/) and [PMC11031815](https://pmc.ncbi.nlm.nih.gov/articles/PMC11031815/).


| Protein Node | Gene Name-Phosphosite | Functional Significance | Source Evidence |
| :--- | :--- | :--- | :--- |
| **Primary Targets** | EGFR-Y1068<br>EGFR-Y1092<br>EGFR-Y1110<br>EGFR-Y1172<br>EGFR-Y1197<br>ERBB2-Y1221<br>ERBB2-Y1248 | RTK receptor C-terminal autophosphorylation | [PMID: 24435321](https://pubmed.ncbi.nlm.nih.gov/24435321/) |
| **Adaptor & Scaffold Proteins** | GAB1-Y627<br>GAB1-Y689<br>SHC1-Y349<br>SHC1-Y350<br>ERRFI1-Y394 | Disruption of SH2/PTB effector docking | [PMC8182571](https://pmc.ncbi.nlm.nih.gov/articles/PMC8182571/) |
| **Downstream Effectors** | STAT5A-Y694<br>AKT1-T308<br>AKT1-S473<br>MAPK1-T185<br>MAPK1-Y187<br>MAPK3-T202<br>MAPK3-Y204 | Simultaneous PI3K/Akt and MAPK axis crash | [PMC10003401](https://pmc.ncbi.nlm.nih.gov/articles/PMC10003401/) / [PMID: 28590019](https://pubmed.ncbi.nlm.nih.gov/28590019/) |
| **Feedback Loop Sites** | EGFR-T669<br>EGFR-T693 | Extinction of downstream feedback loop signals | [PMC9817674](https://pmc.ncbi.nlm.nih.gov/articles/PMC9817674/) |

---C
```{r afatinib primary, message=FALSE}
primary_targs <- c("EGFR-Y1068", "EGFR-Y1092", "EGFR-Y1110", "EGFR-Y1172", "EGFR-Y1197",
  "ERBB2-Y1221", "ERBB2-Y1248")

 plot_features_across_timepoints(experiment,
                                 agent = 'Afa',
                                  features=primary_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct afatinib targs')
 
 
 plot_features_across_timepoints(experiment,
                                 agent = setdiff(experiment$agent,'DMSO'),
                                  features=primary_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct afatinib targs')
```


```{r afatinib adaptor targets, message=FALSE}

adaptor_targs <- c("GAB1-Y627", "GAB1-Y689", "SHC1-Y349", "SHC1-Y350", "ERRFI1-Y394")


 plot_features_across_timepoints(experiment,
                                 agent = 'Afa',
                                  features=adaptor_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Related afatinib targs')

```

```{r afa downstream, message=FALSE}
downstream_targs <- c("STAT5A-Y694", "AKT1-T308", "AKT1-S473", "MAPK1-T185", "MAPK1-Y187", "MAPK3-T202", "MAPK3-Y204")


 plot_features_across_timepoints(experiment,
                                 agent = 'Afa',
                                  features=downstream_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Downstream afatinib targs')
 
```

```{r afa feedback, message=FALSE}


feedback_targs <- c("EGFR-T669", "EGFR-T693")


 plot_features_across_timepoints(experiment,
                                 agent = 'Afa',
                                  features=feedback_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Feedback afatinib targs')
```

### TNO155 SHP2

TNO155 binds the allosteric tunnel interface of the SHP2 phosphatase (PTPN11), locking it into an auto-inhibited, closed conformation. This prevents SHP2 from dephosphorylating its normal target feedback nodes, keeping upstream adapter links severed and suppressing global RTK-to-RAS/MAPK transmission. Detailed proteomic profiling maps are found in [PMID: 33268444](https://nih.gov) and [PMID: 35147814](https://nih.gov).


| Protein Node | Gene Name-Phosphosite | Functional Significance | Source Evidence |
| :--- | :--- | :--- | :--- |
| **Primary Target Autophospho** | PTPN11-Y542<br>PTPN11-Y580 | Allosteric closure restricts access to these baseline regulatory feedback auto-tyrosine sites | [PMID: 33268444](https://nih.gov) / [PMC9495760](https://pmc.ncbi.nlm.nih.gov/articles/PMC9495760/) |
| **Scaffolds & Adaptors** | GAB1-Y627<br>GAB1-Y659<br>FRS2-Y196<br>FRS2-Y436 | Severe depletion of critical tyrosine-anchors required to bind and load GRB2/SOS complexes to the plasma membrane | [PMID: 35147814](https://nih.gov) |
| **Downstream Cascade** | RAF1-S338<br>MAPK1-T185<br>MAPK1-Y187<br>MAPK3-T202<br>MAPK3-Y204 | Collapse of baseline MAPK kinase pathway loading and absolute terminal ERK activation loop signaling | [PMID: 33268444](https://nih.gov) |

```{r tno targets, eval = FALSE}
direct_targs <- c('PTPN11-Y542','PTPN11-Y580')

 plot_features_across_timepoints(experiment,
                                 agent = 'TNO155',
                                  features=direct_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('Direct TNO155 targs')
 
```


```{r tno scaffold targets, eval = FALSE}
adapter_targs <- c('GAB1-Y627','GAB1-Y659','FRS2-Y196','FRS2-Y43')

 plot_features_across_timepoints(experiment,
                                 agent = 'TNO155',
                                  features=adapter_targs,
                                  timepoints = c('8hr','24hr','48hr'),
                                  ftype = 'Gene.With.Phospho.Site',
                                  fmapping = 'Gene.With.Phospho.Site') +
   ggtitle('adapter TNO155 targs')
 
```

```{r putting this here in case we want to harmonize, eval = FALSE}

# Generate an exhaustive lookup dataframe mapping sites to their individual inhibitor signatures
drug_signatures <- tibble::tribble(
  ~site_id,       ~inhibitor,
  # Mirdametinib Group
  "MAPK3-T202",   "Mirdametinib",   "MAPK3-Y204",   "Mirdametinib",
  "MAPK1-T185",   "Mirdametinib",   "MAPK1-Y187",   "Mirdametinib",
  "RPS6KA1-T359", "Mirdametinib",   "RPS6KA1-S363", "Mirdametinib",
  "RPS6KA1-T573", "Mirdametinib",   "RPS6KA3-T365", "Mirdametinib",
  "RPS6KA3-S369", "Mirdametinib",   "RPS6KA5-T581", "Mirdametinib",
  "RPS6KA5-S360", "Mirdametinib",   "FOSL1-S252",   "Mirdametinib",
  "FOSL1-S265",   "Mirdametinib",   "FOS-S362",     "Mirdametinib",
  "FOS-S374",     "Mirdametinib",   "STMN1-S25",    "Mirdametinib",
  "GAB1-S552",    "Mirdametinib",   "MAP2K1-T292",  "Mirdametinib",
  "MAP2K1-T386",  "Mirdametinib",   "SOS1-S1137",   "Mirdametinib",
  "SOS1-S1167",   "Mirdametinib",   "SOS1-S1178",   "Mirdametinib",
  "SOS1-S1193",   "Mirdametinib",   "EGFR-T669",    "Mirdametinib",
  
  # Palbociclib Group
  "RB1-S780",     "Palbociclib",    "RB1-S795",     "Palbociclib",
  "RB1-S807",     "Palbociclib",    "RB1-S811",     "Palbociclib",
  "CDK2-T160",    "Palbociclib",    "CCNE1-S384",   "Palbociclib",
  "E2F1-S364",    "Palbociclib",    "MCM2-S40",     "Palbociclib",
  "MCM2-S108",    "Palbociclib",    "FOXM1-T600",   "Palbociclib",
  "FOXM1-S715",   "Palbociclib",    "CDKN1A-T145",  "Palbociclib",
  "CDKN1B-T157",  "Palbociclib",    "CDKN1B-T187",  "Palbociclib",
  
  # Afatinib Group
  "EGFR-Y1068",   "Afatinib",       "EGFR-Y1092",   "Afatinib",
  "EGFR-Y1110",   "Afatinib",       "EGFR-Y1172",   "Afatinib",
  "EGFR-Y1197",   "Afatinib",       "ERBB2-Y1221",  "Afatinib",
  "ERBB2-Y1248",  "Afatinib",       "GAB1-Y627",    "Afatinib",
  "GAB1-Y689",    "Afatinib",       "SHC1-Y349",    "Afatinib",
  "SHC1-Y350",    "Afatinib",       "ERRFI1-Y394",  "Afatinib",
  "STAT5A-Y694",  "Afatinib",       "AKT1-T308",    "Afatinib",
  "AKT1-S473",    "Afatinib",       "MAPK1-T185",   "Afatinib",
  "MAPK1-Y187",   "Afatinib",       "MAPK3-T202",   "Afatinib",
  "MAPK3-Y204",   "Afatinib",       "EGFR-T669",    "Afatinib",
  "EGFR-T693",    "Afatinib",
  #TNO155 Group
  "PTPN11-Y542",  "TNO155",         "PTPN11-Y580",  "TNO155",
  "GAB1-Y627",    "TNO155",         "GAB1-Y659",    "TNO155",
  "FRS2-Y196",    "TNO155",         "FRS2-Y436",    "TNO155",
  "RAF1-S338",    "TNO155",         "MAPK1-T185",   "TNO155",
  "MAPK1-Y187",   "TNO155",         "MAPK3-T202",   "TNO155",
  "MAPK3-Y204",   "TNO155",
  #Capmatenib Group
  "MET-Y1234",    "Capmatenib",     "MET-Y1235",    "Capmatenib",
  "MET-Y1349",    "Capmatenib",     "MET-Y1356",    "Capmatenib",
  "GAB1-Y307",    "Capmatenib",     "GAB1-Y627",    "Capmatenib",
  "SHC1-Y317",    "Capmatenib",     "AKT1-T308",    "Capmatenib",
  "AKT1-S473",    "Capmatenib",     "MAPK1-T185",   "Capmatenib",
  "MAPK3-T202",   "Capmatenib",     "GSK3B-S9",     "Capmatenib")
  
```

now what? 
