---
title: "timepoint_analysis"

author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    toc: true
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(gridExtra)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


## Data collectiohn and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting the global proteomics abundance data
synapseFileID = "syn74552028"
fileEntity <- synapser::synGet(synapseFileID)
global_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# getting phospho-proteomics abundance data
synapseFileID = "syn74589783"
fileEntity <- synapser::synGet(synapseFileID)
phospho_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}

# cleaning up the column names for the global proteomics data
names(global_dat) <- names(global_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\rawdata\\\\", "", .) %>% 
  gsub("_16Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)

# cleaning up the column names for the phospho-proteomics data
names(phospho_dat) <- names(phospho_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\PhosphoRawData\\\\", "", .) %>% 
  gsub("_20Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)

```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)
```

Finally we generate SummarizedExperiment objects for the global proteomics data
as well as the phospho-proteomics data.

```{r summarized-experiment-creation}
experiment_global <- prep_prot_data(proteomics_data = global_dat,
                                    proteomics_meta = meta_data_global,
                                    missingness_cutoff = 0,
                                    type = 'global')

experiment_phospho <- prep_prot_data(proteomics_data = phospho_dat,
                                     proteomics_meta = meta_data_phospho,
                                     missingness_cutoff = 0,
                                     type = 'phospho')
```



```{r outlier-investigation}

plot <- plot_per_plate_na_overview(experiment_global, "global")
plot

ggsave("../output_figures/outlier_by_plate_global.png",
       plot = plot,
       height = 5,
       width = 5,
       )
```

```{r}
plot <- plot_per_plate_na_overview(experiment_phospho, "phospho")
plot
ggsave("../output_figures/outlier_by_plate_phospho.png",
       plot = plot,
       height = 5,
       width = 5,
       )



```

```{r eval=TRUE}
missingness_cutoff = 0.5
imputation_method = "SampMin"

experiment_global <- prep_prot_data(proteomics_data = global_dat,
                                    proteomics_meta = meta_data_global,
                                    missingness_cutoff = missingness_cutoff,
                                    type = 'global')

experiment_phospho <- prep_prot_data(proteomics_data = phospho_dat,
                                     proteomics_meta = meta_data_phospho,
                                     missingness_cutoff = missingness_cutoff,
                                     type = 'phospho')

experiment_global %<>% 
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method)

experiment_phospho %<>% 
  log_transform(., assay = "proteomics") %>%
  impute_proteomics(., assay = "proteomics", imputation_method)
```


```{r eval=FALSE}
agents <- unique(meta_data$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls]


plates <- unique(colData(experiment_global[, experiment_global$agent ==  agent])$plate)
plates <- append(plates, 10)
ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
sel <- c(ctrl, agent)

experiment_subset <- experiment[, (experiment$plate %in% plates) & experiment$agent %in% sel]

principal_components <- prcomp(t(assay(experiment_subset))) %$%
  cbind(x, colData(experiment_subset))

for(agent in agents){
 plot <- plot_pca(experiment_global, agent, "global", missingness = missingness_cutoff, imputation = imputation_method)
 fname <- paste0("../output_figures/pca_global_by_drug_", agent, ".png")
 ggsave(fname, plot = plot, height = 5, width = 6)
}

for(agent in agents){
  plot <- plot_pca(experiment_phospho, agent, "phospho", missingness = missingness_cutoff, imputation = imputation_method)
 fname <- paste0("../output_figures/pca_phospho_by_drug_", agent, ".png")
 ggsave(fname, plot = plot, height = 5, width = 6)
}
```


```{r eval=FALSE}
for(plate in 1:9){
  plot <- plot_pca_by_plate(experiment_global, plate, "global", missingness = missingness_cutoff, imputation = imputation_method)
  print(plot)
  fname <- paste0("../output_figures/pca_global_by_plate_", plate, ".png")
  ggsave(fname, plot = plot, height = 5, width = 6)  
}

for(plate in 1:9){
  plot <- plot_pca_by_plate(experiment_phospho, plate, "phospho", missingness = missingness_cutoff, imputation = imputation_method)
  print(plot)
  fname <- paste0("../output_figures/pca_phospho_by_plate_", plate, ".png")
  ggsave(fname, plot = plot, height = 5, width = 6)  
}
```


