---
title: "timepoint_analysis"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_notebook:
    toc: true
  html_document:
    toc: true
    df_print: paged
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(gridExtra)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


## Data collectiohn and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting the global proteomics abundance data
synapseFileID = "syn74552028"
fileEntity <- synapser::synGet(synapseFileID)
global_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# getting phospho-proteomics abundance data
synapseFileID = "syn74589783"
fileEntity <- synapser::synGet(synapseFileID)
phospho_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data_raw <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}

# cleaning up the column names for the global proteomics data
names(global_dat) <- names(global_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\rawdata\\\\", "", .) %>% 
  gsub("_16Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)

# cleaning up the column names for the phospho-proteomics data
names(phospho_dat) <- names(phospho_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\PhosphoRawData\\\\", "", .) %>% 
  gsub("_20Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)


prot_data <- list(
  "global" = global_dat,
  "phospho" = phospho_dat
)

```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data_raw %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data_raw %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data_raw %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)

meta_data <- list(
  "global" = meta_data_global,
  "phospho" = meta_data_phospho
)
```

As final perpartion we define a few bespoke helper functions that are only
applicable to this dataset:

```{r bespoke-helper-functions}

prep_experiment_data <- function(
    prot_data,
    meta_data,
    missingness_cutoff=0,
    imputation_method=c("none", 'SampMin', 'means', 'median')
    ){
  
  imputation_method_sel <- match.arg(imputation_method)
  
  if(!is.na(missingness_cutoff) & !(missingness_cutoff >=0 | missingness_cutoff <= 1)){
    stop(
      base::paste(
        "missingness_cutoff needs to be '0 <= missingness_cutoff <=1'",
        "or 'NULL'. Supplied value:", missingness_cutoff),
      call. = FALSE
    )
  }
  
  experiments <- list()
  for(data_type in c('global', 'phospho')){
    
    experiment <- prep_prot_data(proteomics_data = prot_data[[data_type]],
                                 proteomics_meta = meta_data[[data_type]],
                                 missingness_cutoff = missingness_cutoff,
                                 type = data_type)
    
    if(!(imputation_method_sel == 'none')){
      experiment %<>% 
        log_transform(., assay = "proteomics") %>%
        impute_proteomics(., assay = "proteomics", imputation_method_sel)
    }
    experiments[[data_type]] <- experiment
  }
  
  experiments
}


```


## Data quality control

As a first step we are doing some quality control & assessment for the global
and phospho-proteomics data.
To that end we are generating PCA plots for each of the plates that have been
prepared and run through the Mass-spectrometry pipeline.
Each plate contains 60 wells, each containing spheroids treated with a chemical
agent where four wells each were pooled into one sample.

In general each plate contains:

- 4 chemical agents (3 samples each)
- 1 control (DMSO - 3 samples each)
- at one timepoint.

We generate `SummarizedExperiment` objects for the global & phospho-proteomics
data with *NO* missingness cut-off for feature removal and *Sample Minimum*
*imputation*. The imputation step is necessary such that principal components
can be extracted.

```{r prep-data-for-PCA, eval=TRUE}

missingness_cutoff = 0.0
imputation_method = "SampMin"

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  imputation_method
)

```

Next we generate PCA plots

```{r gen-per-plate-pca-plots, eval=TRUE}
plots <- list()

for(prot_type in c('global', 'phospho')){
  for(plate in 1:9){
    experiment <- experiments[[prot_type]]
    plot <- plot_pca_by_plate(experiment,
                              plate,
                              prot_type,
                              missingness = missingness_cutoff,
                              imputation = imputation_method
                              )
    plots[[prot_type]][[plate]] <- plot
    print(plot)
  }
}
```

And save them if desired (set `eval=TRUE`)
```{r save-per-plate-pca-plots, eval=FALSE}
for(prot_type in c('global', 'phospho')){
  for(plate in 1:9){
    fname <- paste0("../output_figures/pca_", prot_type,
                    "_by_plate_", plate, ".png")
    ggsave(fname, plot = plots[[prot_type]][[plate]], height = 5, width = 6)
  }
}
```

### Outlier investigation

```{r outlier-investigation}
missingness_cutoff = 0
imputation_method = 'none'

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  imputation_method
)


plots <- list()
for(prot_type in c('global', 'phospho')){
  experiment <- experiments[[prot_type]]
  plots[[prot_type]] <- plot_per_plate_na_overview(experiment, prot_type)
  print(plots[[prot_type]])
}
```

```{r save-outlier-plots, eval=FALSE}
for(prot_type in c('global', 'phospho')){
  fname <- paste0("../output_figures/outlier_by_plate_", prot_type, ".png")
  ggsave(fname,
         plot = plots[[prot_type]],
         height = 5,
         width = 5,
         )
}
```


## Differential expression analysis

Next we perform a differential expression analysis. To this end we will utilize
some functions sourced from `helper_funcs.R`, specifically:

- `plot_diff_ex` which internally calls a number of functions and calculates the
  differential expression based on limma. The function then returns a plot which
  we save into a named list where plots are accessible by agent and timepoint.
- `plot_diff_ex_grid` which internally uses `ggarrange` from the `ggpubr`
  package to generate plot grids, combining the previously generated volcano
  plots of the differential expression for the three timepoints per agent.

The result is one plot per agent, showing the differential expression for each
of the timepoints.

First we regenerate the `SummarizedExepirment` objects for the global and 
phospho-proteomics data, since we want to perform some missingness removal and
need to impute the missing data, such that `limma` can perform the differential
expression analysis.

```{r data-prep-for-diff-ex, eval=TRUE}
missingness_cutoff = 0.5
imputation_method = "SampMin"

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  imputation_method
)
```

Next we generate all the volcano plots for the differentially expressed proteins
and phosphosites.

```{r differential-expression-volcano-plots, warning=FALSE}

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")


for(prot_type in c('global', 'phospho')){
  
  experiment = experiments[[prot_type]]
  plots <- list()
  plots_agents <- list()
  
  for(agent in agents){
    for(timepoint in timepoints){
      plots[[agent]][[timepoint]] <- plot_diff_ex(experiment, agent, timepoint)
    }
    
    plot_title <- paste('Spheroid', prot_type ,'proteomics:', agent,
                        "vs. Control")
    plots_agents[[agent]] <- plot_diff_ex_grid(plots[[agent]], plot_title)
    print(plots_agents[[agent]])
  }
  
} 
```

Finally we save the plots (if desired - change to `eval=TRUE`)
```{r save-diff-ex-volcano-plots, eval=FALSE}
for(prot_type in c('global', 'phospho')){
  for(agent in agents){
    filename <- paste0("../output_figures/differential_expression_", prot_type,
                       "_", agent, ".png")
    ggsave(filename, plots_agents[[agent]], height = 5, width = 15)
  }
}
```




### Barplots of differentially expressed proteins
TODO: generate bar plots visualizing the number of up & down regulated proteins
and phospho-sites listed in the differential expression volcano plots
```{r eval=FALSE}

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")

prot_type <- 'global'
experiment <- experiments[[prot_type]]

agent <- 'Mirda'

ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
group_1 = paste0(timepoint, "_", ctrl)
group_2 = paste0(timepoint, "_", agent)

plate_sel <- unique(experiment[, experiment$group == group_2]$plate)
experiment_subset <- experiment[, experiment$plate == plate_sel]

align_by_group <- ifelse(test = (prot_type == 'global'), yes = "Protein.Group",
                         no = "Protein.With.Phospho.Site")
diff_ex <- calc_diff_ex(experiment_subset, group_1, group_2,
                        align_by=align_by_group)

cpalette = c('up' = '#74add1', 'down' = '#f46d43')

if(prot_type == "global"){
  dfPlot <- subset(
      as.tibble(rowData(diff_ex)),
      select = c('Genes', 'logFC', 'adj.P.Val'
    )) %>%
    mutate(expression_change = case_when(
      (logFC >= 1 & adj.P.Val < 0.05) ~ 'up',
      (logFC <= -1 & adj.P.Val < 0.05) ~ 'down',
      .default = 'not significant',
    )) %>%
    mutate(labels = case_when(
      expression_change %in% c('up', 'down') ~ Genes,
      .default = NA
    )) %>%
    subset(., expression_change %in% c('up', 'down'))
}

plot <- (
  ggplot(
    data = dfPlot,
    aes(timepoint)
  )
  + geom_bar(aes(fill=expression_change), position = 'dodge2')
  + scale_fill_manual(values = cpalette)
  + theme_minimal()
  + labs(
    title = 'Spheroid phospho proteomics; DMSO vs Mirdametinib', 
    subtitle = paste("Counts of significant differentially expressed proteins", 
                     "( pVal < 0.05; -1  >= logFC >= 1 )"),
    fill = 'expression'
  )
)

print(plot)

```
