---
title: "timepoint_analysis"
author:
- name: Yannick Mahlich
  affiliation: Pacific Northwest National Laboratory
- name: Sara Gosline
  affiliation: Pacific Northwest National Laboratory
date: "`r Sys.Date()`"
output:
 # pdf_document:
#    toc: false
  html_notebook:
    toc: true
  html_document:
    toc: true
    df_print: paged
---

## Library & Data import

The notebook itself relies on a number of R libraries to import and analyze the
proteomics data. Library imports are not shown here, but include packages like:

- synapser
- readxl
- limma

```{r imports, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
library(synapser)
library(readxl)
library(tibble)
library(dplyr)
library(limma)
library(tidyr)
library(matrixStats)
library(ggplot2)
library(leapR)
library(SummarizedExperiment)
library(ggrepel)
library(magrittr)
library(gridExtra)
library(ggalluvial)
library(sva)
```

```{r sourcing-cross-notebook-functions, results='hide'}
source('../src/helper_funcs.R')
```


## Data collection and preparation
We first will collect the global and phospho-proteomics abundance data as well
as the metadata from synapse. The data files are located at:
https://www.synapse.org/Synapse:syn70987742

```{r import-data, results='hide'}
# opening synapse connection
tmp <- synapser::synLogin(silent = TRUE)

# getting the global proteomics abundance data
synapseFileID = "syn74552028"
fileEntity <- synapser::synGet(synapseFileID)
global_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# getting phospho-proteomics abundance data
synapseFileID = "syn74589783"
fileEntity <- synapser::synGet(synapseFileID)
phospho_dat <- readr::read_tsv(
  fileEntity$path,
  na = c("0", "", NA), # the file stored on synapse has '0's instead of NA or ''
  )

# metadata ingest
synapseFileID <- "syn74755705"
fileEntity <- synapser::synGet(synapseFileID)
meta_data_raw <- readxl::read_xlsx(fileEntity$path, sheet = 2)

# closing synpase conncetion
tmp <- synapser::synLogout()
```
Next we do some cleanup of the column names in the global proetomics and
phosphosite abundance data files to align the column names with the sample ids
in the metadata file. This step is necessary such that we can successfully
generate a `SummarizedExperiment` object further down the line.

```{r sample-id-cleanup, eval=TRUE}

# cleaning up the column names for the global proteomics data
names(global_dat) <- names(global_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\rawdata\\\\", "", .) %>% 
  gsub("_16Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)

# cleaning up the column names for the phospho-proteomics data
names(phospho_dat) <- names(phospho_dat) %>% 
  gsub("D:\\\\Data_D\\\\LeDay\\\\spheroid_MN02\\\\PhosphoRawData\\\\", "", .) %>% 
  gsub("_20Apr26_Monty_ES906_2014_50SPD[x]*.raw", "", .) %>% 
  gsub("^[0-9]*_", "", .)


prot_data <- list(
  "global" = global_dat,
  "phospho" = phospho_dat
)

```

Next we clean up and prepare the metadata for `SummarizedExperiment` object
generation. The main 2 steps here are:

- cleaning up the column names
- adding a new coulmns for replicates (to distinguish individual samples of the 
  same type) and combining treatment agent, timepoint and replicate

```{r meta-data-prep}
new_col_names <- c(sample = 'Sample ID \r\n(abbrev)',
                   global_sample_id = "Proteomics Sample ID",
                   phospho_sample_id = "Phosphoproteomics Sample ID",
                   agent = "Drug",
                   timepoint = "Time",
                   plate = "Plate"
                   )

meta_data_raw %<>%
  rowwise() %>%
  mutate(Drug = gsub(" ", "", Drug), Time = paste0(Time, "hr")) %>%
  mutate(group = paste(Time, Drug, sep = "_")) %>%
  group_by(group) %>%
  mutate(replicate = paste0('r', 1:n())) %>%
  mutate(condition = paste(Time, Drug, replicate, sep='_')) %>%
  dplyr::rename(., all_of(new_col_names)) %>%
  ungroup()

meta_data_global <- meta_data_raw %>%
  select(., global_sample_id, condition, plate) %>%
  dplyr::rename(., sample = global_sample_id)

meta_data_phospho <- meta_data_raw %>%
  select(., phospho_sample_id, condition, plate) %>%
  dplyr::rename(., sample = phospho_sample_id)

meta_data <- list(
  "global" = meta_data_global,
  "phospho" = meta_data_phospho
)
```

As final perpartion we define a few bespoke helper functions that are only
applicable to this dataset:

```{r bespoke-helper-functions}

prep_experiment_data <- function(
    prot_data,
    meta_data,
    missingness_cutoff=0,
    samp_cutoff = list(global = 0, phospho = 0),
    imputation_method=c("none", 'SampMin', 'means', 'median'), 
    batch_correct = FALSE
    ){
  
  imputation_method_sel <- match.arg(imputation_method)
  
  if(!is.na(missingness_cutoff) & !(missingness_cutoff >=0 | missingness_cutoff <= 1)){
    stop(
      base::paste(
        "missingness_cutoff needs to be '0 <= missingness_cutoff <=1'",
        "or 'NULL'. Supplied value:", missingness_cutoff),
      call. = FALSE
    )
  }
  
  experiments <- list()
  for(data_type in c('global', 'phospho')){
      
    experiment <- prep_prot_data(proteomics_data = prot_data[[data_type]],
                                 proteomics_meta = meta_data[[data_type]],
                                 missingness_cutoff = missingness_cutoff,
                                 samp_cutoff = samp_cutoff[[data_type]],
                                 type = data_type, 
                                 imputation_method=imputation_method_sel)
    
    if(imputation_method_sel == "none"){
      experiment %<>%
        log_transform(., assay = "proteomics")
    }
    
    if(!(imputation_method_sel == 'none')){
      experiment %<>% 
        impute_proteomics(., assay = "proteomics", imputation_method_sel) %>%
        log_transform(., assay = "proteomics")
      
    }
    
    if(batch_correct && imputation_method_sel != 'none'){
      #first we have to remove zero hour beacuse it is confounded by batch
      #experiment <- experiment[,-which(colData(experiment)$timepoint=='0hr')]
      prot_matrix <- assay(experiment)
      batch_var = colData(experiment)$plate
      cond_var = colData(experiment)$agent
      mod_matrix <- model.matrix(~ as.factor(agent), data = colData(experiment))
      corrected_prot <- sva::ComBat(
             dat = prot_matrix,
             batch = batch_var,
             mod = mod_matrix,
             par.prior = FALSE
              )
    assay(experiment,withDimnames = FALSE) <- corrected_prot  

      
    }
    experiments[[data_type]] <- experiment
  }
  
  experiments
}


```


## Data quality control

As a first step we are doing some quality control & assessment for the global
and phospho-proteomics data.
To that end we are generating PCA plots for each of the plates that have been
prepared and run through the Mass-spectrometry pipeline.
Each plate contains 60 wells, each containing spheroids treated with a chemical
agent where four wells each were pooled into one sample.

In general each plate contains:

- 4 chemical agents (3 samples each)
- 1 control (DMSO - 3 samples each)
- at one timepoint.

We generate `SummarizedExperiment` objects for the global & phospho-proteomics
data with *NO* missingness cut-off for feature removal and *Sample Minimum*
*imputation*. The imputation step is necessary such that principal components
can be extracted.

```{r prep-data-for-PCA, eval=TRUE}

missingness_cutoff = 0
imputation_method = "SampMin"

samp_cutoff = list(global = 0, phospho = 0)

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  samp_cutoff,
  imputation_method,
  # batch_correct = TRUE
)


```

Next we generate PCA plots

```{r gen-per-plate-pca-plots, eval=TRUE}
plots <- list()

for(prot_type in c('global', 'phospho')){
  for(plate in 1:9){
    experiment <- experiments[[prot_type]]
    plot <- plot_pca_by_plate(experiment,
                              plate,
                              prot_type,
                              missingness = missingness_cutoff,
                              imputation = imputation_method
                              )
    plots[[prot_type]][[plate]] <- plot
    print(plot)
  }
}
```

And save them if desired (set `eval=TRUE`)
```{r save-per-plate-pca-plots, eval=FALSE}
for(prot_type in c('global', 'phospho')){
  for(plate in 1:9){
    fname <- paste0("../output_figures/pca_", prot_type,
                    "_by_plate_", plate, ".png")
    ggsave(fname, plot = plots[[prot_type]][[plate]], height = 5, width = 6)
  }
}
```

### Outlier investigation

```{r outlier-investigation}
missingness_cutoff = 0
imputation_method = 'none'

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  samp_cutoff,
  imputation_method
)


plots <- list()
for(prot_type in c('global', 'phospho')){
  experiment <- experiments[[prot_type]]
  plots[[prot_type]] <- plot_per_plate_na_overview(experiment, prot_type)
  print(plots[[prot_type]])
}
```

```{r save-outlier-plots, eval=FALSE}
for(prot_type in c('global', 'phospho')){
  fname <- paste0("../output_figures/outlier_by_plate_", prot_type, ".png")
  ggsave(fname,
         plot = plots[[prot_type]],
         height = 5,
         width = 5,
         )
}
```




## Differential expression analysis

Next we perform a differential expression analysis. To this end we will utilize
some functions sourced from `helper_funcs.R`, specifically:

- `plot_diff_ex` which internally calls a number of functions and calculates the
  differential expression based on limma. The function then returns a plot which
  we save into a named list where plots are accessible by agent and timepoint.
- `plot_diff_ex_grid` which internally uses `ggarrange` from the `ggpubr`
  package to generate plot grids, combining the previously generated volcano
  plots of the differential expression for the three timepoints per agent.

The result is one plot per agent, showing the differential expression for each
of the timepoints.

[SARA SKIPPED THIS] First we regenerate the `SummarizedExepirment` objects for the global and 
phospho-proteomics data, since we want to perform some missingness removal and
need to impute the missing data, such that `limma` can perform the differential
expression analysis.

```{r data-prep-for-diff-ex, eval=FALSE}
missingness_cutoff = 0.5
samp_cutoff= list(global = .9, phospho = .3)
imputation_method = "means"

experiments <- prep_experiment_data(
  prot_data,
  meta_data,
  missingness_cutoff,
  samp_cutoff,
  imputation_method
)
```

Next we generate all the volcano plots for the differentially expressed proteins
and phosphosites.

```{r differential-expression-volcano-plots, warning=FALSE}

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")
log_fc = .5
pval = 0.1
missingness_cutoff = 0
samp_cutoff = list(global = 0, phospho = 0)
imputation_method = 'SampMin'

for(prot_type in c('global', 'phospho')){
  
  experiment = experiments[[prot_type]]
  plots <- list()
  plots_agents <- list()
  
  for(agent in agents){
    for(timepoint in timepoints){
         experiment_subset <- exp_subset(experiment, agent, timepoint) |>
            filter_for_feature_missingness(missingness_cutoff) |>
            filter_for_sample_cutoff(samp_cutoff[[prot_type]]) |> ##do we want to do this first? 
            impute_proteomics(imputation_method = imputation_method) |>
            log_transform()
      plots[[agent]][[timepoint]] <- plot_diff_ex(
        experiment_subset,
        agent,
        timepoint,
        log_fc,
        pval
        )
    }
    
    plot_title <- paste('Spheroid', prot_type ,'proteomics:', agent,
                        "vs. Control")
    plots_agents[[agent]] <- plot_diff_ex_grid(plots[[agent]], plot_title)
    print(plots_agents[[agent]])
  }
  
} 
```

Finally we save the plots (if desired - change to `eval=TRUE`)
```{r save-diff-ex-volcano-plots, eval=FALSE}
for(prot_type in c('global', 'phospho')){
  for(agent in agents){
    filename <- paste0("../output_figures/differential_expression_", prot_type,
                       "_", agent, ".png")
    ggsave(filename, plots_agents[[agent]], height = 5, width = 15)
  }
}
```


### Barplots of differentially expressed proteins

```{r data-prep-for-diff-ex-comparison}

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")
log_fc = 0.5
pval = 0.1

df_diff_ex <- tibble()
all_diffex_stats <- tibble()
for(prot_type in c('global', 'phospho')) {
  experiment <- experiments[[prot_type]]
    
  for(agent in agents){
    for(timepoint in timepoints){
      ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
      group_1 = paste0(timepoint, "_", ctrl)
      group_2 = paste0(timepoint, "_", agent)
      plate_sel <- unique(experiment[, experiment$group == group_2]$plate)
      experiment_subset <- exp_subset(experiment, agent, timepoint) |>#)experiment[, experiment$plate == plate_sel]|>
                       filter_for_sample_cutoff(samp_cutoff[[prot_type]]) |>
            filter_for_feature_missingness(missingness_cutoff) |>
            impute_proteomics(imputation_method = imputation_method) |>
            log_transform()
  
      align_by_group <- ifelse(test = (prot_type == 'global'), yes = "Protein.Group",
                           no = "Protein.With.Phospho.Site")
      diff_ex <- calc_diff_ex(experiment_subset, group_1, group_2,
                          align_by=align_by_group)
      
      if(prot_type == "global"){
        de_tab <- as.tibble(rowData(diff_ex)) |>
          dplyr::select(Feature = 'Genes','logFC','adj.P.Val') 
        
        diff_ex_sub <- diff_ex %>%
          extract_diff_ex_features(., "Genes", log_fc, pval) %>%
          mutate(omics_type = "global proteomics")
      } else if(prot_type == "phospho"){
        de_tab <- as.tibble(rowData(diff_ex)) |>
          dplyr::select(Feature = 'Gene.With.Phospho.Site','logFC','adj.P.Val') 
        
        diff_ex_sub <- diff_ex %>%
          extract_diff_ex_features(., "Gene.With.Phospho.Site", log_fc, pval) %>%
          mutate(omics_type = "phospho-proteomics")
      }
      
      all_diffex_stats <- de_tab|>
        mutate(omics_type = prot_type,
               timepoint = timepoint, 
               agent = agent) |>
        rbind(all_diffex_stats)
      
      diff_ex_sub %<>%
        subset(., expression_change %in% c('up', 'down')) %>%
        mutate(timepoint = timepoint) %>%
        mutate(agent = agent) %>%
        dplyr::count(expression_change, timepoint, agent, omics_type) %>%
        mutate(n = if_else(expression_change == "down", n * -1, n)) %>%
        mutate(timepoint = factor(timepoint, levels = c('8hr', '24hr', '48hr'), labels = c('8h', '24h', '48h')))
      
      df_diff_ex <- bind_rows(df_diff_ex, diff_ex_sub)
  
    }
  }
  write.csv(all_diffex_stats,file='allDiffexFeatures.csv')
}

```

```{r plot-diffex-counts-agent-by-omics-type}
cpalette = c('up' = '#74add1', 'down' = '#f46d43')
plot_title <- paste(
  "Differentially expressed proteins/phospho-sites in MN2",
  "after treatment vs. control"
  )

plot_by_agent <- (
  ggplot(df_diff_ex, aes(x=timepoint, y=n))
  + geom_bar(position="stack", stat="identity", mapping = aes(fill=expression_change)) 
  + scale_fill_manual(values = cpalette) 
  + ggrepel::geom_text_repel(aes(label = n))
  + labs(
      title = plot_title,
      x = "time after treatment",
      y = "# of up/down regulated features",
      fill = 'expression'
    )
  + theme_minimal()
  + theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    )
  + facet_grid(omics_type ~ agent)
)
print(plot_by_agent)
```

```{r plot-diffex-counts-timepoint-by-omics-type}
cpalette = c('up' = '#74add1', 'down' = '#f46d43')
plot_title <- paste(
  "Differentially expressed proteins/phospho-sites in MN2",
  "after treatment vs. control"
  )

plot_by_timepoint <- (
  ggplot(df_diff_ex, aes(x=agent,y=n))
  + geom_bar(aes(fill=expression_change), position="stack", stat="identity")
  + scale_fill_manual(values = cpalette)
  + ggrepel::geom_text_repel(aes(label= n))
  + labs(
      title = plot_title,
      x = "treatment chemical agent",
      y = "# of up/down regulated features",
      fill = 'expression'
    )
  + theme_minimal()
  + theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    )
  + facet_grid(omics_type ~ timepoint)
)
print(plot_by_timepoint)
```

```{r save-diffex-counts-comp, eval=FALSE}
filename <- "../output_figures/differential_expression_counts_agent_by_omic_type.png"
ggsave(filename, plot_by_agent, height = 5, width = 10)

filename <- "../output_figures/differential_expression_counts_timepoint_by_omic_type.png"
ggsave(filename, plot_by_timepoint, height = 5, width = 10)
```

### Alluvial plots of differentially expressed proteins across timepoints

```{r}
agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")
log_fc = 0.5
pval = 0.1


df_diff_ex <- tibble()
for(prot_type in c('global', 'phospho')) {
  experiment <- experiments[[prot_type]]
    
  for(agent in agents){
    for(timepoint in timepoints){
      ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
      group_1 = paste0(timepoint, "_", ctrl)
      group_2 = paste0(timepoint, "_", agent)
      plate_sel <- unique(experiment[, experiment$group == group_2]$plate)
      experiment_subset <- exp_subset(experiment, agent, timepoint) |>
            filter_for_feature_missingness(missingness_cutoff) |>
            filter_for_sample_cutoff(samp_cutoff[[prot_type]]) |> ##do we want to do this first? 
            impute_proteomics(imputation_method = imputation_method) |>
            log_transform()
  
      align_by_group <- ifelse(test = (prot_type == 'global'), yes = "Protein.Group",
                           no = "Protein.With.Phospho.Site")
      diff_ex <- calc_diff_ex(experiment_subset, group_1, group_2,
                          align_by=align_by_group)
      
      if(prot_type == "global"){
        diff_ex_sub <- diff_ex %>%
          extract_diff_ex_features(., "Genes", log_fc, pval) %>%
          mutate(omics_type = "global proteomics")
      } else if(prot_type == "phospho"){
        diff_ex_sub <- diff_ex %>%
          extract_diff_ex_features(., "Gene.With.Phospho.Site", log_fc, pval) %>%
          mutate(omics_type = "phospho-proteomics")
      }
      
      diff_ex_sub %<>%
        subset(., expression_change %in% c('up', 'down')) %>%
        mutate(timepoint = timepoint) %>%
        mutate(agent = agent) %>%
        mutate(timepoint = factor(timepoint, levels = c('8hr', '24hr', '48hr'), labels = c('8h', '24h', '48h')))
      
      df_diff_ex <- bind_rows(df_diff_ex, diff_ex_sub)
      
    }
  }
}

```


```{r}
cpalette = c(
  'up' = '#74add1',
  'down' = '#f46d43',
  'not significant' = 'lightgrey'
  )

for(agent_ in agents){
  df_diff_ex_subset <- df_diff_ex |>
    subset(agent == agent_) |>
    group_by(omics_type) |>
    complete(
      feature, timepoint,
      fill = list(expression_change = "not significant", agent = agent_)
      ) |>
    mutate(labels = feature) |>
    mutate(
      expression_change = factor(
        expression_change,
        levels = c('up', 'down', 'not significant')
        )
      )

  plot <- (
    ggplot(
      df_diff_ex_subset,
      aes(
        x = timepoint,
        stratum = expression_change,
        alluvium = feature,
        fill = expression_change,
        label = expression_change
        )
      )
    + scale_fill_manual(values = cpalette)
    + geom_flow(
        stat = "alluvium",
        # color = "darkgray",
        aes.flow = "forward",
        lode.guidance = "forward",
        )
    + geom_stratum()
    + theme_minimal()
    + labs(
      title = paste(agent_),
      fill = 'expression',
    )
    + theme(
        legend.position = "bottom",
        )
    + facet_wrap(~ omics_type, ncol = 2, scales = "free")
    )
  print(plot)
    
}

```
### Pathway enrichment

```{r load-geneset, results='hide'}
# data("msigdb")
data("ncipid")
data("krbpaths")
```

#### Data prep
Below we (re-)generate the differential expression objects for all combinations of `protein_type`, `agent` and `timepoint`.
The resulting "diffex" objects - really just `SummarizedExeperimet` objects with added information about differential expression - are stored into a named list (with 3 levels) for easier access.
```{r leapR-pathway-enrichment, message=FALSE}

agents <- unique(meta_data_raw$agent)
controls <- c("Water", "DMSO")
agents <- agents[!agents %in% controls & !agents == 'Ola']
timepoints = c("8hr", "24hr", "48hr")


all_diff_ex <- list()
for(prot_type in c('global', 'phospho')) {
  experiment <- experiments[[prot_type]]
    
  for(agent in agents){
    for(timepoint in timepoints){
      ctrl <- ifelse(test = (agent == 'Trab'), yes = "Water", no = "DMSO")
      group_1 = paste0(timepoint, "_", ctrl)
      group_2 = paste0(timepoint, "_", agent)
      plate_sel <- unique(experiment[, experiment$group == group_2]$plate)
      experiment_subset <- exp_subset(experiment, agent, timepoint) |>#)experiment[, experiment$plate == plate_sel]|>
                       filter_for_sample_cutoff(samp_cutoff[[prot_type]]) |>
            filter_for_feature_missingness(missingness_cutoff) |>
            impute_proteomics(imputation_method = imputation_method) |>
            log_transform()
  
      align_by_group <- ifelse(test = (prot_type == 'global'), yes = "Protein.Group",
                           no = "Protein.With.Phospho.Site")
      diff_ex <- calc_diff_ex(experiment_subset, group_1, group_2,
                          align_by=align_by_group)
      
      all_diff_ex[[prot_type]][[agent]][[timepoint]] <- diff_ex
    }
  }
}
```

### Pathway enrichments

#### Global

First we are using the KEGG pathway information to detect enriched KEGG pathways in our datasets.

```{r}
pval <- 0.1
prot_type <- "global"
id_column <- "Genes"
geneset <- krbpaths

for(agent in agents){
  for(timepoint in timepoints){
    # doing a try catch because not for all diffEx sets we have results for the
    # defined pval threshold
    tryCatch({
      enriched_pathways <- leapR(
      geneset = geneset,
      eset = all_diff_ex[[prot_type]][[agent]][[timepoint]],
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "adj.P.Val",
      threshold = pval,
      greaterthan = FALSE
      )
    plot <- plot_leapr_bar(
      enriched_pathways,
      paste(agent, 'at', timepoint)
    )
    print(plot)
    }, error = function(msg){
      message(
        paste0(
          "pval of '", pval,
          "' did not result in differentially expressed proteins for agent '", agent,
          "' at timepoint '", timepoint, "'"
        )
      )
      return(NA)
      }
    )
  }
}

```

Then we repeat the same thing for the pathways in the NCI pathway interaction database:

```{r}
pval <- 0.1
prot_type <- "global"
id_column <- "Genes"
geneset <- ncipid

for(agent in agents){
  for(timepoint in timepoints){
    tryCatch({
      # doing a try catch because not for all diffEx sets we have results for the
      # defined pval threshold
      enriched_pathways <- leapR(
      geneset = geneset,
      eset = all_diff_ex[[prot_type]][[agent]][[timepoint]],
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "adj.P.Val",
      threshold = pval,
      greaterthan = FALSE
      )
    plot <- plot_leapr_bar(
      enriched_pathways,
      paste(agent, 'at', timepoint)
    )
    print(plot)
    }, error = function(msg){
      message(
        paste0(
          "pval of '", pval,
          "' did not result in differentially expressed proteins for agent '", agent,
          "' at timepoint '", timepoint, "'"
        )
      )
      return(NA)
      }
    )
  }
}

```

#### Phospho

```{r load-leapR-gmt}
data("kinasesubstrates")
```

```{r}
pval <- 0.1
prot_type <- "phospho"
id_column <- "Gene.With.Phospho.Site"
geneset <- kinasesubstrates

for(agent in agents){
  for(timepoint in timepoints){
    tryCatch({
      # doing a try catch because not for all diffEx sets we have results for the
      # defined pval threshold
      enriched_pathways <- leapR(
      geneset = geneset,
      eset = all_diff_ex[[prot_type]][[agent]][[timepoint]],
      enrichment_method = "enrichment_in_sets",
      assay_name = 'proteomics',
      id_column = id_column,
      primary_column = "adj.P.Val",
      threshold = pval,
      greaterthan = FALSE
      )
    plot <- plot_leapr_bar(
      enriched_pathways,
      paste(agent, 'at', timepoint)
    )
    print(plot)
    }, error = function(msg){
      message(
        paste0(
          "pval of '", pval,
          "' did not result in differentially expressed proteins for agent '", agent,
          "' at timepoint '", timepoint, "'"
        )
      )
      return(NA)
      }
    )
  }
}

```
