---
title: "DEseq2 data with LRT"
author: "Yang Lyu"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    latex_engine: pdflatex
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    #latex_engine: xelatex
      
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =F,message = F,warning = F,include=F)
```

```{r}
rm(list=ls())
```


```{r}
# Function needed
matrix.please<-function(x) {
  m<-as.matrix(x[,-1])
  rownames(m)<-x[,1]
  m
}
```


```{r}
library(biomaRt) 
library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
```

```{r}
Path <- gsub("/R(/Rcode)?$", "", getwd())

Rawdata_path=file.path(Path,"Rawdata")
dir.create(Rawdata_path)
```


```{r}
# STAR-Salmon
Reference_path="/Volumes/lyu.yang/MAC_1/R/Public_database/Reference"
STAR_path="/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_NF_RNAseq/Nextflow3--star_index/star_salmon"
dir("/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_NF_RNAseq")
Rawcount_path=file.path(Rawdata_path,"salmon.merged.gene_counts.tsv")
TPM_path=file.path(Rawdata_path,"salmon.merged.gene_tpm.tsv")
Raw_counts=readr::read_tsv(Rawcount_path)
TPM_counts=readr::read_tsv(TPM_path)
load(file.path(Reference_path,"Human_GSEA","Protein_coding_genelist.rdata")) # loading Protein_coding_genelist data
Raw_counts_2=Raw_counts%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()%>%mutate_all(.,as.integer)
 
TPM_2=TPM_counts%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()
dir(Rawdata_path)
colnames(Raw_counts_2)
dir(Rawdata_path)
```


```{r,include=T}
Drug_anno=readxl::read_excel(file.path(Rawdata_path,"Drug_annotation.xlsx"))

RNAseq_anno=data.frame(Sample_name=colnames(Raw_counts_2))%>%mutate(individualID="MN2")%>% mutate(Drug =gsub(".*\\_","",Sample_name))%>%
                  mutate(Drug=ifelse(grepl("DMSO",Sample_name),"DMSO",Drug )) %>%
                            mutate(Timepoint=case_when(
                              grepl("0h",Sample_name)~"0h",
                              grepl("8h",Sample_name)~"8h",
                              grepl("24h",Sample_name)~"24h",
                              TRUE ~ NA_character_)) %>%
  mutate(Drug =gsub("\\d","",Drug))%>%left_join(.,Drug_anno,by="Drug")%>%mutate(Time=ifelse(grepl("SEL|TRA$|RIB|PAL|RMC|VOR|TNO",Drug),"12192024","12272024"))%>%mutate(Time=ifelse(grepl("d1",Sample_name),"DMSO_d1",ifelse(grepl("d2",Sample_name),"DMSO_d2",Time )))
                            
Time_drug=RNAseq_anno%>%dplyr::select(Time,Drug)%>%distinct()%>%arrange(Time)

knitr::kable(Time_drug,caption = "Drug and Time")
```

```{r}

PCA_Plot_3<- function (data,Annotation,VAR,Color) {
  # logcountdata row:genes,column: samples
  pca <- prcomp(data) 
  pca_out<-as.data.frame(pca$x)
  df_out<- pca_out %>%tibble::rownames_to_column(var=VAR)  %>% left_join(., Annotation) 
  #df_out<- merge (pca_out,Annotation,by.x=0,by.y=0) 
  
  # label_color<- factor(df_out[,group])
  ggplot(df_out,aes_string(x="PC1",y="PC2")) +geom_point(aes_string(colour = Color))
}

PCA_Plot_3_shape<- function (data,Annotation,VAR,Color,Shape) {
  # logcountdata row:genes,column: samples
  pca <- prcomp(data) 
  pca_out<-as.data.frame(pca$x)
  df_out<- pca_out %>%tibble::rownames_to_column(var=VAR)  %>% left_join(., Annotation) 
  #df_out<- merge (pca_out,Annotation,by.x=0,by.y=0) 
  
  # label_color<- factor(df_out[,group])
  ggplot(df_out,aes_string(x="PC1",y="PC2")) +geom_point(aes_string(colour = Color,shape=Shape),size=5)
}



Drug_count_plot=function(logcount,sample_anno){
  
  logcount_2=logcount%>%t() %>%as.data.frame() %>%mutate(Sample_name=rownames(.))
  Level=sample_anno %>%arrange(Drug)%>%.$Sample_name
  data_mod = reshape2::melt(logcount_2, id.vars="Sample_name", 
                            measure.vars=rownames(logcount)) %>%left_join(sample_anno)%>%arrange(Drug)
  
  data_mod$Sample_name=factor(data_mod$Sample_name,levels =Level )
  #return(logcount_2)
 ggplot(data_mod)+geom_boxplot(aes(x=Sample_name, y=value, color=Drug,fill=Drug))+xlab("Sample")+ylab("log2(Count+1)")+theme(axis.text.x=element_blank())
  
  
}

selectGenes_BY_group <- function(Counts, min.count=10, N=0.90,Sample_Anno){
 Group_list=list()
 for (i in unique(Sample_Anno$Group)) {
  sample_name=Sample_Anno%>%dplyr::filter(Group==i)%>%.$Sample_name
  counts=Counts[, sample_name]
  lib.size <- base::colSums(counts)
  MedianLibSize <- median(lib.size)
  CPM.Cutoff <- min.count / MedianLibSize*1e6
  CPM <- edgeR::cpm(counts,lib.size=lib.size)
  
  min.samples <- round(N * ncol(counts))
 
  f1 <- genefilter::kOverA(min.samples, CPM.Cutoff)
  flist <- genefilter::filterfun(f1)
  Group_list[[i]] <- genefilter::genefilter(CPM, flist)%>%as.data.frame()%>%setNames(paste0(i  ,"_","TF"))
 }
  ## the same as:
  #keep <- apply(CPM, 1, function(x, n = min.samples){
  #  t = sum(x >= CPM.Cutoff) >= n
  #  t
  #})
 
  Output=do.call("cbind",Group_list)
  keep.exprs_2=Output%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum==length(unique(Sample_Anno$Group)))
  myFilt <-Counts[rownames(keep.exprs_2),]
  
}
```

```{r,fig.width=12,fig.height=6,include=T}
Tumor_count_log=log2(Raw_counts_2+1)
Drug_count_plot(Tumor_count_log,RNAseq_anno)

Drug_name = RNAseq_anno$Drug %>% unique()
Drug_name=factor(Drug_name,levels = Drug_name)
# Ensure you have 14 unique Drug names (1 for black, 13 other colors)
if (length(Drug_name) != 14) {
  stop("The number of unique drugs should be 14, including the one colored black.")
}

# Use RColorBrewer to get 13 colors from the Paired palette
paired_colors =RColorBrewer:: brewer.pal(n = 12, name = "Paired")
extra_color =RColorBrewer:: brewer.pal(n = 8, name = "Set3")[1]  # Take one color from another palette (e.g., Set3)

# Combine black and the 13 colors
drug_colors = c("black", paired_colors, extra_color)

# Ensure the color vector matches the order of Drug_name
names(drug_colors) = Drug_name
```
# PCA plot with Drug timepoint and drug class
```{r}
set.seed(12345)


P1=PCA_Plot_3 (t(Tumor_count_log),RNAseq_anno,"Sample_name","Drug")+
  scale_color_manual(values = drug_colors)
P2=PCA_Plot_3  (t(Tumor_count_log),RNAseq_anno,"Sample_name","Timepoint") 
P3=PCA_Plot_3  (t(Tumor_count_log),RNAseq_anno,"Sample_name","Drug_class")

```

```{r,fig.width=12,fig.height=8,include=T}
cowplot::plot_grid(P1,P2,P3,nrow=2)
```

```{r,fig.width=12,fig.height=10,include=T}
set.seed(12345)
P3_2=PCA_Plot_3_shape (t(Tumor_count_log),RNAseq_anno,"Sample_name","Drug","Timepoint")+
  scale_color_manual(values = drug_colors)+facet_wrap(~Drug_class)
P3_2
```

# Method1: PCA with all samples and then split
```{r}


# Regroup and create the PCA plot with custom colors
RNAseq_anno_Regroup = RNAseq_anno %>%
  mutate(New_group = ifelse(Time %in% c("DMSO_d1", "12192024"), "DMSO_d1+12192024", "DMSO_d2+12272024"))

P4 = PCA_Plot_3_shape(t(Tumor_count_log), RNAseq_anno_Regroup, "Sample_name", "Drug", "Timepoint") +
  scale_color_manual(values = drug_colors) +  # Apply the custom colors
  facet_wrap(~New_group)

# Show the plot
P4


```

```{r}
RNAseq_anno_Regroup = RNAseq_anno %>%
  mutate(New_group = ifelse(Time %in% c("DMSO_d2", "12192024"), "DMSO_d2+12192024", "DMSO_d1+12272024"))

P5 = PCA_Plot_3_shape(t(Tumor_count_log), RNAseq_anno_Regroup, "Sample_name", "Drug", "Timepoint") +
  scale_color_manual(values = drug_colors) +  # Apply the custom colors
  facet_wrap(~New_group)

# Show the plot
P5
```
```{r,fig.width=10,fig.height=8,include=T}
cowplot::plot_grid(P4,P5,nrow=2)
```


# Method2: PCA with each subgroup before plotting

## define new group by Experimental date 
```{r,fig.width=12,fig.height=6,include=T}
# Regroup and create the PCA plot with custom colors
RNAseq_anno_Regroup = RNAseq_anno %>%
  mutate(New_group = ifelse(Time %in% c("DMSO_d1", "12192024"), "DMSO_d1+12192024", "DMSO_d2+12272024"))
Plotlist=list()
for (i in unique(RNAseq_anno_Regroup$New_group)){
  Sample=RNAseq_anno_Regroup%>%dplyr::filter(New_group==i)
  data=t(Tumor_count_log[Sample$Sample_name])
 Plotlist[[i]] = PCA_Plot_3_shape(data, RNAseq_anno_Regroup, "Sample_name", "Drug", "Timepoint") +
  scale_color_manual(values = drug_colors)+ggtitle(i)
}



# Show the plot
cowplot::plot_grid(plotlist = Plotlist)


```

```{r,fig.width=12,fig.height=6,include=T}
RNAseq_anno_Regroup = RNAseq_anno %>%
  mutate(New_group = ifelse(Time %in% c("DMSO_d2", "12192024"), "DMSO_d2+12192024", "DMSO_d1+12272024"))
 
Plotlist=list()
for (i in unique(RNAseq_anno_Regroup$New_group)){
  Sample=RNAseq_anno_Regroup%>%dplyr::filter(New_group==i)
  data=t(Tumor_count_log[Sample$Sample_name])
 Plotlist[[i]] = PCA_Plot_3_shape(data, RNAseq_anno_Regroup, "Sample_name", "Drug", "Timepoint") +
  scale_color_manual(values = drug_colors)+ggtitle(i)
}



# Show the plot
cowplot::plot_grid(plotlist = Plotlist)


```



## define new group by timepoints

```{r,fig.width=12,fig.height=6,include=T}

Plotlist=list()
for (i in c("8h","24h")){
  Sample=RNAseq_anno%>%dplyr::filter(Timepoint==i)
  data=t(Tumor_count_log[Sample$Sample_name])
 Plotlist[[i]] = PCA_Plot_3_shape(data, RNAseq_anno, "Sample_name", "Drug", "Time") +
  scale_color_manual(values = drug_colors)+ggtitle(i)
}



# Show the plot
cowplot::plot_grid(plotlist = Plotlist)


```





# Method3: DMSO_d1: switch 8h with 24h, pca with each subtype

## define new group by Experimental date 
```{r,fig.width=10,fig.height=10,include=T}
# Regroup and create the PCA plot with custom colors
RNAseq_anno_swich= RNAseq_anno %>%mutate(Timepoint2 = ifelse(Timepoint=="8h"&grepl("d1",Sample_name),"24h",
                                                              ifelse(Timepoint=="24h"&grepl("d1",Sample_name),"8h",Timepoint)))%>%
  mutate(New_group = ifelse(Time %in% c("DMSO_d1", "12192024"), "DMSO_d1+12192024", "DMSO_d2+12272024"))


Plotlist=list()
for (i in unique(RNAseq_anno_swich$New_group)){
  Sample=RNAseq_anno_swich%>%dplyr::filter(New_group==i)
  data=t(Tumor_count_log[Sample$Sample_name])
 Plotlist[[i]] = PCA_Plot_3_shape(data, RNAseq_anno_swich, "Sample_name", "Drug","Drug_class") +
  scale_color_manual(values = drug_colors)+ggtitle(i)+facet_grid(~Timepoint2)
}



# Show the plot
cowplot::plot_grid(plotlist = Plotlist,ncol=1,rel_widths = c(0.8,1))


```

# DMSO group
```{r,include=T,fig.width=6,fig.height=4}
Control=Tumor_count_log%>%.[,grepl("DMSO",colnames(.)) ]
RNAseq_anno_control=RNAseq_anno%>%dplyr::filter(Drug=="DMSO")%>%mutate(Batch=ifelse(grepl("d1",Sample_name),"Batch1","Batch2"))
PCA_Plot_3_shape (t(Control),RNAseq_anno_control,"Sample_name","Batch","Timepoint")

```



```{r}

DEG_data=list()

for (i in c("d1","d2")){
DMSO_data=RNAseq_anno%>%dplyr::filter(Drug=="DMSO")%>%dplyr::filter(grepl(i,Sample_name))
DMSO_rawcounts=Raw_counts_2[,DMSO_data$Sample_name]
DMSO_anno=DMSO_data%>%.[,c("Sample_name","Drug","Timepoint")]
New_anno=DMSO_anno%>%mutate(Group=paste0(Drug,"_",Timepoint))
New_anno$Timepoint=factor(New_anno$Timepoint,levels = c("0h","8h","24h"))
rownames(New_anno)=New_anno[,1]
filter_counts <- selectGenes_BY_group(DMSO_rawcounts, min.count=10, N=0.90,New_anno)
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =filter_counts, colData =New_anno,design = ~Group)
deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
dds <- DESeq2::DESeq(deseq2Data)

DESeq2::resultsNames(dds)
res_list=list()
Timepoint=c("8h","24h")

res_list=list(DMSO8h_vs_0h=DESeq2::results(dds, name="Group_DMSO_8h_vs_DMSO_0h" ),
                  DMSO24h_vs_0h=DESeq2::results(dds, name="Group_DMSO_24h_vs_DMSO_0h" ))
names(res_list)=c("DMSO8h_vs_0h", "DMSO24h_vs_0h")

DEG_data[[i]]=res_list
}
```


```{r}
create_DEGs_venn <- function(res1, res2, threshold_padj = 0.05, threshold_log2fc = 0.67, 
                             plot_title = "DEGs Venn Diagram", Group=c("Group1","Group2"),
                             category_positions = list(x = c(0.5, 1.5), y = c(1.5, 1.5))) {
  # Load necessary library
  if (!requireNamespace("ggVennDiagram", quietly = TRUE)) {
    stop("Please install the 'ggVennDiagram' package.")
  }
  
  library(ggVennDiagram)
  
  # Extract DEGs based on the specified criteria
  degs1 <- rownames(res1[which(res1$padj < threshold_padj & abs(res1$log2FoldChange) > threshold_log2fc), ])
  degs2 <- rownames(res2[which(res2$padj < threshold_padj & abs(res2$log2FoldChange) > threshold_log2fc), ])
  
  # Extract descriptions from res1 and res2
  Name1 <- res1@elementMetadata %>% as.data.frame() %>% .[2, ] %>% .$description %>% gsub(".*\\:", "", .)%>% gsub("Group ", "", .)
  Name2 <- res2@elementMetadata %>% as.data.frame() %>% .[2, ] %>% .$description %>% gsub(".*\\:", "", .)%>% gsub("Group ", "", .)
  
  # Prepare list for Venn diagram
  degs_list <- list(
    Name1 = degs1,
    Name2 = degs2
  )
  
  # Create the Venn diagram with empty category names
  venn_plot <- ggVennDiagram(degs_list, category.names = c("", ""), label_alpha = 0.5) + 
    scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")  
  
  # Adjust the position of category names
  venn_plot <- venn_plot + 
    theme(legend.position = "none") + 
    geom_text(aes(x = category_positions$x[1], y = category_positions$y[1], label = paste0(Group[1]," ",Name1)), size = 5) +
    geom_text(aes(x = category_positions$x[2], y = category_positions$y[2], label =paste0(Group[2]," ",Name2)), size = 5)
  
  # Print the Venn diagram
  print(venn_plot)
  
  # Optionally save the plot (uncomment the line below if saving is desired)
  # ggsave(paste0(gsub(" ", "_", plot_title), ".png"), venn_plot)
}

# Example usage:
# create_DEGs_venn(res1, res2, category_positions = list(x = c(0.4, 1.6), y = c(2, 2)))

```

```{r}
create_DEGs_venn <- function(res1, res2, threshold_padj = 0.05, threshold_log2fc = 0.67, 
                             plot_title = "DEGs Venn Diagram", Group=c("Group1","Group2"),
                             category_positions = list(x = c(0.5, 1.5), y = c(1.5, 1.5)),
                             circle_colors = c("red", "blue"),
                             change = c("Up", "Down")) {
  # Load necessary library
  if (!requireNamespace("ggVennDiagram", quietly = TRUE)) {
    stop("Please install the 'ggVennDiagram' package.")
  }
  
  library(ggVennDiagram)
  
  # Split DEGs into upregulated and downregulated
  degs1_up <- rownames(res1[which(res1$padj < threshold_padj & res1$log2FoldChange > threshold_log2fc), ])
  degs1_down <- rownames(res1[which(res1$padj < threshold_padj & res1$log2FoldChange < -threshold_log2fc), ])
  
  degs2_up <- rownames(res2[which(res2$padj < threshold_padj & res2$log2FoldChange > threshold_log2fc), ])
  degs2_down <- rownames(res2[which(res2$padj < threshold_padj & res2$log2FoldChange < -threshold_log2fc), ])
  
  # Extract descriptions from res1 and res2
  Name1 <- res1@elementMetadata %>% as.data.frame() %>% .[2, ] %>% .$description %>% gsub(".*\\:", "", .)
  Name1=paste0(Group[1]," ",Name1)
  Name2 <- res2@elementMetadata %>% as.data.frame() %>% .[2, ] %>% .$description %>% gsub(".*\\:", "", .)
  Name2=paste0(Group[2]," ",Name2)
  # Initialize list to store plots
  venn_plot <- NULL
  
  # Create the Venn diagram based on the 'change' variable
  if (change == "Up") {
    degs_list_up <- list(
      batch1 = degs1_up,
      batch2 = degs2_up
    )
    
    venn_plot <- ggVennDiagram(degs_list_up, category.names = c("", ""), label_alpha = 0.5) + 
      scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
      ggtitle(paste0(plot_title, "-Upregulated")) +
      scale_color_manual(values = circle_colors, labels = c("batch1", "batch2")) +
    #  theme(legend.position = "right") +
      geom_text(aes(x = category_positions$x[1], y = category_positions$y[1], label = Name1), size = 3, color = circle_colors[1]) +
      geom_text(aes(x = category_positions$x[2], y = category_positions$y[2], label = Name2), size = 3, color = circle_colors[2])
    
  } else if (change == "Down") {
    degs_list_down <- list(
      batch1 = degs1_down,
      batch2 = degs2_down
    )
    
    venn_plot <- ggVennDiagram(degs_list_down, category.names = c("", ""), label_alpha = 0.5) + 
      scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
      ggtitle(paste0(plot_title, "-Downregulated")) +
      scale_color_manual(values = circle_colors, labels = c("batch1", "batch2")) +
      theme(legend.position = "right") +
      geom_text(aes(x = category_positions$x[1], y = category_positions$y[1], label = Name1), size = 3, color = circle_colors[1]) +
      geom_text(aes(x = category_positions$x[2], y = category_positions$y[2], label = Name2), size = 3, color = circle_colors[2])
  }
  
  # Return the plot
  return(venn_plot)
}

# Example usage:
# Generate the Venn diagram for upregulated DEGs
# up_plot <- create_DEGs_venn(res1, res2, change = "Up")

# Generate the Venn diagram for downregulated DEGs
# down_plot <- create_DEGs_venn(res1, res2, change = "Down")

```

```{r,fig.width=6,fig.height=4}
Plot_list=list()
for (i in c(1,2)) {
  for(t in c(1,2)) {
  Res1=DEG_data[[1]][[i]]
  Res2=DEG_data[[2]][[t]]
  name=paste0(i,"_",t)
  Plot_list[[name]]=create_DEGs_venn (Res1,Res2,Group=c("batch1","batch2"), plot_title = "FC:1.2,q<0.05",change="Up" ,threshold_log2fc = 0.27, category_positions = list(x = c(0.5, 1), y = c(0, 6.5)))
  }
}
```


```{r,fig.width=26,fig.height=6,include=T}
cowplot::plot_grid(plotlist = Plot_list,nrow=1)
```

```{r,fig.width=6,fig.height=4}
Plot_list=list()
for (i in c(1,2)) {
  for(t in c(1,2)) {
  Res1=DEG_data[[1]][[i]]
  Res2=DEG_data[[2]][[t]]
  name=paste0(i,"_",t)
  Plot_list[[name]]=create_DEGs_venn (Res1,Res2,Group=c("batch1","batch2"), plot_title = "FC:1.2,q<0.05",change="Down" ,threshold_log2fc = 0.27, category_positions = list(x = c(0.5, 1), y = c(0, 6.5)))
  }
}
```


```{r,fig.width=26,fig.height=6,include=T}
cowplot::plot_grid(plotlist = Plot_list,nrow=1)
```
```{r}
knitr::knit_exit()
```

```{r}
# GTAC DATA
# Rawcount_path="/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_RNAseq_Kangwen/all.gene_counts.tsv" 
# 
# Rawcount_path="/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_RNAseq_Kangwen/all.gene_counts.tsv"
# Raw_counts=readr::read_tsv(Rawcount_path)%>%dplyr::filter(gene_biotype=="protein_coding")
# #
# Raw_counts_2=Raw_counts%>%.[grepl("sample",colnames(.))]%>%mutate(gene_name=Raw_counts$external_gene_name)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()%>%mutate_all(.,as.integer)
# 
# Raw_counts_2["ACTB",]
```


```{r}
Sample_anno=data.frame(Sample_name=colnames(Raw_counts_2))%>%mutate(Drug=ifelse(grepl("D",Sample_name),"DMSO", ifelse(grepl("M",Sample_name),"Mirda","DMSO") ))%>%mutate(Timepoint=ifelse(grepl("0h",Sample_name),"0hour", ifelse(grepl("8h",Sample_name),"8hour","24hour") ))
```


```{r}
PCA_Plot_3<- function (data,Annotation,VAR,Color,Shape) {
  # logcountdata row:genes,column: samples
  pca <- prcomp(data) 
  pca_out<-as.data.frame(pca$x)
  df_out<- pca_out %>%tibble::rownames_to_column(var=VAR)  %>% left_join(., Annotation) 
  #df_out<- merge (pca_out,Annotation,by.x=0,by.y=0) 
  
  # label_color<- factor(df_out[,group])
  ggplot(df_out,aes_string(x="PC1",y="PC2")) +geom_point(aes_string(colour = Color,shape=Shape),size=3) 
}

```

```{r}
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =Raw_counts_2, colData = Sample_anno,design = ~Timepoint)
deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
```

```{r}
selectGenes <- function(counts, min.count=10, N=0.90){
 
  lib.size <- base::colSums(counts)
  MedianLibSize <- median(lib.size)
  CPM.Cutoff <- min.count / MedianLibSize*1e6
  CPM <- edgeR::cpm(counts,lib.size=lib.size)
  
  min.samples <- round(N * ncol(counts))
 
  f1 <- genefilter::kOverA(min.samples, CPM.Cutoff)
  flist <- genefilter::filterfun(f1)
  keep <- genefilter::genefilter(CPM, flist)
 
  ## the same as:
  #keep <- apply(CPM, 1, function(x, n = min.samples){
  #  t = sum(x >= CPM.Cutoff) >= n
  #  t
  #})
 
  return(keep)
}
 
 
selectGenes_BY_group <- function(Counts, min.count=10, N=0.90,Sample_Anno){
 Group_list=list()
 for (i in unique(Sample_Anno$Group)) {
  sample_name=Sample_Anno%>%dplyr::filter(Group==i)%>%.$Sample_name
  counts=Counts[, sample_name]
  lib.size <- base::colSums(counts)
  MedianLibSize <- median(lib.size)
  CPM.Cutoff <- min.count / MedianLibSize*1e6
  CPM <- edgeR::cpm(counts,lib.size=lib.size)
  
  min.samples <- round(N * ncol(counts))
 
  f1 <- genefilter::kOverA(min.samples, CPM.Cutoff)
  flist <- genefilter::filterfun(f1)
  Group_list[[i]] <- genefilter::genefilter(CPM, flist)%>%as.data.frame()%>%setNames(paste0(i  ,"_","TF"))
 }
  ## the same as:
  #keep <- apply(CPM, 1, function(x, n = min.samples){
  #  t = sum(x >= CPM.Cutoff) >= n
  #  t
  #})
 
  Output=do.call("cbind",Group_list)
}

Volcano_plot_rnaseq_simple=function(All_res_Data,Pathway_list,pathway_name){
  
  cols <- c("DOWN"="#26b3ff", "NO"= "grey","UP"= "#ffad73","up" = "red", "down" = "blue") 
  sizes <- c( "DOWN"=0.5,"UP"=0.5 ,"up" = 3.5, "down" = 3.5, "NO" = 0.5) 
  alphas <- c("UP" = 1, "DOWN" = 1, "NO" = 0.5)
  
  pathway_genes=Pathway_list [[pathway_name]]%>%as.character()
  
  library(ggrepel)
  # Data=All_res_Data%>%dplyr::filter(!gene_symbol%in%pathway_genes)%>%mutate(diffexpressed=ifelse(padj>0.05|is.na(padj),"NO",ifelse(log2FoldChange>0,"UP","DOWN")))%>%mutate(delabel=NA)
    
  pathway_Data=All_res_Data%>%.[.$gene_symbol%in%pathway_genes,]%>%mutate(diffexpressed=ifelse(padj>0.05|is.na(padj),"NO",ifelse(log2FoldChange>0,"up","down")))%>%mutate(delabel=ifelse(diffexpressed!="NO"&(log2FoldChange>0.27|log2FoldChange<as.numeric( "-0.27")),gene_symbol,NA))
  # add a column of NAs
  #de=rbind(Data,pathway_Data)
  de=pathway_Data
  
  MAX_y=max(-log10(de$padj))
  # de$delabel <- NA
  # 
  # de$delabel[de$diffexpressed != "NO"] <- de$gene_symbol[de$diffexpressed != "NO"]
   # plot adding up all layers we have seen so far
  ggplot(data=de, aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed, label=delabel)) +
    geom_point(show.legend = F) + ggtitle(pathway_name)+
    theme_minimal() +
    geom_text_repel(size=3)+scale_color_manual(values=cols)+scale_size_manual(values=sizes)+scale_alpha_manual(values=alphas)+
    geom_vline(xintercept=c(-0.27, 0.27), col="red") +
    geom_hline(yintercept=-log10(0.05), col="red")+theme_bw()+  theme(legend.position = "none", panel.border = element_rect(colour = "black", fill = NA, size= 0.5),    
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) 
  
}

                                           



Volcano_plot_rnaseq_add_label_update=function(All_res_Data,Pathway_list,pathway_name){
  
  cols <- c("DOWN"="#26b3ff", "NO"= "grey","UP"= "#ffad73","up" = "red", "down" = "blue") 
  sizes <- c( "DOWN"=0.5,"UP"=0.5 ,"up" = 3.5, "down" = 3.5, "NO" = 0.5) 
  alphas <- c("UP" = 1, "DOWN" = 1, "NO" = 0.5)
  
  pathway_genes=Pathway_list [[pathway_name]]%>%as.character()
  
  library(ggrepel)
  # Data=All_res_Data%>%dplyr::filter(!gene_symbol%in%pathway_genes)%>%mutate(diffexpressed=ifelse(padj>0.05|is.na(padj),"NO",ifelse(log2FoldChange>0,"UP","DOWN")))%>%mutate(delabel=NA)
    
  pathway_Data=All_res_Data%>%.[.$gene_symbol%in%pathway_genes,]%>%mutate(diffexpressed=ifelse(padj>0.05|is.na(padj),"NO",ifelse(log2FoldChange>0,"up","down")))%>%mutate(delabel=ifelse(diffexpressed!="NO"&(log2FoldChange>0.27|log2FoldChange<as.numeric( "-0.27")),gene_symbol,NA))
  # add a column of NAs
  #de=rbind(Data,pathway_Data)
  de=pathway_Data
  
  MAX_y=max(-log10(de$padj))
  # de$delabel <- NA
  # 
  # de$delabel[de$diffexpressed != "NO"] <- de$gene_symbol[de$diffexpressed != "NO"]
   # plot adding up all layers we have seen so far
  ggplot(data=de, aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed, label=delabel)) +
    geom_point(show.legend = F) + ggtitle(pathway_name)+
    theme_minimal() +
    geom_text_repel(size=3,force_pull= 0,max.overlaps  = Inf,
    data = subset(de, log2FoldChange >2),
    nudge_x= 6 - data$log2FoldChange,
    segment.size= 0.2,
    segment.color= "grey50",
    direction = "y",
    hjust         = 0.5 ) +geom_text_repel(size=3,force_pull= 0,max.overlaps  = Inf,
    data = subset(de, log2FoldChange >0.27&log2FoldChange<2),
    nudge_x= 4.5 - subset(de, log2FoldChange >1&log2FoldChange<2)$log2FoldChange,
    segment.size= 0.2,
    segment.color= "grey50",
    direction = "y",
    hjust         = 0.5 ) +geom_text_repel(size=3,force_pull= 0,max.overlaps  = Inf,
    data = subset(de, log2FoldChange < as.numeric(-0.27)),
    nudge_x= -5 - subset(de, log2FoldChange <as.numeric(-1))$log2FoldChange,
    segment.size= 0.2,
    segment.color= "grey50",
    direction = "y",
    hjust         = 0.5 ) +scale_color_manual(values=cols)+scale_size_manual(values=sizes)+scale_alpha_manual(values=alphas)+
    geom_vline(xintercept=c(-0.27, 0.27), col="red") +
    geom_hline(yintercept=-log10(0.05), col="red")+theme_bw()+  theme(legend.position = "none", panel.border = element_rect(colour = "black", fill = NA, size= 0.5),    
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) 
  
  
  
  
}

Gene_pathway_sig_enrich=function(gene) {
  
 library(msigdbr)
 m_df <- msigdbr(species = "Homo sapiens")
 Cat=c("C2","C4","C5","C6","H")
  emlist=list()
 for (i in Cat) {
 
 m_t2g <- msigdbr(species = "Homo sapiens", category = i)%>% dplyr::select(gs_name, gene_symbol)
 #MSigDb over-presentaton analysis
  em= enricher(gene, TERM2GENE=m_t2g)
 
 emlist[[i]]=em@result%>%as.data.frame()%>%dplyr::filter(pvalue<0.05)
 }
 
 out=do.call("rbind",emlist)
 
}

```




```{r}
# Sample_anno_2=Sample_anno%>%dplyr::filter(Drug!="Control")%>%mutate(Group=paste0(Drug,"_",Timepoint))
# Raw_counts_rm_control=Raw_counts_2%>%.[,Sample_anno_2$Sample_name]
Raw_counts_add_control=Raw_counts_2%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)
Sample_anno_2=data.frame(Sample_name=colnames(Raw_counts_add_control))%>%left_join(.,Sample_anno)%>%mutate(Drug=ifelse(is.na(Drug),"Mirda",Drug))%>%mutate(Timepoint=ifelse(is.na(Timepoint),"0hour",Timepoint))%>%mutate(Group=paste0(Drug,"_",Timepoint))
rownames(Sample_anno_2)=Sample_anno_2$Sample_name
Sample_anno_2$Timepoint=factor(Sample_anno_2$Timepoint,levels = c("0hour","8hour","24hour"))
library(SummarizedExperiment)
library(DESeq2)
keep.exprs <- selectGenes_BY_group(Raw_counts_add_control, min.count=10, N=0.90,Sample_anno_2)
keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
myFilt <- Raw_counts_add_control[rownames(keep.exprs_2),]

dim(myFilt)


```

# Time course analyses with LRT
**The LRT test can be especially helpful when performing time course analyses. We can use the LRT to explore whether there are any significant differences in treatment effect between any of the timepoints.**

**For have an experiment looking at the effect of treatment over time on mice of two different genotypes. We could use a design formula for our ‘full model’ that would include the major sources of variation in our data: genotype, treatment, time, and our main condition of interest, which is the difference in the effect of treatment over time (treatment:time).**
```{r}
#https://rstudio-pubs-static.s3.amazonaws.com/329027_593046fb6d7a427da6b2c538caf601e1.html
deseq2Data_add_control=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt , colData = Sample_anno_2,design = ~Drug+Timepoint+Drug:Timepoint)
deseq2Data_add_control$Timepoint=relevel(deseq2Data_add_control$Timepoint,"0hour")
dds_LRT= DESeq2::DESeq(deseq2Data_add_control,test="LRT", reduced =~Drug+Timepoint)
DESeq2::resultsNames(dds_LRT)
```

```{r}
Mirda_8h_vs_DMSO=results(dds_LRT,name="DrugMirda.Timepoint8hour")%>%as.data.frame()
Mirda_8h_vs_DMSO_sig=Mirda_8h_vs_DMSO%>%dplyr::filter(padj<0.05)
Mirda_24h_vs_DMSO=results(dds_LRT,name="DrugMirda.Timepoint24hour")%>%as.data.frame()
Mirda_24h_vs_DMSO_sig=Mirda_24h_vs_DMSO%>%dplyr::filter(padj<0.05)

```


```{r}
# # Extract results
res_LRT <- results(dds_LRT)
res_LRT_q005=res_LRT %>%as.data.frame()%>%dplyr::filter(padj<0.05)
rld <- rlogTransformation(dds_LRT)

rld_mat <- assay(rld)
cluster_rlog=rld_mat[rownames(res_LRT_q005),]

TPM_counts_clean=TPM_2[rownames(cluster_rlog),]%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)

logTPM=log(TPM_counts_clean+1)
```



# Volcano plot for DEPs

```{r,fig.width=18,fig.height=18,include=T}

DEG_data_1= Mirda_8h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(padj<0.05)

DEG_data_2= Mirda_24h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(padj<0.05)

DEG_list=list("Mirda_vs_DMSO_8hr"=DEG_data_1$gene_symbol,"Mirda_vs_DMSO_24hr"=DEG_data_2$gene_symbol)

Out=Volcano_plot_rnaseq_simple (DEG_data_1,DEG_list,"Mirda_vs_DMSO_8hr")
Out2=Volcano_plot_rnaseq_simple(DEG_data_2,DEG_list,"Mirda_vs_DMSO_24hr") 
cowplot::plot_grid(Out,Out2,ncol=2)
```


```{r,fig.width=12,fig.height=18,include=F}
# Inspired by https://lpantano.github.io/DEGreport/reference/degPatterns.html

data=logTPM
clusters <- DEGreport::degPatterns(data, metadata = Sample_anno_2, time="Timepoint", col="Drug",nClusters =6)
```



```{r}
Cluster_data=clusters[["normalized"]]%>%mutate(cluster=paste0("Cluster",cluster))
Cluster_Genelist=list()
for (i in unique(Cluster_data$cluster)){
 Cluster_Genelist[[i]]=Cluster_data%>%dplyr::filter(cluster==i)%>%.$genes
  
}


sig_plot_list=list()

for (i in names(Cluster_Genelist)) {
  Gene=Cluster_data%>%dplyr::filter(cluster==i)%>%.$genes
  DEG_data_1= Mirda_8h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(gene_symbol%in%Gene)
  DEG_data_2= Mirda_24h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(gene_symbol%in%Gene)

  Out=Volcano_plot_rnaseq_add_label_update(DEG_data_1,Cluster_Genelist,i)
  Out2=Volcano_plot_rnaseq_add_label_update(DEG_data_2,Cluster_Genelist,i) 
  sig_plot_list[[i]]=cowplot::plot_grid(Out,Out2,rnow=1,labels=c("Mirda_vs_DMSO_8h","Mirda_vs_DMSO_24h"),vjust=3.0,hjust=-1,label_size=10)
}
```

```{r}
Data=logTPM%>%t()%>%as.data.frame()%>%tibble::rownames_to_column(var="Sample_name")%>%tidyr::gather(,key="genes",value="Log2TPM",-Sample_name)

cluster_norm=clusters[["normalized"]]%>%dplyr::filter(merge!="Mirda0hour")%>%left_join(.,Data) 
#%>%dplyr::filter(merge!="Mirda0hour")%>%left_join(.,Data)%>%mutate(Cluster=paste0("Cluster",cluster))
# Use the data yourself for custom figures
# P1=ggplot(cluster_norm,
#         aes(Timepoint, Log2TPM, color = Drug,shape=genes)) +
#    geom_point(position = position_jitterdodge(dodge.width = 0.9)) +
#    # change the method to make it smoother
#      geom_smooth(aes(group=Timepoint), method = "lm")
cluster_plot=list()

for (i in unique(cluster_norm$cluster)) {
  data=cluster_norm%>%dplyr::filter(cluster==i)
  cluster_plot[[i]]=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

}

cluster_plot[[i]]
```


# Cluster 1,3 and 6 (Gene expressions were upregulated at 8 and/or 24 hour in DMSO group)

## Cluster1

```{r,fig.width=18,fig.height=15}
i=1
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")
```

```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

## cluster 3

```{r,fig.width=18,fig.height=15}
i=3
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```



## cluster6

```{r,fig.width=18,fig.height=15}
i=6
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

# Cluster 4 and 5 (Gene expressions were downregulated at 8 and/or 24 hour in DMSO group)

## Cluster 4

```{r,fig.width=18,fig.height=15}
i=4
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

## Cluster5

```{r,fig.width=18,fig.height=15}
i=5
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob(out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```


# Cluster 2  (Gene expressions were upregulated at 8 and 24 hour in DMSO and Mirda groups )

```{r,fig.width=18,fig.height=40}
i=2
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=48,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

```{r}
knitr::knit_exit()
```



# ETV5 AND SPRY4 expression (TPM)

```{r,include=F}
Genes=c("ETV5","SPRY4")
Data=logTPM[Genes,]%>%t()%>%as.data.frame()%>%tibble::rownames_to_column(var="Sample_name")%>%tidyr::gather(,key="genes",value="Log2TPM",-Sample_name)

cluster_norm=clusters[["normalized"]]%>%dplyr::filter(grepl("ETV5|SPRY4",genes))%>%dplyr::filter(merge!="Mirda0hour")%>%left_join(.,Data)
# Use the data yourself for custom figures
P1=ggplot(cluster_norm,
        aes(Timepoint, Log2TPM, color = Drug,shape=genes)) +
   geom_point(position = position_jitterdodge(dodge.width = 0.9)) +
   # change the method to make it smoother
     geom_smooth(aes(group=Timepoint), method = "lm")

P2=ggplot(cluster_norm,
        aes(Timepoint, value, color = Drug,shape=genes)) +
   geom_point(position = position_jitterdodge(dodge.width = 0.9)) +ylab("normalized_log2TPM")+
   # change the method to make it smoother
     geom_smooth(aes(group=Timepoint), method = "lm")
```


```{r,fig.width=6,fig.height=3,include=T}
cowplot::plot_grid(P1,P2,nrow=1)
```



# Pathway analysis for each cluster

```{r}
sig_path_list=list()

for (i in 1:6) {
Gene=clusters[["normalized"]]%>%dplyr::filter(cluster==i)%>%.$genes
Cluster_1_sig_path=Gene_pathway_sig_enrich(Gene)
Cluster_1_sig_path_pickup=Cluster_1_sig_path%>%dplyr::filter(Count>2&pvalue<0.05)
sig_path_list[[i]]=Cluster_1_sig_path_pickup%>%mutate(Cluster=i)
}



```


```{r}
sig_path_list[[6]]%>%dplyr::filter(grepl("WP_|HALLMARK|KEGG",ID))
```


```{r}
knitr::knit_exit()
```




```{r}
res=DESeq2::results(Deg_add_control,name="Timepoint_8hour_vs_0hour", alpha = 0.05)%>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
```



# For mirda 24hour vs 0hour
```{r}
# 
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint24hour","Timepoint_24hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))

res=DESeq2::results(Deg_add_control,list(c("Timepoint_24hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))

```



```{r}
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```


# For mirda 8hour vs 0hour
```{r}
# 
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour","Timepoint_8hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))

plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```



# 8hr Mirda vs DMSO

```{r}
# 
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour","Drug_Mirda_vs_DMSO"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]

res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]

```


```{r}

Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```

# 24hr Mirda vs DMSO
```{r}
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint24hour","Drug_Mirda_vs_DMSO"  )), alpha = 0.05) %>%as.data.frame()

res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
genes=c("ETV5")
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```

# for Mirda 8hr vs 0hr
```{r}
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour","Timepoint_8hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()

res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
genes=c("ETV5")
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)+xlab()
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```


```{r}
res=DESeq2::results(Deg_add_control,list(c("Timepoint24hour.DrugMirda","Drug_Mirda_vs_DMSO"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res_q005[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
genes=c("ETV5")
TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 

library(ggplot2)
ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)

```


```{r}
# DMESO
Drug_Mirda_vs_DMSO_res=DESeq2::results(Deg_rm_control,name= "Timepoint8hour.DrugMirda" , alpha = 0.05) %>%as.data.frame()
Drug_Mirda_vs_DMSO_res_q005=Drug_Mirda_vs_DMSO_res%>%dplyr::filter(padj<0.05)

Drug_Mirda_vs_DMSO_res_na=Drug_Mirda_vs_DMSO_res%>%dplyr::filter(is.na(padj))

Drug_Mirda_vs_DMSO_res[c("ETV5", "SPRY4"),]

Raw_counts_rm_control[c("ETV5", "SPRY4"),]

```


```{r}
dds=Deg_add_control
res <- results(dds, name="Timepoint8hour.DrugMirda")%>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
ix = which.min(res$padj) # most significant
ix = which.min(res$log2FoldChange) # most significant

res <- res[order(res$log2FoldChange),] # sort
knitr::kable(res[1:5,-(3:4)])
```


```{r}
ix=rownames( res_q005)[1]
barplot(TPM_2[ix,],las=2, main=rownames(TPM_2)[ix]  )
```


```{r}
# 8hr
dds=Deg_rm_control
res = results(dds, contrast=c("Drug","Mirda","DMSO"))
ix = which.min(res$padj) # most significant
res <- res[order(res$padj),] # sort
knitr::kable(res[1:5,-(3:4)])
```

```{r}
dds=Deg_rm_control
res <- results(dds, list( c("condition_Trt_vs_Ctrl","genotypeMU.conditionTrt") ))
ix = which.min(res$padj) # most significant
res <- res[order(res$padj),] # sort
kable(res[1:5,-(3:4)])
```

 
```{r}
# 24hr
res = results(Deg_rm_control, contrast=c("Drug","Mirda","DMSO"))
ix = which.min(res$padj) # most significant
res <- res[order(res$padj),] # sort
knitr::kable(res[1:5,-(3:4)])
 




```{r}
DMSO_samples=Sample_anno%>%dplyr::filter(Drug%in%c("Control","DMSO"))
DMSO_count=Raw_counts_2[,DMSO_samples$Sample_name]
DMSO_anno=data.frame(Sample_name=colnames(DMSO_count))%>%left_join(.,Sample_anno)
DMSO_fill <- DESeq2:: DESeqDataSetFromMatrix(countData =DMSO_count, colData = DMSO_anno,
                                             design = ~Timepoint)
DMSO_Deg<- DESeq2::DESeq(DMSO_fill)
DESeq2::resultsNames(DMSO_Deg)
DMSO_24h_res <- DESeq2::results(DMSO_Deg,name= "Timepoint_24hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()
DMSO_8h_res <- DESeq2::results(DMSO_Deg,name= "Timepoint_8hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()


```

```{r}
Mirda_samples=Sample_anno%>%dplyr::filter(Drug%in%c("Control","Mirda"))
Mirda_count=Raw_counts_2[,Mirda_samples$Sample_name]
Mirda_anno=data.frame(Sample_name=colnames(Mirda_count))%>%left_join(.,Sample_anno)
Mirda_fill <- DESeq2:: DESeqDataSetFromMatrix(countData =Mirda_count, colData = Mirda_anno,
                                             design = ~Timepoint)
Mirda_Deg<- DESeq2::DESeq(Mirda_fill)
DESeq2::resultsNames(Mirda_Deg)
Mirda_res <- DESeq2::results(Mirda_Deg,  alpha = 0.05) %>%as.data.frame()

Mirda_24h_res <- DESeq2::results(Mirda_Deg,name= "Timepoint_24hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()
Mirda_8h_res <- DESeq2::results(Mirda_Deg,name= "Timepoint_8hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()

```

```{r}

save(deseq2VST,DMSO_8h_res,Mirda_8h_res,DMSO_24h_res,Mirda_24h_res  ,Sample_anno,file=file.path(Rawdata_path,"DEseq2_out.rdata"))
```


```{r}
knitr::knit_exit()
```

 

