---
title: "Evaluate second drug pathways"
author: "Sara Gosline"
date: "2026-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../RNAseq/plottingParameters.R")
source("correlation_functions.R")

```

## Data

We have reformatted the data into three files:
1. sample file describing the samples
2. gene file with gene metadata
3. transcriptomics data with gene expression data
4. combination data with delta AUC values
5. drug data with single AUC values

Most data is now read in by `correlation_functions.R`

```{r data, message=FALSE}

##get the transcript data, this is slow
if(!exists('full_tmat'))
  full_tmat <- get_harmonized_gex()

```

# Evaluate correlations

Here we want to look at correlations between baseline gene expression and drug
response.

### Correlation of AUC and gene expression

Let's first look at the untreated gene expression data and see if we can 
find out features that predict AUC

```{r untreated gene expression}


tmat <- full_tmat[,intersect(colnames(full_tmat), sing_exp$improve_sample_id)]

print('computing gene/aUC correlation')

#first focus on only durgs with 10 data points
#gvals <- md|>subset(count>10)

sing_exp <- sing_exp |> mutate(expr_time='0hr')

cor_res <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_cor(sing_exp,log2(0.0001+tmat),drug,'0hr',48), 
           compute_cor(sing_exp,log2(0.0001+tmat),drug,'0hr',120) )
    })) #end drug iter


cor_res |>
  ggplot(aes(x=corval,col=drug)) + geom_density() +
  scale_color_manual(values=drugCol) + 
  theme_bw() + facet_grid(~ttime)

```

We only have 8 samples here, but across drugs there is an enrichment of genes
that are negatively correlated with AUC, which means they are high when the drug
is MORE effective

### Correlation with known targets

We have a list of known drug targets, we can see if higher expression of those targets
is correlated with AUC

```{r cor targets}
drug.target.info <- read.csv("https://ndownloader.figshare.com/files/20237712") |>
  mutate(name = tolower(name))

sel.targs <- drug.target.info[drug.target.info$name %in% sing_exp$drug,c('name','moa','target')] |>
  unique() |>
  rbind(c('mirdametinib','MEK inhibitor','MAP2K1, MAP2K2')) |>#MEK
  rbind(c('tno155','SHP2 inhibitor','PTPN11')) |>
  rbind(c('trabectedin','','')) |>
  tidyr::separate_longer_delim(target,delim = ', ')


corplots <- lapply(unique(sing_exp$drug), function(d){
  corv <- subset(cor_res, drug == d)
  targs = subset(sel.targs, name == d)
  
  corv$target <- 'Not a target'
  corv$target[which(corv$gene %in%sel.targs$target)] <- 'Drug target'
  ggplot(corv, aes(x=corval, col=target)) + geom_density() + ggtitle(paste0(d, ' target correlation'))
  
})
#subset(cor_res,gene%in%sel.targs$target)|>View()

```

### Other promising correlated features

Now that we have the data we can look at its properties. 

Significance code doesn't work yet
```{r cor sig, eval=FALSE}
print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(gvals$drug), function(drug){
      rbind(compute_sig(exp,tmat,drug,'8hr','48'),
            compute_sig(exp,tmat,drug,'24hr','48'),
            compute_sig(exp,tmat,drug,'8hr','120'),
            compute_sig(exp,tmat,drug,'24hr','120'))
    })) #end drug iter


```

Now we can summarize and calculate how many genes are significantly correlated 
by drug, time point. 

```{r number of genes}

cor_res <- cor_res |>
  mutate(dir=ifelse(corval<0,'neg','pos'))

##here is the count of correlates, based on 0.7 for the moment
cor_counts <- cor_res |>
  subset(abs(corval)>0.7) |> 
  group_by(drug,ttime,etime)|>
  summarize(gene_cor = n_distinct(gene))|>
  arrange(desc(gene_cor))

##let's look at these
ggplot(cor_counts,aes(x=drug,y=gene_cor,fill=etime)) +
         geom_bar(stat='identity',position='dodge') +
  facet_grid(ttime~.)

dir_counts <- cor_res |>
  subset(abs(corval) > 0.75) |> 
  group_by(drug,dir,ttime,etime) |>
  summarize(gene_cor = n_distinct(gene)) |>
  mutate(cor_genes = ifelse(dir == 'pos',gene_cor,-1*gene_cor)) |>
  arrange(desc(cor_genes))

##let's look at these
ggplot(dir_counts,aes(x = drug,y = cor_genes,fill = etime)) +
         geom_bar(stat='identity',position='dodge') +
  facet_grid(ttime~.)

```

## Baseline abundance pathway enrichment

Now we can look at which pathways are correlated with drug efficacy

```{r pathway, warning=FALSE, message=FALSE, eval = FALSE}
library(leapR)
library(SummarizedExperiment)
###move all data into expression set
#first format exp data

cor_mat <- cor_res |>
  dplyr::select(-dir)|>
  tidyr::unite(drug,ttime,etime,sep='_',col='condition') |>
  pivot_wider(names_from=condition,values_from=corval)|>
  tibble::column_to_rownames('gene')

df <- sing_exp |> 
  subset(!is.na(drug)) |>
  tidyr::unite(drug, treat_time, 
               col = 'experiment', sep = '_') |>
  pivot_wider(names_from = experiment,
              values_from = auc) |>
  tibble::column_to_rownames('improve_sample_id')

eset <- SummarizedExperiment(assays = list(cor = cor_mat))
data("ncipid")
data('krbpaths')


cor_enrich <- do.call(rbind,lapply(colnames(cor_mat), function(e){
  print(e)
  suppressMessages(res <- leapR(geneset = krbpaths,
                                  assay_name = 'cor',
          enrichment_method = 'enrichment_in_order',
          primary_columns = e, 
          eset = eset))
    res |>
      mutate(condition = e) |>
      tibble::rownames_to_column('pathway')
})) |>
  tidyr::separate(condition,into=c('drug','treat_time','expr_time'), sep='_')

path_stats <- subset(cor_enrich,BH_pvalue < 0.01) |>
  group_by(drug,treat_time,expr_time) |>
  summarize(numSig=n_distinct(pathway)) 

ggplot(path_stats,aes(x = drug,y = numSig,fill = treat_time)) + 
  geom_bar(position = 'dodge',stat='identity') + facet_grid(expr_time~.)


path_counts <- subset(cor_enrich,BH_pvalue < 0.01) |> 
  group_by(pathway) |>
  summarize(n_drugs=n_distinct(drug))|>arrange(desc(n_drugs))

upaths <- subset(path_counts,n_drugs==1) ## unique pathways

subset(cor_enrich, BH_pvalue < 0.01) |>
  subset(pathway%in%upaths$pathway)


```
Not sure what to do with this - lots of pathways correlated, but they're all the same.

## Correlation of treated data with delta AUC

Here we look at the genes correlated with delta AUC of a particular drug.

### Treated gene expression data

We first read in the transcriptomics data that matches the combination experiments
```{r treated exp, echo=FALSE, warning=FALSE, message=FALSE}
##read in transcritomics, focus on gene name and average counts acros
##gene isoforms

tmat <- full_tmat[,intersect(colnames(full_tmat), exp$improve_sample_id)]


```

```{r compute gene wise correlations, warning=FALSE, message=FALSE}

print('computing gene/aUC correlation')

#first focus on only durgs with 10 data points
gvals <- md |> subset(count > 10)

exp <- exp |>
dplyr::rename(auc = 'delta_auc') |>
  mutate(treat_time = as.numeric(as.character(treat_time)))

treated_cor_res <- do.call(rbind,lapply(unique(gvals$drug), function(drug){
      rbind(compute_cor(exp,tmat,drug,'8hr',48),
            compute_cor(exp,tmat,drug,'24hr',48),
            compute_cor(exp,tmat,drug,'8hr',120),            
            compute_cor(exp,tmat,drug,'24hr',120))
    })) #end drug iter



##now do enrichment
treated_cor_mat <- cor_res |>
#  dplyr::select(-dir)|>
  tidyr::unite(drug,ttime,etime,sep='_',col='condition') |>
  pivot_wider(names_from=condition,values_from=corval)|>
  tibble::column_to_rownames('gene')

treated_df <- exp |> 
  subset(!is.na(drug)) |>
  tidyr::unite(drug, treat_time, 
               col = 'experiment', sep = '_') |>
  pivot_wider(names_from = experiment,
              values_from = auc) |>
  tibble::column_to_rownames('improve_sample_id')

treated_eset <- SummarizedExperiment(assays = list(cor = treated_cor_mat))


treated_cor_enrich <- do.call(rbind,lapply(colnames(treated_cor_mat), function(e){
  print(e)
  suppressMessages(res <- leapR(geneset = krbpaths,
                                  assay_name = 'cor',
          enrichment_method = 'enrichment_in_order',
          primary_columns = e, 
          eset = treated_eset))
    res |>
      mutate(condition = e) |>
      tibble::rownames_to_column('pathway')
})) |>
  tidyr::separate(condition,into=c('drug','treat_time','expr_time'), sep='_')

treated_path_stats <- subset(treated_cor_enrich,BH_pvalue < 0.01) |>
  group_by(drug,treat_time,expr_time) |>
  summarize(numSig=n_distinct(pathway)) 

ggplot(treated_path_stats,aes(x = drug,y = numSig,fill = treat_time)) + 
  geom_bar(position = 'dodge',stat='identity') + facet_grid(expr_time~.)


treated_path_counts <- subset(treated_cor_enrich,BH_pvalue < 0.01) |> 
  group_by(pathway) |>
  summarize(n_drugs=n_distinct(drug))|>arrange(desc(n_drugs))

treated_time_counts <-  subset(treated_cor_enrich,BH_pvalue < 0.01) |> 
  group_by(pathway, drug) |>
  summarize(tp_counts = n())

treated_upaths <- subset(treated_path_counts,n_drugs==1) ## unique pathways

subset(treated_cor_enrich, BH_pvalue < 0.01) |>
  subset(pathway%in%treated_upaths$pathway) 


subset(treated_cor_enrich, BH_pvalue < 0.01) |>
  subset(pathway%in%time_counts$pathway) 

```
Not sure what to do with this? 

## Comparing baseline expression between AUC and delta AUC

Are the same pathways that predict AUC of a drug also predict it's combination efficacy?

```{r plot sig pathways}


#for each drug - how many pathways are correalted across two time points
##how about four (for the combination)

```
