---
title: "ssGSEA correlation"
author: "Sara Gosline"
date: "2026-02-28"
output: html_document
---

---
title: "Evaluate second drug pathways"
author: "Sara Gosline"
date: "2026-01-10"
output: html_document
---

With this markdown we evaluate which pathways are correlated with drug response.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../RNAseq/plottingParameters.R")
source("correlation_functions.R")

```

## Data

We have reformatted the data into three files:
1. sample file describing the samples
2. gene file with gene metadata
3. transcriptomics data with gene expression data
4. combination data with delta AUC values
5. drug data with single AUC values

Most data is now read in by `correlation_functions.R`

```{r data, message=FALSE}

##get the transcript data, this is slow
if(!exists('full_tmat'))
  full_tmat <- get_harmonized_gex()


```



### Try to regress GSEA scores with drug response

Let's first look at the untreated gene expression data and see if we can 
find out features that predict AUC

```{r untreated gene expression}


tmat <- get_gsea_scores(log2(0.001+full_tmat[,intersect(colnames(full_tmat), sing_exp$improve_sample_id)]))


library(glmnet)
#drug = 'mirdametinib'

alpha = 0.2
sig = 0.1

rres <- do.call(rbind, lapply(c(48,120), function(ttime){ #build model for each treatment time
  
  ##build model for each drug in our top counts list
  res <- do.call(rbind, lapply(unique(tcounts$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                  # sing_exp$expr_time == etime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  #df$delta_auc = as.numeric(tvec$delta_auc)

##here we do the LOO for each sample in the drug
  samp_preds <- do.call(rbind,lapply(rownames(df), function(samp){

    st <- setdiff(rownames(df), samp)
    
    dtrain <- as.matrix(df[st,])
  
    model <- glmnet(x = dtrain, y = tvec[st, 'auc'], alpha = alpha)
    coefs <-  setdiff(unique(rownames(which(coef(model) != 0, arr.ind =T))),'(Intercept)')
    preds <- predict(model, newx = df[samp,], s = sig) 
    return(c(preds = preds, features = paste(coefs, collapse=','), training_set = length(st)))
  }
))
  tvec$preds = unlist(samp_preds[,1])
  tvec$features = unlist(samp_preds[,2])
  tvec$training_set = unlist(samp_preds[,3])
  return(tvec)
  }))
 return(res) 
}))

rres$error = (as.numeric(rres$preds)-rres$auc)^2

```



### Model performance
The results show minimal error at 48 hours, with greater error at 120 hrs.

```{r pred results}


##get ranking to make plotting prettier
rs <-  rres |> group_by(drug, .add=TRUE, .drop=FALSE) |> 
  summarize(mse = mean(error)) |>
   arrange(mse)

#join ranking
rres <- rres |> left_join(rs)

##plot
ggplot(rres, aes(x=reorder(drug, mse), y = error, col = drug)) + 
  geom_boxplot() +
  geom_jitter() + 
  facet_grid(~treat_time)+
    scale_color_manual(values=drugCol) + 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 

#subset(rres,drug != 'iag933') 
rres |>
ggplot(aes(x=reorder(drug, mse), y = error, col = drug)) + 
  geom_boxplot() +
  geom_jitter() + 
  facet_grid(~treat_time)+
    scale_color_manual(values=drugCol) + 
    theme_bw()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(x='Drug', y='Cross validation error')

ggsave('fig3a_cv_gsea_error.pdf')
```
Now we can build full models and predict on second time point

```{r full model}

ttime = 48
early_models <- lapply(unique(tcounts$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  model <- glmnet(x = df, y = tvec$auc, alpha = alpha)
  return(model)
})

names(early_models) <- unique(tcounts$drug)

ttime = 120
late_models <- lapply(unique(tcounts$drug), function(drug) {
  ##get the data for a specific treatment, measurement time, and drug
   tvec <- sing_exp[sing_exp$treat_time == ttime &
                   sing_exp$drug == drug,] |>
    drop_na() |>
    arrange(auc) |>
    tibble::column_to_rownames('improve_sample_id')

  samps <- intersect(colnames(tmat), rownames(tvec))
  tvec <- tvec[samps,]
  df = t(tmat[,samps])#data.frame(apply(tmat[, samps],1,log2))
  model <- glmnet(x = df, y = tvec$auc, alpha = alpha)
  return(model)
})

names(late_models) <- unique(tcounts$drug)
```


### Combo gex data

```{r combo gex data}

ttmat <- get_gsea_scores(log2(0.001+full_tmat[,intersect(colnames(full_tmat), exp$improve_sample_id)]))

```

### Combo model evaluation


Now we can try to use these values to predict AUC shift. Keep in mind that
the model predicts AUC, and the AUC shift is between -1 and 1,with 1 meaning
MORE drug sensitivity. 


```{r applying single drug models}

exp <- exp |> subset(!is.na(drug))



early_cpreds <- do.call(rbind, lapply(names(early_models), function(d){
  mod <- early_models[[d]]
  #feats <- intersect(rownames(mod$beta),rownames(stmat))
  pval <- predict(mod, newx = t(ttmat), s = sig)#[feats,])
  
  return(data.frame(drug = d, improve_sample_id = as.numeric(colnames(ttmat)), pred = pval[,1], pred_time = 48))
}) )

late_cpreds <- do.call(rbind, lapply(names(late_models), function(d){
  mod <- late_models[[d]]
  #feats <- intersect(rownames(mod$beta),rownames(stmat))
  pval <- predict(mod, newx = t(ttmat), s = sig)#[feats,])
  
  return(data.frame(drug = d, improve_sample_id = as.numeric(colnames(ttmat)), pred = pval[,1], pred_time = 120))
}) )


dpreds <- rbind(late_cpreds, early_cpreds) |>  left_join(samples) |>
  tidyr::separate(other_id, into=c('samp','initial_drug','time','treat','type'),sep='_') |>
  dplyr::select(drug,improve_sample_id, samp, initial_drug,time,treat, pred, pred_time ) |>
  distinct() |>
  mutate(initial_drug = tolower(initial_drug))

```

We can now plot the results by drug, the time at which they were predicted to have
an effect, and the expression measurement used to predict. 


```{r plot prediction results}
ggplot(dpreds, aes(x = pred, fill = drug)) + 
  geom_histogram(position = 'dodge') +
  scale_fill_manual(values=classCol) + facet_grid(~pred_time)


ggplot(dpreds, aes(x = drug, y = pred, col = initial_drug, shape = time)) + 
  geom_jitter() +
  scale_color_manual(values=classCol) +
  facet_grid( ~ pred_time) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  labs(y='Predicted effect')
ggsave('fig2d_predictedEffect.pdf')

```

### Top combos predicted
```{r sample combos}

dpreds$combination <- apply(dpreds, 1, function(x) paste(sort(c(x[['drug']],x[['initial_drug']])), collapse='+'))

dpreds$oppPred = 1-dpreds$pred


epreds <- subset(dpreds, pred_time == 48)
cmeans <- epreds |> group_by(combination, comboMean = mean(oppPred))

subset(epreds, initial_drug != drug) |>
  left_join(cmeans) |>
  subset(time = '8hr') |>
  ggplot(aes(x = samp, y = reorder(combination,comboMean), fill = oppPred)) + 
  geom_tile() +#facet_grid( ~ pred_time) + 
  labs(y='Combination') + scale_fill_continuous(high='turquoise4', low ='orchid4')


subset(epreds, initial_drug != drug) |>
  subset(initial_drug %in% exp$initial_drug & drug %in% exp$drug)|>
  left_join(cmeans) |>
  subset(time = '8hr') |>
  ggplot(aes(x = samp, y = reorder(combination,comboMean), fill = oppPred)) + 
  geom_tile() +#facet_grid( ~ pred_time) + 
  labs(y='Combination') + scale_fill_continuous(high='turquoise4', low ='orchid4')

ccounts <- epreds |> group_by(combination, pred_time, time) |> 
  summarize(meanAUC = mean(pred), numSamps = n()) |>
  tidyr::separate(combination, remove = FALSE, into=c('drug1','drug2'), sep =',')

pcounts <- epreds |> group_by(initial_drug, drug, time, pred_time) |>
  summarize(meanAUC = median(pred), numSamps = n())

mpcounts <-pcounts |> group_by(drug) |> summarize(mma =mean(meanAUC))

pcounts|> 
  left_join(mpcounts) |># subset(drug1 != drug2) |>
  #subset(numSamps > 2) |>
#  subset(pred_time == 120) |>
  ggplot(aes(x=initial_drug, y=reorder(drug,mma))) +
         scale_fill_viridis_c()+
  geom_tile(aes(fill=meanAUC))+facet_grid( ~ time) +
   theme_minimal()+
      theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) 
```
