---
title: "Gene Expression Summary Stats"
author: "Sara gosline"
date: "2024-12-11"
output: html_document
editor_options: 
  chunk_output_type: console
---


The goal of this document is to pull all RNASeq files from synapse in a metadata-aware fashion and do the basic analysis needed for figure 1 and supplemental figure 1.

Currently we have:

MN-2
| batch | date | treatment |
| 1 | 10/17/23 | DMSO, mirda |
| 2 | 12/19/23 | DMSO, tram, selu, TNO, ribo, palbo, RMC, vorin |
| 2 | 12/27/23 | DMSO, decitabine, olap, trab, doxo, ifo, cap |

JH-2
| batch | date | treatment |
| 3 | A | DMSO, MIR, PAL, RIB, RMC, SEL, TNO, TRAM, vorin |
| 3 | B | DMSO, DEC, IFO, IRI, OLA, TRAB |
| 3 | C | CAP, DOX | # no DMSO?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synapser)
library(dplyr); library(tidyr); library(stringr); library(RColorBrewer)
library(readr);library(readxl)
synLogin()


library(RColorBrewer)
n <- 15
cols=c(brewer.pal(8,'Dark2'),brewer.pal(n-8,'Set2'))
 
pie(rep(1,n), col=sample(cols, n))

drugnames<-list(DMSO='DMSO',MIR='Mirdametinib',CAP='Capmatinib',DEC='Decitabine',IFO='Ifosfamide',
                   OLA='Olaparib',PAL='Palbociclib',RIB='Ribociblib',TNO='TNO155',VOR='Vorinostat',
                   RMC='RMC4630',SEL='Selumetinib',TRA='Trametinib',TRAB='Trabectidin',DOX='Doxorubicin')

names(cols)<-unlist(drugnames)
   
tcolors=c(`8h`='white',`24h`='darkgrey')
  
```

### RNASeq data from synapse.
Created a helper function to get RNASeq data. Currently this uses the file names for metadata. The goal is to update with metadata from synapse directly. 

```{r rnaseq}
##currently this function pulls all data mannually and parses the file names. 
##TODO: udpate using file views
rnaSeqTab<-getRNASeq<-function(){
  ##metadata query first queries file view for all files - in this case i do counts.sf files, but you can also query by data type
  manifest <- synapser::synTableQuery("select * from syn52369043 where name like '%quant.genes.sf'")|>#$asDataFrame()|>
    as.data.frame()|>
    dplyr::select(name,id,specimenID)
  
 ##not sure what the 'samples' are here so isolating them out

 #samples<-manifest$name[grep('sample',manifest$name)]
 #mnt<-manifest$name[grep('MN-2',manifest$name)]

 batch1metadata<-readr::read_csv(synGet('syn64608368')$path)|>
   dplyr::select(specimenID,individualID,Drug,Timepoint,Full_name,Date,Proceeded_id)|>
   distinct()
 batch2metadata<-readr::read_csv(synGet('syn66302373')$path)|>
   dplyr::select(specimenID,individualID,Drug,Timepoint,Full_name,Date,Proceeded_id)|>
   distinct()
 batch3metadata <- read.csv(synGet("syn66050299")$path)|> # was syn6559564 before 20250415
   dplyr::select(specimenID,individualID,Drug,Timepoint,Full_name,Date,Proceeded_id)|>
   distinct()
 batch3metadata$specimenID <- sub("-", "_", batch3metadata$specimenID)
 batch4metadata<-readxl::read_excel(synGet('syn68080392')$path)|>
   dplyr::select(`Label on Cap`,Drug,Timepoint,`Individual ID`)|>
   distinct()
 batch4metadata$individualID <- batch4metadata$`Individual ID`
 batch4metadata$Timepoint <- sub("HR","h",batch4metadata$Timepoint)
 batch4metadata$Timepoint <- sub(" h","h",batch4metadata$Timepoint)
 colnames(batch4metadata)[1] <- "specimenID"
 batch4metadata$name <- paste0(batch4metadata$specimenID,"_quant.genes.sf")
 batch4metadata$Date <- "B4"
 batch4metadata$Full_name <- batch4metadata$Drug
 
 ##METDATA - replace with annotations 
 mn2_b1<-manifest|>#subset(name%in%mnt)|>
    dplyr::right_join(batch1metadata)|>
   dplyr::select(id,specimenID,individualID,Drug,Full_name,Timepoint,Date)
 
 mn2_b2<-manifest|>
   tidyr::separate(name,into=c('Proceeded_id','sf'),sep='_quant.genes.')|>
   dplyr::select(-specimenID)|>
   dplyr::right_join(batch2metadata)|>
    dplyr::select(id,specimenID,individualID,Drug,Full_name,Timepoint,Date)
 
 jh2002_b3 <- manifest|>
   dplyr::filter(stringr::str_starts(name, "B3_"))
 jh2002_b3$specimenID <- sub(".quant.genes.sf","",jh2002_b3$name) # filenames used to end in _quant.genes.sf before April 2, 2025
 jh2002_b3 <- merge(jh2002_b3, batch3metadata, by="specimenID", all=TRUE) |> 
   dplyr::select(id,specimenID,individualID,Drug,Full_name,Timepoint,Date)
 
 wu225_b4 <- manifest|>
   dplyr::filter(stringr::str_starts(name, "B4-"))
 wu225_b4$individualID <- "WU-225"
 wu225_b4$specimenID <- sub("_quant.genes.sf","",wu225_b4$name)
 wu225_b4 <- merge(wu225_b4, as.data.frame(batch4metadata), by=c("name","specimenID","individualID"), all=TRUE) |> 
   dplyr::select(id,specimenID,individualID,Drug,Full_name,Timepoint,Date)
   
 
 allmeta<-rbind(mn2_b1,mn2_b2,jh2002_b3,wu225_b4)
 if (any(is.na(allmeta$id))) {
   warning("missing synIDs for ", paste(allmeta[is.na(allmeta$id),]$specimenID, collapse=" ")) 
   allmeta <- allmeta[!is.na(allmeta$id),]
 }
 
 mtab<-do.call(rbind,lapply(allmeta$id,function(x){
   readr::read_tsv(synapser::synGet(x)$path)|>
      dplyr::mutate(id=x)
  }))|>left_join(allmeta)
 mtab$Drug <- sub(" ","",mtab$Drug)
 
 ##was this removed? 
 

   #mtab$Drug<-unlist(sapply(mtab$Drug,function(x) return(drugnames[[x]])))
  # 
   
  
 return(mtab)

}
   #tidyr::separate(name,sep='_',into=c('sample','tumor','batch','time','drugrep','dbatch'),remove=FALSE)|>
   #tidyr::separate(drugrep,sep='[0-9]',into=c('drug'))|>
   #tidyr::separate(time,sep='h',into=c('time'))
 #mn2$day=''

 # mn2$date<-'12-27-23'
 # mn2$date[mn2$dbatch=='d1']<-'12-19-23' #arbitrary, might need to switch
 # mn2$date[mn2$drug%in%c('TRA','SEL','RIB','RMC','VOR','PAL')]<-'12-19-23'
 # 
 # mmeta<-mn2|>
 #   dplyr::select(id,name,sample,time,drug,date)
 # 
 # 
 # ##METADATA replce with annotations
 # other<-manifest|>
 #   subset(name%in%samples)|>
 #   tidyr::separate(name,into=c('sample','timedrugrep','rest'),remove=FALSE)|>
 #   dplyr::mutate(tumor='mpnst')
 # other$time='0'
 # other$drug<-'DMSO'
 # other$time[grep("8",other$timedrugrep)]<-'8'
 # other$time[grep("24",other$timedrugrep)]<-'24'
 # other$drug[grep("M",other$timedrugrep)]<-'MIRDA'
 # other$date<-'10-17-23'
 # 
 # meta<-other|>
 #   dplyr::select(id,name,sample,time,drug,date)|>
 #   rbind(mmeta)
 # 
 # ##now get data and join with metadata
 #  mtab<-do.call(rbind,lapply(manifest$id,function(x){
 #   readr::read_tsv(synapser::synGet(x)$path)|>
 #     dplyr::mutate(id=x)
 # }))|>left_join(meta)
 #  

   
  


```

### Collect latest data

We can now call our RNASeq helper function defined above to get data and visualize overall. Currently this pulls all data with a `quants.genes.sf` file but can be adjusted. 


```{r data, echo=FALSE,warning=FALSE,message=FALSE}

rnaseqdata<-rnaSeqTab()|>
  tidyr::separate(Name,into=c('ensembl_gene_id','version'),remove=FALSE)

# change "MN2" to "MN-2"
rnaseqdata[rnaseqdata$individualID=="MN2",]$individualID <- "MN-2"

##swap batch 2 issues - DMSO time points
#needswap8<-intersect(intersect(which(rnaseqdata$drug=="DMSO"),which(rnaseqdata$date=='12-19-23')),which(rnaseqdata$time=='8'))
#needswap24<-intersect(intersect(which(rnaseqdata$drug=="DMSO"),which(rnaseqdata$date=='12-19-23')),which(rnaseqdata$time=='24'))

#rnaseqdata[needswap8,'time']='24'
#rnaseqdata[needswap24,'time']='8'

print(paste("Got RNASeq from",length(unique(rnaseqdata$id)),'samples across',length(unique(rnaseqdata$Drug)),'drugs and',length(unique(rnaseqdata$Timepoint)),'timepoints'))

if (file.exists('map.csv')) {
  map <- read.csv('map.csv')
} else {
  library(biomaRt)
  ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
  map<-getBM(attributes=c('ensembl_gene_id','hgnc_symbol'), filters ='ensembl_gene_id', values =unique(rnaseqdata$ensembl_gene_id), mart = ensembl)
  write.csv(map, 'map.csv', row.names = FALSE) 
}


rnaseqgenes<-rnaseqdata|>left_join(map)|>
  subset(hgnc_symbol!="")|>
  dplyr::group_by(hgnc_symbol,Date,id,Drug,Timepoint,individualID)|>
  summarize(TPM=sum(TPM),NumReads=sum(NumReads))

readr::write_csv(rnaseqdata,file='rnaseqcounts.csv')
readr::write_csv(rnaseqgenes,file='rnaseqcounts_hgnc.csv')

synFile <- synapser::File('rnaseqcounts.csv', parent = 'syn64004220')
synapser::synStore(synFile)
synFile <- synapser::File('rnaseqcounts_hgnc.csv', parent = 'syn64004220')
synapser::synStore(synFile)

```


Here we do some pre-processing by removing lowly abundant transcripts and then plot to PCA.

Open questions:
- are there any outliers to remove?
- are there any sample swaps here?

```{r plots and normaliztion}
library(ggplot2)
#rnaseqdata <- read.csv(synapser::synGet("syn64398049")$path)
#rnaseqgenes <- read.csv(synapser::synGet("syn64398051")$path)
zeroes<-rnaseqdata|>
  subset(NumReads<3)|>
  group_by(Name)|>
  summarize(zvals=n())

samps<-length(unique(rnaseqdata$id))
toremove<-subset(zeroes,zvals>samps/2)

reduced.data<-subset(rnaseqdata,!Name%in%toremove$Name)#|>
  #subset(id!='syn64326059') ## possible outlier??

##first do some pretty PCA
tpm<-reduced.data|>
  dplyr::select(Name,id,TPM)|>
  tidyr::pivot_wider(names_from=id,values_from=TPM)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

numReads<-reduced.data|>
  dplyr::select(Name,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

sampmeta<-rnaseqdata|>
  dplyr::select(id,specimenID,Timepoint,Drug,Date,individualID)|>
  distinct()
sampmeta$indDate <- paste0(sampmeta$individualID, ", ",sampmeta$Date)

pcs<-prcomp(log2(0.0001+tpm),scale=TRUE,center=TRUE)

pcvals<-pcs$rotation|>
  as.data.frame()|>
  dplyr::select(PC1,PC2)|>
  tibble::rownames_to_column('id')|>
  left_join(sampmeta)
pcvals$Timepoint <- factor(pcvals$Timepoint, c("0h", "8h", "24h"))

 ggplot(pcvals,aes(x=PC1,y=PC2,col=Drug,shape=Timepoint))+geom_point(size=4)+
   facet_grid(~indDate)+scale_color_discrete()
 ggplot(pcvals,aes(x=PC1,y=PC2,col=Drug,shape=Timepoint))+geom_point(aes(size=4))+facet_grid(~Date)+scale_color_manual(values=cols)

 ggsave('batched_pca.png',width=12,height=6)
 
 ggplot(pcvals,aes(x=PC1,y=PC2,col=Drug,shape=Timepoint))+geom_point(size=4)+
   facet_grid(~individualID)+scale_color_discrete()
 ggsave('batched_pca_noDates.png',width=12,height=6)
 
  ggplot(pcvals,aes(x=PC1,y=PC2,col=Drug,alpha=Timepoint, shape=indDate))+geom_point(size=4)+
   scale_alpha_discrete(range=c(0.1,1))+scale_color_discrete()
 ggsave('batched_pca_noFacet.png',width=8,height=8)
 
 drug.info <- list("Palbociclib" = "CDK inhibitor", "Ribociclib" = "CDK inhibitor",
                  "Trabectedin" = "Chemotherapy", "Ifosfamide" = "DNA alkylating agent",
                  "Decitabine" = "DNMT inhibitor", "Vorinostat" = "HDAC inhibitor",
                  "Mirdametinib" = "MEK inhibitor", "Selumetinib" = "MEK inhibitor",
                  "Trametinib" = "MEK inhibitor", "Capmatinib" = "MET inhibitor",
                  "Olaparib" = "PARP inhibitor", "RMC4630" = "SHP2 inhibitor",
                  "TNO155" = "SHP2 inhibitor", "Doxorubicin" = "TOP inhibitor",
                  "Irinotecan" = "TOP inhibitor", "Verteporfin" = "YAP inhibitor")
 drugCol <- colorRampPalette(RColorBrewer::brewer.pal(length(drug.info),"Set3"))(length(drug.info)) # need color ramp for >12 colors
 drugCol <- c("black", drugCol)
names(drugCol) <- c("DMSO", names(drug.info))
 ggplot(pcvals,aes(x=PC1,y=PC2,col=Drug,alpha=Timepoint, shape=individualID))+geom_point(size=4)+
   scale_alpha_discrete(range=c(0.2,1))+scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol))
 ggsave('batched_pca_noFacet_noDates.png',width=8,height=8)
  ggplot(pcvals,aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol))
 ggsave('batched_pca_noFacet_noDates_sizeTime.png',width=8,height=8)
 
 ggplot(pcvals[pcvals$Date %in% c("B", "C"),],aes(x=PC1,y=PC2,col=Drug,alpha=Timepoint, shape=indDate))+geom_point(size=4)+
   scale_alpha_discrete(range=c(0.1,1))+scale_color_discrete()
 ggsave('batched_pca_noFacet_datesBC.png',width=8,height=6)
 
```

### Batch correction

Clearly there is a batch effect so it will be good to correct for this to enable facile comparisons across datasets. 

TODO: put the SVA code here and upload a corrected file to synapse

```{r batch correct, echo=TRUE, warning=FALSE, message=FALSE}
#library(sva)
##now to remove the batches for date while still correcting for drug and time
library("DESeq2")
library('limma')

counts<-reduced.data|>
  dplyr::select(Name,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()


sampmeta$Date=as.factor(sampmeta$Date)
dds<-DESeqDataSetFromMatrix(countData = round(counts),
                              colData = sampmeta,
                              design = ~ Drug+Date+Timepoint )



dds<-DESeq(dds)

#lets look at both rlog and vsd as transformations
vsd <- vst(dds, blind=FALSE)

mat <- assay(vsd)
mm <- model.matrix(~Timepoint+Drug, colData(vsd))
mat <- limma::removeBatchEffect(mat, batch=vsd$Date, design=mm)
assay(vsd) <- mat


plotPCA(vsd,intgroup='Timepoint')
ggsave('bcorrected_pca_time.pdf')

plotPCA(vsd,intgroup='Date')
ggsave('bcorrected_pca_date.pdf')

cols=c("black",brewer.pal(8,'Dark2'),brewer.pal(15-8,'Set2'))

plotPCA(vsd,intgroup='Drug')+
  scale_color_manual(values=drugCol,breaks=names(drugCol))+theme_minimal()+
  labs(color="Drug")
ggsave('bcorrected_pca_drug.pdf', width=5,height=5)

pca.df <- plotPCA(vsd,intgroup=c('Drug','individualID','Timepoint','Date'),returnData=TRUE)
pca.df$Drug <- droplevels(pca.df$Drug) # didn't work
pca.df$Drug <- factor(pca.df$Drug, levels=c("DMSO",names(drug.info)))

#pca.df[grepl("Trabectidin",pca.df$group),]$Drug <- "Trabectedin"
pca.df[grepl("DMSO",pca.df$group),]$Drug <- "DMSO"
pca.df$Timepoint <- factor(pca.df$Timepoint, c("0h","8h","24h"))
ggplot(pca.df,aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol))
 ggsave('batchCorrected_pca_noFacet_noDates_sizeTime.png',width=8,height=8)
 

 ggplot(pca.df,aes(x=PC1,y=PC2,col=Drug,shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() + facet_wrap(.~Timepoint) +
   scale_color_manual(values=drugCol, breaks=names(drugCol))
 ggsave('batchCorrected_pca_noFacet_noDates_facetTime_diffColors.pdf',width=10,height=8)
 
 ggplot(pca.df[pca.df$Drug!="Irinotecan",],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol)) + scale_x_continuous(limits=c(-40,40)) +
   scale_y_continuous(limits=c(-40,40))
 ggsave('batchCorrected_pca_noFacet_noDates_sizeTime_diffColors.pdf',width=8,height=8)
 
 ggplot(pca.df[pca.df$Timepoint=="0h",],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol)) + scale_x_continuous(limits=c(-40,40)) +
   scale_y_continuous(limits=c(-40,40))
 ggsave('batchCorrected_pca_noFacet_noDates_sizeTime_diffColors_t0.pdf',width=8,height=8)
 
 ggplot(pca.df[pca.df$Timepoint!="24h",],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol)) + scale_x_continuous(limits=c(-40,40)) +
   scale_y_continuous(limits=c(-40,40))
 ggsave('batchCorrected_pca_noFacet_noDates_sizeTime_diffColors_t0or8.pdf',width=8,height=8)
 
 mpnst <- na.omit(unique(pca.df$individualID))
 for (m in mpnst) {
   ggplot(pca.df[pca.df$individualID==m,],aes(x=PC1,y=PC2,col=Drug,shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() + facet_wrap(.~Timepoint) + scale_shape_manual(values=c(16,17), breaks=c("JH-2-002","MN-2")) +
   scale_color_manual(values=drugCol, breaks=names(drugCol))
 ggsave(paste0(m,'_batchCorrected_pca_noFacet_noDates_facetTime_diffColors.pdf'),width=10,height=8)
 
    ggplot(pca.df[pca.df$individualID==m,],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol)) + scale_x_continuous(limits=c(-40,40)) + scale_shape_manual(values=c(16,17), breaks=c("JH-2-002","MN-2")) +
   scale_y_continuous(limits=c(-40,40))
 ggsave(paste0(m,'_batchCorrected_pca_noFacet_noDates_sizeTime_diffColors.pdf'),width=8,height=8)
 
 ggplot(pca.df[pca.df$Timepoint=="0h" & pca.df$individualID==m,],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol)) + scale_x_continuous(limits=c(-40,40)) +scale_shape_manual(values=c(16,17), breaks=c("JH-2-002","MN-2")) +
   scale_y_continuous(limits=c(-40,40))
 ggsave(paste0(m,'_batchCorrected_pca_noFacet_noDates_sizeTime_diffColors_t0.pdf'),width=8,height=8)
 
 ggplot(pca.df[pca.df$Timepoint!="24h" & pca.df$individualID==m,],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
   scale_color_manual(values=drugCol, breaks=names(drugCol)) + scale_x_continuous(limits=c(-40,40)) +scale_shape_manual(values=c(16,17), breaks=c("JH-2-002","MN-2")) +
   scale_y_continuous(limits=c(-40,40))
 ggsave(paste0(m,'_batchCorrected_pca_noFacet_noDates_sizeTime_diffColors_t0or8.pdf'),width=8,height=8)
 }
 # ggplot(pca.df[pca.df$time=="0h",],aes(x=PC1,y=PC2,col=Drug,size=Timepoint, shape=individualID))+geom_point()+
 #   scale_color_discrete() + labs(shape="MPNST") + theme_minimal() +
 #   scale_color_manual(values=cols, breaks=names(drugCol))
 # ggsave('batchCorrected_pca_noFacet_noDates_sizeTime_diffColors.png',width=8,height=8)

#plotPCA(vsd,intgroup='Drug')+scale_color_manual(values=cols)
#ggsave('bcorrected_pca_drug.pdf')


##cortab
md=colData(vsd)[,c('Drug','Timepoint','Date')]
corvsd=cor(assay(vsd),use='pairwise.complete.obs')
pheatmap::pheatmap(corvsd,annotation_row=md,annotation_col=md)

##now add the VSD counts to the table
vtab<-assay(vsd)|>as.data.frame()|>
  tibble::rownames_to_column('Name')|>
  tidyr:::pivot_longer(starts_with('syn'),names_to='id',values_to='VSD_batchCorrected')

rnaseqdata2<-rnaseqdata|>
  left_join(vtab)


rnaseqgenes<-rnaseqdata2|>left_join(map)|>
  subset(hgnc_symbol!="")|>
  dplyr::group_by(hgnc_symbol,Date,id,Drug,Timepoint)|>
  summarize(TPM=sum(TPM),NumReads=sum(NumReads),VSD_batchCorrected=sum(VSD_batchCorrected))


###now we add in the corrected VSD values as well
readr::write_csv(rnaseqdata2,file='rnaseqcounts.csv')
readr::write_csv(rnaseqgenes,file='rnaseqcounts_hgnc.csv')

synFile <- synapser::File('rnaseqcounts.csv', parent = 'syn64004220')
synapser::synStore(synFile)
synFile <- synapser::File('rnaseqcounts_hgnc.csv', parent = 'syn64004220')
synapser::synStore(synFile)

```
### Differential expression

Let's compute the differential expression of each drug/time point compared with DMSO. This analysis is at the core of what we're doing - trying to determine what changes are happening at each time point. 

- Which drugs are causing the greatest changes?
- Which time points are causing the greatest changes?
- TBD: how do these changes vary across samples?

```{r diffex, echo=FALSE,warning=FALSE,message=FALSE}
library("DESeq2")
counts<-reduced.data|>
  dplyr::select(Name,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

diffexCalls<-do.call(rbind,lapply(unique(sampmeta$Date),function(dval){
  cat("Running diffexp for", dval,"\n")
  if (dval == "C") { # use DMSO from group B for group C
    datemeta <- na.omit(sampmeta[sampmeta$Date == dval | 
                           (sampmeta$Drug == "DMSO" & sampmeta$Date == "B"),])
  } else {
   datemeta<-subset(sampmeta,Date==dval) 
  }
  mpnst <- unique(datemeta$individualID) # no need to filter for MPNST because only 1 MPNST per date
  if ("DMSO" %in% datemeta$Drug) {
    temp.drugs <- unique(datemeta$Drug)
    temp.drugs <- temp.drugs[temp.drugs != "DMSO"]
    ##check for each drug at the date
    do.call(rbind,lapply(temp.drugs,function(d_name){
      cat(d_name,"\n")
      ##check for each time point at the drug/date
      do.call(rbind,lapply(c('8h','24h'),function(tval){
        cat(tval,"\n")
        samps <- subset(datemeta, Timepoint==tval)|>
          subset(Drug%in%c(d_name,"DMSO"))|>
          subset(id%in%colnames(counts))
        samps$Drug <- factor(samps$Drug, levels=c("DMSO",d_name)) # make sure contrast is drug vs. dmso per DESeq2 vignette
        cts<-counts[,intersect(samps$id,colnames(counts))]
        dds <- DESeqDataSetFromMatrix(countData = round(cts),
                                      colData = samps,
                                      design = ~ Drug)
        
        dres<-DESeq(dds)|>
          results(contrast=c("Drug",d_name,"DMSO"))|> # double check contrast is drug vs. dmso per DESeq2 vignette
          as.data.frame()|>
          tibble::rownames_to_column('Gene')|>
          dplyr::mutate(Drug=d_name,Timepoint=tval,Date=dval, individualID=mpnst)
        return(dres)
      })) 
    })) 
  }
}))
# note: batch 3 date C doesn't have DMSO so DOX and CAP for JH-2-002 date C cannot be evaluated
# unless we decide to use the DMSO samples from dates A & B

diffex<-diffexCalls|>
  mutate(signif=padj<0.05)|>
  mutate(direction=ifelse(log2FoldChange>0,'up','down'))


dgenes<-diffex|>
  tidyr::separate(Gene,sep="\\.",into=c('ensembl_gene_id','Version'))

diffex_gene<-dgenes|>left_join(map)
##let's save this back to synapse for future reference
readr::write_csv(diffex_gene,file='totalRNASeqDiffExtoDate.csv')
synFile <- synapser::File('totalRNASeqDiffExtoDate.csv', parent = 'syn64004220')
synapser::synStore(synFile)


sumstats<-diffex|>subset(signif)|>
  group_by(Drug,Timepoint,Date,direction,individualID)|>
  summarize(numgenes=n())
sumstats$Timepoint <- factor(sumstats$Timepoint, levels=sort(unique(sumstats$Timepoint), decreasing=TRUE))

print('now have differential expression calls')
ggplot(sumstats,aes(x=Drug,y=numgenes,fill=direction))+facet_grid(individualID~Timepoint)+geom_bar(stat='identity',position='dodge')+coord_flip()+scale_fill_viridis_d()

ggplot(sumstats,aes(x=Drug,y=Timepoint,size=numgenes,col=Drug))+geom_point()+facet_grid(individualID~direction)+
  scale_size(range=c(1,15))+scale_color_discrete()+coord_flip()

ggsave('diffExresults.png',height=8)

##which drugs have more changes at 8 hrs? which at 24?
tp<-sumstats|>tidyr::pivot_wider(names_from='Timepoint',values_from='numgenes')|>
  tidyr::replace_na(list(`24h`=0.0))
tp$maxEffect=ifelse(tp$`24h`>tp$`8h`,'24hrs','8hrs')
tp|>ungroup()|>dplyr::select(Drug,direction,maxEffect)|>
  tidyr::pivot_wider(names_from='direction',values_from='maxEffect')|>
  arrange(Drug)

print(sumstats)
# ggplot(sumstats,aes(x=direction,y=numgenes,fill=Drug))+facet_grid(~Timepoint)+geom_bar(stat='identity',position='dodge')+coord_flip()+scale_fill_manual(values=cols)
 
sumstats<- sumstats|>arrange(desc(numgenes))
levs=sumstats[which(sumstats$direction=='up'&sumstats$Timepoint=='8h'),'Drug']
sumstats$Drug=factor(sumstats$Drug,levels=levs$Drug)
 sumstats|>
   mutate(genedir=ifelse(direction=='down',-1*numgenes,numgenes))|>
 ggplot(aes(x=direction,y=numgenes,fill=Drug))+geom_bar(stat='identity',position='dodge')+scale_fill_manual(values=cols)+coord_flip()+facet_grid(~Timepoint)
 
 ggsave('diffExresults.pdf')
 
 ##which drugs have more changes at 8 hrs? which at 24?
 tp<-sumstats|>tidyr::pivot_wider(names_from='Timepoint',values_from='numgenes')|>
   tidyr::replace_na(list(`24h`=0.0))
  tp$maxEffect=ifelse(tp$`24h`>tp$`8h`,'24hrs','8hrs')
 tp|>ungroup()|>dplyr::select(Drug,direction,maxEffect)|>
   tidyr::pivot_wider(names_from='direction',values_from='maxEffect')|>
   arrange(Drug)
 
 
 ###lets look at correlation across drug patterns and samples
 dmat<-diffex|>dplyr::select(Gene,log2FoldChange,Drug,Timepoint,Date)|>
   tidyr::unite(Drug,Timepoint,Date,col='sample',sep='_')|>
   tidyr::pivot_wider(names_from=sample,values_from=log2FoldChange)|>
   tibble::column_to_rownames('Gene')
 
 cmat<-cor(dmat,use='pairwise.complete.obs',method='spearman')
 meta<-diffex|>dplyr::select(Drug,Timepoint,Date)|>
   distinct()|>
   tidyr::unite(Drug,Timepoint,Date,col='sample',sep='_',remove=FALSE)|>
   tibble::column_to_rownames('sample')
 meta$Date=as.factor(meta$Date)
 
 pheatmap::pheatmap(cmat,annotation_row=meta,annotation_col=meta,annotation_colors = list(Drug=cols,Timepoint=tcolors),file='rnaseqCorHeatmap.pdf')

 
dgenes<-diffex|>
  tidyr::separate(Gene,sep="\\.",into=c('ensembl_gene_id','Version'))

if (file.exists('map.csv')) {
  map <- read.csv('map.csv')
} else {
  library(biomaRt)
  ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
  map<-getBM(attributes=c('ensembl_gene_id','hgnc_symbol'), filters ='ensembl_gene_id', values =unique(rnaseqdata$ensembl_gene_id), mart = ensembl)
  write.csv(map, 'map.csv', row.names = FALSE) 
}
diffex_gene<-dgenes|>left_join(map)
##let's save this back to synapse for future reference
readr::write_csv(diffex_gene,file='totalRNASeqDiffExtoDate.csv')
synFile <- synapser::File('totalRNASeqDiffExtoDate.csv', parent = 'syn64004220')
synapser::synStore(synFile)


drugscounts<-subset(dgenes,signif)|>group_by(ensembl_gene_id,Timepoint,direction)|>summarize(ndrugs=n())|>arrange(desc(ndrugs))|>
  left_join(map)|>
  subset(hgnc_symbol!='')

##most drugs just show up in one or two
ggplot(drugscounts,aes(x=ndrugs,fill=direction))+geom_bar(position='dodge')+ggtitle('Number of drugs with shared differentially expressed genes')

#reduced.genes<-reduced.data|>  
#  tidyr::separate(Name,sep="\\.",into=c('ensembl_gene_id','Version'))|>
#   left_join(map)

topgenes<-subset(drugscounts,ndrugs>9)

##TODO check for duplicates
subset(dgenes,ensembl_gene_id%in%topgenes$ensembl_gene_id)|>
   subset(signif)|>
     left_join(map)|>
     ggplot(aes(col=Drug,y=log2FoldChange,shape=signif,x=hgnc_symbol,size=1))+geom_point()+
  facet_grid(individualID~Timepoint)+coord_flip()+ggtitle("Genes that are differentially expressed across more than 9 drugs")


```
Most genes are only differentially expressed upon a single drug treatment, but some are shared with other drugs. Targeting these canonical responders might be effective, but what if we can find distinct genes changing?

Maybe we want drugs that exhibit inverse behavior for genes like UBASH3B? Uup in mirda/dox, 

Here we ask for genes only changing in one drug - is this a more reliable list?

```{r distinct genes}

ugenes<-subset(drugscounts,ndrugs==1)

subset(dgenes,ensembl_gene_id%in%ugenes$ensembl_gene_id)|>
   subset(signif)|>
  group_by(Drug,Timepoint,direction,individualID)|>
  summarize(ngenes=n())|>
  ggplot()+geom_bar(aes(x=Drug,y=ngenes,fill=direction),stat='identity',position='dodge')+
    facet_grid(individualID~Timepoint)+coord_flip()
```

Lastly, what if we only look at uniquely changing genes that are consistently changing? Are the pathways still the same? 

```{r consistent genes}
##what about genes that are consistent across 8 and 24hr time point/
ggenes<-dgenes|>subset(signif)|>group_by(Drug,direction,ensembl_gene_id)|>summarize(ntime=n())|>subset(ntime==2)

  left_join(ggenes,dgenes)|>
   subset(signif)|>
  group_by(Drug,Timepoint,direction,individualID)|>
  summarize(ngenes=n())|>
  ggplot()+geom_bar(aes(x=Drug,y=ngenes,fill=direction),stat='identity',position='dodge')+
    facet_grid(individualID~Timepoint)+coord_flip()+ggtitle("Consistently changing genes in only one drug")


```

should we target genes that are unique to a drug? and if so, do we want genes that are consistently changed across time points? What are the enrichment statistics for these genes? 
