---
title: "TF Enrichment experiments"
author: "Sara gosline"
date: "2024-12-13"
output: html_document
---

Here we take the differential expression and use it to infer transcrition factors

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#reticulate::use_virtualenv('r-reticulate')
library(synapser)
library(plyr);library(dplyr)
synLogin()

counts<-readr::read_csv(synapser::synGet('syn64398051')$path)
diffex<-read.csv(synapser::synGet('syn64397743')$path) # vector memory error if use readr
if (any(diffex$individualID == "skipped C due to lack of DMSO control\n")) {
  diffex <- diffex[diffex$individualID != "skipped C due to lack of DMSO control\n",]
}
diffex$stat <- as.numeric(diffex$stat)
if (any(diffex$Drug == "Ribociblib")) {
  diffex[diffex$Drug == "Ribociblib",]$Drug <- "Ribociclib"
}
counts_ens<-readr::read_csv(synapser::synGet('syn64398049')$path)

```

## Decoupler analysis 
Here we use the decoupler package to try to infer transcription factor activity from known databases

```{r decoupleR analysis}
library(decoupleR)
library(OmnipathR)
library(tibble)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(ggrepel)

if (file.exists('map.csv')) {
  map <- read.csv('map.csv')
} else {
  library(biomaRt)
  ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
  map<-getBM(attributes=c('ensembl_gene_id','hgnc_symbol'), filters ='ensembl_gene_id', values =unique(counts_ens$ensembl_gene_id), mart = ensembl)
  write.csv(map, 'map.csv', row.names = FALSE) 
}

cmat<- counts_ens|>
  dplyr::select(ensembl_gene_id,id,NumReads)|>
  tidyr::pivot_wider(names_from='id',values_from='NumReads', values_fn=mean)|>
  tibble::column_to_rownames(var='ensembl_gene_id')

design<-counts_ens|>
  dplyr::select(id,Drug,Timepoint,individualID)|>
  distinct()


# net <- decoupleR::get_collectri(organism = 'human', 
#                                 split_complexes = FALSE)
# [OmnipathR] Failed to download `https://omabrowser.org/All/oma-species.txt` (attempt 1/3); error: vector memory exhausted (limit reached?)
# readr::write_csv(net,file='collecTRI_complexes_network.csv')
# net <- read.csv(synapser::synGet("syn64713369")$path)

# net <- decoupleR::get_collectri(organism = 'human', 
#                                 split_complexes = TRUE)
# readr::write_csv(net,file='collecTRI_genes_network.csv')
net <- read.csv(synapser::synGet("syn64713370")$path)

enet<-net|>dplyr::rename(hgnc_symbol='target')|>
  left_join(map)|>
  subset(!is.na(ensembl_gene_id))|>
  dplyr::select(source,target='ensembl_gene_id',mor)

##write net to file and store at syn64693461
# sf<-synapser::File('collecTRI_complexes_network.csv',parentId='syn64693461')
# synapser::synStore(sf)

# sf<-synapser::File('collecTRI_genes_network.csv',parentId='syn64693461')
# synapser::synStore(sf)

sample_acts <- decoupleR::run_ulm(mat = cmat, 
                                  net = enet, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'mor', 
                                  minsize = 5)

##now we have putative transcritpion factors for each condition
tfs<-sample_acts|>
  dplyr::rename(id='condition')|>
  left_join(design)
print(head(tfs))
```
Now we have tf scores
## Most variable transcription factors

Let's start by plotting the most variable transcritpion factors

```{r pressure, echo=FALSE}

vars<-tfs|>group_by(source)|>summarize(var=var(score))|>arrange(desc(var))

tf_mat<-sample_acts|>
  subset(source%in%c(vars$source[1:25]))|>
  dplyr::select(source,condition,score)|>
  pivot_wider(names_from='source',values_from='score')|>
  tibble::column_to_rownames('condition')

ddesign<-design|>
  tibble::column_to_rownames('id')|>
  mutate(Timepoint=as.factor(Timepoint))

pheatmap(tf_mat,annotation_row = ddesign,labels_row=c())


sample_acts|>
  dplyr::rename(id='condition')|>
  mutate(log10p=(-log10(p_value)))|>
  left_join(design)|>
  subset(source%in%vars$source[1:15])|>
  ggplot(aes(x=Drug,y=score,col=source,shape=as.factor(Timepoint),size=log10p))+geom_point()

readr::write_csv(sample_acts,'allTFestimatesBySample.csv')
sf<-synapser::File('allTFestimatesBySample.csv',parentId='syn64693461')
synapser::synStore(sf)

```

I think this plot is more a message of where the outliers lie rather than the actual biological value of each of the TFs. It also shows the need for batch correction. 

## Transcription factors by Drug

Here we use t statistics as input to decoupler, so that we can identify the TFs that are changing positively or negatively in each Drug treatment. 

```{r tfs by Drug,warning=FALSE, message=FALSE}
allcontrast<-do.call(rbind,lapply(unique(diffex$individualID),function(ind){
  do.call(rbind,lapply(unique(diffex$Drug),function(d){
    do.call(rbind,lapply(unique(diffex$Timepoint),function(ti){
      dtab<-subset(diffex,Timepoint==ti)|>
        subset(!is.na(stat))|>
        subset(Drug==d)|>
        subset(individualID==ind)|>
        dplyr::select(ensembl_gene_id,Drug,Timepoint,individualID,stat)|>
        group_by(ensembl_gene_id,Drug,Timepoint,individualID)|>
        summarize(stat=mean(stat,na.rm=TRUE))|>
        tibble::column_to_rownames('ensembl_gene_id')
      dtab <- na.omit(dtab)
      if (nrow(dtab)>0) {
       contrast_acts <- decoupleR::run_ulm(mat = dtab[, 'stat', drop = FALSE],
                                    net = enet,
                                    .source = 'source',
                                    .target = 'target',
                                    .mor='mor',
                                    minsize = 5)|>
        ungroup()|>
        mutate(Drug=d,Timepoint=ti,individualID=ind)
      } else {
        warning("0 rows for ", ti, d, ind)
        contrast_acts <- data.frame()
      }
      return(contrast_acts)
        }))
    }))
}))

allcontrast$padj<-p.adjust(allcontrast$p_value)
allcontrast<-allcontrast|>mutate(signif=(padj<0.01))
subset(allcontrast,padj<0.01)|>
  ggplot(aes(x=source,y=score,size=signif,col=Drug,shape=as.factor(Timepoint)))+
  geom_point()+coord_flip()+facet_grid(Timepoint~individualID)
#allcontrast[allcontrast$Drug=="Ribociblib",]$Drug <- "Ribociclib"
readr::write_csv(allcontrast,file='tfEstimatesEnsGenes.csv')
sf <-synapser::File('tfEstimatesEnsGenes.csv',parentId='syn64693461')
synapser::synStore(sf)
#allcontrast <- read.csv(synapser::synGet("syn64420968")$path)

md <- allcontrast|>
  dplyr::select(Drug,Timepoint,individualID)|>
  distinct()|>
  tidyr::unite(Drug,Timepoint,individualID,col='DrugTimepointMPNST',remove=FALSE)|>
  tibble::column_to_rownames('DrugTimepointMPNST')

mat<-allcontrast|>
  dplyr::select(source,Drug,Timepoint,individualID,score)|>
  tidyr::unite(Drug,Timepoint,individualID,col='DrugTimepointMPNST')|>
  pivot_wider(names_from=DrugTimepointMPNST,values_from=score,values_fill = 0.0)

subset(allcontrast,source%in%c('ZEB1','STAT1','EGR1','RELA'))|>
  ggplot(aes(x=Drug,y=score,size=signif,color=Drug,shape=as.factor(Timepoint)))+
  geom_point()+facet_grid(individualID~source)+coord_flip()
```

We can evaluate how the transcription factor values correlate with one another
```{r tfcors}

corval<-mat|>tibble::column_to_rownames('source')|>cor(use='pairwise.complete.obs')
pheatmap(corval,annotation_row = md,annotation_col=md)

pcs<-mat|>tibble::column_to_rownames('source')|>as.matrix()|>prcomp()

comps<-cbind(pcs$rotation[rownames(md),],md)|>unique()
comps|>
  ggplot(aes(x=PC1,y=PC2,col=Drug,shape=as.factor(Timepoint),size=4))+
  geom_point()+#ggrepel::geom_label_repel(aes(label=Drug))+
  facet_grid(~individualID)
```


### plot number of significant TFs
```{r plot # of results, echo = FALSE}
# prep data
drug.info <- list("Palbociclib" = "CDK inhibitor", "Ribociclib" = "CDK inhibitor",
                  "Trabectidin" = "Chemotherapy", "Ifosfamide" = "DNA alkylating agent",
                  "Decitabine" = "DNMT inhibitor", "Vorinostat" = "HDAC inhibitor",
                  "Mirdametinib" = "MEK inhibitor", "Selumetinib" = "MEK inhibitor",
                  "Trametinib" = "MEK inhibitor", "Capmatinib" = "MET inhibitor",
                  "Olaparib" = "PARP inhibitor", "RMC4630" = "SHP2 inhibitor",
                  "TNO155" = "SHP2 inhibitor", "Doxorubicin" = "TOP inhibitor",
                  "Irinotecan" = "TOP inhibitor", "Verteporfin" = "YAP inhibitor")
base.path <- getwd()
dir.create("TF")
tf.path <- file.path(base.path, "TF")
setwd(tf.path)
#allcontrast <- read.csv(synapser::synGet("syn64420968")$path)
allcontrast$Drug <- factor(allcontrast$Drug, levels=names(drug.info))
timepoints <- c("8h", "24h")
allcontrast$Timepoint <- factor(allcontrast$Timepoint, levels=timepoints)
individualID <- unique(allcontrast$individualID)
all.drugs <- levels(allcontrast$Drug)
all.moas <- sort(unique(unlist(drug.info)))
drugCol <- colorRampPalette(RColorBrewer::brewer.pal(length(all.drugs),"Set3"))(length(all.drugs)) # need color ramp for >12 colors
names(drugCol) <- all.drugs

sig.tf <- allcontrast[allcontrast$padj <= 0.05,]
ntfs <- plyr::ddply(sig.tf, .(individualID, Drug, Timepoint), dplyr::summarize,
                    N = length(na.omit(unique(source))))
ntfs.overall <- plyr::ddply(sig.tf, .(Drug, Timepoint), dplyr::summarize,
                    N = length(na.omit(unique(source))))

# are there more/less enriched tfs at 8h vs. 24h? No (p=0.217)
for (m in individualID) {
 cat(m, "# of tfs enriched at 8h vs. 24h\n")
  t.test(ntfs[ntfs$individualID==m & ntfs$Timepoint=="8h",]$N, 
         ntfs[ntfs$individualID==m & ntfs$Timepoint=="24h",]$N) 
}
cat("Overall", "# of tfs enriched at 8h vs. 24h\n")
t.test(ntfs.overall[ntfs.overall$Timepoint=="8h",]$N, 
       ntfs.overall[ntfs.overall$Timepoint=="24h",]$N)

ntfs$moa <- NA
ntfs.overall$moa <- NA
for (i in all.moas) {
  temp.drugs <- names(drug.info[drug.info %in% i])
  if (any(ntfs$Drug %in% temp.drugs)) {
   ntfs[ntfs$Drug %in% temp.drugs,]$moa <- i 
   ntfs.overall[ntfs.overall$Drug %in% temp.drugs,]$moa <- i
  }
}

# generate plot
ntfs$moa <- factor(ntfs$moa, levels=all.moas)
ntf.plot <- ggplot2::ggplot(ntfs, aes(x=Timepoint, y=N)) + geom_violin(alpha=0) +
  geom_point(aes(color=Drug, shape=moa)) +
  geom_boxplot(width=0.2, alpha = 0) + 
  scale_color_manual(limits=all.drugs, values=drugCol) +
  labs(x="Time (Hours)", y="# of Enriched Transcription Factors",
       color = "Drug Treatment", shape = "Mechanism") +
  scale_shape_manual(values=1:length(all.moas), breaks=all.moas) + 
  theme_classic(base_size=12) +
  theme(legend.box="horizontal") + facet_wrap(~individualID)
ggplot2::ggsave("N_enrichedTFs.pdf", ntf.plot, width=5.5, height=4)

ntfs.overall$moa <- factor(ntfs.overall$moa, levels=all.moas)
ntf.plot.overall <- ggplot2::ggplot(ntfs.overall, aes(x=Timepoint, y=N)) + geom_violin(alpha=0) +
  geom_point(aes(color=Drug, shape=moa)) +
  geom_boxplot(width=0.2, alpha = 0) + scale_color_manual(limits=all.drugs, values=drugCol) +
  labs(x="Time (Hours)", y="# of Enriched Transcription Factors",
       color = "Drug Treatment", shape = "Mechanism") +
  scale_shape_manual(values=1:length(all.moas), breaks=all.moas) + theme_classic(base_size=12) +
  theme(legend.box="horizontal")
ggplot2::ggsave("N_enrichedTFs_overall.pdf", ntf.plot.overall, width=5.5, height=4)
```

### plot all enriched TFs
```{r plot all results, echo = FALSE}
tf.path <- file.path(getwd(),"TF")
setwd(tf.path)
topSets <- unique(allcontrast[allcontrast$signif,]$source)
dot.df <- allcontrast[allcontrast$source %in% topSets,]
mean.results <- plyr::ddply(dot.df, .(source), summarize,
                            score = mean(score, na.rm=TRUE))
setOrder <- mean.results[order(mean.results$score),]$source

dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
dot.df$minusLogFDR <- -log10(dot.df$padj)
maxAbsscore <- ceiling(max(abs(dot.df$score)))
dot.plot <- ggplot2::ggplot(
  na.omit(dot.df),
  ggplot2::aes(
    x = Drug, y = source, color = score,
    size = minusLogFDR
  )
) +
  ggplot2::geom_point() +
  ggplot2::scale_y_discrete(limits = setOrder) +
  ggplot2::scale_x_discrete(limits=all.drugs) +
  scale_color_gradient2(low="blue",high="red", mid="grey", 
                        limits=c(-maxAbsscore, maxAbsscore)) +
  theme_minimal(base_size=12) +
  ggplot2::labs(color = "Score", size = "-log(FDR)") + 
  theme(axis.title=element_blank(), 
        axis.text.x=element_text(angle=45, vjust=1, hjust=1)) +
  geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
  facet_grid(Timepoint ~ individualID)
dot.plot
ggplot2::ggsave("enrichedTFs_dotPlot.pdf", dot.plot, width=12, height=49) 
```


### plot top enriched pathways
```{r plot top results, echo = FALSE}
library(plyr);library(dplyr);library(ggplot2)
tf.path <- file.path(getwd(),"TF")
setwd(tf.path)

moa.info2 <- dplyr::distinct(ntfs[,c("Drug","moa")])
# create tile plot for drug pathway annotations
  moa.plot <- ggplot2::ggplot(moa.info2, aes(x=Drug,y=1, fill=moa)) + 
  geom_tile() + scale_x_discrete(limits=all.drugs) + theme_classic(base_size=12) +
  theme(axis.ticks=element_blank(), axis.text = element_blank(), axis.title=element_blank(),
        axis.line=element_blank()) + labs(fill="Drug Mechanism")
ggplot2::ggsave("MOA_annotations.pdf", moa.plot, width=6,height=3)

n.top.list <- c(1, 2, 3, 5)
for (n.top in n.top.list) {
  ### pathway
  topSigpathway <- data.frame()
  for (i in all.drugs) {
    tempSig <- allcontrast[allcontrast$Drug == i &
                              allcontrast$signif,] %>% slice_max(abs(score), n=n.top)
    topSigpathway <- rbind(topSigpathway, tempSig)
  }
  length(unique(topSigpathway$source)) # 20
  dot.df <- allcontrast[allcontrast$source %in% unique(topSigpathway$source),]
  
  mean.results <- plyr::ddply(dot.df, .(source), summarize,
                              score = mean(score, na.rm=TRUE))
  setOrder <- mean.results[order(mean.results$score),]$source
  drugOrder <- unlist(drug.info)
  
  dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
  dot.df$minusLogFDR <- -log10(dot.df$padj)
  maxAbsscore <- ceiling(max(abs(dot.df$score)))
  dot.plot <- ggplot2::ggplot(
    na.omit(dot.df),
    ggplot2::aes(
      x = Drug, y = source, color = score,
      size = minusLogFDR
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::scale_y_discrete(limits = setOrder) +
    ggplot2::scale_x_discrete(limits=all.drugs) +
    scale_color_gradient2(low="blue",high="red", mid="grey", 
                          limits=c(-maxAbsscore, maxAbsscore)) +
    #viridis::scale_color_viridis() +
    theme_minimal(base_size=12) +
    ggplot2::labs(color = "Score", size = "-log(FDR)") + 
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
          axis.title = element_blank()) +
    geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
    facet_grid(Timepoint~individualID)
  dot.plot
  ggplot2::ggsave(paste0(n.top,"_topEnrichedTFs_dotPlot.pdf"), dot.plot, width=10, height=ifelse(n.top<2,4,2*n.top))
}

```


### plot top enriched pathways for each drug MOA
```{r plot top results per MOA, echo = FALSE}
library(plyr);library(dplyr);library(ggplot2)
tf.path <- file.path(getwd(),"TF")
setwd(tf.path)

n.top.list <- c(1, 2, 3, 5)
for (moa in all.moas) {
  temp.drugs <- names(drug.info[drug.info == moa])
  for (n.top in n.top.list) {
    ### pathway
    topSigpathway <- data.frame()
    for (i in temp.drugs) {
      tempSig <- allcontrast[allcontrast$Drug == i &
                           allcontrast$signif,] %>% slice_max(abs(score), n=n.top)
      topSigpathway <- rbind(topSigpathway, tempSig)
    }
    dot.df <- allcontrast[allcontrast$source %in% unique(topSigpathway$source) &
                        allcontrast$Drug %in% temp.drugs,]
    if (nrow(dot.df) > 0) {
      mean.results <- plyr::ddply(dot.df, .(source), summarize,
                                  score = mean(score, na.rm=TRUE))
      setOrder <- mean.results[order(mean.results$score),]$source
      drugOrder <- unlist(drug.info)
      
      dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
      dot.df$minusLogFDR <- -log10(dot.df$padj)
      maxAbsscore <- ceiling(max(abs(dot.df$score)))
      dot.plot <- ggplot2::ggplot(
        na.omit(dot.df),
        ggplot2::aes(
          x = Drug, y = source, color = score,
          size = minusLogFDR
        )
      ) +
        ggplot2::geom_point() +
        ggplot2::scale_y_discrete(limits = setOrder) +
        ggplot2::scale_x_discrete(limits=temp.drugs) +
        scale_color_gradient2(low="blue",high="red", mid="grey", 
                              limits=c(-maxAbsscore, maxAbsscore)) +
        #viridis::scale_color_viridis() +
        theme_minimal(base_size=12) +
        ggplot2::labs(color = "Score", size = "-log(FDR)") + 
        theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
              axis.title = element_blank()) +
        geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21)
      if (length(unique(dot.df$Timepoint)) > 0 & length(unique(dot.df$individualID)) > 0) {
        dot.plot <- dot.plot + facet_grid(Timepoint~individualID)
      } else if (length(unique(dot.df$Timepoint)) > 0) {
        dot.plot <- dot.plot + facet_grid(Timepoint~.)
      }else if (length(unique(dot.df$individualID)) > 0) {
        dot.plot <- dot.plot + facet_grid(.~individualID)
      }
      dot.plot
      customWidth <- ifelse(length(temp.drugs) < 3, 5, length(temp.drugs)+2)
      ggplot2::ggsave(paste0(moa, "_",n.top,"_topEnrichedTFs_dotPlot.pdf"), 
                      dot.plot, width=customWidth, height=ifelse(n.top<3,3,n.top)) 
    }
  } 
}
```


### upset plots to see overlap across drug treatments
```{r plot overlap of results, echo = FALSE}
for (m in individualID) {
  # generate list of significant results
  sigList <- list()
  for (i in all.drugs) {
    if (nrow(allcontrast[allcontrast$Drug == i & 
                          allcontrast$individualID == m & allcontrast$signif,]) > 0) {
      sigList[[i]] <- na.omit(unique(allcontrast[allcontrast$Drug == i & 
                                                    allcontrast$individualID == m & allcontrast$signif,]$source)) 
    }
  }
  
  if (length(sigList) > 1) {
    # convert to matrix
    myPlot = ComplexHeatmap::make_comb_mat(sigList)
    pdf(file.path(tf.path,paste0(m,"_enrichedTF_upsetPlot.pdf")))
    up <- ComplexHeatmap::UpSet(myPlot)
    ComplexHeatmap::draw(up)
    dev.off()
  }
}
```

```{r STRING physical interaction networks, echo=FALSE}
library(PCSF)
data("STRINGv12")
dir.create("TF_networks")
setwd("TF_networks")
for (m in mpnst) {
  for (t in timepoints) {
    for (d in tested.drugs) {
      temp.tf <- tf[tf$Drug == d & tf$Timepoint == t & tf$individualID == m & tf$signif,]
      pos.tf <- temp.tf[temp.tf$score > 0,]
      neg.tf <- temp.tf[temp.tf$score < 0,]
      
      if (nrow(pos.tf) > 0) {
        pos.edges <- STRINGv12[STRINGv12$from %in% pos.tf$source | 
                               STRINGv12$to %in% pos.tf$source,]
        pos.tf <- pos.tf[pos.tf$source %in% c(pos.edges$from, pos.edges$to),]
        pos.vert <- data.frame(unique(c(pos.edges$from, pos.edges$to)))
        colnames(pos.vert) <- "source"
        pos.vert <- merge(pos.vert, pos.tf, all.x=TRUE)
        pos.vert$type <- "Inferred"
        pos.vert[!is.na(pos.vert$score),]$type <- "Transcription Factor"
        topGraph <- igraph::graph_from_data_frame(pos.edges, directed=FALSE, vertices=pos.vert) 
      tempTitle <- paste0(m,"_",d,"_",t,"_positive_TF")
      RCy3::createNetworkFromIgraph(topGraph, title=tempTitle)
      write.csv(pos.vert, paste0(tempTitle,"_vertices.csv"), row.names=FALSE)
      write.csv(pos.edges, paste0(tempTitle,"_edges.csv"), row.names=FALSE)
      pos.centrality <- data.frame(name = V(topGraph)$name,
                                    degree = igraph::degree(topGraph, mode="all"),
                                    closeness = igraph::closeness(topGraph, mode="all"),
                                    betweenness = igraph::betweenness(topGraph, directed = FALSE),
                                    eigen_centrality = igraph::eigen_centrality(topGraph, directed = FALSE)$vector,
                                    hub_score = igraph::hub_score(topGraph)$vector,
                                    authority_score = igraph::authority_score(topGraph)$vector)
      write.csv(pos.centrality, paste0(tempTitle,"_centrality.csv"), row.names=FALSE)
      }
      
      if (nrow(neg.tf) > 0) {
        neg.edges <- STRINGv12[STRINGv12$from %in% neg.tf$source | 
                               STRINGv12$to %in% neg.tf$source,]
        neg.tf <- neg.tf[neg.tf$source %in% c(neg.edges$from, neg.edges$to),]
        neg.vert <- data.frame(unique(c(neg.edges$from, neg.edges$to)))
        colnames(neg.vert) <- "source"
        neg.vert <- merge(neg.vert, neg.tf, all.x=TRUE)
        neg.vert$type <- "Inferred"
        neg.vert[!is.na(neg.vert$score),]$type <- "Transcription Factor"
       topGraph <- igraph::graph_from_data_frame(neg.edges, directed=FALSE, vertices=neg.vert) 
      tempTitle <- paste0(m,"_",d,"_",t,"_negative_TF")
      RCy3::createNetworkFromIgraph(topGraph, title=tempTitle)
      write.csv(neg.vert, paste0(tempTitle,"_vertices.csv"), row.names=FALSE)
      write.csv(neg.edges, paste0(tempTitle,"_edges.csv"), row.names=FALSE)
      neg.centrality <- data.frame(name = V(topGraph)$name,
                                    degree = igraph::degree(topGraph, mode="all"),
                                    closeness = igraph::closeness(topGraph, mode="all"),
                                    betweenness = igraph::betweenness(topGraph, directed = FALSE),
                                    eigen_centrality = igraph::eigen_centrality(topGraph, directed = FALSE)$vector,
                                    hub_score = igraph::hub_score(topGraph)$vector,
                                    authority_score = igraph::authority_score(topGraph)$vector)
      write.csv(neg.centrality, paste0(tempTitle,"_centrality.csv"), row.names=FALSE) 
      }
    }
  }
}
```