---
title: "Priori TF Enrichment"
author: "Sara Gosline"
date: "2025-01-14"
output: html_document
---


Here we use the Prior package to run TF enrichment of the MPNST data 

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```


```{r load libs}
#library(reticulate)
#reticulate::py_install('priori_regulon-enrichment')
library(synapser)
library(dplyr)
library(readr)
synLogin()
fname=synapser::synGet('syn64398051')$path
counts<-readr::read_csv(fname)

mat <-counts|>
  dplyr::select(features=hgnc_symbol,id,TPM)|>
  tidyr::pivot_wider(names_from=id,values_from=TPM)

readr::write_tsv(mat,file='TPMMat.txt')

mdata <-counts|>
  dplyr::select(id,date,time,drug)|>distinct()

```

## Run priori

irst we use the reticulate package to navigate the python environments and run priori
```{bash priori}
priori TPMMat.txt pr_out
```


## Analyze results
Now we can read in the results from prior

```{r results, echo=FALSE}
weights <-readr::read_tsv('pr_out/priori_activity_score_weights.tsv')
scores <- readr::read_tsv("pr_out/priori_activity_scores.tsv")

print(head(scores))

##let's do PCA of the scores
pcs<-scores|>tibble::column_to_rownames('id')|>as.matrix()|>t()|>prcomp()

pcsdim<-pcs$rotation|>
  as.data.frame()|>
  tibble::rownames_to_column('id')|>
    dplyr::left_join(mdata)

library(ggplot2)

ggplot(pcsdim,aes(x=PC1,y=PC2,col=drug,shape=as.factor(time),size=5))+geom_point()+facet_grid(~date)
```

Now that we have the scores we can join them with the metadata in the initial file and evaluate them.

```{r score summaries}
colnames(scores)[1]<-'id'

lscores<-scores|>
  tidyr::pivot_longer(colnames(scores)[-1],names_to='TF',values_to='activity score')
lscores<-counts|>dplyr::select(date,id,drug,time)|>distinct()|>left_join(lscores)
print(head(lscores))
```

Now we can evaluate significantly altered scores between conditions.

```{r activity score changes}
library(limma)

scoremat<-scores|>
  tibble::column_to_rownames('id')|>
  as.matrix()|>
  t()

##can we run limma just once for all the conditions? 
limmaFromMatrix<-function(mat,samps,d){
  print(d)
  nsamps<-subset(samps,drug==d)
  
  nsamps<-subset(samps,date==nsamps$date[1])|>
    subset(drug=='DMSO')|>
    rbind(nsamps)
  nsamps$drug=as.factor(nsamps$drug)
  nsamps$time=as.factor(nsamps$time)
  
  design <- model.matrix(~0+drug+time+drug*time, data=nsamps)
  res <- lmFit(2^(mat[,nsamps$id]),design)
  res <- eBayes(res)

  fulltab<- topTable(res,coef=paste0('drug',d),number=nrow(mat))|>
      tibble::rownames_to_column('TF')|>
      mutate(drug=d)
#  rtab<-topTable(res,number=nrow(mat))
  return(fulltab)
               
}

res<-do.call(rbind,lapply(setdiff(samps$drug,'DMSO'), function(d) limmaFromMatrix(scoremat,mdata,d)))

res|>subset(adj.P.Val<0.01)|>ggplot(aes(x=drug))+geom_bar()+coord_flip()

res|>subset(adj.P.Val<0.01)|>ggplot(aes(x=TF))+geom_bar()+coord_flip()

res|>subset(adj.P.Val<.01)|>group_by(TF)|>summarize(ndrugs=n())|>arrange(desc(ndrugs))

res|>subset(adj.P.Val<.01)|>group_by(TF)|>summarize(ndrugs=n())|>arrange(ndrugs)
```

MAPK14 and PCBP1 are significant in each experiment and things like EGR3 and CXCL8 are only found in 1 treatment.  HOw do these look?

```{r test plots}


```



