---
title: "Priori TF Enrichment"
author: "Sara Gosline"
date: "2025-01-14"
output: html_document
---


Here we use the Prior package to run TF enrichment of the MPNST data 

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```


```{r load libs}
#library(reticulate)
#reticulate::py_install('priori_regulon-enrichment')
library(synapser)
library(dplyr)
library(readr)
synLogin()
fname=synapser::synGet('syn64398051')$path
counts<-readr::read_csv(fname)

mat <-counts|>
  dplyr::select(features=hgnc_symbol,id,TPM)|>
  tidyr::pivot_wider(names_from=id,values_from=TPM)

readr::write_tsv(mat,file='TPMMat.txt')

mdata <-counts|>
  dplyr::select(id,date,time,drug)|>distinct()

```

## Run priori

irst we use the reticulate package to navigate the python environments and run priori
```{bash priori}
priori TPMMat.txt pr_out
```


## Analyze results
Now we can read in the results from prior

```{r results, echo=FALSE}
weights <-readr::read_tsv('pr_out/priori_activity_score_weights.tsv')
scores <- readr::read_tsv("pr_out/priori_activity_scores.tsv")

print(head(scores))

##let's do PCA of the scores
pcs<-scores|>tibble::column_to_rownames('id')|>as.matrix()|>t()|>prcomp()

pcsdim<-pcs$rotation|>
  as.data.frame()|>
  tibble::rownames_to_column('id')|>
    dplyr::left_join(mdata)

library(ggplot2)

ggplot(pcsdim,aes(x=PC1,y=PC2,col=drug,shape=as.factor(time),size=5))+geom_point()+facet_grid(~date)
```

Now that we have the scores we can join them with the metadata in the initial file and evaluate them.

```{r score summaries}
colnames(scores)[1]<-'id'

lscores<-scores|>
  tidyr::pivot_longer(colnames(scores)[-1],names_to='TF',values_to='activity score')
lscores<-counts|>dplyr::select(date,id,drug,time)|>distinct()|>left_join(lscores)
print(head(lscores))
```

Now we can evaluate significantly altered scores between conditions.

```{r activity score changes}
library(limma)

scoremat<-scores|>
  tibble::column_to_rownames('id')|>
  as.matrix()|>
  t()

limmaFromMatrix<-function(mat,samps){
  
  design <- model.matrix(~0+drug+time+drug*time*date+date, data=samps)
  res <- lmFit(2^(mat),design)
  res <- eBayes(res)
  
  rtab<-topTable(res,number=nrow(mat))|>
    dplyr::select(AveExpr,F,P.Value,adj.P.Val)
  return(rtab)
               
}

diffexCalls<-do.call(rbind,lapply(unique(mdata$date),function(dval){
  datemeta<-subset(mdata,date==dval)|>
    subset(drug!='DMSO')
  ##check for each drug at the date
  do.call(rbind,lapply(unique(datemeta$drug),function(d_name){
    ##check for each time point at the drug/date
    print(d_name)
   do.call(rbind,lapply(c('8','24'),function(tval){
    print(tval)
    samps<-subset(mdata,date==dval)|>
       subset(time==tval)|>
       subset(drug%in%c(d_name,"DMSO"))|>
       tibble::column_to_rownames('id')
    if(length(unique(samps$drug))<2)
        return(NULL)
    mat<-scoremat[,rownames(samps)]
      ##rewrite this for limma
    dds <- limmaFromMatrix(mat,samps)
    dres<-dds|>
       as.data.frame()|>
       tibble::rownames_to_column('Gene')|>
       mutate(drug=d_name,time=tval,date=dval)
    return(dres)
   })) 
  }))
}))



```
