---
title: "Gene Expression Summary Stats"
author: "Sara gosline"
date: "2024-12-11"
output: html_document
---

The goal of this document is to pull all RNASeq files from synapse in a metadata-aware fashion and do the basic analysis needed for figure 1 and supplemental figure 1. 

Currently we have three batches:

|date | treatment |
| --- | --- |
|10/17/23 | DMSO, mirda |
|12/19/23 | DMSO, tram, selu, TNO, ribo, palbo, RMC, vorin |
|12/27/23 | DMSO, decitabine, olap, trab, doxo, ifo, cap |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synapser)
library(dplyr)
synLogin()
```


Created a helper function to get RNASeq data
```{r rnaseq}
##currently this function pulls all data mannually and parses the file names. 
##TODO: udpate using file views
rnaSeqTab<-getRNASeq<-function(){
  ##metadata query first queries file view for all files - in this case i do counts.sf files, but you can also query by data type
  manifest <- synapser::synTableQuery("select * from syn52369043 where name like '%quant.genes.sf'")|>
    as.data.frame()|>
    dplyr::select(name,id)
  
 ##not sure what the 'samples' are here so isolating them out

 samples<-manifest$name[grep('sample',manifest$name)]
 mnt<-manifest$name[grep('MN-2',manifest$name)]

 ##METDATA - replace with anotations 
 mn2<-manifest|>subset(name%in%mnt)|>
   tidyr::separate(name,sep='_',into=c('sample','tumor','batch','time','drugrep','dbatch'),remove=FALSE)|>
   tidyr::separate(drugrep,sep='[0-9]',into=c('drug'))|>
   tidyr::separate(time,sep='h',into=c('time'))
 mn2$day=''

 mn2$date<-'12-27-23'
 mn2$date[mn2$dbatch=='d1']<-'12-19-23' #arbitrary, might need to switch
 mn2$date[mn2$drug%in%c('TRA','SEL','RIB','RMC','VOR','PAL')]<-'12-19-23'
 
 mmeta<-mn2|>
   dplyr::select(id,name,sample,time,drug,date)
 

 ##METADATA replce with annotations
 other<-manifest|>
   subset(name%in%samples)|>
   tidyr::separate(name,into=c('sample','timedrugrep','rest'),remove=FALSE)|>
   dplyr::mutate(tumor='mpnst')
 other$time='0'
 other$drug<-'DMSO'
 other$time[grep("8",other$timedrugrep)]<-'8'
 other$time[grep("24",other$timedrugrep)]<-'24'
 other$drug[grep("M",other$timedrugrep)]<-'MIRDA'
 other$date<-'10-17-23'
 
 meta<-other|>
   dplyr::select(id,name,sample,time,drug,date)|>
   rbind(mmeta)
 
 ##now get data and join with metadata
  mtab<-do.call(rbind,lapply(manifest$id,function(x){
   readr::read_tsv(synapser::synGet(x)$path)|>
     dplyr::mutate(id=x)
 }))|>left_join(meta)
  
  
 
 return(mtab)

}

```

## Geta data from synapse
We can now call our RNASeq helper function defined above to get data and visualize overall

```{r data, echo=FALSE,warning=FALSE,message=FALSE}

rnaseqdata<-rnaSeqTab()

print(paste("Got RNASeq from",length(unique(rnaseqdata$id)),'samples across',length(unique(rnaseqdata$drug)),'drugs and',length(unique(rnaseqdata$time)),'timepoints'))
```

Now first we visualize the samples to assess any batch effects by date.

```{r plots and normaliztion}
library(ggplot2)

zeroes<-rnaseqdata|>
  subset(TPM==0)|>
  group_by(Name)|>
  summarize(zvals=n())

samps<-length(unique(rnaseqdata$id))
toremove<-subset(zeroes,zvals>samps/2)

reduced.data<-subset(rnaseqdata,!Name%in%toremove$Name)

##first do some pretty PCA
tpm<-reduced.data|>
  dplyr::select(Name,id,TPM)|>
  tidyr::pivot_wider(names_from=id,values_from=TPM)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

sampmeta<-rnaseqdata|>
  dplyr::select(id,sample,time,drug,date)|>
  distinct()

pcs<-prcomp(log2(0.0001+tpm),scale=TRUE,center=TRUE)

pcvals<-pcs$rotation|>
  as.data.frame()|>
  dplyr::select(PC1,PC2)|>
  tibble::rownames_to_column('id')|>
  left_join(sampmeta)

 ggplot(pcvals,aes(x=PC1,y=PC2,col=drug,shape=time))+geom_point(aes(size=4))+facet_grid(~date)+scale_color_viridis_d()

```
There seems to be a batch effect so let's correct for that using combat

```{r batch correct}
library(sva)
##now to remove the batches for date while still correcting for drug and time


```
## Differential expression

Let's compute the differential expression of each drug/time point compared with DMSO

```{r diffex, echo=FALSE,warning=FALSE,message=FALSE}
library("DESeq2")

counts<-reduced.data|>
  dplyr::select(Name,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

diffexCalls<-do.call(rbind,lapply(unique(sampmeta$date),function(dval){
  datemeta<-subset(sampmeta,date==dval)|>
    subset(drug!='DMSO')
  ##check for each drug at the date
  do.call(rbind,lapply(unique(datemeta$drug),function(d_name){
    ##check for each time point at the drug/date
   do.call(rbind,lapply(c('8','24'),function(tval){
     samps<-subset(sampmeta,date==dval)|>
       subset(time==tval)|>
       subset(drug%in%c(d_name,"DMSO"))
      cts<-counts[,samps$id]
     dds <- DESeqDataSetFromMatrix(countData = round(cts),
                              colData = samps,
                              design = ~ drug)
     dres<-DESeq(dds)|>
       results()|>
       as.data.frame()|>
       tibble::rownames_to_column('Gene')|>
       mutate(drug=d_name,time=tval,date=dval)
     return(dres)
   })) 
  }))
}))

diffex<-diffexCalls|>
  mutate(signif=padj<0.05)|>
  mutate(direction=ifelse(log2FoldChange>0,'up','down'))

sumstats<-diffex|>subset(signif)|>
  group_by(drug,time,date,direction)|>
  summarize(numgenes=n())

print('now have differential expression calls')
 ggplot(sumstats,aes(x=drug,y=numgenes,fill=direction))+facet_grid(~time)+geom_bar(stat='identity',position='dodge')+coord_flip()+scale_fill_viridis_d()
 
 ggplot(sumstats,aes(x=drug,y=time,size=numgenes,col=drug))+geom_point()+facet_grid(~direction)+scale_size(range=c(1,15))+scale_color_viridis_d()+coord_flip()
```
We can continue to tweak batches/outliers as we move on, but the general barplot should show differential expression.
Next up we can experiment with other visualizations.

```{r alluvial}
if(!require(ggalluvial)){
  install.packages('ggalluvial')
  library('ggalluvial')
}

 ggplot(data=sumstats,aes(axis1=time,axis2=drug,y=numgenes))+geom_alluvium(aes(fill=direction))+
   geom_stratum()+
   scale_x_discrete(limits = c("Time", "Drug"),expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    theme_minimal()+scale_fill_viridis_d()
 


 genestats<-diffex|>dplyr::select(Gene,drug,time,signif,direction)|>
   tidyr::pivot_wider(names_from=time,values_from=signif)|>
   subset(!is.na(`24`))|> #why are there na's?
   subset(!is.na(`8`))|>
   group_by(drug,direction,`8`,`24`)|>
   summarize(numgenes=n())


  ggplot(data=genestats,aes(axis1=`8`,axis2=drug,axis3=`24`,y=numgenes))+geom_alluvium(aes(fill=direction))+
   geom_stratum()+
  scale_x_discrete(limits = c("8 hours", "Drug","24 hours"), expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    theme_minimal() +scale_fill_viridis_d()

```

## Gene based analysis
Let's look at shared trends across the drugs.

```{r gene ranking}

drugscounts<-subset(diffex,signif)|>group_by(Gene,time,direction)|>summarize(ndrugs=n())|>arrange(desc(ndrugs))

library(biomaRt)
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")

dgenes<-diffex|>
  tidyr::separate(Gene,sep="\\.",into=c('ensembl_gene_id','Version'))

map<-getBM(attributes=c('ensembl_gene_id','hgnc_symbol'), filters ='ensembl_gene_id', values =unique(dgenes$ensembl_gene_id), mart = ensembl)

drugscounts<-subset(dgenes,signif)|>group_by(ensembl_gene_id,time,direction)|>summarize(ndrugs=n())|>arrange(desc(ndrugs))|>
  left_join(map)|>
  subset(hgnc_symbol!='')

##most drugs just show up in one or two
ggplot(drugscounts,aes(x=ndrugs,fill=direction))+geom_bar(position='dodge')+ggtitle('Number of drugs with shared differentially expressed genes')

#reduced.genes<-reduced.data|>  
#  tidyr::separate(Name,sep="\\.",into=c('ensembl_gene_id','Version'))|>
#   left_join(map)

topgenes<-subset(drugscounts,ndrugs>8)

subset(dgenes,ensembl_gene_id%in%topgenes$ensembl_gene_id)|>
   subset(signif)|>
     left_join(map)|>
     ggplot(aes(col=drug,y=log2FoldChange,shape=signif,x=hgnc_symbol,size=1))+geom_point()+facet_grid(direction~time)+coord_flip()+ggtitle("Genes that are differentially expressed across more than 8 drugs")

```
Most genes are only differentailly expressed upon a single drug treatment, but some are shared with other drugs. Targeting these canonical responders might be effective, but what if we can find distinct genes changing?

```{r distinct genes}

ugenes<-subset(drugscounts,ndrugs==1)

subset(dgenes,ensembl_gene_id%in%ugenes$ensembl_gene_id)|>
   subset(signif)|>
  group_by(drug,time,direction)|>
  summarize(ngenes=n())|>
  ggplot()+geom_bar(aes(x=drug,y=ngenes,fill=direction),stat='identity',position='dodge')+facet_grid(~time)+coord_flip()

##what about genes that are consistent across 8 and 24hr time point/
ggenes<-dgenes|>subset(signif)|>group_by(drug,direction,ensembl_gene_id)|>summarize(ntime=n())|>subset(ntime==2)

  left_join(ggenes,dgenes)|>
   subset(signif)|>
  group_by(drug,time,direction)|>
  summarize(ngenes=n())|>
  ggplot()+geom_bar(aes(x=drug,y=ngenes,fill=direction),stat='identity',position='dodge')+facet_grid(~time)+coord_flip()+ggtitle("Consistently changing genes in only one drug")


```

should we target genes that are unique to a drug? and if so, do we want genes that are consistently changed across time points? What are the enrichment statistics for these genes? 
