---
title: "Gene Expression Summary Stats"
author: "Sara gosline"
date: "2024-12-11"
output: html_document
---

The goal of this document is to pull all RNASeq files from synapse in a metadata-aware fashion and do the basic analysis needed for figure 1 and supplemental figure 1. 

Currently we have three batches:
|date | treatment |
| --- | --- |
|10/17/23 | DMSO, mirda |
|12/19/23 | DMSO, tram, selu, TNO, ribo, palbo, RMC, vorin |
|12/27/23 | DMSO, decitabine, olap, trab, doxo, ifo, cap |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synapser)
library(dplyr)
synLogin()
```


Created a helper function to get RNASeq data
```{r rnaseq}
##currently this function pulls all data mannually and parses the file names. 
##TODO: udpate using file views
rnaSeqTab<-getRNASeq<-function(){
  ##metadata query first queries file view for all files - in this case i do counts.sf files, but you can also query by data type
  manifest <- synapser::synTableQuery("select * from syn52369043 where name like '%quant.genes.sf'")|>
    as.data.frame()|>
    dplyr::select(name,id)
  
 ##not sure what the 'samples' are here so isolating them out

 samples<-manifest$name[grep('sample',manifest$name)]
 mnt<-manifest$name[grep('MN-2',manifest$name)]

 ##METDATA - replace with anotations 
 mn2<-manifest|>subset(name%in%mnt)|>
   tidyr::separate(name,sep='_',into=c('sample','tumor','batch','time','drugrep','dbatch'))|>
   tidyr::separate(drugrep,sep='[0-9]',into=c('drug'))
 mn2$day=''

 mn2$date<-'122723'
 mn2$date[mn2$dbatch=='d1']<-'121923' #arbitrary, might need to switch
 mn2$date[mn2$drug%in%c('TRA','SEL','RIB','RMC','VOR','PAL')]<-'121923'
 
 mmeta<-mn2|>
   dplyr::select(id,sample,time,drug,date)
 

 ##METADATA replce with annotations
 other<-manifest|>
   subset(name%in%samples)|>
   tidyr::separate(name,into=c('sample','timdrugrep','rest'))|>
   dplyr::mutate(tumor='mpnst')
 other$time='0'
 other$drug<-'DMSO'
 other$time[grep("8",other$timedrugrep)]<-'8'
 other$time[grep("24",other$timedrugrep)]<-'24'
 other$drug[grep("M",other$timedrugrep)]<-'MIRDA'
 other$date<-'101723'
 
 meta<-other|>
   dplyr::select(id,sample,time,drug,date)|>
   rbind(mmeta)
 
 ##now get data and join with metadata
  mtab<-do.call(rbind,lapply(manifest$id,function(x){
   readr::read_tsv(synapser::synGet(x)$path)|>
     dplyr::mutate(id=x)
 }))|>left_join(meta)
  
  
 
 return(mtab)

}

```

## Geta data from synapse
We can now call our RNASeq helper function defined above to get data and visualize overall

```{r data}
library(ggplot2)

rnaseqdata<-rnaSeqTab()

print(paste("Got RNASeq from",length(unique(rnaseqdata$id)),'samples across',length(unique(rnaseqdata$drug)),'drugs and',length(unique(rnaseqdata$time)),'timepoints'))
```


```{r plots and normaliztion}

##first do some pretty PCA
counts<-rnaseqdata|>
  dplyr::select(Name,id,TPM)|>
  tidyr::pivot_wider(names_from=id,values_from=TPM)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

sampmeta<-rnaseqdata|>
  dplyr::select(id,sample,time,drug,date)|>
  distinct()

pcs<-prcomp(counts)

pcvals<-pcs$rotation|>
  as.data.frame()|>
  dplyr::select(PC1,PC2)|>
  tibble::rownames_to_column('id')|>
  left_join(sampmeta)

 ggplot(pcvals,aes(x=PC1,y=PC2,col=drug,shape=time))+geom_point()+facet_grid(~date)

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
