---
title: "TF Enrichment experiments"
author: "Sara gosline"
date: "2024-12-13"
output: html_document
---

Here we take the differential expression and use it to infer transcrition factors

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synapser)
library(dplyr)
synLogin()

counts<-readr::read_csv(synapser::synGet('syn64398051')$path)
diffex<-readr::read_csv(synapser::synGet('syn64397743')$path)
```

## Decoupler analysis 
Here we use the decoupler package to try to infer transcription factor activity from known databases

```{r decoupleR analysis}
library(decoupleR)
library(OmnipathR)
library(tibble)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(ggrepel)

cmat<- counts|>
  dplyr::select(hgnc_symbol,id,NumReads)|>
  tidyr::pivot_wider(names_from='id',values_from='NumReads')|>
  tibble::column_to_rownames(var='hgnc_symbol')

design<-counts|>
  dplyr::select(id,drug,time)|>
  distinct()



net <- decoupleR::get_collectri(organism = 'human', 
                                split_complexes = FALSE)

sample_acts <- decoupleR::run_ulm(mat = cmat, 
                                  net = net, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'mor', 
                                  minsize = 5)

##now we have putative transcritpion factors for each condition
tfs<-sample_acts|>
  dplyr::rename(id='condition')|>
  left_join(design)
print(head(tfs))
```
Now we have tf scores
## Most variable transcription factors

Let's start by plotting the most variable transcritpion factors

```{r pressure, echo=FALSE}

vars<-tfs|>group_by(source)|>summarize(var=var(score))|>arrange(desc(var))

tf_mat<-sample_acts|>
  subset(source%in%c(vars$source[1:25]))|>
  dplyr::select(source,condition,score)|>
  pivot_wider(names_from='source',values_from='score')|>
  tibble::column_to_rownames('condition')

ddesign<-design|>
  tibble::column_to_rownames('id')

pheatmap(tf_mat,annotation_row = ddesign,labels_row=c())


sample_acts|>
  dplyr::rename(id='condition')|>
  mutate(log10p=(-log10(p_value)))|>
  left_join(design)|>
  subset(source%in%vars$source[1:25])|>
  ggplot(aes(x=source,y=score,col=drug,shape=as.factor(time),size=log10p))+geom_point()+coord_flip()
```

I think this plot is more a message of where the outliers lie rather than the actual biological value of each of the TFs. 

## Transcription factors by drug

```{r tfs by drug,warning=FALSE, echo=FALSE}


allcontrast<-do.call(rbind,lapply(unique(diffex$drug),function(d){
    do.call(rbind,lapply(unique(diffex$time),function(ti){
      dtab<-subset(diffex,time==ti)|>
        subset(!is.na(stat))|>
        subset(drug==d)|>
        group_by(hgnc_symbol,drug,time)|>
        summarize(meanStat=mean(stat,na.rm=TRUE))|>
        subset(hgnc_symbol!='')|>
        tibble::column_to_rownames('hgnc_symbol')
      
      contrast_acts <- decoupleR::run_ulm(mat = dtab[, 'meanStat', drop = FALSE], 
                                    net = net, 
                                    .source = 'source', 
                                    .target = 'target',
                                    .mor='mor', 
                                    minsize = 5)|>
        ungroup()|>
        mutate(drug=d,time=ti)
      return(contrast_acts)
        }))
    }))

allcontrast$padj<-p.adjust(allcontrast$p_value)
subset(allcontrast,padj<0.05)|>ggplot(aes(x=source,y=score,col=drug,shape=as.factor(time)))+geom_point()+coord_flip()+facet_grid(~time)
mat<-allcontrast|>dplyr::select(source,drug,time,score)|>tidyr::unite(drug,time,col='drugtime')|>pivot_wider(names_from=drugtime,values_from=score,values_fill = 0.0)


```

We can evaluate how the transcription factor values correlate with one another
```{r tfcors}

corval<-mat|>tibble::column_to_rownames('source')|>cor(use='pairwise.complete.obs')
pheatmap(corval)

pcs<-mat|>tibble::column_to_rownames('source')|>as.matrix()|>prcomp()


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
