---
title: "Combination pathways"
author: "Sara Gosline"
date: "2026-04-01"
output: html_document
---

Here we take a top-down look of what combinations are highly effective and
what pathways the two drugs might share. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../RNAseq/plottingParameters.R")
source("correlation_functions.R")

##get the transcript data, this is slow
if(!exists('full_tmat'))
  full_tmat <- get_harmonized_gex()
```

Here we want to look at the top-ranking drug combinations to see if we 
can use the expression data to understand potential mechanisms.

# Top-ranking drug combos

We can look at the delta AUC scores for the samples for which we have gene expression
data. We see six combinations with a dAUC above 1:

```{r read daucs}
daucs <- readr::read_csv(synGet('syn69978141')$path) |>
  tidyr::separate(Combo, into=c('drug1','drug2'), sep='\\+', remove = FALSE)

sing_gsea <- readr::read_csv("../RNAseq/GSEA_Hallmark_individual/GSEA_results_Hallmark.csv")

##get aucs fro gene xpression samples

daucs |> subset(PDX%in%sing_gsea$MPNST) |>
  group_by(drug1,drug2) |>
  summarize(`mean dAUC` = mean(dAUC,na.rm=T)) |>
  arrange(desc(`mean dAUC`)) |>
  subset(`mean dAUC`>1)

```

## Compute Pathway correlations with single agent AUCs

Now we can look into our baseline gene expression to see if any of these drugs
are anti-correlated with known biological pathway. To do this we 
read in the gene expression data, Then calculate the correlation between untreated gene expression
and single agent AUC scores

```{r untreated gene expression}

nmat <- full_tmat[,intersect(colnames(full_tmat),
                             sing_exp$improve_sample_id)] |>
  as.data.frame()|>
  drop_na()

tmat <- get_gsea_scores(nmat)

print('computing pathway/aUC correlation')

#first focus on only durgs with 10 data points
#gvals <- md|>subset(count>10)

sing_exp <- sing_exp |> mutate(expr_time = '0hr')

cor_res <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_cor(sing_exp,tmat,drug,'0hr',48), 
           compute_cor(sing_exp,tmat,drug,'0hr',120) )
    })) #end drug iter


cor_res |>
  ggplot(aes(x=corval,col=drug)) + geom_density() +
  scale_color_manual(values=drugCol) + 
  theme_bw() + facet_grid(~ttime)

```


## Identify significantly correlated pathways

Here we use a significance test to see if specific correlations are more 
significant.

```{r cor sig, eval=TRUE}

print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_sig(sing_exp,tmat,drug,'0hr',48),
            #compute_sig(sing_exp,tmat,drug,'24hr','48'),
            compute_sig(sing_exp,tmat,drug,'0hr',120))
            #compute_sig(sing_exp,tmat,drug,'24hr','120'))
    })) #end drug iter


merged_res <- cor_res |> left_join(cor_sig) |>
  subset(!drug %in%c('sn-38')) ##removing because we had too few samples


subset(merged_res, sig < 0.05) |>
  head()


```



```{r ssgsea bubble}

bubble_plot(merged_res)
ggsave('t0_ssgseaCorrelation.pdf', height = 6,width=12)

```

## Read in perturbed pathways and check for overlap

Check for overlap between up-regulated pathways and those correlated with drug response.

```{r combine perturbed pathways}

#make sure we ran the RNA seq analysis first
hall_preds <- readr::read_csv('../RNAseq/GSEA_Hallmark/GSEA_results_Hallmark.csv')


##step 1: filter for up-regulated pathways
upregs <- subset(hall_preds, sig == TRUE & NES >0) |>
        mutate(Pathway = stringr::str_replace_all(Feature_set,'HALLMARK_','')) |>
  mutate(DrugTreatment = tolower(DrugTreatment))

##step 2: filter for negatively correlated pathwasy
  combo_preds <- merged_res |>
    dplyr::rename(secondDrug = 'drug','corP' = sig) |>
      mutate(Pathway = stringr::str_replace_all(gene,'HALLMARK_','')) |>
    right_join(upregs) |>
    subset(corP < 0.05) |>
    subset(corval <0) |>
    dplyr::select(Time,DrugTreatment,secondDrug,ttime,Pathway, pathScore='NES',pathwayCor='corval')
  
head(combo_preds)
```

First I create a function to compare this systematically and identify pathways

```{r function to compare}
##for a particular set of drugs, get overlapping pathwas

  
##get a pathway name, plot it across a drug treatment and secondary drug
plot_pathway_with_drugs <- function(genes, 
                                    expr_drug = 'mirdametinib', 
                                    pathwayname='',
                                    ft=''){
  ##get pathway across drug expression changes
  #genes <- gene_sets[[path_name]]
  
  tsamps <- grep(expr_drug, tolower(samples$other_id))
  ##get the DMSO for the same samples
  csamps <- intersect(grep('patient derived xenograft', samples$other_id),
                              grep('3D-MEDS',samples$model_type))

  ## get pathway across untreated samples
  gvals <- full_tmat[genes,intersect(colnames(full_tmat),samples$improve_sample_id[c(tsamps,csamps)])]
#  gvals <- drop_na(as.data.frame(gvals))
  
  meta <- data.frame(drug=c(rep(expr_drug,length(tsamps)),rep('untreated',length(csamps))),
                     sample = c(samples$common_name[tsamps],samples$common_name[csamps]))
  rownames(meta) <- samples$improve_sample_id[c(tsamps,csamps)]
  anncol <- list(drug=c(drugCol[expr_drug],untreated='white'))
  
  pheatmap::pheatmap(gvals, cellwidth = 10,cellheight = 10,annotation_col=meta, 
                    # clustering_distance_rows = 'correlation',
                     clustering_method = 'ward.D2',main = pathwayname,
                    annotation_colors = anncol,
                     filename = paste0(ft,'_heatmap.pdf'))
  
}
##from sing_gsea results (up-regulation?) and merged_res, from other

plot_pathway_correlation <- function(pathname,ft=''){

  cr <- merged_res |> 
    subset(gene == pathname) |>
    mutate(Signif = ifelse(sig < 0.01,'**',ifelse(sig<0.05,'*',"")))
  
  mv <- cr |> group_by(drug)|>
    summarize(mv = mean(corval))|>
    arrange(mv)
  
  cr$drug = factor(cr$drug, levels=unique(mv$drug))
  cr |> ggplot(aes(fill=drug,x=drug, y=corval,group=drug)) + 
     geom_bar(position='dodge',stat='identity') + 
     ggrepel::geom_text_repel(aes(label = Signif),nudge_y=-0.01)+facet_grid(ttime~.) +
    ggtitle(paste("Drug response correlation with",pathname))    +
    theme_bw() +    
    scale_fill_manual(values=drugCol)+
    coord_flip()+
    ylab('AUC correlation') + xlab('Drug')+
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
  ggsave(paste0(ft,'drugCor.pdf'),width=6)
}

##this function takes a combination, finds pathways that
##are common in both, then plots the gene expression and AUCs
##
plot_combo_get_overlap <- function(combo){
  
  combo = tolower(combo)

  ##now we can look for overlap
  c1 <- subset(combo_preds,DrugTreatment==combo[1])|>subset(secondDrug==combo[2])
  c2 <- subset(combo_preds,DrugTreatment==combo[2])|>subset(secondDrug==combo[1])
  
  if(nrow(c1)>0){
    ##plot genes up-regulated?
    apply(c1,1,function(y){
      rv <- intersect(which(hall_preds$Time==y[['Time']]),
              which(hall_preds$Feature_set==paste0('HALLMARK_',y[['Pathway']])))
    rv <- intersect(rv,which(tolower(hall_preds$DrugTreatment)==combo[1]))
    genes <- stringr::str_split(hall_preds$Leading_edge[rv], pattern=', ') |>
      unlist()

    title=paste0(paste0(combo,collapse='_'),'_',y[['Pathway']])
    plot_pathway_with_drugs(genes,combo[1],
                            pathwayname=paste0('HALLMARK_',y[['Pathway']]),
                            ft=title)
    ##plot pathway correlation

    plot_pathway_correlation(paste0('HALLMARK_',y[['Pathway']]),ft=title)

    })
  }
  
    if(nrow(c2)>0){
    ##plot genes up-regulated?
  apply(c2,1,function(y){
      rv <- intersect(which(hall_preds$Time==y[['Time']]),
              which(hall_preds$Feature_set==paste0('HALLMARK_',y[['Pathway']])))
    rv <- intersect(rv,which(tolower(hall_preds$DrugTreatment)==combo[2]))
    genes <- stringr::str_split(hall_preds$Leading_edge[rv], pattern=', ') |>
      unlist()
    
    title=paste0(paste0(rev(combo),collapse='_'),'_',y[['Pathway']])
    plot_pathway_with_drugs(genes,combo[2],
                            pathwayname=paste0('HALLMARK_',y[['Pathway']]),
                            ft=title)
    ##plot pathway correlation

    plot_pathway_correlation(paste0('HALLMARK_',y[['Pathway']]),
                             ft=title)

    })
  }
  #print(paste("We found",nrow(c1),'overlapping pathways from',combo[1],'upregulation and negative corelation to',combo[2]))
  #print(paste("We found",nrow(c2),'overlapping pathways from',combo[2],'upregulation and negative corelation to',combo[1]))

  ##how many pathways are up-regulated and anti-correlated? 
    return(unique(c(c1$Pathway,c2$Pathway)))
}
  

```

Just wanted to confirm, but there is no relationship between overlapping pathways
and dAUC.

# Plotting priority drug combinations

With this data in place we can now move to the data-specific plots.

## Mirda + TNO combination

What stands out from this combination that could give us transcriptional evidence?
It looks like a single pathway


```{r mirda tno}

combo <- c("Mirdametinib",'TNO155')

res<-plot_combo_get_overlap(combo)


plot_pathway_correlation(pathname='HALLMARK_KRAS_SIGNALING_UP',ft='kras_only_up')

##plot genes in pathway in treated samples?


```


## Mirda + vorin combination

```{r mirda vorin}

combo <- c("Mirdametinib",'Vorinostat')
res <- plot_combo_get_overlap(combo)


```
## Vorin + Capmatinib


```{r vorin cap}

combo <- c("Capmatinib",'Vorinostat')

res<-plot_combo_get_overlap(combo)

```
Nothing overlaps here! Maybe we can reverse the sign and see if there is overlap
in pathways?


## TNO155 and doxorubicin

```{r tno dox}
combo <- c("TNO155",'Doxorubicin')

res<-plot_combo_get_overlap(combo)


```

## Mechanisms of overlap
What pathways are the most commonly involved in drug effective combinations? 

Lets see what pathways show up the most frequently

```{r, eval =FALSE}


combo_preds <- readr::read_csv('SuppTable5_combinationScores.csv')


nmat <- full_tmat[,intersect(colnames(full_tmat),
                             sing_exp$improve_sample_id)] |>
  as.data.frame()|>
  drop_na()

tmat <- get_gsea_scores(nmat)

pcounts <- combo_preds |> 
  group_by(Pathway) |>
  summarize(drugsUpregulating = n_distinct(DrugTreatment),correlatedDrugs = n_distinct(secondDrug)) |>
  mutate(total = drugsUpregulating + correlatedDrugs) |>
  tidyr::pivot_longer(c(drugsUpregulating,correlatedDrugs), names_to='source', values_to='drugs') |>
  arrange(desc(total))

ggplot(pcounts, aes(x=reorder(Pathway,total), y = drugs, fill = source)) + 
  geom_bar(position = 'dodge',stat = 'identity') +
  xlab('Pathway') +
  theme(axis.text.x=element_text(angle = 45, vjust=1, hjust=1))

ggsave('commonPathways.pdf')

```

