---
title: "Combination pathways"
author: "Sara Gosline"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../RNAseq/plottingParameters.R")
source("correlation_functions.R")

```

Here we want to look at the top-ranking drug combinations to see if we 
can use the expression data to understand potential mechanisms.

## Mechanisms of overlap
What pathways are the most commonly involved in drug effective combinations? 

Lets see what pathways show up the most frequently

```{r}
combo_preds <- readr::read_csv('SuppTable5_combinationScores.csv')


nmat <- full_tmat[,intersect(colnames(full_tmat),
                             sing_exp$improve_sample_id)] |>
  drop_na()

tmat <- get_gsea_scores(log2(0.001 + nmat))

pcounts <- combo_preds |> 
  group_by(Pathway) |>
  summarize(drugsUpregulating = n_distinct(DrugTreatment),correlatedDrugs = n_distinct(secondDrug)) |>
  mutate(total = drugsUpregulating + correlatedDrugs) |>
  tidyr::pivot_longer(c(drugsUpregulating,correlatedDrugs), names_to='source', values_to='drugs') |>
  arrange(desc(total))

ggplot(pcounts, aes(x=reorder(Pathway,total), y = drugs, fill = source)) + 
  geom_bar(position = 'dodge',stat = 'identity') +
  xlab('Pathway') +
  theme(axis.text.x=element_text(angle = 45, vjust=1, hjust=1))

ggsave('commonPathways.pdf')

```
Top pathways include angiogenesis, KRAS signaling, and interferon alpha response. 

## Visualizing top pathways

We can now look at the top three pathways to see what proteins
are doing both at the correlative level and expression level?

```{r how to plot pathways}
##first get the pathways

m_df <- msigdbr::msigdbr(db_species = 'HS', collection = 'H')
gene_sets <- split(x = m_df$gene_symbol, f = m_df$gs_name)
  
##get a pathway name, plot it across a drug treatment and secondary drug
plot_pathway_with_drugs <- function(path_name = 'HALLMARK_KRAS_SIGNALING_DN', 
                                    expr_drug = 'mirdametinib', 
                                    cor_drug = 'tno155'){
  ##get pathway across drug expression changes
  genes <- gene_sets[[path_name]]
  
  tsamps <- samples[grep(expr_drug, tolower(samples$other_id)),]
  ##get the DMSO for the same samples
  csamps <- samples[grep('DMSO', samples$other_id),]

  ## get pathway across untreated samples
  gvals <- log2(0.001 + full_tmat[genes,intersect(colnames(full_tmat),sing_exp$improve_sample_id)])
  
}



#define function to plot specific pathway correlation and heatmaps? 
get_path_drug_cor <- function(pathway = 'KRAS_SIGNALING_DN', drug){
  
  path_vals <- tmat[paste0('HALLMARK_',pathway),]

  evals <- data.frame(sing_exp[sing_exp$drug %in% drug,]) 
  evals$pathway_score <- path_vals[as.character(evals$improve_sample_id)]

  ggplot(evals, aes(x = pathway_score, y = auc, col = drug)) +
    geom_point(size = 3) +
    #ggrepel::geom_label_repel(aes(label=common_name))+
    facet_grid(~treat_time) +
    geom_smooth(method = 'lm', se = FALSE, stroke = 2) + 
     theme_minimal(base_size=12) + scale_color_manual(values = classCol)  +
    labs(x = paste0(pathway,' pathway score'), y = 'AUC')
}


## here we look at pathway scores of the untreated sample and the drug-induced sample
get_double_drug_path_cor <- function(drug = 'mirdametinib', drug2 = 'vorinostat', paths){
  
  sing_gsea <- readr::read_csv("../RNAseq/GSEA_Hallmark_individual/GSEA_results_Hallmark.csv")
  
  path_vals <- tmat[paste0('HALLMARK_',paths),] |>
    as.data.frame()|>
    tibble::rownames_to_column('Pathway') |>
    tidyr::pivot_longer(c(-Pathway),names_to='improve_sample_id',values_to='pathway score')
  
  path_fc <- subset(sing_gsea, Feature_set %in% paste0('HALLMARK_',paths)) |>
    subset(DrugTreatment == drug2) |>
    dplyr::select(common_name = 'MPNST',Time,NES,Pathway = 'Feature_set')

  evals <- data.frame(sing_exp[sing_exp$drug %in% drug,]) |>
    mutate(improve_sample_id=as.character(improve_sample_id))|>
    inner_join(path_vals)
  
  p1<-ggplot(evals, aes(y = `pathway score`, x = auc, col = Pathway)) +
    geom_point(size = 3) +
    ggrepel::geom_label_repel(aes(label=common_name))+
    facet_grid(~treat_time) +
    geom_smooth(method = 'lm', se = FALSE, stroke = 1) + 
     theme_minimal(base_size=12) + #scale_color_manual(values = classCol)  +
    labs(x = paste0(drug,' AUC'), y = 'Pathway score')
  
  p2<- evals|>inner_join(path_fc) |> 
    ggplot(aes(x=`NES`, size = `pathway score`,  y = auc,col = Pathway, shape = common_name)) + 
    geom_point() + 
    labs(x=paste(drug2, 'induced NES'), y = paste(drug,'auc')) + 
    facet_grid(treat_time~Time)
  
  return(list(p1,p2))
}


```

### Angiogenesis

```{r angio}

adrugs <- combo_preds |> subset(Pathway == 'ANGIOGENESIS') |>
  arrange(desc(pathScore))

adrugs
get_path_drug_cor('ANGIOGENESIS',unique(adrugs$secondDrug))

```

### KRAS Signaling

```{r kras}

kdrugs <- combo_preds |> subset(Pathway == 'KRAS_SIGNALING_DN')

kdrugs
get_path_drug_cor('KRAS_SIGNALING_DN',unique(kdrugs$secondDrug))

```


### Interforn alpha response

```{r ialpha}

idrugs <- combo_preds |> subset(Pathway=='INTERFERON_ALPHA_RESPONSE')


idrugs
get_path_drug_cor('INTERFERON_ALPHA_RESPONSE',unique(idrugs$secondDrug))

```

## Focusing on mirdametinib - vorinostat

Mirdametinib and vorinostat seem to be the most potent combination across the board

```{r}

mv <- combo_preds |> 
  subset(DrugTreatment %in% c('mirdmetinib','vorinostat')) |>
  subset(secondDrug %in% c('mirdametinib','vorinostat'))


mv_plots<- lapply(unique(mv$Pathway), function(p) get_path_drug_cor(p,unique(mv$secondDrug)))

cowplot::plot_grid(plotlist = mv_plots,nrow=2)
```


```{r extra}
 ksd <- gvals[intersect(genes,rownames(gvals)),] |>
  tibble::rownames_to_column('gene') |> 
  tidyr::pivot_longer(c(-gene),names_to = 'improve_sample_id', values_to = 'expr') |>
  left_join(mutate(sing_exp[sing_exp$drug %in% mdrugs,],
                   improve_sample_id = as.character(improve_sample_id)))

 
```
Now that we can see that mirdametinib has multiple combinations, we
can dive into the specific pathway that is activated. 


We can also go back to the genes identified in the diffex and see how they behave
### Heatmap of pathway genes

Do the heatmaps of genes look at all interesting? 
```{r pathgenes}

gvals <- log2(0.001 + full_tmat[,intersect(colnames(full_tmat),sing_exp$improve_sample_id)])


m_df <- msigdbr::msigdbr(species = "Homo sapiens", category = "H")
gene_sets <- split(x = m_df$gene_symbol, f = m_df$gs_name)

##get mirda leading edge genes see what we see
genes <- gene_sets$HALLMARK_KRAS_SIGNALING_DN
genes <- gene_sets$HALLMARK_HEDGEHOG_SIGNALING

fset = 'HALLMARK_KRAS_SIGNALING_DN'
#fset = 'HALLMARK_HEDGEHOG_SIGNALING'
le_genes <- subset(hall_preds,Feature_set==fset) |>
  subset(DrugTreatment=='Vorinostat')
le_early_genes <- stringr::str_split(le_genes[which(le_genes$Time=='8h'),]$Leading_edge,', ') |> unlist()
le_late_genes <- stringr::str_split(le_genes[which(le_genes$Time=='24h'),]$Leading_edge,', ') |> unlist()


evals <- data.frame(sing_exp[sing_exp$drug %in% c('mirdametinib'),])  |>
  dplyr::select(improve_sample_id, drug, treat_time, auc)|>
  distinct() |>
  tidyr::unite(drug,treat_time, col='drug_treat') |>
  pivot_wider(names_from = drug_treat, values_from = auc) |> tibble::column_to_rownames('improve_sample_id')

evals <- evals[which(rownames(evals)%in%colnames(gvals)),]
e_evals <- evals[,grep('48',colnames(evals))]
l_evals <- evals[,grep('120',colnames(evals))]


pheatmap::pheatmap(gvals[which(rownames(gvals)%in%genes),order(rowMeans(evals))], 
                   annotation_col = evals, cluster_cols = FALSE, 
                   cellwidth = 10,
                   clustering_method = 'ward.D2',
                   clustering_distance_rows = 'correlation',
                   cellheight = 10, main = 'All KRAS_SIGNALING_DN genes',
                   filename = 'ksd_genes.pdf')


pheatmap::pheatmap(gvals[which(rownames(gvals)%in%le_early_genes),order(rowMeans(l_evals))], 
                   annotation_col = l_evals, cluster_cols = FALSE, 
                                      clustering_distance_rows = 'correlation',
                   cellwidth = 10,
                                      clustering_method = 'ward.D2',
                   cellheight = 10, main = 'Early KRAS_SIGNALING_DN genes',
                   filename = 'early_ksd_genes.pdf')

pheatmap::pheatmap(gvals[which(rownames(gvals)%in%le_late_genes),order(rowMeans(l_evals))], 
                   annotation_col = l_evals, 
                   cluster_cols = FALSE, 
                                      clustering_distance_rows = 'correlation',
                                      clustering_method = 'ward.D2',
                   cellwidth = 10,
                   cellheight = 10, main = 'Late KRAS_SIGNALING_DN genes',
                   filename = 'late_ksd_genes.pdf')

pheatmap::pheatmap(gvals[which(rownames(gvals)%in%union(le_late_genes, le_early_genes)),
                         order(rowMeans(l_evals))], 
                   annotation_col = l_evals, cluster_cols = FALSE,
                                      clustering_distance_rows = 'correlation',
                                      clustering_method = 'ward.D2',
                   cellwidth = 10,
                   cellheight = 10, main = 'All Leading edge KRAS_SIGNALING_DN genes',
                   filename = 'le_ksd_genes.pdf')


pheatmap::pheatmap(gvals[which(rownames(gvals)%in%intersect(le_late_genes,le_early_genes)),
                         order(rowMeans(l_evals))], 
                   annotation_col = l_evals, cluster_cols = FALSE, 
                                      clustering_distance_rows = 'correlation',
                                      clustering_method = 'ward.D2',
                   cellwidth = 10,
                   cellheight = 10, main = 'Shared leading edge KRAS_SIGNALING_DN genes',
                   filename = 'shared_le_ksd_genes.pdf')
##now get TNO genes


```
