---
title: "ssGSEA correlation"
author: "Sara Gosline"
date: "2026-02-28"
output: html_document
---

With this markdown we evaluate which pathways are correlated with drug response
and compare with those that are up-regulated.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../RNAseq/plottingParameters.R")
source("correlation_functions.R")

```

## Data

We have reformatted the data into three files:
1. sample file describing the samples
2. gene file with gene metadata
3. transcriptomics data with gene expression data
4. combination data with delta AUC values
5. drug data with single AUC values

Most data is now read in by `correlation_functions.R`

```{r data, message=FALSE}

##get the transcript data, this is slow
if(!exists('full_tmat'))
  full_tmat <- get_harmonized_gex()

```

Now we have all the transcriptional data. There
are some missing values, but we'll keep them in for now.

## Correlation of AUC and pathway scores

Let's first look at the untreated gene expression data and see if we can 
find out features that predict AUC

```{r untreated gene expression}

nmat <- full_tmat[,intersect(colnames(full_tmat),
                             sing_exp$improve_sample_id)] |>
  drop_na()

tmat <- get_gsea_scores(log2(0.001 + nmat))

print('computing pathway/aUC correlation')

#first focus on only durgs with 10 data points
#gvals <- md|>subset(count>10)

sing_exp <- sing_exp |> mutate(expr_time = '0hr')

cor_res <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_cor(sing_exp,tmat,drug,'0hr',48), 
           compute_cor(sing_exp,tmat,drug,'0hr',120) )
    })) #end drug iter


cor_res |>
  ggplot(aes(x=corval,col=drug)) + geom_density() +
  scale_color_manual(values=drugCol) + 
  theme_bw() + facet_grid(~ttime)

```

We only have 8 samples here, but across drugs there is an enrichment of genes
that are negatively correlated with AUC, which means they are high when the drug
is MORE effective


### Identify significant pathways

Here we use a significance test to see if specific correlations are more 
significant.

```{r cor sig, eval=TRUE}

print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(sing_exp$drug), function(drug){
      rbind(compute_sig(sing_exp,tmat,drug,'0hr',48),
            #compute_sig(sing_exp,tmat,drug,'24hr','48'),
            compute_sig(sing_exp,tmat,drug,'0hr',120))
            #compute_sig(sing_exp,tmat,drug,'24hr','120'))
    })) #end drug iter


merged_res <- cor_res |> left_join(cor_sig)


subset(merged_res, sig < 0.05) |>
  head()


```


While we have very few pathways that past multiple testing correction, we can 
still dive further to see if these pathway scores exist in the combination data.

# Plot ssGSEA scores by drug

Here we want to re-create the bubble plots with pathway significance and 
correlation.

```{r ssgsea bubble}

bubble_plot(merged_res)
ggsave('t0_ssgseaCorrelation.pdf', height = 6,width=12)

```

## Correlation with AUC Shift scores

Now we can do the same with our AUC shift scores, which are more powerful.

```{r auc shift correlation}

nmat <- full_tmat[,intersect(colnames(full_tmat), exp$improve_sample_id)] |>
  drop_na()
tmat <- get_gsea_scores(log2(0.001+nmat))


gvals <- md |> subset(count > 10)

exp <- exp |>
  dplyr::rename(auc = 'delta_auc') |>
  mutate(treat_time = as.numeric(as.character(treat_time))) |>
  subset(!is.na(drug))

treated_cor_res <- do.call(rbind,lapply(unique(gvals$drug), function(drug){
      rbind(compute_cor(exp,tmat,drug,'8hr',48),
            compute_cor(exp,tmat,drug,'24hr',48),
            compute_cor(exp,tmat,drug,'8hr',120),            
            compute_cor(exp,tmat,drug,'24hr',120))
    })) #end drug iter


print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(gvals$drug), function(drug){
      rbind(compute_sig(exp,tmat,drug,'8hr',48),
            compute_sig(exp,tmat,drug,'24hr',48),
            compute_sig(exp,tmat,drug,'8hr',120),
            compute_sig(exp,tmat,drug,'24hr',120))
    })) #end drug iter

full_res <- treated_cor_res |>
  left_join(cor_sig)

subset(cor_sig, sig < 0.01) |> head()


bubble_plot(full_res)
ggsave('treated_ssgseaCorrelation.pdf', height = 12, width = 12)
```



## Drug-specific inquiry 

For each drug of interest, can we 
1. name the differentially expressed pathwasy
2. ask which drugs correlate with drug response?
3. confirm in the treated samples? 

```{r drug-spec}

#make sure we ran the RNA seq analysis first
hall_preds <- readr::read_csv('../RNAseq/GSEA_Hallmark/GSEA_results_Hallmark.csv')


##step 1: filter for up-regulated pathways
upregs <- subset(hall_preds, sig == TRUE & NES >0) |>
        mutate(Pathway = stringr::str_replace_all(Feature_set,'HALLMARK_','')) |>
  mutate(DrugTreatment = tolower(DrugTreatment))

##step 2: filter for negatively correlated pathwasy
  combo_preds <- merged_res |>
    dplyr::rename(secondDrug = 'drug','corP' = sig) |>
      mutate(Pathway = stringr::str_replace_all(gene,'HALLMARK_','')) |>
    right_join(upregs) |>
    subset(corP < 0.05) |>
    subset(corval <0) |>
    dplyr::select(Time,DrugTreatment,secondDrug,ttime,Pathway, pathScore='NES',pathwayCor='corval')
  
head(combo_preds)

readr::write_csv(combo_preds,file='SuppTable5_combinationScores.csv')
```

To view the combination predictions and select those agents that exhibit the most 
cross talk in our dataset we use a sankey plot. This connects drugs with activated
pathways to drugs with correlated pathways. 

```{r}
library(ggalluvial)

plotSankey <- function(){

  sumvals <- combo_preds |>
    select(Time,DrugTreatment,secondDrug,Pathway) |>
    distinct() |>
    #group_by(DrugTreatment,Time,secondDrug) |> 
    summarize(pathways = n_distinct(Pathway),.by = c(DrugTreatment,Time,secondDrug))|>
    ungroup()

  early <- sumvals|>subset(Time =='8h')|>
    ggplot( aes(y=pathways, axis1=DrugTreatment, axis2=secondDrug))+
    ggalluvial::geom_alluvium(aes(fill=DrugTreatment)) +
    scale_fill_manual(values = classCol)+
    guides(fill='none') +
    ggalluvial::geom_stratum(alpha = 0.7, width = 1/3) +
    scale_x_continuous(breaks = 1:2, 
                       labels = c("Single drug pathway activation",
                                  "Secondary drug pathway sensitivity"))  +
    geom_text(stat="stratum", aes(label = after_stat(stratum))) +
    ggtitle("Pathways at 8 hours")
  
  late <- sumvals|>subset(Time =='24h')|>
    ggplot( aes(y=pathways, axis1=DrugTreatment, axis2=secondDrug))+
    ggalluvial::geom_alluvium(aes(fill=DrugTreatment)) +
    scale_fill_manual(values = classCol)+
    guides(fill='none') +
    ggalluvial::geom_stratum(alpha = 0.7, width = 1/3) +
    scale_x_continuous(breaks = 1:2, 
                       labels = c("Single drug pathway activation",
                                  "Secondary drug pathway sensitivity"))  +
    geom_text(stat="stratum", aes(label = after_stat(stratum))) +
    ggtitle("Pathways at 24 hours")
 
  cowplot::plot_grid(early,late) 
}

plotDrugFromDiffEx <- function(drug){
 #drug = 'Mirdametinib'
  
  sig_feats <- subset(hall_preds, DrugTreatment==drug) |>
    subset(p_value < 0.05)|>
    subset(FDR_q_value < 0.05) |>
    mutate(Pathway = stringr::str_replace_all(Feature_set,'HALLMARK_',''))
  
  t0_preds <- merged_res |>
    dplyr::rename(corDrug = 'drug','corP' = sig) |>
      mutate(Pathway = stringr::str_replace_all(gene,'HALLMARK_','')) |>
    right_join(sig_feats) |>
    subset(corP < 0.05) |>
    mutate(Regulation = 'decrease with drug treatment')
  t0_preds$Regulation[t0_preds$NES>0] <-'increase with drug treatment'
  
  ggplot(t0_preds, aes(x=Pathway, y =corval, fill = corDrug )) +
    geom_bar(position = 'dodge', stat = 'identity') + 
    scale_fill_manual(values = drugCol) + 
    facet_grid(ttime~Regulation) + 
    coord_flip() + 
    theme_bw() +    
    theme(axis.title=element_blank(), axis.text.x=element_text(angle=45, vjust=1, hjust=1))+
    labs(x = 'Altered pathway', y = 'AUC correlation', 
    title = paste0('AUC correlations of ',drug,'-altering pathways'))
}

#plotDrugFromDiffEx('Vorinostat')
#plotDrugFromDiffEx("Mirdametinib")

#ggsave('mirdaComboSuggs.pdf')
plotDrugFromDiffEx('TNO155')
#ggsave('tnoComboSuggs.pdf')
plotSankey()

ggsave('alluvialDrugPreds.pdf', width=12,height=8)


```
We can also evaluate how many drugs share pathways with each other.

```{r drugCounts}

c1_counts <- combo_preds |>
   group_by(DrugTreatment) |> 
   summarize(`sharedPathwayWithDrugAssociatedPath` = n_distinct(secondDrug)) |>
    arrange(desc(sharedPathwayWithDrugAssociatedPath)) |>
  dplyr::rename(drug = 'DrugTreatment')


c2_counts <- combo_preds |>
    group_by(secondDrug) |> 
    summarize(drugsThatAlterSharedPathway = n_distinct(DrugTreatment)) |>
   arrange((desc(drugsThatAlterSharedPathway))) |>
  dplyr::rename(drug = 'secondDrug')

combo_counts <- c1_counts |> full_join(c2_counts) |>
  replace_na(list(sharedPathwayWithDrugAssociatedPath=0,drugsThatAlterSharedPathway=0))|>
  mutate(total = sharedPathwayWithDrugAssociatedPath + drugsThatAlterSharedPathway) |>
  arrange(desc(total))

combo_counts

tested_drugs <- c('mirdametinib','vorinostat','tno155','capmatinib','palbociclib','verteporfin',
                  'iag933','doxorubicin','trabectedin','olaparib', 'pexidartinib')

```
Should this stay as a table or should we try to plot it?

```{r drug plot}

face <- rep('plain',length(combo_counts$drug))
names(face)<-combo_counts$drug

face[tested_drugs]<-'bold'
            
combo_counts$drug <- factor(combo_counts$drug,levels=combo_counts$drug)
combo_counts |> tidyr::pivot_longer(c(drugsThatAlterSharedPathway,sharedPathwayWithDrugAssociatedPath),
                                    names_to='source',
                                    values_to='number of drugs with overlapping pathways') |>
  ggplot(aes(x=drug, y = `number of drugs with overlapping pathways`, fill = source)) +
  geom_bar(position='dodge', stat = 'identity') +
  xlab('drug names')+
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1, 
                                 color=classCol[as.character(combo_counts$drug)],
                                 face = face                   ))

ggsave('drugsWithOverlappingPaths.pdf')
```

Now we can move the top drugs to the next phase of analysis.
