---
title: "Contaminant identification"
output: html_document
date: "2024-08-22"
---

# overview
# 1. load data
# 2. for each sample, calculate % of abundance from each protein
# 3. look at distribution of protein %'s across all samples
# 4. calculate log2(abundance in sample / ratio in samples)
# 5. look at distribution of proteins with higher % in samples than ref
# 6. define outlier proteins at top 5% of proteins with higher % in samples than ref
# 7. look at homology in outlier proteins - if marked as human but also in mouse/rat, then also add mouse/rat versions to list

# 1. load data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(synapser)
library(plyr)
library(dplyr)
library(reshape2)
library(ggplot2)
# load nonrelative data
synapser::synLogin()
global.df <- read.table(synapser::synGet("syn62378517")$path, sep = "\t", header = T)
global.long <- na.omit(reshape2::melt(global.df, variable.name = "Sample", value.name = "Abundance"))
```

# 2. for each sample, calculate % of abundance from each protein
```{r calculate % protein abundance}
sampleSums <- colSums(global.df[2:ncol(global.df)], na.rm = TRUE)
global.long$Percent_Abundance <- NA
for (i in 1:length(colnames(global.df)[2:ncol(global.df)])) {
  temp.sample <- colnames(global.df)[i+1]
  global.long[global.long$Sample == temp.sample,]$Percent_Abundance <- 100*global.long[global.long$Sample == temp.sample,]$Abundance/sampleSums[i]
}
hist(global.long$Percent_Abundance, xlab = "% Abundance")
hist(global.long[global.long$Percent_Abundance < 1,]$Percent_Abundance, xlab = "% Abundance", main = "% Abundance < 1") # most proteins seem to represent < 0.1% of their sample
hist(global.long[global.long$Percent_Abundance < 0.1,]$Percent_Abundance, xlab = "% Abundance", main = "% Abundance < 0.1") # most proteins seem to represent < 0.01% of their sample
hist(global.long[global.long$Percent_Abundance < 0.01,]$Percent_Abundance, xlab = "% Abundance", main = "% Abundance < 0.01") # most proteins seem to represent < 0.002% of their sample
hist(global.long[global.long$Percent_Abundance > 1,]$Percent_Abundance, xlab = "% Abundance", main = "% Abundance > 1") # a few hundred proteins represent a majority of their samples
```
# 4. calculate log2(abundance in sample / ratio in samples)
```{r logRatio}
rownames(global.df) <- global.df$Specie
global.ratio <- log(global.df[,2:ncol(global.df)]/global.df$ref, 2)
global.ratio$Specie <- rownames(global.ratio)
ratio.long <- reshape2::melt(global.ratio, variable.name = "Sample", value.name = "Log2Ratio")
ratio.long <- ratio.long[ratio.long$Sample != "ref",]

hist(ratio.long$Log2Ratio) # most proteins have lower abundance in samples than ref
hist(ratio.long[ratio.long$Log2Ratio > 0,]$Log2Ratio)
log2cutoff <- quantile(ratio.long$Log2Ratio, 0.95, na.rm = TRUE)
```

# 5. what if we remove proteins with > 1% of abundance
```{r logRatio}
abundant.prots <- na.omit(unique(global.long[global.long$Percent_Abundance > 1,]$Specie)) # 23
percent.abundant.prots <- 100*length(abundant.prots)/length(na.omit(unique(global.long$Specie))) # 0.22%
filtered.global <- global.df[!(global.df$Specie %in% abundant.prots),]
rownames(filtered.global) <- filtered.global$Specie
global.ratio <- log(filtered.global[,2:ncol(filtered.global)]/filtered.global$ref, 2)
global.ratio$Specie <- rownames(global.ratio)
ratio.long <- reshape2::melt(global.ratio, variable.name = "Sample", value.name = "Log2Ratio")
ratio.long <- ratio.long[ratio.long$Sample != "ref",]

hist(ratio.long$Log2Ratio) # most proteins have lower abundance in samples than ref
hist(ratio.long[ratio.long$Log2Ratio > 0,]$Log2Ratio)
log2cutoff2 <- quantile(ratio.long$Log2Ratio, 0.95, na.rm = TRUE)
# didn't change distribution much

prot.summary <- ddply(global.long, .(Specie), summarize,
                      mean_abundance = mean(Abundance, na.rm = TRUE),
                      median_abundance = median(Abundance, na.rm = TRUE),
                      max_abundance = max(Abundance, na.rm = TRUE),
                      min_abundance = min(Abundance, na.rm = TRUE),
                      sd_abundance = sd(Abundance, na.rm = TRUE),
                      mean_perc_abundance = mean(Percent_Abundance, na.rm = TRUE),
                      median_perc_abundance = median(Percent_Abundance, na.rm = TRUE),
                      max_perc_abundance = max(Percent_Abundance, na.rm = TRUE),
                      min_perc_abundance = min(Percent_Abundance, na.rm = TRUE),
                      sd_perc_abundance = sd(Percent_Abundance, na.rm = TRUE))
prot.summary$median_perc_abundance_percentile <- 100*dplyr::percent_rank(prot.summary$median_perc_abundance)
prot.summary$max_perc_abundance_percentile <- 100*dplyr::percent_rank(prot.summary$max_perc_abundance)
hist(prot.summary$median_perc_abundance)
hist(prot.summary$max_perc_abundance)
perc.cutoff <- 0.99
median_perc_cutoff <- quantile(prot.summary$median_perc_abundance, perc.cutoff) # 0.0219 for 0.95 percentile, 0.0741 for 0.99
write.csv(prot.summary, "Protein_summary.csv", row.names = FALSE)

# what about just look at the top 5% of proteins based on median_perc_abundance
#bigProts <- prot.summary %>% slice_max(median_perc_abundance, n=ceiling(0.05*nrow(prot.summary))) # 523 proteins more abundant in most samples than 95% of proteins
bigProts <- prot.summary[prot.summary$median_perc_abundance_percentile > perc.cutoff, ] # 523 proteins more abundant in most samples than 95% of proteins, 105 for 99%
write.csv(bigProts, "Proteins with higher median percent abundance than 99 percent of proteins.csv", row.names = FALSE)
bigProts <- read.csv("Proteins with higher median percent abundance than 99 percent of proteins[61].csv")
mouse.bigProts <- bigProts[endsWith(bigProts$Specie, "_MOUSE"),] # 11: collagen, laminin, etc.
rat.bigProts <- bigProts[endsWith(bigProts$Specie, "_RAT"),] # 2: CO1A1, CO1A2 (alpha 1/2 chains of rat collagen type i)
# some of these big proteins marked as human may be mouse/rat too but we assumed they're human

global.long$Species <- "Human"
global.long[endsWith(global.long$Specie, "_MOUSE"),]$Species <- "Mouse"
global.long[endsWith(global.long$Specie, "_RAT"),]$Species <- "Rat"
outlier.prots <- global.long[global.long$Specie %in% bigProts$Specie,] # 105
global.long <- global.long|>
  tidyr::separate(Sample,into=c('PDX','Drug','Timepoint (hrs)','Unit','Replicate'),sep='_',remove=FALSE)
global.long.w.ref <- global.long
global.long <- global.long[global.long$Sample != "ref",]
global.long$`Timepoint (hrs)` <- factor(global.long$`Timepoint (hrs)`, levels = c("0", "0.5", "3", "24"))
global.long$Replicate <- substr(global.long$Replicate, 4,4)
for (i in unique(outlier.prots$Species)) {
  n.outlier.prots <- length(unique(outlier.prots[outlier.prots$Species == i,]$Specie))
  print(paste(i, "has", n.outlier.prots, "proteins with median % abundance percentile >", perc.cutoff))
  temp.width <- ifelse(n.outlier.prots > 20, 40, n.outlier.prots*3)
  
  # boxplot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ Specie)
  ggsave(paste0("Contaminants_boxplot_", i, ".pdf"), 
         width = temp.width)
  
  # scatter plot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Abundance`, color = Drug, shape = Replicate)) +
    geom_point() + facet_grid(PDX ~ Specie)
  ggsave(paste0("Contaminants_scatterPlot_", i, ".pdf"), 
         width = temp.width)
  
  # boxplot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Percent_Abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ Specie) + ylab("% Abundance")
  ggsave(paste0("Contaminants_percent_boxplot_", i, ".pdf"), 
         width = temp.width)
  
  # scatter plot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Percent_Abundance`, color = Drug, shape = Replicate)) +
    geom_point() + facet_grid(PDX ~ Specie) + ylab("% Abundance")
  ggsave(paste0("Contaminants_percent_scatterPlot_", i, ".pdf"), 
         width = temp.width)
  
  # how many in each sample?
    ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Replicate`, fill = Drug)) + ylab("# of Contaminants") +
    geom_bar(stat="count", position="dodge") + facet_grid(PDX ~ `Timepoint (hrs)`)
  ggsave(paste0("Contaminants_barPlot_", i, ".pdf"), 
         width = temp.width)
}

# main culprit seems to be rat collagen type i (CO1A1 and CO1A2)
hist(global.long[endsWith(global.long$Specie, "CO1A1_RAT"),]$Percent_Abundance)
hist(global.long[endsWith(global.long$Specie, "CO1A2_RAT"),]$Percent_Abundance)
CO1A1.cutoff <- quantile(global.long[endsWith(global.long$Specie, "CO1A1_RAT"),]$Percent_Abundance, 0.95) # 2.73
CO1A2.cutoff <- quantile(global.long[endsWith(global.long$Specie, "CO1A2_RAT"),]$Percent_Abundance, 0.95) # 2.91
CO1A1.outliers <- distinct(global.long[endsWith(global.long$Specie, "CO1A1_RAT") & global.long$Percent_Abundance > CO1A1.cutoff,])
CO1A2.outliers <- distinct(global.long[endsWith(global.long$Specie, "CO1A2_RAT") & global.long$Percent_Abundance > CO1A2.cutoff,])
CO1A.outliers <- unique(c(CO1A1.outliers$Sample, CO1A2.outliers$Sample))

filtered.prot <- global.long[!(global.long$Sample %in% CO1A.outliers),]

# try our own cutoff because CO1A1 has some separation between samples
CO1A1.outliers <- distinct(global.long[endsWith(global.long$Specie, "CO1A1_RAT") & global.long$Percent_Abundance > 1.5,])
filtered.prot <- global.long[!(global.long$Sample %in% CO1A1.outliers$Sample),]

# need to decide which samples to remove based on contamination score... calculate contamination score first


# after outliers are removed
for (i in unique(outlier.prots$Species)) {
  n.outlier.prots <- length(unique(outlier.prots[outlier.prots$Species == i,]$Specie))
  temp.width <- ifelse(n.outlier.prots > 20, 40, n.outlier.prots*3)
  
  # boxplot
  ggplot(filtered.prot[filtered.prot$Species == i &
                    filtered.prot$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ Specie)
  ggsave(paste0("Contaminants_boxplot_noCO1A1Outliers_", i, ".pdf"), 
         width = temp.width)
  
  # scatter plot
  ggplot(filtered.prot[filtered.prot$Species == i &
                    filtered.prot$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Abundance`, color = Drug, shape = Replicate)) +
    geom_point() + facet_grid(PDX ~ Specie)
  ggsave(paste0("Contaminants_scatterPlot_noCO1A1Outliers_", i, ".pdf"), 
         width = temp.width)
  
  # boxplot
  ggplot(filtered.prot[filtered.prot$Species == i &
                    filtered.prot$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Percent_Abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ Specie) + ylab("% Abundance")
  ggsave(paste0("Contaminants_percent_boxplot_noCO1A1Outliers_", i, ".pdf"), 
         width = temp.width)
  
  # scatter plot
  ggplot(filtered.prot[filtered.prot$Species == i &
                    filtered.prot$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Percent_Abundance`, color = Drug, shape = Replicate)) +
    geom_point() + facet_grid(PDX ~ Specie) + ylab("% Abundance")
  ggsave(paste0("Contaminants_percent_scatterPlot_noCO1A1Outliers_", i, ".pdf"), 
         width = temp.width)
  
  # how many in each sample?
    ggplot(filtered.prot[filtered.prot$Species == i &
                    filtered.prot$Specie %in% outlier.prots$Specie,],
         aes(x=`Replicate`, fill = Drug)) + ylab("# of Contaminants") +
    geom_bar(stat="count", position="dodge") + facet_grid(PDX ~ `Timepoint (hrs)`)
  ggsave(paste0("Contaminants_barPlot_noCO1A1Outliers_", i, ".pdf"), 
         width = temp.width)
}

# also look at max percent abundance
biggestProts <- bigProts[bigProts$max_perc_abundance_percentile > 99.9,] # 11
outlier.prots <- global.long[global.long$Specie %in% biggestProts$Specie,]

for (i in unique(outlier.prots$Species)) {
  n.outlier.prots <- length(unique(outlier.prots[outlier.prots$Species == i,]$Specie))
  print(paste(i, "has", n.outlier.prots, "proteins with median % abundance percentile >", perc.cutoff))
  temp.width <- ifelse(n.outlier.prots > 20, 40, n.outlier.prots*3)
  
  # boxplot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ Specie)
  ggsave(paste0("Biggest_contaminants_boxplot_", i, ".pdf"), 
         width = temp.width)
  
  # scatter plot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Abundance`, color = Drug, shape = Replicate)) +
    geom_point() + facet_grid(PDX ~ Specie)
  ggsave(paste0("Biggest_contaminants_scatterPlot_", i, ".pdf"), 
         width = temp.width)
  
  # boxplot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Percent_Abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ Specie) + ylab("% Abundance")
  ggsave(paste0("Biggest_contaminants_percent_boxplot_", i, ".pdf"), 
         width = temp.width)
  
  # scatter plot
  ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Timepoint (hrs)`, y=`Percent_Abundance`, color = Drug, shape = Replicate)) +
    geom_point() + facet_grid(PDX ~ Specie) + ylab("% Abundance")
  ggsave(paste0("Biggest_contaminants_percent_scatterPlot_", i, ".pdf"), 
         width = temp.width)
  
  # how many in each sample?
    ggplot(global.long[global.long$Species == i &
                    global.long$Specie %in% outlier.prots$Specie,],
         aes(x=`Replicate`, fill = Drug)) + ylab("# of Contaminants") +
    geom_bar(stat="count", position="dodge") + facet_grid(PDX ~ `Timepoint (hrs)`)
  ggsave(paste0("Biggest_contaminants_barPlot_", i, ".pdf"), 
         width = temp.width)
}

CO1A1.cutoff <- quantile(global.long[endsWith(global.long$Specie, "CO1A1_RAT"),]$Percent_Abundance, 0.95)# boxplots of these proteins for each condition

# try to normalize after removing these biggest proteins
global.df.wo.bigProts <- global.df[!(global.df$Specie %in% bigProts$Specie),]
rownames(global.df) <- global.df$Specie
global.ratio <- log(global.df[,2:ncol(global.df)]/global.df$ref, 2)
global.ratio$Specie <- rownames(global.ratio)
ratio.long <- reshape2::melt(global.ratio, variable.name = "Sample", value.name = "Log2Ratio")
ratio.long <- ratio.long[ratio.long$Sample != "ref",]

hist(ratio.long$Log2Ratio) # most proteins have lower abundance in samples than ref
hist(ratio.long[ratio.long$Log2Ratio > 0,]$Log2Ratio)
log2cutoff <- quantile(ratio.long$Log2Ratio, 0.95, na.rm = TRUE)
write.csv(global.ratio[,-"Specie"], "Global_original_without_")

# revisit distribution without these abundant proteins
filtered.global2 <- global.long[!(global.long$Specie %in% bigProts$Specie),]
filtered.prot.summary2 <- prot.summary[!(prot.summary$Specie %in% bigProts),]
hist(global.long[!(global.long$Specie %in% bigProts$Specie),]$Abundance)
hist(global.long[!(global.long$Specie %in% bigProts$Specie),]$Percent_Abundance)
hist(prot.summary[!(prot.summary$Specie %in% bigProts),]$median_perc_abundance)
hist(prot.summary[!(prot.summary$Specie %in% bigProts),]$max_perc_abundance)

# plot 

```

# 6. define outlier proteins at top 5% of proteins with higher % in samples than ref
# 7. look at homology in outlier proteins - if marked as human but also in mouse/rat, then also add mouse/rat versions to list
```{r sum CO1A1/2}
global.long.w.ref <- global.long
global.long <- global.long[global.long$Sample != "ref",]|>
  tidyr::separate(Sample,into=c('PDX','Drug','Timepoint (hrs)','Unit','Replicate'),sep='_',remove=FALSE)
global.long$`Timepoint (hrs)` <- factor(global.long$`Timepoint (hrs)`, levels = c("0", "0.5", "3", "24"))
global.long$Replicate <- substr(global.long$Replicate, 4,4)
CO1A1.long <- global.long[global.long$Sample != "ref" & (grepl("CO1A1", global.long$Specie) | global.long$Specie == "NP_000079.2"),]
CO1A2.long <- global.long[global.long$Sample != "ref" & (grepl("CO1A2", global.long$Specie) | global.long$Specie == "NP_000080.2"),]
CO1A.long <- rbind(CO1A1.long, CO1A2.long)

# boxplot
collagen.list <- list("CO1A1" = CO1A1.long, "CO1A2" = CO1A2.long, "Collagen Type I" = CO1A.long)
for (i in 1:length(collagen.list)) {
  temp.collagen.df <- ddply(collagen.list[[i]], .(Sample, PDX, Drug, `Timepoint (hrs)`, Replicate), summarize,
                  sum_abundance = sum(Abundance, na.rm = TRUE),
                  sum_percent_abundance = sum(Percent_Abundance, na.rm = TRUE))
  write.csv(temp.collagen.df, paste0(names(collagen.list)[i], "_percent_abundance.csv"), row.names = FALSE)
  
  pdf(paste0(names(collagen.list)[i], "_percent_histogram.pdf"))
  hist(temp.collagen.df$sum_percent_abundance, xlab = "% Abundance of Collagen Type I", main="")
  dev.off()
  
  ggplot(temp.collagen.df,
         aes(x=`Timepoint (hrs)`, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(cols=vars(PDX)) + ylab(paste0("% Abundance of ", names(collagen.list)[i]))
  ggsave(paste0(names(collagen.list)[i], "_percent_boxplot.pdf"))
  
  # scatter plot
  ggplot(temp.collagen.df,
         aes(x=Replicate, y=`sum_percent_abundance`, color = Drug)) +
    geom_point() + facet_grid(PDX ~ `Timepoint (hrs)`) + ylab(paste0("% Abundance of ", names(collagen.list)[i]))
  ggsave(paste0(names(collagen.list)[i], "_percent_scatterPlot.pdf")) 
}
t.test(temp.collagen.df[temp.collagen.df$PDX=="JH.2.002",]$sum_percent_abundance, temp.collagen.df[temp.collagen.df$PDX=="MN.2",]$sum_percent_abundance)
# JH-2-002 samples are more contaminated than MN-2 with p = 4.646E-10
  
  
# repeat but for each protein name (sum across species)
human.bigProts.map <- readxl::read_excel("Human_top1Percent_abundant_proteins_idmapping_2024_08_23.xlsx")

#load protein names
unique.bigProts <- read.csv("Contamination_protein_names_UniProt.csv")
all.summed <- list()
for (i in unique.bigProts[,1]) {
  # get data for this protein across all species
  temp.refseq <- na.omit(unique(human.bigProts.map[human.bigProts.map$`generic protein name` == i,]$From))
  temp.long <- global.long[global.long$Sample != "ref" & 
                              (grepl(i, global.long$Specie) | global.long$Specie %in% temp.refseq),]
  
  # sum data across species
  temp.df <- ddply(temp.long, .(Sample, PDX, Drug, `Timepoint (hrs)`, Replicate), summarize,
                  sum_abundance = sum(Abundance, na.rm = TRUE),
                  sum_percent_abundance = sum(Percent_Abundance, na.rm = TRUE))
  all.summed[[i]] <- temp.df
  write.csv(temp.df, paste0(i, "_percent_abundance.csv"), row.names = FALSE)
  
  pdf(paste0(i, "_percent_histogram.pdf"))
  hist(temp.df$sum_percent_abundance, xlab = paste("% Abundance of", i), main="")
  dev.off()
  
  ggplot(temp.df,
         aes(x=`Timepoint (hrs)`, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(cols=vars(PDX)) + ylab(paste("% Abundance of", i))
  ggsave(paste0(i, "_percent_boxplot.pdf"))
  
  # scatter plot
  ggplot(temp.df,
         aes(x=Replicate, y=`sum_percent_abundance`, color = Drug)) +
    geom_point() + facet_grid(PDX ~ `Timepoint (hrs)`) + ylab(paste("% Abundance of", i))
  ggsave(paste0(i, "_percent_scatterPlot.pdf")) 
}
all.summed.df <- data.table::rbindlist(all.summed, use.names = TRUE, idcol = "Protein")

ggplot(all.summed.df,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ `Timepoint (hrs)`) + ylab("% Abundance")
  ggsave("Top1PercentAbundantProteins_percent_boxplot.pdf", width = 40)
  
  ggplot(all.summed.df,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(rows=vars(PDX)) + ylab("% Abundance") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave("Top1PercentAbundantProteins_PDX_percent_boxplot.pdf", width = 40)
  
  # need to reduce list further
  top.bigProts <- bigProts %>% slice_max(median_perc_abundance, n=50)
  top.bigProts$Name <- top.bigProts$Specie
  top.names <- c()
  for (i in 1:nrow(top.bigProts)) {
    if (grepl("_MOUSE", top.bigProts$Specie[i]) | grepl("_RAT", top.bigProts$Specie[i])) {
      top.bigProts$Name[i] <- strsplit(top.bigProts$Specie[i], "[|]")[[1]][2]
    } else {
      top.bigProts$Name[i] <- paste0(na.omit(unique(human.bigProts.map[human.bigProts.map$From == top.bigProts$Specie[i],]$`generic protein name`)), 
                                     collapse = " ")
    }
    top.names <- unique(c(top.names, top.bigProts$Name[i]))
  }
  top.all.summed.df <- all.summed.df[all.summed.df$Protein %in% top.names,] # 6 if pick 20, 14 if pick 50
  median.top.all.summed.df <- ddply(top.all.summed.df, .(Protein), summarize,
                                    median_sum_perc_abundance = median(sum_percent_abundance, na.rm = TRUE),
                                    max_sum_perc_abundance = max(sum_percent_abundance, na.rm = TRUE))
  top.all.summed.df <- merge(top.all.summed.df, median.top.all.summed.df, all.x = TRUE)
  top.all.summed.df <- top.all.summed.df[order(top.all.summed.df$median_sum_perc_abundance, decreasing = TRUE),]
  
  ggplot(top.all.summed.df,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ `Timepoint (hrs)`) + ylab("% Abundance")
  ggsave("Top14_Top1PercentAbundantProteins_PDX_time_percent_boxplot.pdf")
  
  ggplot(top.all.summed.df,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(rows=vars(PDX)) + ylab("% Abundance") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave("Top14_Top1PercentAbundantProteins_PDX_percent_boxplot.pdf")
  
  ggplot(top.all.summed.df,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + ylab("% Abundance") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave("Top14_Top1PercentAbundantProteins_percent_boxplot.pdf")

  # rename and compile before selection
  median.all.summed.df <- ddply(all.summed.df, .(Protein), summarize,
                                    median_sum_perc_abundance = median(sum_percent_abundance, na.rm = TRUE),
                                    max_sum_perc_abundance = max(sum_percent_abundance, na.rm = TRUE))
  all.summed.df.v2 <- merge(all.summed.df, median.all.summed.df, all.x = TRUE)
  n.top <- 20
  top.prots <- median.all.summed.df %>% slice_max(median_sum_perc_abundance, n=20)
  top.all.summed.df.v2 <- all.summed.df.v2[all.summed.df.v2$Protein %in% top.prots$Protein,]
  top.all.summed.df.v2 <- top.all.summed.df.v2[order(top.all.summed.df.v2$median_sum_perc_abundance, decreasing = TRUE),]
  
  ggplot(top.all.summed.df.v2,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(PDX ~ `Timepoint (hrs)`) + ylab("% Abundance") + scale_x_discrete(limit = unique(top.all.summed.df.v2$Protein)) + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave(paste0("Top_", n.top,"_Top1PercentAbundantProteins_PDX_time_percent_boxplot.pdf"))
  
  ggplot(top.all.summed.df.v2,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + facet_grid(rows=vars(PDX)) + ylab("% Abundance") + scale_x_discrete(limit = unique(top.all.summed.df.v2$Protein)) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave(paste0("Top_", n.top, "_Top1PercentAbundantProteins_PDX_percent_boxplot.pdf"))
  
  ggplot(top.all.summed.df.v2,
         aes(x=Protein, y=`sum_percent_abundance`, fill = Drug)) +
    geom_boxplot() + ylab("% Abundance") + scale_x_discrete(limit = unique(top.all.summed.df.v2$Protein)) + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave(paste0("Top_", n.top, "_Top1PercentAbundantProteins_percent_boxplot.pdf"))

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
