---
title: "DEseq2 data with LRT"
author: "Kuangying Yang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    latex_engine: pdflatex
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    #latex_engine: xelatex
      
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =F,message = F,warning = F,include=F)
```

```{r}
rm(list=ls())
```

```{r}
library(biomaRt) 
library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
```


```{r}
source(file.path(getwd(),"Rfunction.R" ))
Path <- gsub("/R(/Rcode)?$", "", getwd())
Rawdata_path=file.path(Path,"Rawdata")
dir.create(Rawdata_path)
Process_path=file.path(Path,"Processed_data")
dir.create(Process_path)
```




```{r}
# STAR-Salmon
Reference_path="/Users/hirbea/Desktop/Kuangying/Thesis/RNAseq/MPNST_PDX_drug/Reference"
STAR_path="/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_NF_RNAseq/Nextflow3--star_index/star_salmon"
dir("/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_NF_RNAseq")
Rawcount_path=file.path(Rawdata_path,"salmon.merged.gene_counts.tsv")
TPM_path=file.path(Rawdata_path,"salmon.merged.gene_tpm.tsv")
Raw_counts=readr::read_tsv(Rawcount_path)
TPM_counts=readr::read_tsv(TPM_path)
load(file.path(Reference_path,
               #"Human_GSEA",
               "Protein_coding_genelist.rdata")) # loading Protein_coding_genelist data
Raw_counts_2=Raw_counts%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()%>%mutate_all(.,as.integer)
Tumor_count_log=log2(Raw_counts_2+1)
TPM_2=TPM_counts%>%.[,c(2:ncol(.))]%>%dplyr::filter(gene_name%in%Protein_coding_genelist)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()
Drug_anno=readxl::read_excel(file.path(Rawdata_path,"Drug_annotation.xlsx"))
Drug_class_name="DNMTi"

```

# Switch 8h with 24h in DMSO_d1; remove MN.2_MPNST_b2_0h_DMSO3_d2 in DMSO_d2

## Original
```{r}
RNAseq_anno_org=data.frame(Sample_name=colnames(Raw_counts_2))%>%mutate(individualID="MN2")%>% mutate(Drug =gsub(".*\\_","",Sample_name))%>%
                  mutate(Drug=ifelse(grepl("DMSO",Sample_name),"DMSO",Drug )) %>%
                            mutate(Timepoint=case_when(
                              grepl("0h",Sample_name)~"0h",
                              grepl("8h",Sample_name)~"8h",
                              grepl("24h",Sample_name)~"24h",
                              TRUE ~ NA_character_)) %>%
  mutate(Drug =gsub("\\d","",Drug))%>%left_join(.,Drug_anno,by="Drug")%>%mutate(Time=ifelse(grepl("SEL|TRA$|RIB|PAL|RMC|VOR|TNO",Drug),"12192024","12272024"))%>%mutate(Time=ifelse(grepl("d1",Sample_name),"DMSO_d1",ifelse(grepl("d2",Sample_name),"DMSO_d2",Time )))
Drug_name = RNAseq_anno_org$Drug %>% unique()
Drug_name=factor(Drug_name,levels = Drug_name)
# Ensure you have 14 unique Drug names (1 for black, 13 other colors)
if (length(Drug_name) != 14) {
  stop("The number of unique drugs should be 14, including the one colored black.")
}

# Use RColorBrewer to get 13 colors from the Paired palette
paired_colors =RColorBrewer:: brewer.pal(n = 12, name = "Paired")
extra_color =RColorBrewer:: brewer.pal(n = 8, name = "Set3")[1]  # Take one color from another palette (e.g., Set3)

# Combine black and the 13 colors
drug_colors = c("black", paired_colors, extra_color)
set.seed(12345)
RNAseq_DMSO_D1=RNAseq_anno_org%>%dplyr::filter(Drug=="DMSO")
P1=PCA_Plot_3_shape (t(Tumor_count_log),RNAseq_DMSO_D1,"Sample_name","Drug","Timepoint")+
  scale_color_manual(values = drug_colors)+facet_wrap(~Time)

```

# After switching and removing
```{r}

RNAseq_anno=RNAseq_anno_org%>%mutate(Timepoint=ifelse(Time=="DMSO_d1"&Timepoint=="8h","24hc",ifelse(Time=="DMSO_d1"&Timepoint=="24h","8hc",Timepoint )))%>%mutate(Timepoint=gsub("c","",Timepoint))
Drug_name = RNAseq_anno$Drug %>% unique()
Drug_name=factor(Drug_name,levels = Drug_name)
# Ensure you have 14 unique Drug names (1 for black, 13 other colors)
if (length(Drug_name) != 14) {
  stop("The number of unique drugs should be 14, including the one colored black.")
}

# Use RColorBrewer to get 13 colors from the Paired palette
paired_colors =RColorBrewer:: brewer.pal(n = 12, name = "Paired")
extra_color =RColorBrewer:: brewer.pal(n = 8, name = "Set3")[1]  # Take one color from another palette (e.g., Set3)

# Combine black and the 13 colors
drug_colors = c("black", paired_colors, extra_color)
set.seed(12345)
RNAseq_DMSO_D1=RNAseq_anno%>%dplyr::filter(Drug=="DMSO")%>%dplyr::filter(Sample_name!="MN.2_MPNST_b2_0h_DMSO3_d2")
P2=PCA_Plot_3_shape (t(Tumor_count_log),RNAseq_DMSO_D1,"Sample_name","Drug","Timepoint")+
  scale_color_manual(values = drug_colors)+facet_wrap(~Time)

                            
```
                            
```{r,include=T,fig.width=8,fig.height=8}
cowplot::plot_grid(P1,P2, labels = c("Original","Corrected"),nrow=2)
```

# Select DMSO_D1 as control

```{r}

Drug_pickup=RNAseq_anno%>%dplyr::filter(grepl(Drug_class_name,Drug_class))

# Select one DMSO group (There are two DMSO group)
DMSO_data=RNAseq_anno%>%dplyr::filter(Drug=="DMSO")%>%dplyr::filter(grepl("d1",Sample_name))

All_data=bind_rows(Drug_pickup,DMSO_data)%>%arrange(Drug,Timepoint)
```

```{r,include=T}
knitr::kable(All_data[,1:6],caption = "Sample information")%>%kableExtra::kable_styling(latex_options = c("scale_down","hold_position"))
```
## Data reorganization: adding 0 timepoint in Raw count Data and TPM data for drug group
## Deseq2 vocanal plot

```{r}
if ( !file.exists(file.path(Process_path,Drug_class_name,"Deseq2_OUT_vs_DMSO_d1.rdata" ))) {
 Drug_reslist=list()

  for (i in Drug_pickup$Drug) {
  DMSO_rawcounts=Raw_counts_2[,DMSO_data$Sample_name]
  DMSO_anno=DMSO_data%>%.[,c("Sample_name","Drug","Timepoint")]
  Drug_samples=RNAseq_anno%>%dplyr::filter(Drug==i)%>%.[,c("Sample_name","Drug","Timepoint")]
  Drug_rawcounts=Raw_counts_2[,Drug_samples$Sample_name]
  Drug_0h=DMSO_rawcounts[,grepl("0h",colnames(DMSO_rawcounts))]
  colnames(Drug_0h)=gsub("DMSO",i,colnames(Drug_0h))
  Drug_0h_anno=data.frame(Sample_name=colnames(Drug_0h),Drug=i,Timepoint="0h")
  New_counts=cbind(DMSO_rawcounts,Drug_0h,Drug_rawcounts)
  New_anno=rbind(DMSO_anno,Drug_0h_anno,Drug_samples)%>%mutate(Group=paste0(Drug,"_",Timepoint))
  New_anno$Drug=factor(New_anno$Drug,levels = c("DMSO",i))
  New_anno$Timepoint=factor(New_anno$Timepoint,levels = c("0h","8h","24h"))
  rownames(New_anno)=New_anno[,1]
  filter_counts <- selectGenes_BY_group(New_counts, min.count=10, N=0.90,New_anno)
  deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =filter_counts, colData =New_anno,design = ~Group)
  deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
  #dds_LRT= DESeq2::DESeq(deseq2Data,test="LRT", reduced =~Drug+Timepoint)
  #res_LRT = DESeq2::results(dds_LRT)
  dds <- DESeq2::DESeq(deseq2Data)
  res_list=list()
  Timepoint=c("8h","24h")
  for (t in Timepoint) {
    drug_0h=paste0(i,"_","0h")
    DMSO_t=paste0("DMSO","_",t)
    group1=paste0(i,"_",t)
    res_list[[t]]=list(drug_vs_0h=DESeq2::results(dds, contrast = c("Group", group1, drug_0h)),
                      drug_vs_DMSO=DESeq2::results(dds, contrast = c("Group", group1, DMSO_t)))
    
    names(res_list[[t]])=c(paste0(i,"_",t,"_vs_",i ,"_0h"), paste0(i,"_",t,"_vs_","DMSO_",t))
  }
 
Drug_reslist[[i]]=list(Counts=New_counts,Anno=New_anno,deseq2VST,DDS=dds,VST=deseq2VST,RES=res_list)
  
}


names(Drug_reslist)
dir.create(file.path(Process_path,Drug_class_name))

save(Drug_reslist,file=file.path(Process_path,Drug_class_name,"Deseq2_OUT_vs_DMSO_d1.rdata" ))
} 

load(file.path(Process_path,Drug_class_name,"Deseq2_OUT_vs_DMSO_d1.rdata" ))
```

```{r,include=T,fig.width=12,fig.height=12}
plot_list=list()
for (d in unique(Drug_pickup$Drug)) {
 Drug= Drug_pickup$Drug[d]
 Data=Drug_reslist[[d]]
 RES_all=Data$RES


 for (i in 1:length(RES_all)) {
   RES_t=RES_all[[i]]
   time=names(RES_all)[i]
    for (t in names(RES_t)){
      RES1=RES_t[[t]]%>%as.data.frame()%>%mutate(gene_symbol=rownames(.))
      #list_name=t
      plot_list[[t]]=Volcano_plot_rnaseq_simple_2(RES1,top_n = 20)+ggtitle(t)
   
  }
  
  
}
}

cowplot::plot_grid(plotlist=plot_list,ncol=4)
```

## PCA plot based on VST data
```{r,include=T,fig.width=8,fig.height=6}
plot_list=list()
for (d in unique(Drug_pickup$Drug)) {
 Drug= Drug_pickup$Drug[d]
 Data=Drug_reslist[[d]]
 RES_all=Data$VST%>%as.data.frame()
 plot_list[[d]]=PCA_Plot_3_shape (t(RES_all),All_data,"Sample_name","Drug","Timepoint")+
  scale_color_manual(values = drug_colors)
 colnames(RES_all)
}


cowplot::plot_grid(plotlist=plot_list,ncol=1)
```


## DEGs numbers 

```{r,include=T,fig.width=12,fig.height=12}
plot_list=list()
for (d in unique(Drug_pickup$Drug)) {
 Drug= Drug_pickup$Drug[d]
 Data=Drug_reslist[[d]]
 RES_all=Data$RES


 for (i in 1:length(RES_all)) {
   RES_t=RES_all[[i]]
   time=names(RES_all)[i]
    for (t in names(RES_t)){
      RES1=RES_t[[t]]%>%as.data.frame()%>%mutate(gene_symbol=rownames(.))
      #list_name=t
      plot_list[[t]]=Bar_plot_RES (RES1, FC_cutoff = 0.67)+ggtitle(t)
   
  }
  
  
}
}

cowplot::plot_grid(plotlist=plot_list,ncol=4)
```

## 10 MAPK pathway genes signature

```{r,include=T}
# MAPK_genes=c("PHLDA1","SPRY2", "SPRY4", "ETV4", "ETV5", "DUSP4", "DUSP6", "CCND1", "EPHA2", "EPHA4") 
# print(MAPK_genes)
```

## DNMT pathway genes signature
```{r}
DNMT_genes=c("DNMT1","DNMT3A","DNMT3B", "MGMT", "TET1", "TET2", "IDH1", "IDH2", "EZH2", "KMT2A", 
             "IFNA2", "IFNA14", "IFNAR1", "IFNAR2", "UBR5",
             "EED", "SUZ12") 
print(DNMT_genes)
```


```{r,include=F}

out=list()

for (i in names(Drug_reslist)){
  Data=Drug_reslist[[i]]
  Sample_list=Data$Counts%>%colnames(.)
  Genes=rownames(Data$RES[[1]][[1]])
  TPM_pickup_2=TPM_2%>%.[,colnames(.) %in%Sample_list]
  Annotion_file=Data$Anno%>%arrange(Timepoint,Drug)%>%dplyr::filter(Sample_name%in%colnames(TPM_pickup_2))
  Annotion_file_2=Annotion_file%>%.[,c("Timepoint","Drug")]
  TPM_pickup_3=TPM_pickup_2%>%as.data.frame()%>%.[Annotion_file$Sample_name] 
  rownames(TPM_pickup_3)=rownames(TPM_pickup_2)
  Color_Set1=RColorBrewer::brewer.pal(5,"Set1")[1:2]
  names(Color_Set1)=unique(Annotion_file$Drug)
  Color_Set2=RColorBrewer::brewer.pal(5,"Set1")[2:4]
  names(Color_Set2)=unique(Annotion_file$Timepoint)
 
  Color_list=list(Drug=Color_Set1,Timepoint=Color_Set2)
  number= Annotion_file_2%>%dplyr::filter(Timepoint=="8h")%>%nrow()
 out[[i]]=TPM_heatmap(TPM_pickup_3,DNMT_genes,Annotation_col=Annotion_file_2,Annotation_color=Color_list,gap_set = c(3,6,9,12,1) ,Title="DNMT_genes")%>%.$gtable #c(3,3+number,3+number)

}

```

```{r,fig.width=6,fig.height=6,include=T}
cowplot::plot_grid(plotlist=out,ncol=1)
```
```{r}
#knitr::knit_exit()
```


```{r,include=F}
GMT_files=list.files(file.path(Reference_path,"Human_GSEA"),pattern="all.v7.5.1.symbols.gmt.txt",full.names = T)%>%.[grepl("c2|c5|h",.)]

FGSEA_list=list()
FGSEA_q005_list=list()
for (d in unique(Drug_pickup$Drug)) {
  Data_drug=Drug_reslist[[d]]$RES
  for (i in 1:length(Data_drug)) {
     data_drug_time=Data_drug[[i]]
     names(data_drug_time)
     for (j in 1:length(data_drug_time)) {
      data_drug_time_group=data_drug_time[[j]]%>%as.data.frame() 
      name2=names(data_drug_time)[j]
      print(name2)
      res_q005=data_drug_time_group%>%dplyr::filter(padj<0.05)
      res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
      ranks <- tibble::deframe(res_q005_rank)
      out_list=list()
      set.seed(12345)
      for (x in 1:length(GMT_files)){
        Data=fgsea::gmtPathways(GMT_files[x])
        set.seed(12345)
        out_list[[x]] <- fgsea::fgseaMultilevel(Data, ranks,
                           minSize=10,
                           maxSize=500,nPermSimple = 10000)
     }
    FGSEA_list[[name2]]=do.call("rbind",out_list)
    FGSEA_q005_list[[name2]]=do.call("rbind",out_list)%>%as.data.frame()%>%dplyr::filter(padj<0.05)
   }
   }
 }
```

```{r}
Pickup_groups=names(FGSEA_q005_list)%>%.[grepl("DMSO",.)]
```

```{r,include=T}
FGSEA_q005_list_pickup=FGSEA_q005_list[Pickup_groups]
FGSEA_q005_list_pickup_DNMT=lapply(FGSEA_q005_list_pickup, function(x) out=x%>%dplyr::filter(grepl("DNMT",pathway)) )
Out=do.call("rbind",FGSEA_q005_list_pickup_DNMT)%>%tibble::rownames_to_column(var="Group")%>%.[,c(1,2,3,7)]
Out$Group=gsub("\\..*","",Out$Group)
knitr::kable(x=Out,caption = "DNMT pathways (padj<0.05)")%>%kableExtra::kable_styling(latex_options = c("scale_down","hold_position"))%>%kableExtra:: column_spec(column=c(1:2),width="5cm")
```
```{r}
#knitr::knit_exit()
```



```{r,}
output_plot=list()
for (i in names(FGSEA_q005_list_pickup )){
 
  data=FGSEA_q005_list_pickup[[i]]%>%dplyr::filter(grepl("KEGG|WP_|HALLMARK|REACTOME|GO",pathway))
  up_NES=data%>%dplyr::filter(NES>0)%>%arrange(desc(NES))%>%.[1:10,]%>%na.omit()
  Dn_NES=data%>%dplyr::filter(NES<0)%>%arrange(NES)%>%.[1:10,]%>%na.omit()
  Outlist=list()
  if (nrow(up_NES)>0) {
    
    Outlist[["up"]]=Enrich_dotplot_Single(up_NES)+ggtitle(i)
  }
  if (nrow(Dn_NES)>0) {
    
    Outlist[["DN"]]=Enrich_dotplot_Single(Dn_NES)+ggtitle(i)
  }
  

  output_plot[[i]]=cowplot::plot_grid(plotlist =  Outlist,nrow=1)
}

```

# TOP 10 up/DN pathways in each drug vs DMSO 
**(KEGG|WP_|HALLMARK|REACTOME|GO)**


```{r,fig.width=28,fig.height=12,include=T}
output_plot[[1]]
```
\newpage
```{r,fig.width=40,fig.height=20,include=T}
output_plot[[2]]
```

\newpage
```{r,fig.width=28,fig.height=10,include=T}
# output_plot[[3]]
```
\newpage
```{r,fig.width=14,fig.height=6,include=T}
# output_plot[[4]]
```

```{r}
knitr::knit_exit()
```












#----------------- end here--------------




```{r}
FGSEA_list=list()
Res_list_8h=res_list[[1]]
 
 FGSEA_q005_list=list()
 for (i in 1:length(Res_list_8h)){
  data=Res_list_8h[[i]]
  name2=names(Res_list_8h)[[i]]
  res_q005=data%>%as.data.frame()%>%dplyr::filter(padj<0.05)
  
  res_q005_rank=res_q005%>%arrange(log2FoldChange) %>%.[c("log2FoldChange")]%>%as.data.frame() %>%tibble::rownames_to_column()
  ranks <- tibble::deframe(res_q005_rank)
  out_list=list()
  for (x in 1:length(GMT_files)){
 
     data=fgsea::gmtPathways(GMT_files[x])
    set.seed(12345)
    out_list[[x]] <- fgsea::fgseaMultilevel(data, ranks,
                           minSize=10,
                           maxSize=500,nPermSimple = 10000)
  }
                           
  FGSEA_list[[name2]]=do.call("rbind",out_list)
  FGSEA_q005_list[[name2]]=do.call("rbind",out_list)%>%as.data.frame()%>%dplyr::filter(padj<0.05)

}

all_sig_pathway_name=lapply(FGSEA_q005_list,function(x) x$pathway )%>%unlist()%>%unique()


NES_sig_pathway_impute=list()
for(i in names(FGSEA_list)){
  Nes_data=FGSEA_list[[i]]
  NES_sig_pathway_impute[[i]]=data.frame(pathway=all_sig_pathway_name)%>%left_join(., Nes_data)%>%mutate_at("NES",~tidyr::replace_na(.,0))%>%mutate_at("padj", ~tidyr::replace_na(.,1))%>%mutate(group=i)
  
  
}

 Pathwaylist_sig_all=do.call("rbind",NES_sig_pathway_impute)
 Pathwaylist_sig_all$group=factor(Pathwaylist_sig_all$group,c(names(NES_sig_pathway_impute)))
 
# if (!file.exists(file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))) {
#save(Pathwaylist_sig_all,FGSEA_q005_list,file=file.path(Rawdata_path,"2_Flipped_rm_one_sample_GSEA_data.rdata" ))}
```



# Select DMSO D2 as control

```{r}

Drug_pickup=RNAseq_anno%>%dplyr::filter(grepl(Drug_class_name,Drug_class))

# Select one DMSO group (There are two DMSO group)
DMSO_data=RNAseq_anno%>%dplyr::filter(Drug=="DMSO")%>%dplyr::filter(grepl("d2",Sample_name))

All_data=bind_rows(Drug_pickup,DMSO_data)%>%arrange(Drug,Timepoint)
```

```{r,include=T}
knitr::kable(All_data[,1:6],caption = "Sample information")%>%kableExtra::kable_styling(latex_options = c("scale_down","hold_position"))
```

## Data reorganization: adding 0 timepoint in Raw count Data and TPM data for drug group  
## Deseq2 

```{r}
Drug_reslist=list()

for (i in Drug_pickup$Drug) {
  
  DMSO_rawcounts=Raw_counts_2[,DMSO_data$Sample_name]
  DMSO_anno=DMSO_data%>%.[,c("Sample_name","Drug","Timepoint")]
  Drug_samples=RNAseq_anno%>%dplyr::filter(Drug==i)%>%.[,c("Sample_name","Drug","Timepoint")]
  Drug_rawcounts=Raw_counts_2[,Drug_samples$Sample_name]
  Drug_0h=DMSO_rawcounts[,grepl("0h",colnames(DMSO_rawcounts))]
  colnames(Drug_0h)=gsub("DMSO",i,colnames(Drug_0h))
  Drug_0h_anno=data.frame(Sample_name=colnames(Drug_0h),Drug=i,Timepoint="0h")
  New_counts=cbind(DMSO_rawcounts,Drug_0h,Drug_rawcounts)
  New_anno=rbind(DMSO_anno,Drug_0h_anno,Drug_samples)%>%mutate(Group=paste0(Drug,"_",Timepoint))
  New_anno$Drug=factor(New_anno$Drug,levels = c("DMSO",i))
  New_anno$Timepoint=factor(New_anno$Timepoint,levels = c("0h","8h","24h"))
  rownames(New_anno)=New_anno[,1]
 

  filter_counts <- selectGenes_BY_group(New_counts, min.count=10, N=0.90,New_anno)
  deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =filter_counts, colData =New_anno,design = ~Group)
  deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
  #dds_LRT= DESeq2::DESeq(deseq2Data,test="LRT", reduced =~Drug+Timepoint)
  #res_LRT = DESeq2::results(dds_LRT)
  dds <- DESeq2::DESeq(deseq2Data)
  res_list=list()
  Timepoint=c("8h","24h")
  for (t in Timepoint) {
    drug_0h=paste0(i,"_","0h")
    DMSO_t=paste0("DMSO","_",t)
    group1=paste0(i,"_",t)
    res_list[[t]]=list(drug_vs_0h=DESeq2::results(dds, contrast = c("Group", group1, drug_0h)),
                      drug_vs_DMSO=DESeq2::results(dds, contrast = c("Group", group1, DMSO_t)))
    
    names(res_list[[t]])=c(paste0(i,"_",t,"_vs_",i ,"_0h"), paste0(i,"_",t,"_vs_","DMSO_",t))
  }
 
Drug_reslist[[i]]=list(Counts=New_counts,Anno=New_anno,deseq2VST,DDS=dds,VST=deseq2VST,RES=res_list)
  
}


names(Drug_reslist)
save(Drug_reslist,file=file.path(Process_path,"Deseq2_OUT_vs_DMSO_d2.rdata" ))
 
```




```{r,include=T,fig.width=12,fig.height=12}
plot_list=list()
for (d in unique(Drug_pickup$Drug)) {
 Drug= Drug_pickup$Drug[d]
 Data=Drug_reslist[[d]]
 RES_all=Data$RES


 for (i in 1:length(RES_all)) {
   RES_t=RES_all[[i]]
   time=names(RES_all)[i]
    for (t in names(RES_t)){
      RES1=RES_t[[t]]%>%as.data.frame()%>%mutate(gene_symbol=rownames(.))
      #list_name=t
      plot_list[[t]]=Volcano_plot_rnaseq_simple_2(RES1,top_n = 20)+ggtitle(t)
   
  }
  
  
}
}

cowplot::plot_grid(plotlist=plot_list,ncol=4)
```


```{r,include=T,fig.width=16,fig.height=4}
plot_list=list()
for (d in unique(Drug_pickup$Drug)) {
 Drug= Drug_pickup$Drug[d]
 Data=Drug_reslist[[d]]
 RES_all=Data$VST%>%as.data.frame()
 plot_list[[d]]=PCA_Plot_3_shape (t(RES_all),All_data,"Sample_name","Drug","Timepoint")
 colnames(RES_all)
}


cowplot::plot_grid(plotlist=plot_list,ncol=4)
```


# Stat


```{r,include=T,fig.width=12,fig.height=12}
plot_list=list()
for (d in unique(Drug_pickup$Drug)) {
 Drug= Drug_pickup$Drug[d]
 Data=Drug_reslist[[d]]
 RES_all=Data$RES


 for (i in 1:length(RES_all)) {
   RES_t=RES_all[[i]]
   time=names(RES_all)[i]
    for (t in names(RES_t)){
      RES1=RES_t[[t]]%>%as.data.frame()%>%mutate(gene_symbol=rownames(.))
      #list_name=t
      plot_list[[t]]=Bar_plot_RES (RES1, FC_cutoff = 0.67)+ggtitle(t)
   
  }
  
  
}
}

cowplot::plot_grid(plotlist=plot_list,ncol=4)
```


```{r}
#file_name=paste0(Drug_class_name,"_Deseq2.rdata")
#save(Drug_reslist,file=file.path(Process_path, file_name))
```


```{r}
knitr::knit_exit()
```


```{r,fig.width=12,fig.height=18}
for (d in seq_along(Drug_pickup$Drug)) {
  
  # Retrieve the current drug
  Drug <- Drug_pickup$Drug[d]
  
  # Access the corresponding results list for the current drug
  Data <- Drug_reslist[[d]]
  RES_all <- Data$RES
  
  for (i in seq_along(RES_all)) {
    
    # Retrieve the list of results for the current index
    RES_t <- RES_all[[i]]
    
    for (t in names(RES_t)) {
      
      # Access the specific result by name
      RES1 <- RES_t[[t]]
      
      # Load the DESeq2 library if not already loaded
      if (!requireNamespace("DESeq2", quietly = TRUE)) {
        library(DESeq2)
      }
      
      # Plot the MA plot for the current result
      plotMA(RES1, ylim = c(-5, 5), main = paste(Drug, "-", t))
    }
  }
}
```


```{r}
# Sample_anno_2=Sample_anno%>%dplyr::filter(Drug!="Control")%>%mutate(Group=paste0(Drug,"_",Timepoint))
# Raw_counts_rm_control=Raw_counts_2%>%.[,Sample_anno_2$Sample_name]
Raw_counts_add_control=Raw_counts_2%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)
Sample_anno_2=data.frame(Sample_name=colnames(Raw_counts_add_control))%>%left_join(.,Sample_anno)%>%mutate(Drug=ifelse(is.na(Drug),"Mirda",Drug))%>%mutate(Timepoint=ifelse(is.na(Timepoint),"0hour",Timepoint))%>%mutate(Group=paste0(Drug,"_",Timepoint))
rownames(Sample_anno_2)=Sample_anno_2$Sample_name
Sample_anno_2$Timepoint=factor(Sample_anno_2$Timepoint,levels = c("0hour","8hour","24hour"))
library(SummarizedExperiment)
library(DESeq2)

keep.exprs <- selectGenes_BY_group(Raw_counts_add_control, min.count=10, N=0.90,Sample_anno_2)
keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
myFilt <- Raw_counts_add_control[rownames(keep.exprs_2),]
dim(myFilt)
}


```


```{r}
# Raw_counts_add_control=Raw_counts_2%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)
# Sample_anno_2=data.frame(Sample_name=colnames(Raw_counts_add_control))%>%left_join(.,Sample_anno)%>%mutate(Drug=ifelse(is.na(Drug),"Mirda",Drug))%>%mutate(Timepoint=ifelse(is.na(Timepoint),"0hour",Timepoint))%>%mutate(Group=paste0(Drug,"_",Timepoint))
# rownames(Sample_anno_2)=Sample_anno_2$Sample_name
# Sample_anno_2$Timepoint=factor(Sample_anno_2$Timepoint,levels = c("0hour","8hour","24hour"))
# library(SummarizedExperiment)
# library(DESeq2)
#   keep.exprs <- selectGenes_BY_group(Raw_counts_add_control, min.count=10, N=0.90,Sample_anno_2)
# keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
# myFilt <- Raw_counts_add_control[rownames(keep.exprs_2),]
# 
# dim(myFilt)



```
 

```{r,fig.width=12,fig.height=6,include=T}
Tumor_count_log=log2(Raw_counts_2+1)
Drug_count_plot(Tumor_count_log,RNAseq_anno)
```

```{r}
set.seed(12345)


P1=PCA_Plot_3  (t(Tumor_count_log),RNAseq_anno,"Sample_name","Drug")
P2=PCA_Plot_3  (t(Tumor_count_log),RNAseq_anno,"Sample_name","Timepoint")
P3=PCA_Plot_3  (t(Tumor_count_log),RNAseq_anno,"Sample_name","Drug_class")

```

```{r,fig.width=12,fig.height=8,include=T}
cowplot::plot_grid(P1,P2,P3,nrow=2)
```

```{r,fig.width=12,fig.height=10,include=T}
set.seed(12345)
P3_2=PCA_Plot_3_shape (t(Tumor_count_log),RNAseq_anno,"Sample_name","Drug","Timepoint")+facet_wrap(~Drug_class)
P3_2
```



```{r}
knitr::knit_exit()
```

```{r}
# GTAC DATA
# Rawcount_path="/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_RNAseq_Kangwen/all.gene_counts.tsv" 
# 
# Rawcount_path="/Volumes/Hirbe_lab_Drive_5/Hirbe_SR003350_RNAseq_Kangwen/all.gene_counts.tsv"
# Raw_counts=readr::read_tsv(Rawcount_path)%>%dplyr::filter(gene_biotype=="protein_coding")
# #
# Raw_counts_2=Raw_counts%>%.[grepl("sample",colnames(.))]%>%mutate(gene_name=Raw_counts$external_gene_name)%>%group_by(gene_name)%>%summarise_all(mean)%>%as.data.frame()%>%matrix.please()%>%as.data.frame()%>%mutate_all(.,as.integer)
# 
# Raw_counts_2["ACTB",]
```


```{r}
Sample_anno=data.frame(Sample_name=colnames(Raw_counts_2))%>%mutate(Drug=ifelse(grepl("D",Sample_name),"DMSO", ifelse(grepl("M",Sample_name),"Mirda","DMSO") ))%>%mutate(Timepoint=ifelse(grepl("0h",Sample_name),"0hour", ifelse(grepl("8h",Sample_name),"8hour","24hour") ))
```


```{r}
PCA_Plot_3<- function (data,Annotation,VAR,Color,Shape) {
  # logcountdata row:genes,column: samples
  pca <- prcomp(data) 
  pca_out<-as.data.frame(pca$x)
  df_out<- pca_out %>%tibble::rownames_to_column(var=VAR)  %>% left_join(., Annotation) 
  #df_out<- merge (pca_out,Annotation,by.x=0,by.y=0) 
  
  # label_color<- factor(df_out[,group])
  ggplot(df_out,aes_string(x="PC1",y="PC2")) +geom_point(aes_string(colour = Color,shape=Shape),size=3) 
}

```

```{r}
deseq2Data=DESeq2:: DESeqDataSetFromMatrix(countData =Raw_counts_2, colData = Sample_anno,design = ~Timepoint)
deseq2VST <- DESeq2::vst(deseq2Data)%>%SummarizedExperiment::assay()
```

```{r}
# Sample_anno_2=Sample_anno%>%dplyr::filter(Drug!="Control")%>%mutate(Group=paste0(Drug,"_",Timepoint))
# Raw_counts_rm_control=Raw_counts_2%>%.[,Sample_anno_2$Sample_name]
Raw_counts_add_control=Raw_counts_2%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)
Sample_anno_2=data.frame(Sample_name=colnames(Raw_counts_add_control))%>%left_join(.,Sample_anno)%>%mutate(Drug=ifelse(is.na(Drug),"Mirda",Drug))%>%mutate(Timepoint=ifelse(is.na(Timepoint),"0hour",Timepoint))%>%mutate(Group=paste0(Drug,"_",Timepoint))
rownames(Sample_anno_2)=Sample_anno_2$Sample_name
Sample_anno_2$Timepoint=factor(Sample_anno_2$Timepoint,levels = c("0hour","8hour","24hour"))
library(SummarizedExperiment)
library(DESeq2)
keep.exprs <- selectGenes_BY_group(Raw_counts_add_control, min.count=10, N=0.90,Sample_anno_2)
keep.exprs_2=keep.exprs%>%mutate(Sum=apply(.,1,sum))%>%dplyr::filter(Sum>2)
myFilt <- Raw_counts_add_control[rownames(keep.exprs_2),]

dim(myFilt)


```

# Time course analyses with LRT
**The LRT test can be especially helpful when performing time course analyses. We can use the LRT to explore whether there are any significant differences in treatment effect between any of the timepoints.**

**For have an experiment looking at the effect of treatment over time on mice of two different genotypes. We could use a design formula for our ‘full model’ that would include the major sources of variation in our data: genotype, treatment, time, and our main condition of interest, which is the difference in the effect of treatment over time (treatment:time).**
```{r}
#https://rstudio-pubs-static.s3.amazonaws.com/329027_593046fb6d7a427da6b2c538caf601e1.html
deseq2Data_add_control=DESeq2:: DESeqDataSetFromMatrix(countData =myFilt , colData = Sample_anno_2,design = ~Drug+Timepoint+Drug:Timepoint)
deseq2Data_add_control$Timepoint=relevel(deseq2Data_add_control$Timepoint,"0hour")
dds_LRT= DESeq2::DESeq(deseq2Data_add_control,test="LRT", reduced =~Drug+Timepoint)
DESeq2::resultsNames(dds_LRT)
```

```{r}
Mirda_8h_vs_DMSO=results(dds_LRT,name="DrugMirda.Timepoint8hour")%>%as.data.frame()
Mirda_8h_vs_DMSO_sig=Mirda_8h_vs_DMSO%>%dplyr::filter(padj<0.05)
Mirda_24h_vs_DMSO=results(dds_LRT,name="DrugMirda.Timepoint24hour")%>%as.data.frame()
Mirda_24h_vs_DMSO_sig=Mirda_24h_vs_DMSO%>%dplyr::filter(padj<0.05)

```


```{r}
# # Extract results
res_LRT <- results(dds_LRT)
res_LRT_q005=res_LRT %>%as.data.frame()%>%dplyr::filter(padj<0.05)
rld <- rlogTransformation(dds_LRT)

rld_mat <- assay(rld)
cluster_rlog=rld_mat[rownames(res_LRT_q005),]

TPM_counts_clean=TPM_2[rownames(cluster_rlog),]%>%mutate(sample_rep.0h1=sample.0h1,sample_rep.0h2=sample.0h2,sample_rep.0h3=sample.0h3)

logTPM=log(TPM_counts_clean+1)
```



# Volcano plot for DEPs

```{r,fig.width=18,fig.height=18,include=T}

DEG_data_1= Mirda_8h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(padj<0.05)

DEG_data_2= Mirda_24h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(padj<0.05)

DEG_list=list("Mirda_vs_DMSO_8hr"=DEG_data_1$gene_symbol,"Mirda_vs_DMSO_24hr"=DEG_data_2$gene_symbol)

Out=Volcano_plot_rnaseq_simple (DEG_data_1,DEG_list,"Mirda_vs_DMSO_8hr")
Out2=Volcano_plot_rnaseq_simple(DEG_data_2,DEG_list,"Mirda_vs_DMSO_24hr") 
cowplot::plot_grid(Out,Out2,ncol=2)
```


```{r,fig.width=12,fig.height=18,include=F}
# Inspired by https://lpantano.github.io/DEGreport/reference/degPatterns.html

data=logTPM
clusters <- DEGreport::degPatterns(data, metadata = Sample_anno_2, time="Timepoint", col="Drug",nClusters =6)
```



```{r}
Cluster_data=clusters[["normalized"]]%>%mutate(cluster=paste0("Cluster",cluster))
Cluster_Genelist=list()
for (i in unique(Cluster_data$cluster)){
 Cluster_Genelist[[i]]=Cluster_data%>%dplyr::filter(cluster==i)%>%.$genes
  
}


sig_plot_list=list()

for (i in names(Cluster_Genelist)) {
  Gene=Cluster_data%>%dplyr::filter(cluster==i)%>%.$genes
  DEG_data_1= Mirda_8h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(gene_symbol%in%Gene)
  DEG_data_2= Mirda_24h_vs_DMSO%>%mutate(gene_symbol=rownames(.))%>%dplyr::filter(gene_symbol%in%Gene)

  Out=Volcano_plot_rnaseq_add_label_update(DEG_data_1,Cluster_Genelist,i)
  Out2=Volcano_plot_rnaseq_add_label_update(DEG_data_2,Cluster_Genelist,i) 
  sig_plot_list[[i]]=cowplot::plot_grid(Out,Out2,rnow=1,labels=c("Mirda_vs_DMSO_8h","Mirda_vs_DMSO_24h"),vjust=3.0,hjust=-1,label_size=10)
}
```

```{r}
Data=logTPM%>%t()%>%as.data.frame()%>%tibble::rownames_to_column(var="Sample_name")%>%tidyr::gather(,key="genes",value="Log2TPM",-Sample_name)

cluster_norm=clusters[["normalized"]]%>%dplyr::filter(merge!="Mirda0hour")%>%left_join(.,Data) 
#%>%dplyr::filter(merge!="Mirda0hour")%>%left_join(.,Data)%>%mutate(Cluster=paste0("Cluster",cluster))
# Use the data yourself for custom figures
# P1=ggplot(cluster_norm,
#         aes(Timepoint, Log2TPM, color = Drug,shape=genes)) +
#    geom_point(position = position_jitterdodge(dodge.width = 0.9)) +
#    # change the method to make it smoother
#      geom_smooth(aes(group=Timepoint), method = "lm")
cluster_plot=list()

for (i in unique(cluster_norm$cluster)) {
  data=cluster_norm%>%dplyr::filter(cluster==i)
  cluster_plot[[i]]=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

}

cluster_plot[[i]]
```


# Cluster 1,3 and 6 (Gene expressions were upregulated at 8 and/or 24 hour in DMSO group)

## Cluster1

```{r,fig.width=18,fig.height=15}
i=1
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")
```

```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

## cluster 3

```{r,fig.width=18,fig.height=15}
i=3
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```



## cluster6

```{r,fig.width=18,fig.height=15}
i=6
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

# Cluster 4 and 5 (Gene expressions were downregulated at 8 and/or 24 hour in DMSO group)

## Cluster 4

```{r,fig.width=18,fig.height=15}
i=4
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

## Cluster5

```{r,fig.width=18,fig.height=15}
i=5
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=20,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob(out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```


# Cluster 2  (Gene expressions were upregulated at 8 and 24 hour in DMSO and Mirda groups )

```{r,fig.width=18,fig.height=40}
i=2
data=cluster_norm%>%dplyr::filter(cluster==i)
out1=DEGreport::degPlotCluster(data, "Timepoint", "Drug")

```


```{r,fig.width=18,fig.height=48,include=T}
cowplot::plot_grid(gridExtra::arrangeGrob( out1,nullGrob(), ncol=1),sig_plot_list[[i]],ncol=2,rel_widths = c(0.5,1))
```

```{r}
knitr::knit_exit()
```



# ETV5 AND SPRY4 expression (TPM)

```{r,include=F}
Genes=c("ETV5","SPRY4")
Data=logTPM[Genes,]%>%t()%>%as.data.frame()%>%tibble::rownames_to_column(var="Sample_name")%>%tidyr::gather(,key="genes",value="Log2TPM",-Sample_name)

cluster_norm=clusters[["normalized"]]%>%dplyr::filter(grepl("ETV5|SPRY4",genes))%>%dplyr::filter(merge!="Mirda0hour")%>%left_join(.,Data)
# Use the data yourself for custom figures
P1=ggplot(cluster_norm,
        aes(Timepoint, Log2TPM, color = Drug,shape=genes)) +
   geom_point(position = position_jitterdodge(dodge.width = 0.9)) +
   # change the method to make it smoother
     geom_smooth(aes(group=Timepoint), method = "lm")

P2=ggplot(cluster_norm,
        aes(Timepoint, value, color = Drug,shape=genes)) +
   geom_point(position = position_jitterdodge(dodge.width = 0.9)) +ylab("normalized_log2TPM")+
   # change the method to make it smoother
     geom_smooth(aes(group=Timepoint), method = "lm")
```


```{r,fig.width=6,fig.height=3,include=T}
cowplot::plot_grid(P1,P2,nrow=1)
```



# Pathway analysis for each cluster

```{r}
sig_path_list=list()

for (i in 1:6) {
Gene=clusters[["normalized"]]%>%dplyr::filter(cluster==i)%>%.$genes
Cluster_1_sig_path=Gene_pathway_sig_enrich(Gene)
Cluster_1_sig_path_pickup=Cluster_1_sig_path%>%dplyr::filter(Count>2&pvalue<0.05)
sig_path_list[[i]]=Cluster_1_sig_path_pickup%>%mutate(Cluster=i)
}



```


```{r}
sig_path_list[[6]]%>%dplyr::filter(grepl("WP_|HALLMARK|KEGG",ID))
```


```{r}
knitr::knit_exit()
```




```{r}
res=DESeq2::results(Deg_add_control,name="Timepoint_8hour_vs_0hour", alpha = 0.05)%>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
```



# For mirda 24hour vs 0hour
```{r}
# 
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint24hour","Timepoint_24hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))

res=DESeq2::results(Deg_add_control,list(c("Timepoint_24hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))

```



```{r}
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```


# For mirda 8hour vs 0hour
```{r}
# 
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour","Timepoint_8hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))

plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```



# 8hr Mirda vs DMSO

```{r}
# 
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour","Drug_Mirda_vs_DMSO"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]

res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]

```


```{r}

Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```

# 24hr Mirda vs DMSO
```{r}
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint24hour","Drug_Mirda_vs_DMSO"  )), alpha = 0.05) %>%as.data.frame()

res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
genes=c("ETV5")
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```

# for Mirda 8hr vs 0hr
```{r}
res=DESeq2::results(Deg_add_control,list(c("DrugMirda.Timepoint8hour","Timepoint_8hour_vs_0hour"  )), alpha = 0.05) %>%as.data.frame()

res_q005=res%>%dplyr::filter(padj<0.05)%>%arrange(desc(log2FoldChange))
res_q005[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
genes=c("ETV5")
plist=list()
library(ggplot2)
for (i in 1:5){
  genes=rownames (res_q005)[i]
  TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 
 plist[[i]]= ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)+xlab()
  
}

cowplot::plot_grid(plotlist = plist,ncol=3)
```


```{r}
res=DESeq2::results(Deg_add_control,list(c("Timepoint24hour.DrugMirda","Drug_Mirda_vs_DMSO"  )), alpha = 0.05) %>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
res_q005[c("ETV5", "SPRY4"),]
Drug_Mirda_vs_DMSO_res_na=res_q005%>%dplyr::filter(is.na(padj))
genes=c("ETV5")
TPM_data=TPM_2%>%t()%>%as.data.frame()%>%.[genes]%>%setNames("symbol")%>%mutate(Sample_name=rownames(.)) 

library(ggplot2)
ggplot(TPM_data,aes(x=Sample_name,y=symbol))+geom_col()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab(genes)

```


```{r}
# DMESO
Drug_Mirda_vs_DMSO_res=DESeq2::results(Deg_rm_control,name= "Timepoint8hour.DrugMirda" , alpha = 0.05) %>%as.data.frame()
Drug_Mirda_vs_DMSO_res_q005=Drug_Mirda_vs_DMSO_res%>%dplyr::filter(padj<0.05)

Drug_Mirda_vs_DMSO_res_na=Drug_Mirda_vs_DMSO_res%>%dplyr::filter(is.na(padj))

Drug_Mirda_vs_DMSO_res[c("ETV5", "SPRY4"),]

Raw_counts_rm_control[c("ETV5", "SPRY4"),]

```


```{r}
dds=Deg_add_control
res <- results(dds, name="Timepoint8hour.DrugMirda")%>%as.data.frame()
res_q005=res%>%dplyr::filter(padj<0.05)
ix = which.min(res$padj) # most significant
ix = which.min(res$log2FoldChange) # most significant

res <- res[order(res$log2FoldChange),] # sort
knitr::kable(res[1:5,-(3:4)])
```


```{r}
ix=rownames( res_q005)[1]
barplot(TPM_2[ix,],las=2, main=rownames(TPM_2)[ix]  )
```


```{r}
# 8hr
dds=Deg_rm_control
res = results(dds, contrast=c("Drug","Mirda","DMSO"))
ix = which.min(res$padj) # most significant
res <- res[order(res$padj),] # sort
knitr::kable(res[1:5,-(3:4)])
```

```{r}
dds=Deg_rm_control
res <- results(dds, list( c("condition_Trt_vs_Ctrl","genotypeMU.conditionTrt") ))
ix = which.min(res$padj) # most significant
res <- res[order(res$padj),] # sort
kable(res[1:5,-(3:4)])
```

 
```{r}
# 24hr
res = results(Deg_rm_control, contrast=c("Drug","Mirda","DMSO"))
ix = which.min(res$padj) # most significant
res <- res[order(res$padj),] # sort
knitr::kable(res[1:5,-(3:4)])
 




```{r}
DMSO_samples=Sample_anno%>%dplyr::filter(Drug%in%c("Control","DMSO"))
DMSO_count=Raw_counts_2[,DMSO_samples$Sample_name]
DMSO_anno=data.frame(Sample_name=colnames(DMSO_count))%>%left_join(.,Sample_anno)
DMSO_fill <- DESeq2:: DESeqDataSetFromMatrix(countData =DMSO_count, colData = DMSO_anno,
                                             design = ~Timepoint)
DMSO_Deg<- DESeq2::DESeq(DMSO_fill)
DESeq2::resultsNames(DMSO_Deg)
DMSO_24h_res <- DESeq2::results(DMSO_Deg,name= "Timepoint_24hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()
DMSO_8h_res <- DESeq2::results(DMSO_Deg,name= "Timepoint_8hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()


```

```{r}
Mirda_samples=Sample_anno%>%dplyr::filter(Drug%in%c("Control","Mirda"))
Mirda_count=Raw_counts_2[,Mirda_samples$Sample_name]
Mirda_anno=data.frame(Sample_name=colnames(Mirda_count))%>%left_join(.,Sample_anno)
Mirda_fill <- DESeq2:: DESeqDataSetFromMatrix(countData =Mirda_count, colData = Mirda_anno,
                                             design = ~Timepoint)
Mirda_Deg<- DESeq2::DESeq(Mirda_fill)
DESeq2::resultsNames(Mirda_Deg)
Mirda_res <- DESeq2::results(Mirda_Deg,  alpha = 0.05) %>%as.data.frame()

Mirda_24h_res <- DESeq2::results(Mirda_Deg,name= "Timepoint_24hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()
Mirda_8h_res <- DESeq2::results(Mirda_Deg,name= "Timepoint_8hour_vs_0hour" , alpha = 0.05) %>%as.data.frame()

```

```{r}

save(deseq2VST,DMSO_8h_res,Mirda_8h_res,DMSO_24h_res,Mirda_24h_res  ,Sample_anno,file=file.path(Rawdata_path,"DEseq2_out.rdata"))
```


```{r}
knitr::knit_exit()
```

 

