---
title: "Compare Pathways to CCLE"
author: "Sara Gosline"
date: "2026-04-30"
output: html_document
---



Here we take a top-down look of what combinations are highly effective and
what pathways the two drugs might share. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(synapser)
synLogin()
source("../pathwayChanges/plottingParameters.R")
source("correlation_functions.R")

bs3 <- brewer.pal(4, "Set3") 


```


# Top-ranking drug combos

We can look at the delta AUC scores for the samples for which we have gene expression
data. We see five combinations with a dAUC above 1:

```{r read daucs}
daucs <- readr::read_csv(synGet('syn69978141')$path) |>
  tidyr::separate(Combo, into=c('drug1','drug2'), sep='\\+', remove = FALSE)

##read in GSEA from treated samples
sing_gsea <- readr::read_csv("../pathwayChanges/GSEA_Hallmark_individual/GSEA_results_Hallmark.csv")|>
  mutate(Feature_set = tolower(gsub("_",' ',Feature_set)))

##get aucs fro gene xpression samples

daucs |> subset(PDX%in%sing_gsea$MPNST) |>
  group_by(drug1,drug2) |>
  summarize(md = mean(dAUC,na.rm=T)) |>
  subset(md > 1) |>
  arrange(desc(md))

```

##Load in diffex pathways

We can now ask if pathways up-regulated by one of the combinations of interest might be of interest to other
combinations in other cell lines. 

```{r pathways}

drugs <- c('Mirdametinib','TNO155','Vorinostat','Capmatinib','Trabectedin','Doxorubicin')
hall_preds <- readr::read_csv('../pathwayChanges/GSEA_Hallmark/GSEA_results_Hallmark.csv')|>
  mutate(Feature_set = tolower(gsub("_",' ',Feature_set))) |>
  mutate(Feature_set = gsub("hallmark ","",Feature_set))

sig_paths <- subset(hall_preds, DrugTreatment%in%drugs)|> 
  subset(NES > 0) |>
  subset(FDR_q_value < 0.25) |>
  subset(p_value<0.05)


ggplot(sig_paths,aes(x=Feature_set,y=NES,fill=tolower(DrugTreatment))) +
  geom_bar(position='dodge',stat='identity') +
  scale_fill_manual(values=drugCol) +
  coord_flip()
```
These drugs cause a lot of pathways to be up-regulated. 

## Score CCLE Pathways

CCLE has a large set of matched gene expression and drug response data. While many of 
these drugs do not overlap the drugs we used, we can determine if there our approach
is valid outside of MPNST for identifying novel drugs in combination.

```{r read in ccle data}

#TODO: put these files on synapse
r1 <- readRDS(synGet('syn74935645')$path)
r2 <- readRDS(synGet('syn74935650')$path)

cmat <- rbind(r1,r2) |>
  as.data.frame() |>
  tibble::remove_rownames() |>
  tibble::column_to_rownames('CCLE_ID') |>
  t()

cmat <- cmat[-which(duplicated(rownames(cmat))),]

gmat <- get_gsea_scores(cmat)
rownames(gmat) <- sapply(rownames(gmat), function(x) tolower(gsub("HALLMARK ","",gsub("_"," ",x))))

##get MOA for later
##is it worth it to incorporate drug class? 
gmt.drug <- leapR::read_gene_sets(synGet('syn74955107')$path)

gmt.drug$matrix <- t(gmt.drug$matrix)

moa.df <- do.call(rbind,
                  lapply(c(1:nrow(gmt.drug$matrix)), function(i){
  dl <- setdiff(gmt.drug$matrix[i,],c('',NULL,NA))
  data.frame(moa = rep(gmt.drug$names[i]),drug=dl)
}))

```
## Correlate CCLE scores with drug response

Now we can do the same thing we did in the MPNST and look at
correlations with various pathways and drugs

```{r drug scores}

prism <- read.csv(file = synGet('syn74955119')$path, check.names=FALSE) ##these are the names that map to moa

p_exp <- prism[,2:ncol(prism)] |>
  tidyr::pivot_longer(-CCLE_ID,names_to='drug',values_to='auc')|>
  dplyr::rename(improve_sample_id = 'CCLE_ID')|>
  mutate(treat_time = 120, expr_time = '0hr')

cor_res <- do.call(rbind,lapply(unique(p_exp$drug), function(drug){
      compute_cor(p_exp,gmat,drug,'0hr',120)
                        })) #e  

cor_res$drug <- sapply(cor_res$drug,function(x) gsub('-','.',x))

##and calculate significance
print('computing gene/auc significance')
cor_sig <- do.call(rbind,lapply(unique(p_exp$drug), function(drug){
      compute_sig(p_exp,gmat,drug,'0hr',120)
    })) #end drug iter

cor_sig$drug <- sapply(cor_sig$drug,function(x) gsub('-','.',x))


merged_res <- cor_res |> left_join(cor_sig) |>
  left_join(moa.df)
```

How do we visualize this many correlations? Let's do it one pathway at 
a time.

```{r visualize}

plot_pathway_correlation <- function(pathname,drugclass= NA, ft = ''){

  cr <- merged_res |> 
    subset(gene == pathname) |>
    mutate(Signif = ifelse(qval < 0.01,'**',ifelse(qval<0.05,'*',"")))
  
  if(!is.na(drugclass))
    cr <- subset(cr,moa==drugclass)
  if(!any(cr$qval<0.01))
    return('No significant drugs in this class/pathway')
  mv <- cr |> group_by(drug)|>
    summarize(mv = mean(corval,na.rm=T))|>
    arrange(mv)
  
  cr$drug = factor(cr$drug, levels=unique(mv$drug))
  if(is.na(drugclass)){
    cr |> ggplot(aes(fill=moa,x=drug, y=corval,group=moa)) + 
     geom_bar(position='dodge',stat='identity') + 
     #ggrepel::geom_text_repel(aes(label = Signif),nudge_y=-0.01)+
    #facet_grid(ttime~.) +
    ggtitle(paste("Drug response correlation with",pathname))    +
    theme_bw() +    
   # scale_fill_manual(values=moaCol)+
    coord_flip()+
    ylab('AUC correlation') + xlab('Drug')+
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
  }else{
        cr |> ggplot(aes(fill=drug,x=drug, y=corval)) + 
     geom_bar(position='dodge',stat='identity') + 
     ggrepel::geom_text_repel(aes(label = Signif),nudge_y=-0.01)+
    #facet_grid(ttime~.) +
    ggtitle(paste("Drug response correlation with",gsub('HALLMARK_','',pathname)))    +
    theme_bw() +    
   # scale_fill_manual(values=moaCol)+
    coord_flip()+
    ylab('AUC correlation') + xlab('Drug')+
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
  }
 
  ggsave(paste0(ft,'CCLE_drugCor.pdf'),width=6)
}



```

Now can we look to see if other vorinostat up-regulated pathways are anti-correlated
with MEK sensitivity?
```{r vorin mek check}


vpaths <- sig_paths[sig_paths$DrugTreatment=='Vorinostat','Feature_set'] |> unique()

res<-lapply(vpaths$Feature_set, function(x) {
  plot_pathway_correlation(x,'MEK inhibitor',paste0('fig6a_',x,'_correlation_MEKi'))
})
#plot_pathway_correlation("HALLMARK_KRAS_SIGNALING_UP",'kras_only')
```
## Identify MOA that are enriched for each pathway

Now we can use the MOA list to see if there are specific drug classes that are
more sensitive to specific pathways

```{r leapR enrichment}

library(leapR)

cormat <- cor_res[,c('gene','drug','corval')] |> 
  pivot_wider(names_from='gene', values_from='corval') |>
  tibble::column_to_rownames('drug') |>
  as.matrix()


##now we can run a KS test on each pathway to see
##what drug mechanisms are more negatively correlated with each
##class
eres <- do.call(rbind,lapply(colnames(cormat),function(p){
    leapR(gmt.drug,'enrichment_in_order',
       eset=SummarizedExperiment::SummarizedExperiment(assays=list(cors=cormat)), 
       assay_name='cors',
       primary_columns=p) |>
    tibble::rownames_to_column('drugClass')|>
    mutate(pathway=p)}))


```



## now rank and check for overlap

What pathways are up-regulated in our own MPNST?

```{r}

upreg_paths <- eres |>
  subset(pathway %in% sig_paths$Feature_set) |>
  subset(BH_pvalue < 0.01) |> 
  subset(ingroup_mean < -1.0)  |>
  mutate(pathname = stringr::str_replace(pathway,"HALLMARK_",""))

##of course the ingroup mean is a z-score, so we need to re-compoute? 


ggplot(upreg_paths,aes(x=reorder(pathname,ingroup_mean),y=ingroup_mean,fill=drugClass)) +
  geom_bar(position='dodge',stat='identity') +
  xlab('Pathways up-regulated') + ylab("Mean correlation between AUC and pathway score in drug class")+
  coord_flip()

#ggsave('fig_6c_drugClassCorrelation.pdf',width=12,height=8)
```

## Figure Plots

### Fig 6b: MPNST Up-reg

First we can plot the CCLE correlation of pathways up-regulated in MPNST. 

```{r mpnst upreg}

# Add more colors to this palette :

mekis <- upreg_paths[upreg_paths$drugClass=='MEK inhibitor',]
coul <- colorRampPalette(bs3)(length(unique(mekis$pathname)))

sig_paths$Pathway <- sig_paths$Feature_set

names(coul)<- unique(mekis$pathname)

subset(sig_paths,Feature_set%in%mekis$pathway) |>
  ggplot(aes(fill=Pathway, y = NES,x = DrugTreatment)) + 
  #scale_fill_manual(values=drugCol) + 
  geom_bar(stat='identity', position='dodge') + 
  xlab('Drug') + ylab("Expression change in MPNST") + coord_flip() +
  ggtitle("MEKi correlated pathways") +
  facet_grid(~Time)+
  scale_fill_manual(values=coul)

ggsave('fig6b_mekiCorPathsInMPNST.pdf')


```

### Fig 6A: pathways associated with MEKi
More broadly we can check to see what pathways are anticorrelated with MEKi

```{r plot meki drug paths}


md <- subset(merged_res, moa=='MEK inhibitor')|>
  subset(corval <0)|>
  subset(qval<0.05)

md$Pathway = gsub('HALLMARK_','',md$gene)
#coul <- colorRampPalette(bs3)(length(unique(md$gene)))

ggplot(md,aes(x=reorder(drug, corval), y = corval, fill = Pathway)) +
  geom_bar(position='dodge',stat='identity') +
  coord_flip() +
  scale_fill_manual(values=coul) +
  xlab('Drug')+ylab('Correlation')

ggsave('fig6a_meki_pathcor.pdf')

```

```{r validation perhaps}



#tikis <- upreg_paths[upreg_paths$drugClass=='topoisomerase inhibitor',]

#subset(sig_paths,Feature_set%in%tikis$pathway) |>
#  ggplot(aes(x=Feature_set, y = NES, fill = tolower(DrugTreatment))) + 
#  scale_fill_manual(values=drugCol) + geom_bar(stat='identity', position='dodge') + coord_flip() +
#  ggtitle("TOPOIi correlated pathways") +
#  facet_grid(~Time)

plot_pathway_drug_class <- function(path='HALLMARK_KRAS_SIGNALING_UP',
                                 drugclass = 'MEK inhibitor'){
  pscore <- data.frame(pathScore = gmat[path,], 
                       improve_sample_id = colnames(gmat))
                       
  drugs <- setdiff(gmt.drug$matrix[which(gmt.drug$names == drugclass),],c(NA,NULL,''))
  
  dvals <- p_exp |>
    subset(drug %in% drugs) |>
    left_join(pscore) 
  
  ggplot(dvals,aes(x=pathScore, y = auc, col=drug)) +
    geom_point() + 
    geom_smooth(method = "lm", se = TRUE) +
    ggtitle(paste(drugclass,'drugs and',path,'scores'))
}

plot_pathway_drug_class()

#ggsave('kras_up_meki.pdf',width=6)

plot_pathway_drug_class('kras signaling dn','EGFR inhibitor')
#ggsave('kras_down_egfr.pdf',width=6)

plot_pathway_correlation("kras signaling dn",'EGFR inhibitor','kras_only_egfr')
```

### Fig 6c: MOA class scores by drug
We can focus on pathways that are-regulated at both time point AND anti-correlated with

```{r}

combined <- eres |>
  dplyr::mutate(Pathway = gsub("HALLMARK_",'',pathway)) |>
  right_join(sig_paths) |>
  subset(BH_pvalue < (0.01)) |>
  subset(ingroup_mean < (-1))

ggplot(combined, aes(x=sort_by(Pathway, ingroup_mean), y =ingroup_mean, fill = drugClass)) +
  geom_bar(position='dodge',stat='identity') +
  coord_flip() +
  facet_grid(~DrugTreatment) +
  scale_fill_manual(values =colorRampPalette(bs3)(length(unique(combined$drugClass))))+
  xlab('Pathway') + ylab('Mean correlation with AUC across class') + 
  theme_bw() + 
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))

ggsave('fig_6c_enrichmentStats.pdf',width=12)
``` 

