---
title: "Data Integration"
author: "Sara gosline"
date: "2024-12-18"
output: html_document
---

This file collects existing data - to date it is transcritpion factor and kinase enrichment stats from various drugs and time points, and experiments with ways to incorporate them across different studies.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(synapser)
library(dplyr)
synLogin()

tfs<-readr::read_csv(synapser::synGet('syn64420968')$path)|>
  mutate(sample='MN.2')

kins<-readr::read_csv(synapser::synGet('syn64322657')$path)
library(GGally)

#ggtable(tips, "smoker", c("day", "time", "sex"))




```

## Data summary
First we need to collect where we're at with all the different types of analyses and time points


```{r data}

##figure out how many changes we have in each

kinsum<-subset(kins,p.value<0.05)|>
  dplyr::select(drug,sample,Gene='Kinase.Gene',score='z.score',time)|>
  mutate(geneType='Kinase',time=as.factor(time))|>
  mutate(direction=ifelse(score>0,'up','down'))

tfsum<-subset(tfs,padj<0.05)|>
  dplyr::select(drug,sample,time,Gene='source',score)|>
  mutate(geneType='TF',time=as.factor(time))|>
  mutate(direction=ifelse(score>0,'up','down'))

fullsum<-rbind(kinsum,tfsum)

ggtable(kinsum,'drug',c('sample','time','direction'))+ggtitle('Kinases active by drug and time')

ggtable(tfsum,'drug',c('sample','time','direction'))+ggtitle("TFs active by drug and time")

```

## trying more canonical plots
Bar charts can also show changes by direction

```{r pressure, echo=FALSE}

tcounts<-fullsum|>group_by(drug,sample,time,geneType,direction)|>summarize(count=n())
tcounts<-tcounts|>mutate(regulators=ifelse(direction=='up',count,-1*count))

ggplot(tcounts,aes(x=drug,y=regulators,fill=geneType))+geom_bar(stat='identity',position='dodge')+facet_grid(sample~time)+coord_flip()

ggsave('totalRegulatorCounts.png')

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
