---
title: "Gene Expression Summary Stats"
author: "Sara gosline"
date: "2024-12-11"
output: html_document
editor_options: 
  chunk_output_type: console
---


The goal of this document is to pull all RNASeq files from synapse in a metadata-aware fashion and do the basic analysis needed for figure 1 and supplemental figure 1.

Currently we have:

MN-2
| batch | date | treatment |
| 1 | 10/17/23 | DMSO, mirda |
| 2 | 12/19/23 | DMSO, tram, selu, TNO, ribo, palbo, RMC, vorin |
| 2 | 12/27/23 | DMSO, decitabine, olap, trab, doxo, ifo, cap |

JH-2
| batch | date | treatment |
| 3 | A | DMSO, MIR, PAL, RIB, RMC, SEL, TNO, TRAM, vorin |
| 3 | B | DMSO, DEC, IFO, IRI, OLA, TRAB |
| 3 | C | CAP, DOX | # no DMSO?

WU-225
| batch | date | Treatment
| 4 | | xxx

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synapser)
library(dplyr); library(tidyr); library(stringr); library(RColorBrewer)
library(readr);library(readxl); library(ggplot2)
synLogin()


  
```

### RNASeq data from synapse.
Created a helper function to get RNASeq data. Currently this uses the file names for metadata. The goal is to update with metadata from synapse directly. 

```{r rnaseq}
##currently this function pulls all data manually and parses the file names. 
source('downloadCorrectRNASeq.R')
```

### Collect latest data

We can now call our RNASeq helper function defined above to get data and visualize overall. Currently this pulls all data with a `quants.genes.sf` file but can be adjusted. 


```{r data, echo=FALSE,warning=FALSE,message=FALSE}
rnaseqdata<-read.csv(synapser::synGet('syn64398049')$path)

zeroes<-rnaseqdata|>
  subset(NumReads<3)|>
  group_by(ensembl_gene_id)|>
  summarize(zvals=n())

samps<-length(unique(rnaseqdata$id))
toremove<-subset(zeroes,zvals>samps/2)

reduced.data<-subset(rnaseqdata,!ensembl_gene_id%in%toremove$ensembl_gene_id)

```

### Differential expression across all MPNST

Let's compute the differential expression of each drug/time point compared with DMSO. This analysis is at the core of what we're doing - trying to determine what changes are happening at each time point. 

- Which drugs are causing the greatest changes?
- Which time points are causing the greatest changes?


```{r diffex, echo=FALSE,warning=FALSE,message=FALSE}
library("DESeq2")
counts<-reduced.data|>
  dplyr::select(ensembl_gene_id,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads, values_fn = mean)|>
  tibble::column_to_rownames('ensembl_gene_id')|>
  as.matrix()


sampmeta <- reduced.data |>
  dplyr::select(id,specimenID, individualID, Drug,Date,Timepoint)|>
  distinct()

sampmeta$Date=as.factor(sampmeta$Date)

temp.drugs <- unique(sampmeta$Drug)
temp.drugs <- setdiff(temp.drugs, c('DMSO','untreated'))
diffexCalls<-do.call(rbind,lapply(temp.drugs,function(d_name){
  cat(d_name,"\n")
  batches <- subset(sampmeta, Drug == d_name)
  ##check for each time point at the drug/date
  do.call(rbind,lapply(c('8h','24h'),function(tval){
    cat(tval,"\n")
    samps <- subset(sampmeta, Timepoint==tval)|>
      subset(Drug%in%c(d_name,"DMSO"))|>
      subset(id%in%colnames(counts))|>
      subset(Date%in%batches$Date) # SG Make sure we only look at DMSO from that date
    samps$Drug <- factor(samps$Drug, levels=c("DMSO",d_name)) # make sure contrast is drug vs. dmso per DESeq2 vignette
    cts<-counts[,intersect(samps$id,colnames(counts))]
    dds <- DESeqDataSetFromMatrix(countData = round(cts),
                                  colData = samps,
#                                  design = ~ Drug + individualID + Date) #date was NOT corrected for in this data, but restricting dates avoids this
                                  design = ~ Drug + individualID) # use MPNST as batch factor; already corrected for date factor
    
    dres<-DESeq(dds)|>
      results(contrast=c("Drug",d_name,"DMSO"))|> # double check contrast is drug vs. dmso per DESeq2 vignette
      as.data.frame()|>
      tibble::rownames_to_column('Gene')|>
      dplyr::mutate(Drug=d_name,Timepoint=tval)
    return(dres)
  })) 
})) 

diffex<-diffexCalls|>
  mutate(signif=padj<0.05)|>
  mutate(direction=ifelse(log2FoldChange>0,'up','down'))


dgenes<-diffex|>
  tidyr::separate(Gene,sep="\\.",into=c('ensembl_gene_id','Version'))

diffex_gene<-dgenes|>left_join(map)
##let's save this back to synapse for future reference
readr::write_csv(diffex_gene,file='totalRNASeqDiffExtoDate.csv')
synFile <- synapser::File('totalRNASeqDiffExtoDate.csv', parent = 'syn64004220')
synapser::synStore(synFile)
```

## Do some plots

```{r plot pooled differential expression}

sumstats<-diffex|>subset(signif)|>
  group_by(Drug,Timepoint,direction)|>
  summarize(numgenes=n())

print('now have differential expression calls')
ggplot(sumstats,aes(x=Drug,y=numgenes,fill=direction))+geom_bar(stat='identity',position='dodge')+coord_flip()+scale_fill_viridis_d()

ggplot(sumstats,aes(x=Drug,y=Timepoint,size=numgenes,col=Drug))+geom_point()+facet_grid(.~direction)+
  scale_size(range=c(1,15))+scale_color_discrete()+coord_flip()

ggsave('diffExresults.png',height=8)

##which drugs have more changes at 8 hrs? which at 24?
tp<-sumstats|>tidyr::pivot_wider(names_from='Timepoint',values_from='numgenes')|>
  tidyr::replace_na(list(`24h`=0.0))
tp$maxEffect=ifelse(tp$`24h`>tp$`8h`,'24hrs','8hrs')
tp|>ungroup()|>dplyr::select(Drug,direction,maxEffect)|>
  tidyr::pivot_wider(names_from='direction',values_from='maxEffect')|>
  arrange(Drug)

print(sumstats)
 
sumstats<- sumstats|>arrange(desc(numgenes))
levs=sumstats[which(sumstats$direction=='up'&sumstats$Timepoint=='8h'),'Drug']
sumstats$Drug=factor(sumstats$Drug,levels=levs$Drug)
 sumstats|>
   mutate(genedir=ifelse(direction=='down',-1*numgenes,numgenes))|>
 ggplot(aes(x=direction,y=numgenes,fill=Drug))+geom_bar(stat='identity',position='dodge')+
   #scale_fill_manual(values=cols)+
   coord_flip()+facet_grid(~Timepoint)
 
 ggsave('diffExresults.pdf')
 
 ###lets look at correlation across drugs
 dmat<-diffex|>dplyr::select(Gene,log2FoldChange,Drug,Timepoint)|>
   tidyr::unite(Drug,Timepoint,col='sample',sep='_')|>
   tidyr::pivot_wider(names_from=sample,values_from=log2FoldChange)|>
   tibble::column_to_rownames('Gene')
 
 cmat<-cor(dmat,use='pairwise.complete.obs',method='spearman')
 meta<-diffex|>dplyr::select(Drug,Timepoint)|>
   distinct()|>
   tidyr::unite(Drug,Timepoint,col='sample',sep='_',remove=FALSE)|>
   tibble::column_to_rownames('sample')

drugscounts<-subset(dgenes,signif)|>group_by(ensembl_gene_id,Timepoint,direction)|>
  summarize(ndrugs=n())|>arrange(desc(ndrugs))|>
  left_join(map)|>
  subset(hgnc_symbol!='')

##most drugs just show up in one or two
ggplot(drugscounts,aes(x=ndrugs,fill=direction))+geom_bar(position='dodge')+ggtitle('Number of drugs with shared differentially expressed genes')

```


