---
title: "Gene Expression Summary Stats"
author: "Sara gosline"
date: "2024-12-11"
output: html_document
editor_options: 
  chunk_output_type: console
---


The goal of this document is to pull all RNASeq files from synapse in a metadata-aware fashion and do the basic analysis needed for figure 1 and supplemental figure 1.

Currently we have:

MN-2
| batch | date | treatment |
| 1 | 10/17/23 | DMSO, mirda |
| 2 | 12/19/23 | DMSO, tram, selu, TNO, ribo, palbo, RMC, vorin |
| 2 | 12/27/23 | DMSO, decitabine, olap, trab, doxo, ifo, cap |

JH-2
| batch | date | treatment |
| 3 | A | DMSO, MIR, PAL, RIB, RMC, SEL, TNO, TRAM, vorin |
| 3 | B | DMSO, DEC, IFO, IRI, OLA, TRAB |
| 3 | C | CAP, DOX | # no DMSO?

WU-225
| batch | date | Treatment
| 4 | | xxx
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synapser)
library(dplyr); library(tidyr); library(stringr); library(RColorBrewer)
library(readr);library(readxl)
library(ggplot2)
synLogin()
library("DESeq2")
library('limma')


```

### RNASeq data from synapse.
Created a helper function to get RNASeq data. Currently this uses the file names for metadata. The goal is to update with metadata from synapse directly. 

```{r rnaseq}
##currently this function pulls all data manually and parses the file names. 
source('downloadCorrectRNASeq.R')
```

### Collect latest data

We can now call our RNASeq helper function defined above to get data and visualize overall. Currently this pulls all data with a `quants.genes.sf` file but can be adjusted. 


```{r collect data, echo=FALSE,warning=FALSE,message=FALSE}

  rnaseqdata <- rnaSeqTab()
  rnaseqgenes <- addGeneNames(rnaseqdata)

```


Here we do some pre-processing by removing lowly abundant transcripts 

Open questions:
- are there any outliers to remove?
- are there any sample swaps here?

```{r clean up data}
library(ggplot2)
zeroes<-rnaseqdata|>
  subset(NumReads<3)|>
  group_by(Name)|>
  summarize(zvals=n())

samps<-length(unique(rnaseqdata$id))
toremove<-subset(zeroes,zvals>samps/2)

reduced.data<-subset(rnaseqdata,!Name%in%toremove$Name)

##first do some pretty PCA
tpm<-reduced.data|>
  dplyr::select(Name,id,TPM)|>
  tidyr::pivot_wider(names_from=id,values_from=TPM)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

numReads<-reduced.data|>
  dplyr::select(Name,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()


 
```

### Batch correction

Clearly there is a batch effect so it will be good to correct for this to enable facile comparisons across datasets. 
This will upload two files to synapse: one with the full batch corrected data,
another with the values collapse by gene name

```{r batch correct, echo=TRUE, warning=FALSE, message=FALSE}
#library(sva)
##now to remove the batches for date while still correcting for drug and time


counts<-reduced.data|>
  dplyr::select(Name,id,NumReads) |>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name') |>
  as.matrix()

sampmeta <- reduced.data |>
  dplyr::select(id,specimenID, individualID, Drug,Date,Timepoint)|>
  distinct()

sampmeta$Date=as.factor(sampmeta$Date)
dds<-DESeqDataSetFromMatrix(countData = round(counts),
                              colData = sampmeta,
                              design = ~ Drug+Date+Timepoint )



dds<-DESeq(dds)

#lets look at both rlog and vsd as transformations
vsd <- vst(dds, blind=FALSE)

mat <- assay(vsd)
mm <- model.matrix(~Timepoint+Drug, colData(vsd))
mat <- limma::removeBatchEffect(mat, batch=vsd$Date, design=mm)
assay(vsd) <- mat



##cortab
md=colData(vsd)[,c('Drug','Timepoint','Date')]
corvsd=cor(assay(vsd),use='pairwise.complete.obs')
pheatmap::pheatmap(corvsd,annotation_row=md,annotation_col=md)

##now add the VSD counts to the table
vtab<-assay(vsd)|>as.data.frame()|>
  tibble::rownames_to_column('Name')|>
  tidyr:::pivot_longer(starts_with('syn'),names_to='id',values_to='VSD_batchCorrected')

rnaseqdata2<-rnaseqdata|>
  left_join(vtab)|>
  tidyr::separate(Name, into=c('ensembl_gene_id','ver'),sep='\\.')


rnaseqgenes<-rnaseqdata2|>
  left_join(map)|>
  subset(hgnc_symbol!="")|>
  dplyr::group_by(hgnc_symbol,Date,id,Drug,Timepoint)|>
  summarize(TPM=sum(TPM),NumReads=sum(NumReads),VSD_batchCorrected=sum(VSD_batchCorrected))

###now we add in the corrected VSD values as well
readr::write_csv(rnaseqdata2,file='rnaseqcounts.csv')
readr::write_csv(rnaseqgenes,file='rnaseqcounts_hgnc.csv')

synFile <- synapser::File('rnaseqcounts.csv', parent = 'syn64004220')
synapser::synStore(synFile)
synFile <- synapser::File('rnaseqcounts_hgnc.csv', parent = 'syn64004220')
synapser::synStore(synFile)

```

### Differential expression

Let's compute the differential expression of each drug/time point compared with DMSO. 
This analysis is at the core of what we're doing - trying to determine what 
changes are happening at each time point. 

```{r diffex, echo=FALSE,warning=FALSE,message=FALSE}
library("DESeq2")
counts<-reduced.data|>
  dplyr::select(Name,id,NumReads)|>
  tidyr::pivot_wider(names_from=id,values_from=NumReads)|>
  tibble::column_to_rownames('Name')|>
  as.matrix()

diffexCalls<-do.call(rbind,lapply(as.character(unique(sampmeta$Date)),function(dval){
  cat("Running diffexp for", dval,"\n")
  if (dval == "C") { # use DMSO from group B for group C
    datemeta <- na.omit(sampmeta[sampmeta$Date == dval | 
                           (sampmeta$Drug == "DMSO" & sampmeta$Date == "B"),])
  } else {
   datemeta<-subset(sampmeta,Date==dval) 
  }
  datemeta <- subset(datemeta,Drug!='untreated')
  mpnst <- unique(datemeta$individualID) # no need to filter for MPNST because only 1 MPNST per date
  if ("DMSO" %in% datemeta$Drug) {
    temp.drugs <- unique(datemeta$Drug)
    temp.drugs <- temp.drugs[temp.drugs != "DMSO"]

    ##check for each drug at the date
    do.call(rbind,lapply(temp.drugs,function(d_name){
      cat(d_name,"\n")
      ##check for each time point at the drug/date
      do.call(rbind,lapply(c('8h','24h'),function(tval){
        cat(tval,"\n")
        samps <- subset(datemeta, Timepoint==tval)|>
          subset(Drug%in%c(d_name,"DMSO"))|>
          subset(id%in%colnames(counts))
        samps$Drug <- factor(samps$Drug, levels=c("DMSO",d_name)) # make sure contrast is drug vs. dmso per DESeq2 vignette
        cts<-counts[,intersect(samps$id,colnames(counts))]
        dds <- DESeqDataSetFromMatrix(countData = round(cts),
                                      colData = samps,
                                      design = ~ Drug)
        
        dres<-DESeq(dds)|>
          results(contrast=c("Drug",d_name,"DMSO"))|> # double checks contrast is drug vs. dmso per DESeq2 vignette
          as.data.frame()|>
          tibble::rownames_to_column('Gene')|>
          dplyr::mutate(Drug=d_name,Timepoint=tval,Date=dval, individualID=mpnst)
        return(dres)
      })) 
    })) 
  }
}))

diffex<-diffexCalls|>
  mutate(signif=padj<0.05)|>
  mutate(direction=ifelse(log2FoldChange>0,'up','down'))

dgenes<-diffex|>
  tidyr::separate(Gene,sep="\\.",into=c('ensembl_gene_id','Version'))


diffex_gene<-dgenes|>left_join(map)
##let's save this back to synapse for future reference
dir.create("individual")
setwd("individual")
readr::write_csv(diffex_gene,file='totalRNASeqDiffExtoDate.csv')
synFile <- synapser::File('totalRNASeqDiffExtoDate.csv', parent = 'syn69910862')
synapser::synStore(synFile)


sumstats<-diffex|>subset(signif)|>
  group_by(Drug,Timepoint,direction,individualID)|>
  summarize(numgenes=n())

print('now have differential expression calls')
ggplot(sumstats,aes(x=Drug,y=numgenes,fill=direction))+facet_grid(individualID~Timepoint)+geom_bar(stat='identity',position='dodge')+coord_flip()+scale_fill_viridis_d()

ggplot(sumstats,aes(x=Drug,y=Timepoint,size=numgenes,col=Drug))+geom_point()+facet_grid(individualID~direction)+
  scale_size(range=c(1,15))+scale_color_discrete()+coord_flip()

# ggsave('diffExresults.pdf')
 
 ###lets look at correlation across drugs
 dmat<-diffex|>dplyr::select(Gene,log2FoldChange,Drug,Timepoint)|>
   tidyr::unite(Drug,Timepoint,col='sample',sep='_')|>
   tidyr::pivot_wider(names_from=sample,values_from=log2FoldChange, values_fn=mean)|>
   tibble::column_to_rownames('Gene')
 
 cmat<-cor(dmat,use='pairwise.complete.obs',method='spearman')
 meta<-diffex|>dplyr::select(Drug,Timepoint)|>
   distinct()|>
   tidyr::unite(Drug,Timepoint,col='sample',sep='_',remove=FALSE)|>
   tibble::column_to_rownames('sample')
 


drugscounts<-subset(dgenes,signif)|>group_by(ensembl_gene_id,Timepoint,direction)|>summarize(ndrugs=n())|>arrange(desc(ndrugs))|>
  left_join(map)|>
  subset(hgnc_symbol!='')

##most drugs just show up in one or two
ggplot(drugscounts,aes(x=ndrugs,fill=direction))+geom_bar(position='dodge')+ggtitle('Number of drugs with shared differentially expressed genes')

```
Most genes are only differentially expressed upon a single drug treatment, but some are shared with other drugs. Targeting these canonical responders might be effective, but what if we can find distinct genes changing?

