---
title: "Compare RNASeq Summaries"
author: "Sara gosline"
date: "2025-01-27"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr);library(dplyr);library(tidyverse); library(synapser); library(pheatmap); library(RColorBrewer)
source("plottingParameters.R")

```

Here we want to compare the various RNAseq analyses. We want to ask:
1. Which drugs elicit the most similar response? 
  - check RNASeq
  - check TF responses
  - check pathways
2. which drugs elicit the most distinct responses? 
  - check RNASeq
  - check TF responses
  - check pathways
3. What are the most profound changes caused by each drug? Do these predict other drugs? 
  - check RNASEq
  - check TF response
  - check pathways

## Load data, plot results

Load data, plot total differentially expressed genes for Panel 3A

```{r load data}
synLogin()
#all_exp<-readr::read_csv(synGet('syn64398051')$path)
diffex_genes<-readr::read_csv(synGet('syn64397743')$path)|>
  select(-Version)
#diffex_tfs<-readr::read_csv(synGet('syn64420968')$path)
#diff_tf_scores<-readr::read_csv(synGet('syn64695280')$path)

diffex_genes <- diffex_genes[diffex_genes$Drug != "Irinotecan",] # since irinotecan isn't metabolically active

##now we have three different ways of comparing the drugs
genemat<-diffex_genes|>
  dplyr::select(ensembl_gene_id,log2FoldChange,Drug,Timepoint)|>
  subset(!is.na(log2FoldChange))|>
  tidyr::unite(Timepoint,Drug,col='condition')|>
  tidyr::pivot_wider(names_from=condition,values_from=log2FoldChange,values_fn=mean,values_fill=0.0)|>
  tibble::column_to_rownames('ensembl_gene_id')|>
  as.matrix()

```

Next we can quickly evaluate how differential expression values are

```{r gene correlation}

genecor<-cor(genemat,use='pairwise.complete.obs')
write.csv(genecor, "geneCorr.csv")

annot <- as.data.frame(rownames(genecor)) |> 
  tidyr::separate('rownames(genecor)',into=c('Timepoint','Drug'),sep='_')
drug.info.df <- as.data.frame(t(as.data.frame(drug.info)))
drug.info.df$Drug <- rownames(drug.info.df)
colnames(drug.info.df)[1] <- "MOA"
annot <- merge(annot, drug.df, by="Drug", all.x=TRUE)
rownames(annot) <- paste0(annot$Timepoint, "_", annot$Drug)
annot <- annot[rownames(genecor),c('Timepoint','Drug','MOA')]
annot$Timepoint <- factor(annot$Timepoint, levels=c("8h","24h"))
annot$MOA <- factor(annot$MOA, levels=unique(annot$MOA))
annot$Drug <- factor(annot$Drug, levels=unique(annot$Drug))

#for this plot we rename moaCol to index by moa name
names(moaCol)<-unlist(drug.info)

annot.colors = list(MOA = moaCol,
                    Drug = drugCol,
                    Timepoint = c("8h" = "#FFFFFF", "24h" = "#000000"))

graphics.off()
pdf("geneCorr.pdf",width=6,height=4.5)
pheatmap(genecor, annotation_row=annot, annotation_col=annot,
         annotation_colors=annot.colors,
         show_colnames=FALSE, show_rownames=FALSE)
dev.off()

genemat<-diffex_genes[diffex_genes$padj < 0.05,]|>
  dplyr::select(ensembl_gene_id,log2FoldChange,Drug,Timepoint)|>
  subset(!is.na(log2FoldChange))|> 
  tidyr::unite(Timepoint,Drug,col='condition')|>
  tidyr::pivot_wider(names_from=condition,values_from=log2FoldChange,values_fn=mean,values_fill=0.0)|>
  tibble::column_to_rownames('ensembl_gene_id')|>
  as.matrix()

genecor<-cor(genemat,use='pairwise.complete.obs')
write.csv(genecor,"degCorr.csv")

graphics.off()
pdf("degCorr.pdf",width=6,height=4.5)
pheatmap(genecor, annotation_row=annot, annotation_col=annot, annotation_colors=annot.colors,
         show_colnames=FALSE, show_rownames=FALSE)
dev.off()

subset(diffex_genes,padj<0.05)|>
  group_by(Drug,Timepoint,direction)|>
  summarize(ngenes=n_distinct(hgnc_symbol))|>
  mutate(dgenes=ifelse(direction=='down',-log10(ngenes),log10(ngenes)))|>
  ggplot(aes(x=Drug,fill=direction,
             y=dgenes))+geom_bar(stat='identity',position='dodge',alpha=0.75)+
  scale_y_continuous(breaks=c(-4,-2,0,2,4), labels=c(-10^4, -10^2, 0, 10^2, 10^4)) +
  facet_grid(.~Timepoint) + scale_fill_manual(values=c("blue","red")) +
  labs(y='# of Differentially Expressed Genes (q < 0.05)', fill="Direction")+coord_flip()
ggsave("Ndegs_log10Scale.pdf", width=5.5, height=5)

subset(diffex_genes,padj<0.05)|>group_by(Drug,Timepoint,direction)|>
  summarize(ngenes=n_distinct(hgnc_symbol))|>
  mutate(dgenes=ifelse(direction=='down',-ngenes,ngenes))|>
  ggplot(aes(x=Drug,fill=direction,
             y=dgenes))+geom_bar(stat='identity',position='dodge',alpha=0.75)+
  facet_grid(.~Timepoint) + scale_fill_manual(values=c("blue","red")) +
  labs(y='# of Differentially Expressed Genes (q < 0.05)', fill="Direction")+coord_flip()
ggsave("Ndegs.pdf", width=5.5, height=5)


timepoints <- c("8h","24h")
diffex_genes$log2FoldChange <- as.numeric(diffex_genes$log2FoldChange)
diffex_genes <- diffex_genes[!is.na(diffex_genes$hgnc_symbol),]
n.top.list <- c(5)
n.top = 5
##for (n.top in n.top.list) {
  ### pathway
  topSigpathway <- data.frame()
  for (i in drugnames) {
    tempSig <- diffex_genes[diffex_genes$Drug == i &
                              diffex_genes$signif,] %>% slice_max(abs(log2FoldChange), n=n.top)
    topSigpathway <- rbind(topSigpathway, tempSig)
  }
  length(unique(topSigpathway$hgnc_symbol)) # 20
  dot.df <- na.omit(diffex_genes[diffex_genes$hgnc_symbol %in% unique(topSigpathway$hgnc_symbol),])
  
  mean.results <- plyr::ddply(dot.df, .(hgnc_symbol), summarize,
                              Log2FC = mean(log2FoldChange, na.rm=TRUE))
  setOrder <- mean.results[order(mean.results$Log2FC),]$hgnc_symbol
  drugOrder <- unlist(drug.info)
  
  dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
  dot.df$minusLogFDR <- -log10(dot.df$padj)
  maxAbsscore <- ceiling(max(abs(dot.df$log2FoldChange)))
  dot.plot <- ggplot2::ggplot(
    dot.df,
    ggplot2::aes(
      x = Drug, y = hgnc_symbol, color = log2FoldChange,
      size = minusLogFDR
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::scale_y_discrete(limits = setOrder) +
    ggplot2::scale_x_discrete(limits=drugnames) +
    scale_color_gradient2(low="blue",high="red", mid="grey", 
                          limits=c(-maxAbsscore, maxAbsscore)) +
    theme_minimal(base_size=12) +
    ggplot2::labs(color = "Log2FC", size = "-log(FDR)") + 
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
          axis.title = element_blank()) +
    geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
    facet_grid(.~Timepoint)
  dot.plot
  ggplot2::ggsave(paste0(n.top,"_topDEGs_dotPlot.pdf"), dot.plot, width=10, height=ifelse(n.top<2,4,2*n.top))
#}
# there were ensembl ids which mapped to the same gene

n.top.list <- c(10, 20, 30, 50)
for (n.top in n.top.list) {
  dot.df <- na.omit(diffex_genes) %>% slice_max(abs(log2FoldChange), n=n.top)
  
  mean.results <- plyr::ddply(dot.df, .(hgnc_symbol), summarize,
                              Log2FC = mean(log2FoldChange, na.rm=TRUE))
  setOrder <- mean.results[order(mean.results$Log2FC),]$hgnc_symbol
  drugOrder <- unlist(drug.info)
  
  dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
  dot.df$minusLogFDR <- -log10(dot.df$padj)
  maxAbsscore <- ceiling(max(abs(dot.df$log2FoldChange)))
  dot.plot <- ggplot2::ggplot(
    dot.df,
    ggplot2::aes(
      x = Drug, y = hgnc_symbol, color = log2FoldChange,
      size = minusLogFDR
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::scale_y_discrete(limits = setOrder) +
    ggplot2::scale_x_discrete(limits=drugnames) +
    scale_color_gradient2(low="blue",high="red", mid="grey", 
                          limits=c(-maxAbsscore, maxAbsscore)) +
    theme_minimal(base_size=12) +
    ggplot2::labs(color = "Log2FC", size = "-log(FDR)") + 
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
          axis.title = element_blank()) +
    geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
    facet_grid(.~Timepoint)
  dot.plot
  ggplot2::ggsave(paste0(n.top,"_topDEGsOverall_dotPlot.pdf"), dot.plot, width=10, height=ifelse(n.top<16,4,n.top/4))
}
# 
# #drug.targets <- read.csv("comboPrediction_usingTop500Sig/target_diffexp.csv")
# dot.df <- na.omit(diffex_genes[diffex_genes$hgnc_symbol %in% drug.targets$hgnc_symbol,]) #%>% slice_max(abs(log2FoldChange), n=n.top)
# 
# mean.results <- plyr::ddply(dot.df, .(hgnc_symbol), summarize,
#                             Log2FC = mean(log2FoldChange, na.rm=TRUE))
# setOrder <- mean.results[order(mean.results$Log2FC),]$hgnc_symbol
drugOrder <- rownames(drug.df)
# timepoints <- c("8h","24h")
# dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
# dot.df$minusLogFDR <- -log10(dot.df$padj)
# maxAbsscore <- ceiling(max(abs(dot.df$log2FoldChange)))
# dot.plot <- ggplot2::ggplot(
#   dot.df,
#   ggplot2::aes(
#     x = Drug, y = hgnc_symbol, color = log2FoldChange,
#     size = minusLogFDR
#   )
# ) +
#   ggplot2::geom_point() +
#   ggplot2::scale_y_discrete(limits = setOrder) +
#   ggplot2::scale_x_discrete(limits=drugOrder) +
#   scale_color_gradient2(low="blue",high="red", mid="grey", 
#                         limits=c(-maxAbsscore, maxAbsscore)) +
#   theme_minimal(base_size=12) +
#   ggplot2::labs(color = "Log2FC", size = "-log(FDR)") + 
#   theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
#         axis.title = element_blank()) +
#   geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
#   facet_grid(.~Timepoint)
# dot.plot
# ggplot2::ggsave("targetDiffExp_dotPlot.pdf", dot.plot, width=10, height=ifelse(length(setOrder)<16,4,length(setOrder)/4))
# 
# dot.plot <- ggplot2::ggplot(
#   dot.df,
#   ggplot2::aes(
#     x = Drug, y = hgnc_symbol, color = log2FoldChange,
#     size = minusLogFDR
#   )
# ) +
#   ggplot2::geom_point() +
#   ggplot2::scale_x_discrete(limits=drugOrder) +
#   scale_color_gradient2(low="blue",high="red", mid="grey", 
#                         limits=c(-maxAbsscore, maxAbsscore)) +
#   theme_minimal(base_size=12) +
#   ggplot2::labs(color = "Log2FC", size = "-log(FDR)") + 
#   theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
#         axis.title = element_blank()) +
#   geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
#   facet_grid(.~Timepoint)
# dot.plot
# ggplot2::ggsave("targetDiffExp_dotPlot_alphabeticalOrder.pdf", dot.plot, width=10, height=ifelse(length(setOrder)<16,4,length(setOrder)/2.67))
# 
# ggplot2::ggsave("targetDiffExp_dotPlot_alphabeticalOrder_shorter.pdf", dot.plot, width=10, height=ifelse(length(setOrder)<16,4,length(setOrder)/3.5))
# 
# dot.df <- na.omit(diffex_genes[diffex_genes$hgnc_symbol %in% drug.targets$hgnc_symbol,]) #%>% slice_max(abs(log2FoldChange), n=n.top)
# sig.dot.df <- dot.df[dot.df$signif,]
# dot.df <- dot.df[dot.df$hgnc_symbol %in% sig.dot.df$hgnc_symbol,]
# 
# drugOrder <- rownames(drug.info)
# timepoints <- c("8h","24h")
# dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
# dot.df$minusLogFDR <- -log10(dot.df$padj)
# maxAbsscore <- ceiling(max(abs(dot.df$log2FoldChange)))
# dot.plot <- ggplot2::ggplot(
#   dot.df,
#   ggplot2::aes(
#     x = Drug, y = hgnc_symbol, color = log2FoldChange,
#     size = minusLogFDR
#   )
# ) +
#   ggplot2::geom_point() +
#   ggplot2::scale_x_discrete(limits=drugOrder) +
#   scale_color_gradient2(low="blue",high="red", mid="grey", 
#                         limits=c(-maxAbsscore, maxAbsscore)) +
#   theme_minimal(base_size=12) +
#   ggplot2::labs(color = "Log2FC", size = "-log(FDR)") + 
#   theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
#         axis.title = element_blank()) +
#   geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
#   facet_grid(.~Timepoint)
# dot.plot
# ggplot2::ggsave("sigTargetDiffExp_dotPlot_alphabeticalOrder.pdf", dot.plot, width=10, height=ifelse(length(setOrder)<16,4,length(setOrder)/2.67))
# 
# ggplot2::ggsave("sigTargetDiffExp_dotPlot_alphabeticalOrder_shorter.pdf", dot.plot, width=10, height=ifelse(length(setOrder)<16,4,length(setOrder)/3.5))
# 
# diffex_genes$Timepoint <- factor(diffex_genes$Timepoint, levels=c("8h","24h"))
#Nbar <- ggplot2::ggplot(na.omit(diffex_genes[diffex_genes$signif,]), aes(x=Drug, fill=Timepoint)) + 
#  geom_bar(stat="count") + theme_classic() + 
#  scale_x_discrete(limits=drugOrder) + theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
#                                             axis.title.x=element_blank()) + 
#  ylab("# of Differentially\nExpressed Genes")
#ggsave("nSigDEGs.pdf", Nbar, width=10, height=2)

drugOrder = names(drug.class)
Ndegs <- plyr::ddply(na.omit(diffex_genes[diffex_genes$signif,]), .(Timepoint, Drug),summarize,
                     n=length(unique(hgnc_symbol)))
Ndegs$Timepoint <- factor(Ndegs$Timepoint, levels=c("8h","24h"))
Nbar <- ggplot2::ggplot(Ndegs, aes(x=Drug, y=n, fill=Timepoint)) + 
  geom_bar(stat="identity", position="dodge") + 
  theme_classic() + scale_y_continuous(trans="log10") + 
  scale_x_discrete(limits=drugOrder) + 
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1,
                                 color=classCol[drugOrder]),
        axis.title.x=element_blank()) + 
  ylab("# of Differentially\nExpressed Genes")
ggsave("fig3a_nSigDEGs_log10.pdf", Nbar, width=5, height=2)
write.csv(Ndegs, "nSigDEGs.csv", row.names=FALSE)

# upset plot
# generate list of significant results
sigList <- list()
for (i in unique(diffex_genes$Drug)) {
  if (nrow(diffex_genes[diffex_genes$Drug == i & 
                        diffex_genes$signif,]) > 0) {
    sigList[[i]] <- na.omit(unique(diffex_genes[diffex_genes$Drug == i & 
                                                  diffex_genes$signif,]$source)) 
  }
}

if (length(sigList) > 1) {
  # convert to matrix
  myPlot = ComplexHeatmap::make_comb_mat(sigList)
  pdf(file.path("DE_upsetPlot.pdf"))
  up <- ComplexHeatmap::UpSet(myPlot)
  ComplexHeatmap::draw(up)
  dev.off()
}

```

Many changes happening, often at 24 hours.

There is still some duplicate code here, and we just dont need this figure.
```{r box and whisker, echo=FALSE, eval = FALSE}
# prep data

diffex_genes <- read.csv(synapser::synGet('syn64397743')$path)|>
  select(-Version)
diffex_genes$Drug <- factor(diffex_genes$Drug, levels=names(drug.info))
timepoints <- c("8h", "24h")
diffex_genes$Timepoint <- factor(diffex_genes$Timepoint, levels=timepoints)


sig.tf <- na.omit(diffex_genes[diffex_genes$padj < 0.05,])
ntfs.overall <- plyr::ddply(sig.tf, .(Drug, Timepoint), dplyr::summarize,
                    N = length(na.omit(unique(hgnc_symbol))))

# are there more/less enriched tfs at 8h vs. 24h? No, p = 0.4687
cat("Overall", "# of DEGs at 8h vs. 24h\n")
t.test(ntfs.overall[ntfs.overall$Timepoint=="8h",]$N, 
       ntfs.overall[ntfs.overall$Timepoint=="24h",]$N)

ntfs.overall$moa <- NA
for (i in all.moas) {
  temp.drugs <- names(drug.info[drug.info %in% i])
  if (any(ntfs.overall$Drug %in% temp.drugs)) {
   ntfs.overall[ntfs.overall$Drug %in% temp.drugs,]$moa <- i
  }
}

# generate plot
ntfs.overall$moa <- factor(ntfs.overall$moa, levels=all.moas)
ntf.plot.overall <- ggplot2::ggplot(ntfs.overall, aes(x=Timepoint, y=N)) + geom_violin(alpha=0) +
  geom_point(aes(color=Drug, shape=moa)) +
  geom_boxplot(width=0.2, alpha = 0) + scale_color_manual(limits=drugnames, values=drugCol) +
  labs(x="Time (Hours)", y="# of Differentially Expressed Genes (q < 0.05)",
       color = "Drug Treatment", shape = "Mechanism") +
  scale_shape_manual(values=1:length(all.moas), breaks=all.moas) + theme_classic(base_size=12) +
  theme(legend.box="horizontal")
ggplot2::ggsave("N_DEGs_overall.pdf", ntf.plot.overall, width=6, height=4.5)
`
