---
title: "TF Enrichment experiments"
author: "Sara gosline"
date: "2024-12-13"
output: html_document
---

Here we take the differential expression and use it to infer transcription factors

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#reticulate::use_virtualenv('r-reticulate')
library(synapser)
library(plyr);library(dplyr)
synLogin()

source("plottingParameters.R")

counts<-readr::read_csv(synapser::synGet('syn64398051')$path)|>
  subset(Drug!='untreated')
diffex<-read.csv(synapser::synGet('syn64397743')$path)|> # vector memory error if use readr
  select(-Version)
diffex$stat <- as.numeric(diffex$stat)

counts_ens<-readr::read_csv(synapser::synGet('syn64398049')$path) |>
  subset(Drug!='untreated')



```

## Decoupler analysis 
Here we use the decoupler package to try to infer transcription factor activity from known databases

```{r decoupleR analysis}
library(decoupleR)
library(OmnipathR)
library(tibble)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(ggrepel)

if (file.exists('map.csv')) {
  map <- read.csv('map.csv')
} else {
  library(biomaRt)
  ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
  map<-getBM(attributes=c('ensembl_gene_id','hgnc_symbol'), filters ='ensembl_gene_id', values =unique(counts_ens$ensembl_gene_id), mart = ensembl)
  write.csv(map, 'map.csv', row.names = FALSE) 
}

cmat<- counts_ens|>
  dplyr::select(ensembl_gene_id,id,NumReads)|>
  tidyr::pivot_wider(names_from='id',values_from='NumReads', values_fn=mean)|>
  tibble::column_to_rownames(var='ensembl_gene_id')

design<-counts_ens|>
  dplyr::select(id,Drug,Timepoint,individualID)|>
  distinct()


# net <- decoupleR::get_collectri(organism = 'human', 
#                                 split_complexes = FALSE)
# [OmnipathR] Failed to download `https://omabrowser.org/All/oma-species.txt` (attempt 1/3); error: vector memory exhausted (limit reached?)
# readr::write_csv(net,file='collecTRI_complexes_network.csv')
# net <- read.csv(synapser::synGet("syn64713369")$path)

# net <- decoupleR::get_collectri(organism = 'human', 
#                                 split_complexes = TRUE)
# readr::write_csv(net,file='collecTRI_genes_network.csv')
net <- read.csv(synapser::synGet("syn64713370")$path)

enet<-net|>dplyr::rename(hgnc_symbol='target')|>
  left_join(map)|>
  subset(!is.na(ensembl_gene_id))|>
  dplyr::select(source,target='ensembl_gene_id',mor)

##write net to file and store at syn64693461
# sf<-synapser::File('collecTRI_complexes_network.csv',parentId='syn64693461')
# synapser::synStore(sf)

# sf<-synapser::File('collecTRI_genes_network.csv',parentId='syn64693461')
# synapser::synStore(sf)

sample_acts <- decoupleR::run_ulm(mat = cmat, 
                                  net = enet, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'mor', 
                                  minsize = 5)

##now we have putative transcritpion factors for each condition
tfs<-sample_acts|>
  dplyr::rename(id='condition')|>
  left_join(design)
print(head(tfs))
```
Now we have tf scores
## Most variable transcription factors

Let's start by plotting the most variable transcritpion factors, we dont need this for

```{r pressure, echo=FALSE, eval = FALSE}

vars<-tfs|>group_by(source)|>summarize(var=var(score))|>arrange(desc(var))

tf_mat<-sample_acts|>
  subset(source%in%c(vars$source[1:25]))|>
  dplyr::select(source,condition,score)|>
  pivot_wider(names_from='source',values_from='score')|>
  tibble::column_to_rownames('condition')

ddesign<-design|>
  tibble::column_to_rownames('id')|>
  mutate(Timepoint=as.character(Timepoint))

tcolors=c("white",'lightgrey','darkgrey')
names(tcolors)<-c('0h','8h','24h')
pheatmap(tf_mat,annotation_row = ddesign,labels_row=c(),annotation_colors=list(Timepoint=tcolors,Drug=cols))


sample_acts|>
  dplyr::rename(id='condition')|>
  mutate(log10p=(-log10(p_value)))|>
  left_join(design)|>
  subset(source%in%vars$source[1:15])|>
  ggplot(aes(x=source,y=score,col=Drug,size=log10p))+geom_point()+facet_grid(~Timepoint)+scale_color_manual(values=cols)+coord_flip()

readr::write_csv(sample_acts,'allTFestimatesByTime.csv')
sf<-synapser::File('allTFestimatesByTime.csv',parentId='syn64693461')
synapser::synStore(sf)

```

I think this plot is more a message of where the outliers lie rather than the actual biological value of each of the TFs. It also shows the need for batch correction. 

## Transcription factors by Drug

Here we use t statistics as input to decoupler, so that we can identify the TFs that are changing positively or negatively in each Drug treatment. 

```{r tfs by Drug,warning=FALSE, message=FALSE}
allcontrast<-do.call(rbind,lapply(unique(diffex$Drug),function(d){
  do.call(rbind,lapply(unique(diffex$Timepoint),function(ti){
    dtab<-subset(diffex,Timepoint==ti)|>
      subset(!is.na(stat))|>
      subset(Drug==d)
    if (nrow(dtab) > 0) {
      dtab <- dtab |> dplyr::select(ensembl_gene_id,Drug,Timepoint,stat)|>
        group_by(ensembl_gene_id,Drug,Timepoint)|>
        summarize(stat=mean(stat,na.rm=TRUE))|>
        tibble::column_to_rownames('ensembl_gene_id')
      dtab <- na.omit(dtab) 
    }
    if (nrow(dtab)>0) {
      contrast_acts <- decoupleR::run_ulm(mat = dtab[, 'stat', drop = FALSE],
                                          net = enet,
                                          .source = 'source',
                                          .target = 'target',
                                          .mor='mor',
                                          minsize = 5)|>
        ungroup()|>
        mutate(Drug=d,Timepoint=ti)
    } else {
      warning("0 rows for ", ti, d)
      contrast_acts <- data.frame()
    }
    return(contrast_acts)
  }))
}))

allcontrast$padj<-p.adjust(allcontrast$p_value)
allcontrast<-allcontrast|>mutate(signif=(padj<0.01))
subset(allcontrast,padj<0.01)|>
  ggplot(aes(x=source,y=score,size=signif,col=Drug,shape=as.factor(Timepoint)))+
  geom_point()+coord_flip()+facet_grid(.~Timepoint)+scale_color_manual(values=drugCol)

readr::write_csv(allcontrast,file='tfEstimatesEnsGenesByTime.csv')
sf <-synapser::File('tfEstimatesEnsGenesByTime.csv',parentId='syn64693461')
synapser::synStore(sf)
#allcontrast <- read.csv(synapser::synGet("syn64420968")$path)

md <- allcontrast|>
  dplyr::select(Drug,Timepoint)|>
  distinct()|>
  tidyr::unite(Drug,Timepoint,col='DrugTimepoint',remove=FALSE)|>
  tibble::column_to_rownames('DrugTimepoint')

mat<-allcontrast|>
  dplyr::select(source,Drug,Timepoint,score)|>
  tidyr::unite(Drug,Timepoint,col='DrugTimepoint')|>
  pivot_wider(names_from=DrugTimepoint,values_from=score,values_fill = 0.0)
```

We can evaluate how the transcription factor values correlate with one another
but don't use this analysis in the paper.
```{r tfcors, eval=FALSE}
corval<-mat|>tibble::column_to_rownames('source')|>cor(use='pairwise.complete.obs')

cortab<-corval|>
  as.data.frame()|>
 tibble::rownames_to_column('drug_exp1')|>
  tidyr::pivot_longer(contains('h'),names_to='drug_exp2',values_to='cor')|>
  arrange(cor)

print(cortab)
tcolors=list(`8h`='lightgrey',`24h`='darkgrey')
#md$Date=as.factor(md$Date)
pheatmap(corval,annotation_row = md)#,annotation_col=md,annotation_colors=list(Drug=drugCol,Timepoint=tcolors))

pcs<-mat|>tibble::column_to_rownames('source')|>as.matrix()|>prcomp()

comps<-cbind(pcs$rotation[rownames(md),],md)|>unique()
comps|>
  ggplot(aes(x=PC1,y=PC2,col=Drug,shape=as.factor(Timepoint),size=4))+
  geom_point()

```


### plot number of significant TFs

```{r plot # of results, echo = FALSE}
# prep data

base.path <- getwd()
dir.create("TF")
tf.path <- file.path(base.path, "TF")
setwd(tf.path)
#allcontrast <- read.csv(synapser::synGet("syn69911032")$path) # was syn64420968
allcontrast <- allcontrast[allcontrast$Drug != "Irinotecan",] # since irinotecan isn't metabolically active
#allcontrast$Drug <- factor(allcontrast$Drug, levels=names(drug.info))
timepoints <- c("8h", "24h")
allcontrast$Timepoint <- factor(allcontrast$Timepoint, levels=timepoints)
individualID <- unique(allcontrast$individualID)
all.drugs <- unique(allcontrast$Drug)
all.moas <- sort(unique(unlist(drug.info)))


sig.tf <- allcontrast[allcontrast$padj <= 0.05,]
ntfs.overall <- plyr::ddply(sig.tf, .(Drug, Timepoint), dplyr::summarize,
                    N = length(na.omit(unique(source))))

# are there more/less enriched tfs at 8h vs. 24h? More at 8h: two-sided (p=0.04375196); 8h greater 1-sided (p=0.02187598)
cat("Overall", "# of tfs enriched at 8h vs. 24h\n")
t.test(ntfs.overall[ntfs.overall$Timepoint=="8h",]$N, 
       ntfs.overall[ntfs.overall$Timepoint=="24h",]$N)

ntfs.overall$moa <- NA
for (i in all.moas) {
  temp.drugs <- names(drug.info[drug.info %in% i])
  if (any(ntfs.overall$Drug %in% temp.drugs)) {
   ntfs.overall[ntfs.overall$Drug %in% temp.drugs,]$moa <- i
  }
}

# generate plot
ntfs.overall$moa <- factor(ntfs.overall$moa, levels=all.moas)
ntf.plot.overall <- ggplot2::ggplot(ntfs.overall, aes(x=Timepoint, y=N)) + geom_violin(alpha=0) +
  geom_point(aes(color=Drug, shape=moa)) +
  geom_boxplot(width=0.2, alpha = 0) + scale_color_manual(limits=all.drugs, values=drugCol) +
  labs(x="Time (Hours)", y="# of Enriched Transcription Factors",
       color = "Drug Treatment", shape = "Mechanism") +
  scale_shape_manual(values=1:length(all.moas), breaks=all.moas) + theme_classic(base_size=12) +
  theme(legend.box="horizontal")

ntf.plot.overall <- ggplot2::ggplot(ntfs.overall, aes(x=Timepoint, y=N)) + #geom_violin(alpha=0) +
  geom_bar(position='dodge',stat='identity', aes(fill=Drug)) +
  #geom_boxplot(width=0.2, alpha = 0) + 
  scale_fill_manual(limits=all.drugs, values=drugCol) +
  labs(x="Time (Hours)", y="# of Enriched Transcription Factors",
       color = "Drug Treatment", shape = "Mechanism") +
  scale_shape_manual(values=1:length(all.moas), breaks=all.moas) + theme_classic(base_size=12) +
  theme(legend.box="horizontal")

ggplot2::ggsave("N_enrichedTFs_overall.pdf", ntf.plot.overall, width=5.5, height=4)
```

### plot all enriched TFs
```{r plot all results, echo = FALSE}

moaCol <- classCol[intersect(names(classCol),allcontrast$Drug)]
tf.path <- file.path(getwd(),"TF")
setwd(tf.path)
topSets <- unique(allcontrast[allcontrast$signif,]$source)
dot.df <- allcontrast[allcontrast$source %in% topSets,]
mean.results <- plyr::ddply(dot.df, .(source), summarize,
                            score = mean(score, na.rm=TRUE))
setOrder <- mean.results[order(mean.results$score),]$source

dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
dot.df$minusLogFDR <- -log10(dot.df$padj)
maxAbsscore <- ceiling(max(abs(dot.df$score)))
dot.plot <- ggplot2::ggplot(
  na.omit(dot.df),
  ggplot2::aes(
    x = Drug, y = source, color = score,
    size = minusLogFDR
  )
) +
  ggplot2::geom_point() +
  ggplot2::scale_y_discrete(limits = setOrder) +
  ggplot2::scale_x_discrete(limits=names(moaCol)) +
  scale_color_gradient2(low="blue",high="red", mid="grey", 
                        limits=c(-maxAbsscore, maxAbsscore)) +
  theme_minimal(base_size=12) +
  ggplot2::labs(color = "Score", size = "-log(FDR)") + 
  theme(axis.title=element_blank(), 
        axis.text.x=element_text(angle=45, vjust=1, hjust=1,colour=moaCol)) +
  geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
  facet_grid(.~Timepoint)
dot.plot
#etfs <- length(unique)
ggplot2::ggsave("enrichedTFs_dotPlot.pdf", dot.plot, width=8, height=12) 

red.df <- dot.df[-grep('_',dot.df$source,),]

dot.plot <- ggplot2::ggplot(
  na.omit(red.df),
  ggplot2::aes(
    x = Drug, y = source, color = score,
    size = minusLogFDR
  )
) +
  ggplot2::geom_point() +
#ggplot2::scale_y_discrete(limits = setOrder) +
  ggplot2::scale_x_discrete(limits=names(moaCol)) +
  scale_color_gradient2(low="blue",high="red", mid="grey", 
                        limits=c(-maxAbsscore, maxAbsscore)) +
  theme_minimal(base_size=12) +
  ggplot2::labs(color = "Score", size = "-log(FDR)") + 
  theme(axis.title=element_blank(), 
        axis.text.x=element_text(angle=45, vjust=1, hjust=1,color=moaCol)) +
  geom_point(data = subset(red.df, signif), col = "black", stroke = 1.5, shape = 21) +
  facet_grid(.~Timepoint)

etfs<-length(unique(red.df$source))
dot.plot
ggsave(paste0('enrichedNonComplex_',etfs,'_tfs.pdf'), dot.plot, width=8, height=8)

```
Now we have a visualization of the transcriptional mechanisms behind these plots


### plot top enriched pathways
dont need this analysis

```{r plot top results, echo = FALSE}
library(plyr);library(dplyr);library(ggplot2);library(poolr)
tf.path <- file.path(getwd(),"TF")
setwd(tf.path)

# #moa.info2 <- dplyr::distinct(ntfs[,c("Drug","moa")])
# moa.info2 <- as.data.frame(t(as.data.frame(drug.info[names(drug.info) %in% unique(allcontrast$Drug)])))
# colnames(moa.info2) <- "moa"
# moa.info2$Drug <- rownames(moa.info2)
# moaPal <- scales::hue_pal()(length(unique(moa.info2$moa)))
# names(moaPal) <- unique(moa.info2$moa)
# # create tile plot for drug pathway annotations
#   moa.plot <- ggplot2::ggplot(moa.info2, aes(x=Drug,y=1, fill=moa)) + 
#   geom_tile() + scale_x_discrete(limits=all.drugs) + theme_classic(base_size=12) +
#   theme(axis.ticks=element_blank(), axis.text = element_blank(), axis.title=element_blank(),
#         axis.line=element_blank()) + labs(fill="Drug Mechanism") + scale_color_manual(values=moaPal, breaks=names(moaPal))
# ggplot2::ggsave("MOA_annotations.pdf", moa.plot, width=6,height=3)

n.top.list <- c(1, 2, 3, 5)
tested.drugs <- all.drugs[all.drugs %in% allcontrast$Drug]
for (n.top in n.top.list) {
  ### pathway
  topSigpathway <- data.frame()
  for (i in tested.drugs) {
    tempSig <- allcontrast[allcontrast$Drug == i &
                              allcontrast$signif,] %>% slice_max(abs(score), n=n.top)
    topSigpathway <- rbind(topSigpathway, tempSig)
  }
  length(unique(topSigpathway$source)) # 20
  dot.df <- allcontrast[allcontrast$source %in% unique(topSigpathway$source),]
  com <-grep('_',dot.df$source)
  if(length(com)>0)
    dot.df <- dot.df[-com,]

  mean.results <- plyr::ddply(dot.df, .(source), summarize,
                              score = mean(score, na.rm=TRUE))
  setOrder <- mean.results[order(mean.results$score),]$source
  drugOrder <- intersect(names(classCol), dot.df$Drug)
  dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
  dot.df$minusLogFDR <- -log10(dot.df$padj)
  maxAbsscore <- ceiling(max(abs(dot.df$score)))
  dot.plot <- ggplot2::ggplot(
    na.omit(dot.df),
    ggplot2::aes(
      x = Drug, y = source, color = score,
      size = minusLogFDR
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::scale_y_discrete(limits = setOrder) +
    ggplot2::scale_x_discrete(limits=drugOrder) +
    scale_color_gradient2(low="blue",high="red", mid="grey", 
                          limits=c(-maxAbsscore, maxAbsscore)) +
    #viridis::scale_color_viridis() +
    theme_minimal(base_size=12) +
    ggplot2::labs(color = "Score", size = "-log(FDR)") + 
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1,
                                   color=classCol[drugOrder]),
          axis.title = element_blank()) +
    geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21) +
    facet_grid(.~Timepoint)
  dot.plot
  ggplot2::ggsave(paste0(n.top,"_topEnrichedTFs_dotPlot.pdf"), dot.plot, width=8, height=ifelse(n.top<2,4,n.top))
}
```


### plot top enriched pathways for each drug MOA

here we plot Tfs for each MOA, also dont need to evaluate

```{r plot top results per MOA, echo = FALSE, eval=FALSE}
library(plyr);library(dplyr);library(ggplot2)
tf.path <- file.path(getwd(),"TF")
setwd(tf.path)

n.top.list <- c(1, 2, 3, 5)
for (moa in all.moas) {
  temp.drugs <- names(drug.info[drug.info == moa])
  for (n.top in n.top.list) {
    ### pathway
    topSigpathway <- data.frame()
    for (i in temp.drugs) {
      tempSig <- allcontrast[allcontrast$Drug == i &
                           allcontrast$signif,] %>% slice_max(abs(score), n=n.top)
      topSigpathway <- rbind(topSigpathway, tempSig)
    }
    dot.df <- allcontrast[allcontrast$source %in% unique(topSigpathway$source) &
                        allcontrast$Drug %in% temp.drugs,]
    if (nrow(dot.df) > 0) {
      mean.results <- plyr::ddply(dot.df, .(source), summarize,
                                  score = mean(score, na.rm=TRUE))
      setOrder <- mean.results[order(mean.results$score),]$source
      drugOrder <- unlist(drug.info)
      
      dot.df$Timepoint <- factor(dot.df$Timepoint, levels=timepoints)
      dot.df$minusLogFDR <- -log10(dot.df$padj)
      maxAbsscore <- ceiling(max(abs(dot.df$score)))
      dot.plot <- ggplot2::ggplot(
        na.omit(dot.df),
        ggplot2::aes(
          x = Drug, y = source, color = score,
          size = minusLogFDR
        )
      ) +
        ggplot2::geom_point() +
        ggplot2::scale_y_discrete(limits = setOrder) +
        ggplot2::scale_x_discrete(limits=temp.drugs) +
        scale_color_gradient2(low="blue",high="red", mid="grey", 
                              limits=c(-maxAbsscore, maxAbsscore)) +
        theme_minimal(base_size=12) +
        ggplot2::labs(color = "Score", size = "-log(FDR)") + 
        theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
              axis.title = element_blank()) +
        geom_point(data = subset(dot.df, signif), col = "black", stroke = 1.5, shape = 21)
      if (length(unique(dot.df$Timepoint)) > 0) {
        dot.plot <- dot.plot + facet_grid(.~Timepoint)
      }
      dot.plot
      customWidth <- ifelse(length(temp.drugs) < 3, 5, length(temp.drugs)+2)
      ggplot2::ggsave(paste0(moa, "_",n.top,"_topEnrichedTFs_dotPlot.pdf"), 
                      dot.plot, width=customWidth, height=ifelse(n.top<3,3,n.top)) 
    }
  } 
}
```


### upset plots to see overlap across drug treatments
```{r plot overlap of results, echo = FALSE, eval=FALSE}
# generate list of significant results
sigList <- list()
for (i in all.drugs) {
  if (nrow(allcontrast[allcontrast$Drug == i & 
                       allcontrast$signif,]) > 0) {
    sigList[[i]] <- na.omit(unique(allcontrast[allcontrast$Drug == i & 
                                                 allcontrast$signif,]$source)) 
  }
}

if (length(sigList) > 1) {
  # convert to matrix
  myPlot = ComplexHeatmap::make_comb_mat(sigList)
  pdf(file.path(tf.path,"enrichedTF_upsetPlot.pdf"))
  up <- ComplexHeatmap::UpSet(myPlot)
  ComplexHeatmap::draw(up)
  dev.off()
}
```

