---
title: "Kinase Enrichment from Phosphoproteomic data using KSEA" 
author: "Ruby Fore"
date: "2024-11-19"
output: html_document
---

```{r}
# load required libraries
library(ggplot2)
library(dplyr)
library(httr)
library(jsonlite)
library(KSEAapp)

```



Obtain the phosphosite data from Synapse 

```{r}
# login to Synapse
synapser::synLogin()
```


```{r}
# normalized phosphoproteomics data
fileEntity <- synapser::synGet('syn61695519')
phosphosites <- read.table(fileEntity$path, header = T)
```


Check for level of missingness in phosphosite data


```{r}
dim(phosphosites)
sum(is.na(phosphosites))
# number of proteins present in at least 50% of samples (should be 0 
# thanks to preprocessing)
sum(rowSums(is.na(phosphosites))/66 > 0.5)
```

Now we can impute the remaining missing values to the mean of each protein/row. 

```{r}
# get the indices of NA's
na_indices <- which(is.na(phosphosites), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
phosphosites[na_indices] <- rowMeans(phosphosites, na.rm=T)[na_indices[,1]]
# check that all NA's have been imputed
sum(is.na(phosphosites))
```


Now we want to create a data frame with the sample info, seperated into columns.

```{r}
#make factor info from the column names
factorInfo <- data.frame(condition = colnames(phosphosites))

factorInfo<-factorInfo|>
  tidyr::separate(condition, into = c("sample", 'drug-time-rep'), sep = '_', remove=F, extra='merge') |>
  tidyr::separate(`drug-time-rep`, into=c("drug", "time-rep"), sep = "_", extra='merge') |>
  tidyr::separate(`time-rep`, into=c("time", "rep"), sep=-8)

```

The function below runs differential gene expression in limma and returns the 'topTable' of genes, p-vals and log fold changes. 

```{r}
# function to get limma topTable for specified data, formula and contrast
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  
  factorInfo <- data.frame(condition = colnames(data))
factorInfo<-factorInfo|>
  tidyr::separate(condition, into = c("sample", 'drug-time-rep'), sep = '_', remove=F, extra='merge') |>
  tidyr::separate(`drug-time-rep`, into=c("drug", "time-rep"), sep = "_", extra='merge') |>
  tidyr::separate(`time-rep`, into=c("time", "rep"), sep=-8)



factorInfo$drug <-as.factor(factorInfo$drug)

factorInfo$time <- as.factor(factorInfo$time)


expSet <- Biobase::ExpressionSet(as.matrix(data))

# log2 transformation if distribution isn't normal
ex <- Biobase::exprs(expSet)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                          na.rm = TRUE
))
LogC <- (qx[5] > 100) ||
  (qx[6] - qx[1] > 50 && qx[2] > 0)
#print(LogC)
if (LogC) {
  ex[which(ex <= 0)] <- NaN
  Biobase::exprs(proSet) <- log2(ex)
}

# print commands left to aid the user in double checking the correct contrast
# is being ran for each combo of sample, drug, time
#print('before assigning factors')
expSet$sample <- factorInfo$sample
#print(factorInfo$drug)
expSet$drug <- factorInfo$drug
#print(factorInfo$time)
expSet$time <- factorInfo$time
#print('before model matrix')
design <- stats::model.matrix(formula, expSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
fit <- limma::lmFit(expSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)

fit <- limma::eBayes(fit)
deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
return(list(deg, nDiffExpressed))
}
```


Using `dplyr` to reformat the data to make running all the contrasts a bit more logical and easy to keep track of. Takes a second or three. 
```{r}
# do we need this only for phosphorylation ID?
phosphosites$p_site_ids <- rownames(phosphosites)
# change -67 to -c(p_site_ids) ???
samps<-colnames(phosphosites)[-67]

long_phosphosites<-phosphosites|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio') |>
  tidyr::separate(condition, into = c("sample", 'drug-time-rep'), sep = '_', remove=F, extra='merge') |>
  tidyr::separate(`drug-time-rep`, into=c("drug", "time-rep"), sep = "_", extra='merge') |>
  tidyr::separate(`time-rep`, into=c("time", "rep"), sep="_hr_")

```


Running the differential expression analyses

```{r}
ddat<-data.frame()
for (s in c("JH.2.002", "MN.2")){
  for (d in c("Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat")){
    for (tv in c(0.5, 3, 24)){
      
       tmat<-long_phosphosites|>
         subset(sample==s) |>
        subset(drug==d)|>
        subset(time==tv)|>
        select(p_site_ids,condition,logRatio)|>
        dplyr::distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('p_site_ids')
      
       if(nrow(tmat)==0)##there is no data for this
        next
       #print(names(tmat))
       cmat<-long_phosphosites|>
         subset(sample==s) |>
        subset(drug=="DMSO")|>
        subset(time==tv)|>
        select(p_site_ids,condition,logRatio)|>
        dplyr::distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('p_site_ids')

       
       #print(names(cbind(tmat, cmat)))
        res<-getNdiffExpressedGenes(cbind(tmat,cmat), formula = ~ drug  + 0, analysisContrast =paste0('drug', d,'-drugDMSO'))
        tres<-res[[1]]|>mutate(time=tv,drug=d, sample=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
    }
  }
}
```

Make a table of number of differentially expressed phosphosites

```{r}

dcounts = ddat|>subset(adj.P.Val<0.05)|>group_by(time,drug,sample)|>summarize(DiffEx=n())
# table
knitr::kable(dcounts)
# barchart
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=sample))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)+
  ggtitle("Number of differentially expressed phosphorylated proteins \n compared to DMSO at same timepoint") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

Here we run KSEA without peptide or protein information, but these could be added to obtain potentially richer results. 
```{r}
# already created - dont run twice
# dataFolder <- synapser::Folder("Kinase Enrichment", parent = 'syn61841025')
# dataFolder <- synapser::synStore(dataFolder)
# set parent folder as Kinase Enrichment 
dataFolder <- synapser::synGet("syn64096845")

samps <-c('JH.2.002','MN.2')
drugs <-c("Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat")
times <- c(0.5, 3, 24)

##now run through all combinations but bind them together in a single table

ksea_res<-do.call(rbind,lapply(samps,function(s){
   dres<-do.call(rbind, lapply(drugs,function(d){
      tres<-do.call(rbind,lapply(times,function(tv){
     #     print(paste(tv,d,s))
           kseadat <- ddat |>
              subset(drug == d) |>
              subset(time ==tv) |>
              subset(sample == s) |>
              rename(rawGeneName = Gene)|>
    # choose the last element when splitting on dashes
          mutate(Residue.Both = stringr::str_split_i(rawGeneName, "-", -1)) |>
    # split on second dash
          tidyr::separate(rawGeneName, remove=F, into = c('Gene', 'blankcol2'), sep="[^ -][^-]*$") # |>
 

          if (nrow(kseadat) < 2)
                  return(data.frame(Kinase.Gene=c(),log2FC=c(),mS=c(),Enrichment=c(),m=c(),z.score=c(),p.value=c(),drug=c(),time=c(),sample=c()))
          #replace lowercase letters with semicolon
          kseadat$Residue.Both <- gsub('([[:lower:]])', ';', kseadat$Residue.Both)
          print(dim(kseadat))
          # remove trailing semicolons
          kseadat$Residue.Both <- stringr::str_sub(kseadat$Residue.Both, end = -2)
          #remove trailing dash from gene/protein names
          kseadat$Gene <- stringr::str_sub(kseadat$Gene, end = -2)
          #prepare data for ksea
          kseadat <- kseadat[, c("Gene", "Residue.Both", "P.Value", "logFC")]
          # KSEA is finicky about names
          kseadat$FC <- 2^kseadat$logFC
          names(kseadat)[names(kseadat) =='P.Value'] <- 'p'
          
          kseadat <- data.frame(kseadat, Protein = rep(NA, nrow(kseadat)), Peptide = rep(NA, nrow(kseadat)))
          
          kseadat$Peptide <- "NULL"
          kseadat$Protein <- "NULL"
          kseadat <- kseadat[, c("Protein", "Gene", "Peptide", "Residue.Both", "p", "FC")]
          
          
          KSEA.Complete(KSData, kseadat, NetworKIN = FALSE, m.cutoff = 1, p.cutoff = .1)
          
          
          KSEAKinaseScores <- read.csv(file="KSEA Kinase Scores.csv",row.names=NULL)|>
            mutate(drug=d,time=tv,sample=s)
          KSEALinks <- read.csv(file='Kinase-Substrate Links.csv',row.names=NULL)|>
            mutate(drug=d,time=tv,sample=s)
          return(KSEAKinaseScores)

      }))
      
      return(tres)
  }))
   return(dres)
}))

#for (s in c("JH.2.002", "MN.2")){
#  for (d in c("Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat")){
#    for (tv in c(0.5, 3, 24)){
      
# filterdat <- ddat |> subset(adj.P.Val < 0.05) |>
#   subset(drug == d) |>
#   subset(time ==tv) |>
#   subset(sample == s)
#    if (nrow(filterdat) < 10)
#            next
#       problemrows <-  c(770, 845, 1536, 1552, 1931, 2348, 2922, 2946, 3237,
# 4461, 4631, 4930, 4998, 5279, 5428, 5879, 5899, 6115, 6118, 6187)

write.csv(ksea_res, file=paste0('KSEA-Kinase-scores.csv'))
mySynFile1 <- synapser::File('KSEA-Kinase-scores.csv', parent = dataFolder)
synapser::synStore(mySynFile1)

#write.csv(KSEALinks, file=paste0("./", d, "_", tv, "_", s, "_", 'KSEA-Kinase-Links.csv'))
#mySynFile2 <- synapser::File(paste0("./", d, "_", tv, "_", s, "_", 'KSEA-Kinase-Links.csv'), parent = dataFolder)
#synapser::synStore(mySynFile2)

#    }
#  }
#}
```
