---
title: "MIT phosphosite kinase enrichment"
author: "Ruby Fore"
date: "2024-12-02"
output: html_document
---

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(synapser)
```

```{r}
synLogin()
fileEntity <- synapser::synGet('syn63836182')

phosphosites <- read.table(fileEntity$path, header = T)
```


Check for level of missingness in phosphosite data


```{r}
dim(phosphosites)
sum(is.na(phosphosites))
# number of proteins present in at least 50% of samples
sum(rowSums(is.na(phosphosites))/66 > 0.5)



```
Rows with high levels of missingness have been removed. 

Now we can impute the remaining missing values to the mean of each protein/row. 

### impute for NAs  - change to row mean
```{r}
# get the indices of NA's
na_indices <- which(is.na(phosphosites), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
phosphosites[na_indices] <- rowMeans(phosphosites, na.rm=T)[na_indices[,1]]
#na_indices
sum(is.na(phosphosites))

```

```{r}
#make factor info
factorInfo <- data.frame(condition = colnames(phosphosites))

factorInfo<-factorInfo|>
  tidyr::separate(condition, into = c("sample", 'drug-time-rep'), sep = '_', remove=F, extra='merge') |>
  tidyr::separate(`drug-time-rep`, into=c("drug", "time-rep"), sep = "_", extra='merge') |>
  tidyr::separate(`time-rep`, into=c("time", "rep"), sep=-8)

```
The function below runs differential gene expression in limma and returns the 'topTable' of genes, p-vals and log fold changes.




```{r}
# function to get limma topTable for specified data, formula and contrast
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  
  factorInfo <- data.frame(condition = colnames(data))
factorInfo<-factorInfo|>
  tidyr::separate(condition, into = c("sample", 'drug-time-rep'), sep = '_', remove=F, extra='merge') |>
  tidyr::separate(`drug-time-rep`, into=c("drug", "time-rep"), sep = "_", extra='merge') |>
  tidyr::separate(`time-rep`, into=c("time", "rep"), sep=-8)



factorInfo$drug <-as.factor(factorInfo$drug)

factorInfo$time <- as.factor(factorInfo$time)


expSet <- Biobase::ExpressionSet(as.matrix(data))

# log2 transformation if distribution isn't normal
ex <- Biobase::exprs(expSet)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                          na.rm = TRUE
))
LogC <- (qx[5] > 100) ||
  (qx[6] - qx[1] > 50 && qx[2] > 0)
#print(LogC)
if (LogC) {
  ex[which(ex <= 0)] <- NaN
  Biobase::exprs(proSet) <- log2(ex)
}

# print commands left to aid the user in double checking the correct contrast
# is being ran for each combo of sample, drug, time
#print('before assigning factors')
expSet$sample <- factorInfo$sample
#print(factorInfo$drug)
expSet$drug <- factorInfo$drug
#print(factorInfo$time)
expSet$time <- factorInfo$time
#print('before model matrix')
design <- stats::model.matrix(formula, expSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
fit <- limma::lmFit(expSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)

fit <- limma::eBayes(fit)
deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
return(list(deg, nDiffExpressed))
}
```


Using `dplyr` to reformat the data to make running all the contrasts a bit more logical and easy to keep track of. Takes a second or three. 
```{r}
# do we need this only for phosphorylation ID?
phosphosites$p_site_ids <- rownames(phosphosites)
# change -67 to -c(p_site_ids) ???
samps<-colnames(phosphosites)[-67]

long_phosphosites<-phosphosites|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio') |>
  tidyr::separate(condition, into = c("sample", 'drug-time-rep'), sep = '_', remove=F, extra='merge') |>
  tidyr::separate(`drug-time-rep`, into=c("drug", "time-rep"), sep = "_", extra='merge') |>
  tidyr::separate(`time-rep`, into=c("time", "rep"), sep="_hr_")

```

Running the differential expression analyses

```{r}
ddat<-data.frame()
for (s in c("JH.2.002", "MN.2")){
  for (d in c("Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat")){
    for (tv in c(0.5, 3, 24)){
      
       tmat<-long_phosphosites|>
         subset(sample==s) |>
        subset(drug==d)|>
        subset(time==tv)|>
        select(p_site_ids,condition,logRatio)|>
        dplyr::distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('p_site_ids')
      if(nrow(tmat)==0)##there is no data for this
        next
       #print(names(tmat))
       cmat<-long_phosphosites|>
         subset(sample==s) |>
        subset(drug=="DMSO")|>
        subset(time==tv)|>
        select(p_site_ids,condition,logRatio)|>
        dplyr::distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('p_site_ids')
      
       
       
       #print(names(cbind(tmat, cmat)))
        res<-getNdiffExpressedGenes(cbind(tmat,cmat), formula = ~ drug  + 0, analysisContrast =paste0('drug', d,'-drugDMSO'))
        tres<-res[[1]]|>mutate(time=tv,drug=d, sample=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
    }
  }
}
```

Make a table of number of differentially expressed phosphosites

```{r}

dcounts = ddat|>subset(adj.P.Val<0.05)|>group_by(time,drug,sample)|>summarize(DiffEx=n())
# table
knitr::kable(dcounts)
# barchart
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=sample))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)+
  ggtitle("Number of differentially expressed phosphorylated proteins \n compared to DMSO at same timepoint") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```


```{r}



 for (s in c("JH.2.002", "MN.2")){
   for (d in c("Mirdametinib", "Selumetinib", "Trametinib", "Vorinostat")){
     for (tv in c(0.5, 3, 24)){
# for downregulated phosphosites
mitforeground <- ddat |> subset(adj.P.Val < 0.05) |>
  subset(drug == d) |>
  subset(time ==tv) |>
  subset(sample == s) 
if (nrow(mitforeground) < 9)
          next
parentDir = "~/Projects/NF_Projects/"
foreground <- stringr::str_split_fixed(mitforeground$Gene, "\\.", 3)[,2]
  names(foreground) <- c()
write.table(foreground, row.names=F, col.names=F, 
          quote=F, 
          file=paste0(parentDir, d, "_", tv, "_", s, "_", 'foreground-set-for-MIT-kinase-enrichment.csv'))
  
  

mitbackground <- ddat |> subset(adj.P.Val >= 0.05) |>
  subset(drug == d) |>
  subset(time ==tv) |>
  subset(sample == s) 

background <- stringr::str_split_fixed(mitbackground$Gene, "\\.", 3)[,2]
names(background) <- c()
write.table(background, row.names=F, col.names=F,
          quote=F, 
          file=paste0(parentDir, d, "_", tv, "_", s, "_", 'background-set-for-MIT-kinase-enrichment.csv'))
     }
   }
}

```
Note that there is no way to retrieve MIT kinase enrichment results programatically, so uploading the results to Synapse must be done my hand. 

Also be sure to, when running the analysis at WEBSITELINK to select under 'run enrichment analysis on' option select seronin/threonine AND tyrosine for complete results

```{r}

dataFolder <- synapser::synGet("syn64096845")
filenames <- c('mit-enrichment-analysis-result-table-Mirdametinib-24-JH.2.002.txt',
               "mit-enrichment-analysis-result-table-Trametinib-0.5-JH.2.002.txt",
               'mit-enrichment-analysis-result-table Vorinostat-24-JH.2.002.txt',
               'mit-enrichment-analysis-result-table-Vorinostat-24-MN2.txt')

for (i in 1:length(filenames)){
  mySynFile <- synapser::File(paste0("~/Downloads/", filenames[i]), parent = dataFolder)
  synapser::synStore(mySynFile)
}

```