---
title: "Phosphoproteomics enrichment"
author: "Sara gosline"
date: "2024-10-21"
output: html_document
---
```{r, echo=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggfortify)
synapser::synLogin()
#corrected global proteomics: syn61695507
# corrected phospho proteomics: 'syn61695519'
fileEntity <- synapser::synGet('syn61695519')
origPhospho <- read.table(fileEntity$path, header = T)
```


### impute for NAs  - change to row mean 


```{r}
# get the indices of NA's
correctedProteomics<-origPhospho

na_indices <- which(is.na(correctedProteomics), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
correctedProteomics[na_indices] <- rowMeans(correctedProteomics, na.rm=T)[na_indices[,1]]
#na_indices
sum(is.na(correctedProteomics))
dim(correctedProteomics)

samps<-colnames(correctedProteomics)
longProt<-correctedProteomics|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)
```



## Differential expression

We create a single function to compute differential expression using LIMMA.
```{r}
# function to get N genes, adj.P.value < 0.05
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  

factorInfo <- data.frame(stringr::str_split_fixed(names(data), "_", 5))

names(factorInfo) <- c('sample', 'trt', 'time','hr', 'rep')
factorInfo$sample <- as.factor(factorInfo$sample)
factorInfo$trt <-as.factor(factorInfo$trt)

factorInfo$time <- as.factor(factorInfo$time)

# rownames have already been assigned
proSet <- Biobase::ExpressionSet(as.matrix(data))

 # log2 transformation if distribution isn't normal
 ex <- Biobase::exprs(proSet)
 qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                           na.rm = TRUE
 ))
 LogC <- (qx[5] > 100) ||
   (qx[6] - qx[1] > 50 && qx[2] > 0)
 if (LogC) {
   ex[which(ex <= 0)] <- NaN
   Biobase::exprs(proSet) <- log2(ex)
 }


proSet$sample <- factorInfo$sample
proSet$trt <- factorInfo$trt
proSet$time <- factorInfo$time
 #browser()
design <- stats::model.matrix(formula, proSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
#print(cont.matrix)
fit <- limma::lmFit(proSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)
#browser()
fit <- limma::eBayes(fit)
deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
return(list(deg, nDiffExpressed))
}
```

### Compare all time points and conditions

A loop through the data frame calculates all the conditions and adds them to a new data frame.


```{r}
##created for loop to calculate all diff ex stats
t0dat<-data.frame()
ddat<-data.frame()
for(d in c('DMSO','Mirdametinib','Selumetinib','Vorinostat')){
  for(tv in c('0.5','3','24')){
    for(s in c('JH.2.002','MN.2')){
      ##get matrix of values of sample at condition
      tmat<-longProt|>
        subset(sample==s)|>
        subset(drug==d)|>
        subset(time==tv)|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
      if(nrow(tmat)==0)##there is no data for this
        next
      ##matrix of values at DMSO 0
      cmat<-longProt|>
        subset(sample==s)|>
        subset(drug=='DMSO')|>
        subset(time=='0')|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
  
      if(d!='DMSO'){    
        #matrix of values at DMSO same time
        dmat<-longProt|>
          subset(sample==s)|>
          subset(drug=='DMSO')|>
          subset(time==tv)|>
          select(gene,condition,logRatio)|>
          distinct()|>
          tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
        ## get result compared to DMSO
        res<-getNdiffExpressedGenes(cbind(dmat,tmat), formula = ~ trt  + 0, analysisContrast =paste0('trt',d,'-trtDMSO')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
      }
           ##first get restult compare to t0
        res<-getNdiffExpressedGenes(cbind(cmat,tmat), formula = ~ time  + 0, analysisContrast =paste0('time',tv,'-time0')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        t0dat<-rbind(t0dat,tres)
      
    }
  }
}

t0counts = t0dat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(t0counts)
ggplot(t0counts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)
```

Again we see large changes that occur during DMSO treatment, but not as much, particularly at 3 hours.

```{r other changes}

dcounts = ddat|>subset(adj.P.Val<0.05)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(dcounts)
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)

```
Here we see a few changes at .5 hours and more at 24. the 3 hour time point is not great. Let's do an upset plot to see if any proteins are shared across conditions.

```{r upset}
library(UpSetR)
ddat|>
  subset(adj.P.Val<0.1)|>
  select(Gene,drug,time,samp)|>
  tidyr::unite(drug,time,samp,col='condition')|>
  mutate(present=1)|>
  tidyr::pivot_wider(names_from=condition,values_from=present,values_fill=0)|>
  tibble::column_to_rownames('Gene')|>
#  as.matrix()|>
    upset()
```

There are 30 phosphosites shared during mirdametinib treatment, and 19 shared under vorinostat treatment. One phosphosites is shared across all treatments?

```{r shared changes}

ddat|>
  subset(adj.P.Val<0.1)|>
  group_by(Gene)|>summarize(count=n())|>
  arrange(desc(count))
```

MAPK3 and MAPK1 exhibit changes in many conditions.

## Kinase enrichment

We can use KSEA to do kinase enrichment

```{r}
dirname='KinaseEnrichResults-forSara-2024-10-23/'
ksea_files=list.files(dirname,pattern='KSEA-Kinase-scores')

allkins<-do.call(rbind,lapply(ksea_files,function(x){
  tab<-readr::read_csv(paste(dirname,x,sep='/'))
  nl <- unlist(strsplit(x,split='_'))
  tab$Drug=nl[1]
  tab$time=nl[2]
  tab$sample=nl[3]
  tab$AdjP<- (-log10(p.adjust(tab$p.value)))
  return(tab)
}))|>
  dplyr::rename(Kinase='Kinase.Gene')

sigs<-subset(allkins,AdjP>1)
ggplot(sigs,aes(x=Drug,y=Kinase,col=z.score,size=AdjP))+geom_point()+facet_grid(sample~time)+scale_color_gradient(low='darkblue',high='gold')+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave("kinaseEnrichmentKSEA.png",height=8)
```


Now. we can try to build a network

## 24 hr networks
Here we can plot the results and see what the network is!

```{r pcsf}
library(PCSF)
data(STRING)
ppi<-construct_interactome(STRING)

lapply(unique(allkins$Drug),function(d){
  print(d)
  mtargs<-subset(allkins,Drug==d)|>subset(p.value<0.01)|>
    subset(time=='24')
  if(nrow(mtargs)<2)
    return(NULL)
  mvals<-mtargs$z.score
  names(mvals)<-mtargs$Kinase
  #print(mvals)
  uprand<-PCSF_rand(ppi,terminals=mvals[mvals>0])
#  write_graph(uprand,file=paste0(d,'_24hr_up',format='gml'))

  res=enrichment_analysis(uprand)
  plot.PCSFe(res$subnet)
  downrand<-PCSF_rand(ppi,terminals=abs(mvals[mvals<0]))
 # write_graph(downrand,file=paste0(d,'_24hr_up',format='gml'))

  res=enrichment_analysis(downrand)
  plot.PCSFe(res$subnet)
})
```

##30 min networks
here are the networks for 30 min

```{r other network}

lapply(unique(allkins$Drug),function(d){
  print(d)
  mtargs<-subset(allkins,Drug==d)|>subset(p.value<0.05)|>
    subset(time=='0.5')
  if(nrow(mtargs)<2)
    return(NULL)
  mvals<-mtargs$z.score
  
  names(mvals)<-mtargs$Kinase
  uprand<-PCSF_rand(ppi,terminals=mvals[mvals>0])
  res=enrichment_analysis(uprand)
#  pdf(paste0(d,'_24hrs_up.pdf'))
  #write_graph(uprand,file=paste0(d,'_30min_up',format='gml'))
  plot.PCSFe(res$subnet)
  #dev.off()
  downrand<-PCSF_rand(ppi,terminals=abs(mvals[mvals<0]))
 # pdf(paste0(d,'_24hrs_down.pdf'))
  res=enrichment_analysis(downrand)
  plot.PCSFe(res$subnet)
  #write_graph(downrand,file=paste0(d,'_30min_down',format='gml'))

  #devoff()
})
 
```


