---
title: "Phosphoproteomics network analysis"
author: "Sara gosline"
date: "2024-11-14"
output: html_document
---
```{r, echo=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggfortify)
synapser::synLogin()
#corrected global proteomics: syn61695507
# corrected phospho proteomics: 'syn61695519'
fileEntity <- synapser::synGet('syn61695519')
origPhospho <- read.table(fileEntity$path, header = T)
```


### impute for NAs  - change to row mean 


```{r}
# get the indices of NA's
correctedProteomics<-origPhospho

na_indices <- which(is.na(correctedProteomics), arr.ind = T)
# replace unique row by row with the rowMeans(..., na.rm=T)
correctedProteomics[na_indices] <- rowMeans(correctedProteomics, na.rm=T)[na_indices[,1]]
#na_indices
sum(is.na(correctedProteomics))
dim(correctedProteomics)

samps<-colnames(correctedProteomics)
longProt<-correctedProteomics|>
  tibble::rownames_to_column('gene')|>
  tidyr::pivot_longer(cols=all_of(samps),names_to='condition',values_to='logRatio')|>
  tidyr::separate(condition,into=c('sample','drug','time','unit','rep'),sep='_',remove=FALSE)
```



## Differential expression

We create a single function to compute differential expression using LIMMA.
```{r}
# function to get N genes, adj.P.value < 0.05
getNdiffExpressedGenes <- function(data, formula, analysisContrast){
  

factorInfo <- data.frame(stringr::str_split_fixed(names(data), "_", 5))

names(factorInfo) <- c('sample', 'trt', 'time','hr', 'rep')
factorInfo$sample <- as.factor(factorInfo$sample)
factorInfo$trt <-as.factor(factorInfo$trt)

factorInfo$time <- as.factor(factorInfo$time)

# rownames have already been assigned
proSet <- Biobase::ExpressionSet(as.matrix(data))

 # log2 transformation if distribution isn't normal
 ex <- Biobase::exprs(proSet)
 qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0),
                           na.rm = TRUE
 ))
 LogC <- (qx[5] > 100) ||
   (qx[6] - qx[1] > 50 && qx[2] > 0)
 if (LogC) {
   ex[which(ex <= 0)] <- NaN
   Biobase::exprs(proSet) <- log2(ex)
 }


proSet$sample <- factorInfo$sample
proSet$trt <- factorInfo$trt
proSet$time <- factorInfo$time
 #browser()
design <- stats::model.matrix(formula, proSet)

cont.matrix <- limma::makeContrasts(contrasts=analysisContrast, levels = design)
#print(design)
#print(cont.matrix)
fit <- limma::lmFit(proSet, design)
fit <- limma::contrasts.fit(fit, cont.matrix)
#browser()
fit <- limma::eBayes(fit)
deg <- limma::topTable(fit,
                                   adjust = "fdr",
                                   number = nrow(fit), 
                                sort.by = 'p')

nDiffExpressed <- sum(deg$adj.P.Val <= 0.05)
return(list(deg, nDiffExpressed))
}
```

### Compare all time points and conditions

A loop through the data frame calculates all the conditions and adds them to a new data frame.


```{r}
##created for loop to calculate all diff ex stats
t0dat<-data.frame()
ddat<-data.frame()
for(d in c('DMSO','Mirdametinib','Selumetinib','Vorinostat')){
  for(tv in c('0.5','3','24')){
    for(s in c('JH.2.002','MN.2')){
      ##get matrix of values of sample at condition
      tmat<-longProt|>
        subset(sample==s)|>
        subset(drug==d)|>
        subset(time==tv)|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
      if(nrow(tmat)==0)##there is no data for this
        next
      ##matrix of values at DMSO 0
      cmat<-longProt|>
        subset(sample==s)|>
        subset(drug=='DMSO')|>
        subset(time=='0')|>
        select(gene,condition,logRatio)|>
        distinct()|>
        tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
  
      if(d!='DMSO'){    
        #matrix of values at DMSO same time
        dmat<-longProt|>
          subset(sample==s)|>
          subset(drug=='DMSO')|>
          subset(time==tv)|>
          select(gene,condition,logRatio)|>
          distinct()|>
          tidyr::pivot_wider(names_from='condition',values_from='logRatio')|>tibble::column_to_rownames('gene')
        ## get result compared to DMSO
        res<-getNdiffExpressedGenes(cbind(dmat,tmat), formula = ~ trt  + 0, analysisContrast =paste0('trt',d,'-trtDMSO')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        ddat<-rbind(ddat,tres)
      }
           ##first get restult compare to t0
        res<-getNdiffExpressedGenes(cbind(cmat,tmat), formula = ~ time  + 0, analysisContrast =paste0('time',tv,'-time0')) 
        tres<-res[[1]]|>mutate(time=tv,drug=d,samp=s)|>
          tibble::rownames_to_column('Gene')
        t0dat<-rbind(t0dat,tres)
      
    }
  }
}

t0counts = t0dat|>subset(adj.P.Val<0.1)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(t0counts)
ggplot(t0counts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)
```

Again we see large changes that occur during DMSO treatment, but not as much, particularly at 3 hours.

```{r other changes}

dcounts = ddat|>subset(adj.P.Val<0.05)|>group_by(time,drug,samp)|>summarize(DiffEx=n())
print(dcounts)
ggplot(dcounts,aes(x=drug,y=DiffEx,fill=samp))+geom_bar(stat='identity',position='dodge')+facet_grid(~time)

```
Here we see a few changes at .5 hours and more at 24. the 3 hour time point is not great. Let's do an upset plot to see if any proteins are shared across conditions.



## Networks
I dont think networks will be the solution, but i tried below

```{r networks}

library(PCSF)
data("STRINGv12")
data("PSP_NK")



allnets<-lapply(unique(ddat$Drug),function(d){
  print(d)
  mtargs<-subset(ddat,drug==d)|>subset(P.Value<0.001)|>
    subset(time=='24')
  if(nrow(mtargs)<2)
    return(NULL)
  mvals<-mtargs$logFC
  names(mvals)<-sapply(mtargs$Gene,function(x) gsub('s$|t$|y$','',x))
  #print(mvals)
  uprand<-PCSF_phos(STRINGv12,c(),PSP_NK,phosterminals=mvals[mvals>0],fname=paste0(d,'_24hrs'))
#  write_graph(uprand,file=paste0(d,'_24hr_up',format='gml'))

  upres=enrichment_analysis(uprand)
  #plot.PCSFe(res$subnet)
  down<-PCSF_phos(STRINGv12,c(),PSP_NK,phosterminals=abs(mvals[mvals>0]),fname=paste0(d,'_24hrs'))<
 # write_graph(downrand,file=paste0(d,'_24hr_up',format='gml'))

  res=enrichment_analysis(downrand)
  #plot.PCSFe(res$subnet)
  return(list(upreg=upres$subnet,downreg=res$subnet))
})

jaccard <- function(a, b) {
  na<-names(V(a))
  nb<-names(V(b))
    intersection = length(intersect(na, nb))
    union = length(na) + length(nb) - intersection
    return (intersection/union)
}

```

##30 min networks
here are the networks for 30 min

```{r other network}

lapply(unique(ddat$Drug),function(d){
  printd
  mtargs<-subset(ddat,Drug==d)|>subset(P.Value<0.05)|>
    subset(time=='0.5')
  if(nrow(mtargs)<2)
    return(NULL)
   mvals<-mtargs$logFC
  names(mvals)<-sapply(mtargs$Gene,function(x) gsub('s$|t$|y$','',x))
  #print(mvals)
  uprand<-PCSF_phos(STRINGv12,c(),PSP_NK,phosterminals=mvals[mvals>0],fname=paste0(d,'_30min'))

  res=enrichment_analysis(uprand)
#  pdf(paste0(d,'_24hrs_up.pdf'))
  #write_graph(uprand,file=paste0(d,'_30min_up',format='gml'))
  plot.PCSFe(res$subnet)
  #dev.off()
  down<-PCSF_phos(STRINGv12,c(),PSP_NK,phosterminals=abs(mvals[mvals>0]),fname=paste0(d,'_30min'))<
 # pdf(paste0(d,'_24hrs_down.pdf'))
  res=enrichment_analysis(downrand)
  plot.PCSFe(res$subnet)
  #write_graph(downrand,file=paste0(d,'_30min_down',format='gml'))

  #devoff()
})
 
```


